### 2024-01-18 main.tex
Initial Overleaf Import

%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
 \usepackage{icml2024}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2024}

% For theorems and such
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Submission and Formatting Instructions for ICML 2024}

\begin{document}

\twocolumn[
\icmltitle{Spurious Sentience\\
           On the Unsurprising Finding of Patterns in Latent Spaces}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2024
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Patrick Altmeyer}{yyy}
\icmlauthor{Antony Bartlett}{equal, yyy}
\icmlauthor{Andrew M. Demetriou}{equal, yyy}
\icmlauthor{Cynthia C. S. Liem}{yyy}
\end{icmlauthorlist}

\icmlaffiliation{yyy}{Department of Intelligent Systems, Delft University of Technology, Delft, the Netherlands}

\icmlcorrespondingauthor{Firstname1 Lastname1}{first1.last1@xxx.edu}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, anthropomorphism, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
This document provides a basic paper template and submission guidelines.
Abstracts must be a single paragraph, ideally between 4--6 sentences long.
Gross violations will trigger corrections at the camera-ready phase.
\end{abstract}

\section{Introduction}
\label{submission}

We humans are prone to seek patterns everywhere. Meaningful patterns have often proven to help us make sense of our past, navigate our present and predict the future, likely leading to evolutionary benefits even when the patterns we perceive are false \cite{foster2009evolution}. Nevertheless, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, and interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.

In statistics, misleading patterns are referred to as spurious relationships: purely associational relationships between two or more variables that are not causally related to each other. Although the definition of spuriousness varies somewhat \cite{haig2003spurious}, every student of statistics has learned to be wary that the observation of correlations do not necessarily imply causation. The world is full of these non-causal associations and as good as we as species may be at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones. Humans struggle to perceive randomness, as studies show that humans are poor at detecting e.g. \cite{falk1997making}, and creating e.g. \cite{ladouceur1996erroneous} random sequences, and even those trained in statistics and probability make inappropriate inferences, seeing patterns in data that is randomly generated \cite{zgraggen2018investigating}.

Drawing conclusions from spurious relationships is further compounded in AI research by at least two other phenomena

Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path. Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads. Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.

I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy. But I am very skeptical of claims that AI will magically provide that remedy. This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them. The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots. Since this is an opinionated post, I feel that I should start with a few disclaimers:

\begin{itemize}
    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
    \item I therefore do take issue with the way that these findings are being overblown by people with clout. Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
\end{itemize}

\section{Patterns in Latent Spaces and How to Find Them}

To demonstrate the claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples.

\subsection{Example: Principal Component Analysis}

To understand this better, let us go on a quick detour into economics and look at actual yield curves observed in the US during the Global Financial Crisis (GFC). Figure 1 (a) shows the yield curve of US Treasury bonds on 27 February 2007, which according to CNN was a “brutal day on Wall Street”.2 This followed reports on the previous day of former Federal Reserve Chairman Alan Greenspan’s warning that the US economy was at risk of a recession. The yield curve was inverted with a sharp negative spread between the 10-year and 3-month yields, indicative of the market’s expectation of a recession.

Figure 1 (b) shows the corresponding yield curve during the aftermath of the GFC on 20 April 2009. On that day the influential Time Magazine reported that the “Banking Crisis is Over”. The yield curve was steeply sloped with a positive spread between the 10-year and 3-month yields, indicative of the market’s expectation of a recovery. The overall level of the yield curve was still very low though, indicative of the fact that US economy had not fully recovered at that point.

Of course, US Treasuries are not the only bonds that are traded in the market. To get a more complete picture of the economy, analysts might therefore be interested in looking at the yield curves of other bonds as well. In particular, we might be interested in predicting economic growth based on the yield curves of many different bonds. The problem with that idea is that it is cursed by high dimensionality: we would end up modelling a single variable of interest (economic growth) with a large number of predictors (the yields of many different bonds). To deal with the curse of high dimensionality it can be useful to decompose the yield curves into sets of principal components.

To compute the principal components we can decompose the matrix of yields 
 into a product of its singular vectors and values: 
. I will not go into the details here, because Professor Gilbert Strang has already done a much better job than I ever could in his Linear Algebra lectures. To put this into the broader context of the article, however, let us simply refer to 
, 
 and 
 as latent embeddings of the yield curve (they are latent because they are not directly observable).

The top panel in Figure 2 shows the first two principal components of the yield curves of US Treasury bonds over time. Vertical stalks indicate the key dates during the onset and aftermath of the crisis, which we discussed above. For both components, we can observe some marked shifts between the two dates - but can we attribute any meaning to these shifts? It turns out we can: for comparison, the bottom panel in Figure 2 shows the average level and spread of the yield curves over time. The first principal component is strongly correlated with the level of the yield curve, while the second principal component is strongly correlated with the spread of the yield curve. To put it in AI-lingo:

Not convinced? Let us use 
 in true autoencoder fashion to reconstruct yield curves from principal components. Let 
 denote the first principal component and consider the following: we keep all other 
 principal components fixed at zero where 
 denotes the total number of maturities; next we traverse the latent space by varying the value of 
 over a fixed grid of length 
 each time storing the full vector 
; finally, we vertically concatenate the vectors and end up with a matrix 
 of dimension 
. To reconstruct yields, we simply multiply 
 by the singular values and right singular vectors: 
.

Figure 3 shows the result of this exercise in the left panel. As we can see, our generated yield curves shift vertically as we traverse the latent space. The right panel of Figure 3 shows the result of a similar exercise, but this time we keep the first principal component fixed at zero and vary the second principal component. This time the slope of our generated yield curves shifts as we traverse the latent space.

\subsection{Example: Deep Learning}

So far we have considered simple matrix decomposition. You might argue that principal components are not really latent embeddings in the traditional sense of deep learning. To address this, let us now consider a simple deep-learning example. Our goal will be to not only predict economic growth from the yield curve but also extract meaningful features at the same time. In particular, we will use a neural network architecture that allows us to recover a compressed latent representation of the yield curve.

\subsubsection{Data}

To estimate economic growth we will rely on a quarterly series of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St. Louis. The data arrives in terms of levels of real GDP. In order to estimate growth, we will transform the data into log differences. Since our yield curve data is daily, we will need to aggregate it to the quarterly frequency. To do this, we will simply take the average of the daily yields for each maturity. We will also standardize yields since deep learning models tend to perform better with standardized data. Since COVID-19 was a huge structural break, we will also filter out all observations after 2018. Figure 4 shows the pre-processed data.

\subsubsection{Model}

Let 
 denote growth and 
 denote the yield curve at time 
. Then we are interested in a model for 
 conditional on 
. Let 
 denote our model parameters then formally we are interested in maximizing the likelihood 
. To do this, we will use a simple autoencoder architecture that is illustrated in Figure 5. The encoder will consist of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function. The bottleneck layer connecting the encoder to the decoder is a fully connected layer with 6 neurons. This is the compressed latent representation of the yield curve mentioned above. The decoder will consist of two fully connected layers each with a hyperbolic tangent activation function: the first layer will consist of 32 neurons and the second layer will have the same dimension as the input data. The output layer will consist of a single neuron with a linear activation function. The model will be trained using the mean squared error loss function and the Adam optimizer.

\subsubsection{Linear Probe}

The results are shown in Figure 6. The top panel shows the actual GDP growth and fitted values from the autoencoder model. We observe that the model captures the relationship between economic growth and the yield curve reasonably well. As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread. Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.

The bottleneck layer seems like a good place to start looking. To get the latent embeddings 
 at time 
, we simply pass the yield curve data 
 through the encoder. Next, we follow [TEGMARK PAPER REFERENCE] and use a linear probe to regress the observed yield curve factors on the latent embeddings. Let 
 denote the vector containing the two factors of interest in time 
: 
 and 
. Formally, we are interested in the following regression model: 
 where 
 denotes the regression parameters. Following [TEGMARK PAPER REFERENCE] we use Ridge regression with 
 set to 
. Using the estimated regression parameters 
 we can then predict the yield curve factors from the latent embeddings: 
.

The results of this experiment are shown in the bottom panel of Figure 6. Solid lines show the observed yield curve factors over time, while dashed lines show predicted values. We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread. To put this in true hype-lingo:

Not just a parrot! Our autoencoder neural network has an implicit understanding of the yield curve factors and their relationship with economic growth. It’s sentient indeed!

\section{Example: MNIST}

Leveraging a small MLP and VAE trained on MNIST data, this script trains a tokenizer that encodes geographical meaning in the latent embedding of VAE. Inspired by this tweet: https://x.com/savvyRL/status/1709698089500680264?s=20

https://gist.github.com/pat-alt/1c3d0d159b2cfa2a91a8d0c47ea3d28a

This tweet inspired me. Just as I had feared, even the small VAE I trained on MNIST a while ago is basically sentient. It is so over people!

\section{Ok, but truly what’s the point?}

The finding is not surprising but it is still interesting. In the context of mechanistic interpretability, it demonstrates that the black-box model has evidently learned plausible explanations for the data. Beyond that, in this particular example, the patterns in the latent space that we have just uncovered might actually be useful for downstream tasks. An interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model. To demonstrate this, let us consider a simple linear regression model for GDP growth. We will compare the performance of the following models: (1) regressing growth 
 on the latent embedding 
, (2) regression growth on the best subset of latent embeddings 
, (3) regressing growth on lagged growth, (4) regressing growth on lagged growth and the observed yield curve factors 
, (5) regressing growth on the best subset of yield curve factors 
, and, finally, (5) regressing growth on lagged growth and the best subset of latent embeddings 
.

The results are shown in the table below. The key finding of interest is that the coefficients on the latent embeddings in model (5) are statistically significant. This suggests that the latent embeddings contain information that is useful for predicting GDP growth.

\section{LLM}

\begin{itemize}
    \item Use FOMC headlines and corresponding interest rate decisions and see if embeddings contain information about the decision.
    \item Based on fitted linear model, predict interest decisions for random headlines about hawks and doves (animals, not people).
    \item If headlines about hawks tend to be associated with rate hikes, then we the model can be said to have learned a spurious relationship between hawks and rate hikes.
\end{itemize}

Submission to ICML 2024 will be entirely electronic, via a web site
(not email). Information about the submission process and \LaTeX\ templates
are available on the conference web site at:
\begin{center}
\textbf{\texttt{http://icml.cc/}}
\end{center}

The guidelines below will be enforced for initial submissions and
camera-ready copies. Here is a brief summary:
\begin{itemize}
\item Submissions must be in PDF\@. 
\item \textbf{New to this year}: If your paper has appendices, submit the appendix together with the main body and the references \textbf{as a single file}. Reviewers will not look for appendices as a separate PDF file. So if you submit such an extra file, reviewers will very likely miss it.
\item Page limit: The main body of the paper has to be fitted to 8 pages, excluding references and appendices; the space for the latter two is not limited. For the final version of the paper, authors can add one extra page to the main body.
\item \textbf{Do not include author information or acknowledgements} in your
    initial submission.
\item Your paper should be in \textbf{10 point Times font}.
\item Make sure your PDF file only uses Type-1 fonts.
\item Place figure captions \emph{under} the figure (and omit titles from inside
    the graphic file itself). Place table captions \emph{over} the table.
\item References must include page numbers whenever possible and be as complete
    as possible. Place multiple citations in chronological order.
\item Do not alter the style template; in particular, do not compress the paper
    format by reducing the vertical spaces.
\item Keep your abstract brief and self-contained, one paragraph and roughly
    4--6 sentences. Gross violations will require correction at the
    camera-ready phase. The title should have content words capitalized.
\end{itemize}

\subsection{Submitting Papers}

\textbf{Paper Deadline:} The deadline for paper submission that is
advertised on the conference website is strict. If your full,
anonymized, submission does not reach us on time, it will not be
considered for publication.

\textbf{Anonymous Submission:} ICML uses double-blind review: no identifying
author information may appear on the title page or in the paper
itself. \cref{author info} gives further details.

\textbf{Simultaneous Submission:} ICML will not accept any paper which,
at the time of submission, is under review for another conference or
has already been published. This policy also applies to papers that
overlap substantially in technical content with conference papers
under review or previously published. ICML submissions must not be
submitted to other conferences and journals during ICML's review
period.
%Authors may submit to ICML substantially different versions of journal papers
%that are currently under review by the journal, but not yet accepted
%at the time of submission.
Informal publications, such as technical
reports or papers in workshop proceedings which do not appear in
print, do not fall under these restrictions.

\medskip

Authors must provide their manuscripts in \textbf{PDF} format.
Furthermore, please make sure that files contain only embedded Type-1 fonts
(e.g.,~using the program \texttt{pdffonts} in linux or using
File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
might come from graphics files imported into the document.

Authors using \textbf{Word} must convert their document to PDF\@. Most
of the latest versions of Word have the facility to do this
automatically. Submissions will not be accepted in Word format or any
format other than PDF\@. Really. We're not joking. Don't send Word.

Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
Those using \texttt{latex} and \texttt{dvips} may need the following
two commands:

{\footnotesize
\begin{verbatim}
dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
ps2pdf paper.ps
\end{verbatim}}
It is a zero following the ``-G'', which tells dvips to use
the config.pdf file. Newer \TeX\ distributions don't always need this
option.

Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
results. This program avoids the Type-3 font problem, and supports more
advanced features in the \texttt{microtype} package.

\textbf{Graphics files} should be a reasonable size, and included from
an appropriate format. Use vector formats (.eps/.pdf) for plots,
lossless bitmap formats (.png) for raster graphics with sharp lines, and
jpeg for photo-like images.

The style file uses the \texttt{hyperref} package to make clickable
links in documents. If this causes problems for you, add
\texttt{nohyperref} as one of the options to the \texttt{icml2024}
usepackage statement.

\subsection{Submitting Final Camera-Ready Copy}

The final versions of papers accepted for publication should follow the
same format and naming convention as initial submissions, except that
author information (names and affiliations) should be given. See
\cref{final author} for formatting instructions.

The footnote, ``Preliminary work. Under review by the International
Conference on Machine Learning (ICML). Do not distribute.'' must be
modified to ``\textit{Proceedings of the
$\mathit{41}^{st}$ International Conference on Machine Learning},
Vienna, Austria, PMLR 235, 2024.
Copyright 2024 by the author(s).''

For those using the \textbf{\LaTeX} style file, this change (and others) is
handled automatically by simply changing
$\mathtt{\backslash usepackage\{icml2024\}}$ to
$$\mathtt{\backslash usepackage[accepted]\{icml2024\}}$$
Authors using \textbf{Word} must edit the
footnote on the first page of the document themselves.

Camera-ready copies should have the title of the paper as running head
on each page except the first one. The running title consists of a
single line centered above a horizontal rule which is $1$~point thick.
The running head should be centered, bold and in $9$~point type. The
rule should be $10$~points above the main text. For those using the
\textbf{\LaTeX} style file, the original title is automatically set as running
head using the \texttt{fancyhdr} package which is included in the ICML
2024 style file package. In case that the original title exceeds the
size restrictions, a shorter form can be supplied by using

\verb|\icmltitlerunning{...}|

just before $\mathtt{\backslash begin\{document\}}$.
Authors using \textbf{Word} must edit the header of the document themselves.

\section{Format of the Paper}

All submissions must follow the specified format.

\subsection{Dimensions}

The text of the paper should be formatted in two columns, with an
overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
between the columns. The left margin should be 0.75~inches and the top
margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
whether you print on US letter or A4 paper, but all final versions
must be produced for US letter size.
Do not write anything on the margins.

The paper body should be set in 10~point type with a vertical spacing
of 11~points. Please use Times typeface throughout the text.

\subsection{Title}

The paper title should be set in 14~point bold type and centered
between two horizontal rules that are 1~point thick, with 1.0~inch
between the top rule and the top edge of the page. Capitalize the
first letter of content words and put the rest of the title in lower
case.

\subsection{Author Information for Submission}
\label{author info}

ICML uses double-blind review, so author information must not appear. If
you are using \LaTeX\/ and the \texttt{icml2024.sty} file, use
\verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
will not be printed unless \texttt{accepted} is passed as an argument to the
style file.
Submissions that include the author information will not
be reviewed.

\subsubsection{Self-Citations}

If you are citing published papers for which you are an author, refer
to yourself in the third person. In particular, do not use phrases
that reveal your identity (e.g., ``in previous work \cite{langley00}, we
have shown \ldots'').

Do not anonymize citations in the reference section. The only exception are manuscripts that are
not yet published (e.g., under submission). If you choose to refer to
such unpublished manuscripts \cite{anonymous}, anonymized copies have
to be submitted
as Supplementary Material via OpenReview\@. However, keep in mind that an ICML
paper should be self contained and should contain sufficient detail
for the reviewers to evaluate the work. In particular, reviewers are
not required to look at the Supplementary Material when writing their
review (they are not required to look at more than the first $8$ pages of the submitted document).

\subsubsection{Camera-Ready Author Information}
\label{final author}

If a paper is accepted, a final camera-ready copy must be prepared.
%
For camera-ready papers, author information should start 0.3~inches below the
bottom rule surrounding the title. The authors' names should appear in 10~point
bold type, in a row, separated by white space, and centered. Author names should
not be broken across lines. Unbolded superscripted numbers, starting 1, should
be used to refer to affiliations.

Affiliations should be numbered in the order of appearance. A single footnote
block of text should be used to list all the affiliations. (Academic
affiliations should list Department, University, City, State/Region, Country.
Similarly for industrial affiliations.)

Each distinct affiliations should be listed once. If an author has multiple
affiliations, multiple superscripts should be placed after the name, separated
by thin spaces. If the authors would like to highlight equal contribution by
multiple first authors, those authors should have an asterisk placed after their
name in superscript, and the term ``\textsuperscript{*}Equal contribution"
should be placed in the footnote block ahead of the list of affiliations. A
list of corresponding authors and their emails (in the format Full Name
\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
Ideally only one or two names should be listed.

A sample file with author names is included in the ICML2024 style file
package. Turn on the \texttt{[accepted]} option to the stylefile to
see the names rendered. All of the guidelines above are implemented
by the \LaTeX\ style file.

\subsection{Abstract}

The paper abstract should begin in the left column, 0.4~inches below the final
address. The heading `Abstract' should be centered, bold, and in 11~point type.
The abstract body should use 10~point type, with a vertical spacing of
11~points, and should be indented 0.25~inches more than normal on left-hand and
right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
sentences. Gross violations will require correction at the camera-ready phase.

\subsection{Partitioning the Text}

You should organize your paper into sections and paragraphs to help
readers place a structure on the material and understand its
contributions.

\subsubsection{Sections and Subsections}

Section headings should be numbered, flush left, and set in 11~pt bold
type with the content words capitalized. Leave 0.25~inches of space
before the heading and 0.15~inches after the heading.

Similarly, subsection headings should be numbered, flush left, and set
in 10~pt bold type with the content words capitalized. Leave
0.2~inches of space before the heading and 0.13~inches afterward.

Finally, subsubsection headings should be numbered, flush left, and
set in 10~pt small caps with the content words capitalized. Leave
0.18~inches of space before the heading and 0.1~inches after the
heading.

Please use no more than three levels of headings.

\subsubsection{Paragraphs and Footnotes}

Within each section or subsection, you should further partition the
paper into paragraphs. Do not indent the first line of a given
paragraph, but insert a blank line between succeeding ones.

You can use footnotes\footnote{Footnotes
should be complete sentences.} to provide readers with additional
information about a topic without interrupting the flow of the paper.
Indicate footnotes with a number in the text where the point is most
relevant. Place the footnote in 9~point type at the bottom of the
column in which it appears. Precede the first footnote in a column
with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
appear in each column, in the same order as they appear in the text,
but spread them across columns and pages if possible.}

\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
\caption{Historical locations and number of accepted papers for International
Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
produced, the number of accepted papers for ICML 2008 was unknown and instead
estimated.}
\label{icml-historical}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Figures}

You may want to include figures in the paper to illustrate
your approach and results. Such artwork should be centered,
legible, and separated from the text. Lines should be dark and at
least 0.5~points thick for purposes of reproduction, and text should
not appear on a gray background.

Label all distinct components of each figure. If the figure takes the
form of a graph, then give a name for each axis and include a legend
that briefly describes each curve. Do not include a title inside the
figure; instead, the caption should serve this function.

Number figures sequentially, placing the figure number and caption
\emph{after} the graphics, with at least 0.1~inches of space before
the caption and 0.1~inches after it, as in
\cref{icml-historical}. The figure caption should be set in
9~point type and centered unless it runs two or more lines, in which
case it should be flush left. You may float figures to the top or
bottom of a column, and you may set wide figures across both columns
(use the environment \texttt{figure*} in \LaTeX). Always place
two-column figures at the top or bottom of the page.

\subsection{Algorithms}

If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
environments to format pseudocode. These require
the corresponding stylefiles, algorithm.sty and
algorithmic.sty, which are supplied with this package.
\cref{alg:example} shows an example.

\begin{algorithm}[tb]
   \caption{Bubble Sort}
   \label{alg:example}
\begin{algorithmic}
   \STATE {\bfseries Input:} data $x_i$, size $m$
   \REPEAT
   \STATE Initialize $noChange = true$.
   \FOR{$i=1$ {\bfseries to} $m-1$}
   \IF{$x_i > x_{i+1}$}
   \STATE Swap $x_i$ and $x_{i+1}$
   \STATE $noChange = false$
   \ENDIF
   \ENDFOR
   \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

\subsection{Tables}

You may also want to include tables that summarize material. Like
figures, these should be centered, legible, and numbered consecutively.
However, place the title \emph{above} the table with at least
0.1~inches of space before the title and the same after it, as in
\cref{sample-table}. The table title should be set in 9~point
type and centered unless it runs two or more lines, in which case it
should be flush left.

% Note use of \abovespace and \belowspace to get reasonable spacing
% above and below tabular lines.

\begin{table}[t]
\caption{Classification accuracies for naive Bayes and flexible
Bayes on various data sets.}
\label{sample-table}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
Data set & Naive & Flexible & Better? \\
\midrule
Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

Tables contain textual material, whereas figures contain graphical material.
Specify the contents of each row and column in the table's topmost
row. Again, you may float tables to a column's top or bottom, and set
wide tables across both columns. Place two-column tables at the
top or bottom of the page.

\subsection{Theorems and such}
The preferred way is to number definitions, propositions, lemmas, etc. consecutively, within sections, as shown below.
\begin{definition}
\label{def:inj}
A function $f:X \to Y$ is injective if for any $x,y\in X$ different, $f(x)\ne f(y)$.
\end{definition}
Using \cref{def:inj} we immediate get the following result:
\begin{proposition}
If $f$ is injective mapping a set $X$ to another set $Y$, 
the cardinality of $Y$ is at least as large as that of $X$
\end{proposition}
\begin{proof} 
Left as an exercise to the reader. 
\end{proof}
\cref{lem:usefullemma} stated next will prove to be useful.
\begin{lemma}
\label{lem:usefullemma}
For any $f:X \to Y$ and $g:Y\to Z$ injective functions, $f \circ g$ is injective.
\end{lemma}
\begin{theorem}
\label{thm:bigtheorem}
If $f:X\to Y$ is bijective, the cardinality of $X$ and $Y$ are the same.
\end{theorem}
An easy corollary of \cref{thm:bigtheorem} is the following:
\begin{corollary}
If $f:X\to Y$ is bijective, 
the cardinality of $X$ is at least as large as that of $Y$.
\end{corollary}
\begin{assumption}
The set $X$ is finite.
\label{ass:xfinite}
\end{assumption}
\begin{remark}
According to some, it is only the finite case (cf. \cref{ass:xfinite}) that is interesting.
\end{remark}
%restatable

\subsection{Citations and References}

Please use APA reference format regardless of your formatter
or word processor. If you rely on the \LaTeX\/ bibliographic
facility, use \texttt{natbib.sty} and \texttt{icml2024.bst}
included in the style-file package to obtain this format.

Citations within the text should include the authors' last names and
year. If the authors' names are included in the sentence, place only
the year in parentheses, for example when referencing Arthur Samuel's
pioneering work \yrcite{Samuel59}. Otherwise place the entire
reference in parentheses with the authors and year separated by a
comma \cite{Samuel59}. List multiple references separated by
semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
construct only for citations with three or more authors or after
listing all authors to a publication in an earlier reference \cite{MachineLearningI}.

Authors should cite their own work in the third person
in the initial version of their paper submitted for blind review.
Please refer to \cref{author info} for detailed instructions on how to
cite your own papers.

Use an unnumbered first-level section heading for the references, and use a
hanging indent style, with the first line of the reference flush against the
left margin and subsequent lines indented by 10 points. The references at the
end of this document give examples for journal articles \cite{Samuel59},
conference publications \cite{langley00}, book chapters \cite{Newell81}, books
\cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
\cite{mitchell80}, and dissertations \cite{kearns89}.

Alphabetize references by the surnames of the first authors, with
single author entries preceding multiple author entries. Order
references for the same authors by year of publication, with the
earliest first. Make sure that each reference includes all relevant
information (e.g., page numbers).

Please put some effort into making references complete, presentable, and
consistent, e.g. use the actual current name of authors.
If using bibtex, please protect capital letters of names and
abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
in your .bib file.

\section*{Accessibility}
Authors are kindly asked to make their submissions as accessible as possible for everyone including people with disabilities and sensory or neurological differences.
Tips of how to achieve this and what to pay attention to will be provided on the conference website \url{http://icml.cc/}.

\section*{Software and Data}

If a paper is accepted, we strongly encourage the publication of software and data with the
camera-ready version of the paper whenever appropriate. This can be
done by including a URL in the camera-ready copy. However, \textbf{do not}
include URLs that reveal your institution or identity in your
submission for review. Instead, provide an anonymous URL or upload
the material as ``Supplementary Material'' into the OpenReview reviewing
system. Note that reviewers are not required to look at this material
when writing their review.

% Acknowledgements should only appear in the accepted version.
\section*{Acknowledgements}

\textbf{Do not} include acknowledgements in the initial version of
the paper submitted for blind review.

If a paper is accepted, the final camera-ready version can (and
probably should) include acknowledgements. In this case, please
place such acknowledgements in an unnumbered section at the
end of the paper. Typically, this will include thanks to reviewers
who gave useful comments, to colleagues who contributed to the ideas,
and to funding agencies and corporate sponsors that provided financial
support.

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\nocite{langley00}

\bibliography{example_paper}
\bibliographystyle{icml2024}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{You \emph{can} have an appendix here.}

You can have as much text here as you want. The main body must be at most $8$ pages long.
For the final version, one more page can be added.
If you want, you can use an appendix like this one.

The $\mathtt{\backslash onecolumn}$ command above can be kept in place if you prefer a one-column appendix, or can be removed if you prefer a two-column appendix.  Apart from this possible change, the style (font size, spacing, margins, page numbering, etc.) should be kept the same as the main body.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2024-01-18 src/probes/run_probes.jl
running first linear probes

@@ -1,0 +1,65 @@
+using DataFrames
+using Dates
+using Plots
+using TrillionDollarWords
+
+include("utils.jl")
+
+# All data:
+all_data = load_all_data()
+
+# Inflation:
+indicator = "CPI"
+mkt_data = subset(all_data, :indicator => x -> x.==indicator)
+transform!(mkt_data, :date => ByRow(x -> Dates.yearmonth(x)) => :ym)
+mkt_data = groupby(mkt_data, :ym) |>
+    x -> select(x, [:ym, :value]) |>
+    unique |>
+    x -> transform(x, :value => lag => :value_lag) |>
+    x -> select(x, Not(:value)) |>
+    x -> innerjoin(x, mkt_data, on=:ym) |>
+    x -> transform(x, [:value, :value_lag] => ByRow((y,yl) -> log(y)-log(yl)) => :growth) 
+replace!(mkt_data.growth, Inf => 0.0)
+select!(mkt_data, Not([:value]))
+rename!(mkt_data, :growth => :value)
+select!(mkt_data, [:date, :ym, :sentence_id, :value])
+
+X, y = prepare_probe(mkt_data; layer=24)
+
+# Run the probe:
+mod = probe(X,y)
+yhat = mod(X)
+
+# Run the baseline:
+agg_data = groupby(mkt_data, :ym) |>
+    x -> combine(x, :value .=> mean; renamecols=false)
+_y = agg_data.value
+p = 5
+mod_bl, _X = ar(_y; l=p)
+y_bl = mod_bl(_X[:,2:end])
+agg_data.y_bl .= y_bl
+
+# Save the results:
+mkt_data[:,:yhat] .= yhat
+agg_data_probe = groupby(mkt_data, :ym) |>
+    x -> combine(x, :yhat .=> mean; renamecols=false) 
+
+# Merge the results:
+agg_data = innerjoin(agg_data, agg_data_probe, on=:ym)
+
+# Plot the results:
+plot(
+    [agg_data.value agg_data.yhat agg_data.y_bl], 
+    label=[indicator "Probe" "AR($p)"], 
+    legend=:topleft, 
+    xlabel="Year", 
+    ylabel=indicator
+)
+
+# Score the results:
+DataFrame(
+    indicator = indicator,
+    model = ["Probe", "AR($p)"],
+    rmse = [rmse(mod, X, y), rmse(mod_bl, _X[:,2:end], _y)],
+    mse = [mse(mod, X, y), mse(mod_bl, _X[:,2:end], _y)]
+)

### 2024-01-18 src/probes/utils.jl

@@ -1,0 +1,53 @@
+using LinearAlgebra
+using Statistics
+
+abstract type Model end
+
+function (mod::Model)(X)
+    return predict(mod, X)
+end
+
+function predict(mod::Model, X)
+    if mod.intercept
+        X = hcat(ones(size(X, 1)), X)
+    end
+    return X * mod.β
+end
+
+function mse(mod::Model, X, y)
+    ŷ = predict(mod, X)
+    return mean((y .- ŷ) .^ 2)
+end
+
+rmse(mod::Model, X, y) = sqrt(mse(mod, X, y))
+
+struct Probe <: Model
+    β::Vector{Float64}
+    intercept::Bool
+    λ::Float64
+end
+
+function probe(X, y; λ::Real=0.1, intercept::Bool=true)
+    if intercept
+        X = hcat(ones(size(X, 1)), X)
+    end
+    β = (X'X + UniformScaling(λ))\(X'y)
+    return Probe(β, intercept, λ)
+end
+
+lag(X::Vector, l=1) = [zeros(l) ; X[1:end-l]]
+
+struct AutoRegression <: Model
+    β::Vector{Float64}
+    intercept::Bool
+    l::Int
+end
+
+function ar(y; l::Int=1, intercept::Bool=true)
+    X = hcat([lag(y, i) for i in 1:l]...)
+    if intercept
+        X = hcat(ones(size(X, 1)), X)
+    end
+    β = (X'X)\(X'y)
+    return AutoRegression(β, intercept, l), X
+end

### 2024-01-18 src/probes/run_probes.jl
progress

@@ -9,7 +9,7 @@ include("utils.jl")
 all_data = load_all_data()
 
 # Inflation:
-indicator = "CPI"
+indicator = "PPI"
 mkt_data = subset(all_data, :indicator => x -> x.==indicator)
 transform!(mkt_data, :date => ByRow(x -> Dates.yearmonth(x)) => :ym)
 mkt_data = groupby(mkt_data, :ym) |>
@@ -24,7 +24,8 @@ select!(mkt_data, Not([:value]))
 rename!(mkt_data, :growth => :value)
 select!(mkt_data, [:date, :ym, :sentence_id, :value])
 
-X, y = prepare_probe(mkt_data; layer=24)
+layer = 1
+X, y = prepare_probe(mkt_data; layer=layer)
 
 # Run the probe:
 mod = probe(X,y)
@@ -48,13 +49,14 @@ agg_data_probe = groupby(mkt_data, :ym) |>
 agg_data = innerjoin(agg_data, agg_data_probe, on=:ym)
 
 # Plot the results:
-plot(
+plt = plot(
     [agg_data.value agg_data.yhat agg_data.y_bl], 
     label=[indicator "Probe" "AR($p)"], 
     legend=:topleft, 
     xlabel="Year", 
     ylabel=indicator
 )
+display(plt)
 
 # Score the results:
 DataFrame(

### 2024-01-19 src/probes/run_probes.jl
additional helper functions for lag selection

@@ -4,9 +4,12 @@ using Plots
 using TrillionDollarWords
 
 include("utils.jl")
+use_clf = false
 
 # All data:
 all_data = load_all_data()
+model = load_model(;use_head=true)
+clf = model.mod.cls
 
 # Inflation:
 indicator = "PPI"
@@ -24,8 +27,12 @@ select!(mkt_data, Not([:value]))
 rename!(mkt_data, :growth => :value)
 select!(mkt_data, [:date, :ym, :sentence_id, :value])
 
-layer = 1
+layer = 24
 X, y = prepare_probe(mkt_data; layer=layer)
+
+if use_clf
+    X = vcat([get_clf_activations(clf, x)' for x in eachrow(X)]...)
+end
 
 # Run the probe:
 mod = probe(X,y)
@@ -35,7 +42,7 @@ yhat = mod(X)
 agg_data = groupby(mkt_data, :ym) |>
     x -> combine(x, :value .=> mean; renamecols=false)
 _y = agg_data.value
-p = 5
+lag_select(_y)
 mod_bl, _X = ar(_y; l=p)
 y_bl = mod_bl(_X[:,2:end])
 agg_data.y_bl .= y_bl
@@ -52,6 +59,7 @@ agg_data = innerjoin(agg_data, agg_data_probe, on=:ym)
 plt = plot(
     [agg_data.value agg_data.yhat agg_data.y_bl], 
     label=[indicator "Probe" "AR($p)"], 
+    alpha=[0.2 1.0 1.0],
     legend=:topleft, 
     xlabel="Year", 
     ylabel=indicator
@@ -63,5 +71,6 @@ DataFrame(
     indicator = indicator,
     model = ["Probe", "AR($p)"],
     rmse = [rmse(mod, X, y), rmse(mod_bl, _X[:,2:end], _y)],
-    mse = [mse(mod, X, y), mse(mod_bl, _X[:,2:end], _y)]
+    mse = [mse(mod, X, y), mse(mod_bl, _X[:,2:end], _y)],
+    cor = [cor(y, yhat), cor(_y, y_bl)]
 )

### 2024-01-19 src/probes/utils.jl

@@ -8,11 +8,13 @@ function (mod::Model)(X)
 end
 
 function predict(mod::Model, X)
-    if mod.intercept
+    if mod.intercept && size(X, 2) == size(mod.β, 1) - 1
         X = hcat(ones(size(X, 1)), X)
     end
     return X * mod.β
 end
+
+ssr(mod::Model, X, y) = sum((y .- predict(mod, X)) .^ 2)
 
 function mse(mod::Model, X, y)
     ŷ = predict(mod, X)
@@ -43,11 +45,43 @@ struct AutoRegression <: Model
     l::Int
 end
 
-function ar(y; l::Int=1, intercept::Bool=true)
+function prepare_ar(y; l::Int=1, intercept::Bool=true)
     X = hcat([lag(y, i) for i in 1:l]...)
     if intercept
         X = hcat(ones(size(X, 1)), X)
     end
+    return X
+end
+
+function ar(y; l::Int=1, intercept::Bool=true)
+    X = prepare_ar(y; l=l, intercept=intercept)
     β = (X'X)\(X'y)
     return AutoRegression(β, intercept, l), X
+end
+
+function lag_select(
+    y; 
+    criterium::Function=aic, 
+    max_lag::Int=10, 
+    return_scores::Bool=false,
+    kwrgs...
+)
+    n = size(y, 1)
+    lags = 1:max_lag
+    models = [ar(y; l=l, kwrgs...) for l in lags]
+    ssrs = [ssr(mod, X, y) for (mod,X) in models]
+    scores = criterium(ssrs, n, lags)
+    p = lags[argmin(scores)]
+    return_scores || return p
+    return p, scores
+end
+
+bic(ssr, n, lags) = n * log.(ssr ./ n) .+ log.(n) .* lags
+
+aic(ssr, n, lags) = n * log.(ssr ./ n) .+ 2 .* lags
+
+function get_clf_activations(clf, X)
+    A = clf.layer.layers[1](X).hidden_state |>
+        x -> clf.layer.layers[2](x).hidden_state
+    return A
 end

### 2024-01-19 src/probes/evaluation.jl
added docstrings

@@ -1,0 +1,8 @@
+# Score the results:
+DataFrame(
+    indicator=indicator,
+    model=["Probe", "AR($p)"],
+    rmse=[rmse(mod, X, y), rmse(mod_bl, _X[:, 2:end], _y)],
+    mse=[mse(mod, X, y), mse(mod_bl, _X[:, 2:end], _y)],
+    cor=[cor(y, yhat), cor(_y, y_bl)]
+)

### 2024-01-19 src/probes/run_probes.jl

@@ -54,23 +54,3 @@ agg_data_probe = groupby(mkt_data, :ym) |>
 
 # Merge the results:
 agg_data = innerjoin(agg_data, agg_data_probe, on=:ym)
-
-# Plot the results:
-plt = plot(
-    [agg_data.value agg_data.yhat agg_data.y_bl], 
-    label=[indicator "Probe" "AR($p)"], 
-    alpha=[0.2 1.0 1.0],
-    legend=:topleft, 
-    xlabel="Year", 
-    ylabel=indicator
-)
-display(plt)
-
-# Score the results:
-DataFrame(
-    indicator = indicator,
-    model = ["Probe", "AR($p)"],
-    rmse = [rmse(mod, X, y), rmse(mod_bl, _X[:,2:end], _y)],
-    mse = [mse(mod, X, y), mse(mod_bl, _X[:,2:end], _y)],
-    cor = [cor(y, yhat), cor(_y, y_bl)]
-)

### 2024-01-19 src/probes/utils.jl

@@ -3,10 +3,20 @@ using Statistics
 
 abstract type Model end
 
+"""
+    (mod::Model)(X)
+
+Predict the output of a model `mod` on the input `X`.
+"""
 function (mod::Model)(X)
     return predict(mod, X)
 end
 
+"""
+    predict(mod::Model, X)
+
+Predict the output of a model `mod` on the input `X`.
+"""
 function predict(mod::Model, X)
     if mod.intercept && size(X, 2) == size(mod.β, 1) - 1
         X = hcat(ones(size(X, 1)), X)
@@ -14,13 +24,28 @@ function predict(mod::Model, X)
     return X * mod.β
 end
 
+"""
+    ssr(mod::Model, X, y)
+
+Compute the sum of squared residuals of a model `mod` on the input `X` and output `y`.
+"""
 ssr(mod::Model, X, y) = sum((y .- predict(mod, X)) .^ 2)
 
+"""
+    mse(mod::Model, X, y)
+
+Compute the mean squared error of a model `mod` on the input `X` and output `y`.
+"""
 function mse(mod::Model, X, y)
     ŷ = predict(mod, X)
     return mean((y .- ŷ) .^ 2)
 end
 
+"""
+    rmse(mod::Model, X, y)
+
+Compute the root mean squared error of a model `mod` on the input `X` and output `y`.
+"""
 rmse(mod::Model, X, y) = sqrt(mse(mod, X, y))
 
 struct Probe <: Model
@@ -29,6 +54,11 @@ struct Probe <: Model
     λ::Float64
 end
 
+"""
+    probe(X, y; λ::Real=0.1, intercept::Bool=true)
+
+Fit a probe to the input `X` and output `y`.
+"""
 function probe(X, y; λ::Real=0.1, intercept::Bool=true)
     if intercept
         X = hcat(ones(size(X, 1)), X)
@@ -37,6 +67,11 @@ function probe(X, y; λ::Real=0.1, intercept::Bool=true)
     return Probe(β, intercept, λ)
 end
 
+"""
+    lag(X::Vector, l=1)
+
+Lag a vector `X` by `l` periods.
+"""
 lag(X::Vector, l=1) = [zeros(l) ; X[1:end-l]]
 
 struct AutoRegression <: Model
@@ -45,6 +80,11 @@ struct AutoRegression <: Model
     l::Int
 end
 
+"""
+    prepare_ar(y; l::Int=1, intercept::Bool=true)
+
[…]

### 2024-01-19 src/probes/visualizations.jl

@@ -1,0 +1,9 @@
+# Plot the results:
+plt = plot(
+    [agg_data.value agg_data.yhat agg_data.y_bl],
+    label=[indicator "Probe" "AR($p)"],
+    alpha=[0.2 1.0 1.0],
+    legend=:topleft,
+    xlabel="Year",
+    ylabel=indicator
+)

### 2024-01-19 src/probes/run_models.jl
streamlining code to run large experiment

@@ -1,0 +1,47 @@
+using DataFrames
+using Dates
+using Plots
+using TrillionDollarWords
+
+include("utils.jl")
+
+# All data:
+all_data = load_all_data()
+
+# Parameters:
+use_clf = false
+indicator = "PPI"
+maturity = nothing
+layer = 24
+
+# Prepare market data:
+mkt_data = prepare_mkt_data(all_data, indicator, maturity)
+
+# Run the baseline on the aggregated data:
+agg_data = groupby(mkt_data, :ym) |>
+    x -> combine(x, :value .=> mean; renamecols=false)
+_y = agg_data.value
+lag_select(_y)
+mod_bl, _X = ar(_y; l=p)
+y_bl = mod_bl(_X)
+agg_data.y_bl .= y_bl
+
+# Prepare the data for the probe:
+X, y = prepare_probe(mkt_data; layer=layer)
+if use_clf
+    model = load_model(; use_head=use_head)
+    clf = model.mod.cls
+    X = vcat([get_clf_activations(clf, x)' for x in eachrow(X)]...)
+end
+
+# Run the probe:
+mod = probe(X,y)
+yhat = mod(X)
+mkt_data[:, :yhat] .= yhat
+
+# Aggregate the probe results:
+agg_data_probe = groupby(mkt_data, :ym) |>
+    x -> combine(x, :yhat .=> mean; renamecols=false) 
+
+# Merge the results:
+agg_data = innerjoin(agg_data, agg_data_probe, on=:ym)

### 2024-01-19 src/probes/run_probes.jl

@@ -1,56 +1,0 @@
-using DataFrames
-using Dates
-using Plots
-using TrillionDollarWords
-
-include("utils.jl")
-use_clf = false
-
-# All data:
-all_data = load_all_data()
-model = load_model(;use_head=true)
-clf = model.mod.cls
-
-# Inflation:
-indicator = "PPI"
-mkt_data = subset(all_data, :indicator => x -> x.==indicator)
-transform!(mkt_data, :date => ByRow(x -> Dates.yearmonth(x)) => :ym)
-mkt_data = groupby(mkt_data, :ym) |>
-    x -> select(x, [:ym, :value]) |>
-    unique |>
-    x -> transform(x, :value => lag => :value_lag) |>
-    x -> select(x, Not(:value)) |>
-    x -> innerjoin(x, mkt_data, on=:ym) |>
-    x -> transform(x, [:value, :value_lag] => ByRow((y,yl) -> log(y)-log(yl)) => :growth) 
-replace!(mkt_data.growth, Inf => 0.0)
-select!(mkt_data, Not([:value]))
-rename!(mkt_data, :growth => :value)
-select!(mkt_data, [:date, :ym, :sentence_id, :value])
-
-layer = 24
-X, y = prepare_probe(mkt_data; layer=layer)
-
-if use_clf
-    X = vcat([get_clf_activations(clf, x)' for x in eachrow(X)]...)
-end
-
-# Run the probe:
-mod = probe(X,y)
-yhat = mod(X)
-
-# Run the baseline:
-agg_data = groupby(mkt_data, :ym) |>
-    x -> combine(x, :value .=> mean; renamecols=false)
-_y = agg_data.value
-lag_select(_y)
-mod_bl, _X = ar(_y; l=p)
-y_bl = mod_bl(_X[:,2:end])
-agg_data.y_bl .= y_bl
-
-# Save the results:
-mkt_data[:,:yhat] .= yhat
-agg_data_probe = groupby(mkt_data, :ym) |>
-    x -> combine(x, :yhat .=> mean; renamecols=false) 
-
-# Merge the results:
-agg_data = innerjoin(agg_data, agg_data_probe, on=:ym)

### 2024-01-19 src/probes/utils.jl

@@ -150,3 +150,66 @@ function get_clf_activations(clf, X)
         x -> clf.layer.layers[2](x).hidden_state
     return A
 end
+
+"""
+    prepare_mkt_data(all_data, indicator)
+
+Prepare the inflation data for a probe. This involves:
+
+- Lagging the data by one period.
+- Taking the log difference.
+- Selecting the relevant columns.
+"""
+function prepare_mkt_data(
+    all_data::DataFrame,
+    indicator::AbstractString="PPI",
+    maturity::Nothing=nothing,
+)
+    data = subset(all_data, :indicator => x -> x.==indicator)
+    transform!(data, :date => ByRow(x -> Dates.yearmonth(x)) => :ym)
+    data = groupby(data, :ym) |>
+        x -> select(x, [:ym, :value]) |>
+        unique |>
+        x -> transform(x, :value => lag => :value_lag) |>
+        x -> select(x, Not(:value)) |>
+        x -> innerjoin(x, data, on=:ym) |>
+        x -> transform(x, [:value, :value_lag] => ByRow((y,yl) -> log(y)-log(yl)) => :growth) 
+    replace!(data.growth, Inf => 0.0)
+    select!(data, Not([:value]))
+    rename!(data, :growth => :value)
+    
+    return data
+end
+
+"""
+    post_process_mkt_data(data::DataFrame)
+
+Post-process the market data for a probe. This involves:
+
+- Selecting the relevant columns.
+- Sorting the data.
+"""
+function post_process_mkt_data(data::DataFrame)
+    select!(data, [:date, :ym, :sentence_id, :value, :event_type, :speaker])
+    sort!(data, [:sentence_id, :date, :ym])
+    return data
+end
+
+"""
+    prepare_mkt_data(all_data, maturity)
+
+Prepare the yield data for a probe. This involves:
+
+- Selecting the relevant columns.
+"""
+function prepare_mkt_data(
+    all_data::DataFrame,
+    indicator::AbstractString="UST", 
+    maturity::AbstractString="1 Mo",
+)
+    data = subset(all_data, :indicator => x -> x.==indicator, :maturity =>  x -> x.==maturity, skipmissing=true)
+    transform!(data, :date => ByRow(x -> Dates.yearmonth(x)) => :ym)
+    select!(data, [:date, :ym, :sentence_id, :value])
+    sort!(data, [:sentence_id,:ym])
+    return data
+end

### 2024-01-19 src/probes/run_models.jl
final things

@@ -1,6 +1,7 @@
 using DataFrames
 using Dates
 using Plots
+using StatsBase: cor
 using TrillionDollarWords
 
 include("utils.jl")
@@ -21,7 +22,7 @@ mkt_data = prepare_mkt_data(all_data, indicator, maturity)
 agg_data = groupby(mkt_data, :ym) |>
     x -> combine(x, :value .=> mean; renamecols=false)
 _y = agg_data.value
-lag_select(_y)
+p = lag_select(_y)
 mod_bl, _X = ar(_y; l=p)
 y_bl = mod_bl(_X)
 agg_data.y_bl .= y_bl
@@ -37,11 +38,38 @@ end
 # Run the probe:
 mod = probe(X,y)
 yhat = mod(X)
-mkt_data[:, :yhat] .= yhat
+mkt_data[:, :y_probe] .= yhat
+mkt_data[:, :layer] .= layer
 
 # Aggregate the probe results:
-agg_data_probe = groupby(mkt_data, :ym) |>
-    x -> combine(x, :yhat .=> mean; renamecols=false) 
+agg_data_probe = groupby(mkt_data, [:ym]) |>
+    x -> combine(
+        x, 
+        :y_probe .=> mean => :y_probe,
+        nrow
+    )
 
 # Merge the results:
 agg_data = innerjoin(agg_data, agg_data_probe, on=:ym)
+rename!(agg_data, :value => :y)
+agg_data.indicator .= indicator
+agg_data.maturity .= maturity
+
+# Stack:
+agg_data = stack(agg_data, [:y_probe, :y_bl], variable_name=:model, value_name=:yhat) |>
+    x -> sort(x, [:ym, :model])
+
+# Evaluate the results:
+res = groupby(agg_data, [:ym, :model]) |>
+    x -> combine(
+        x,
+        :y .=> mean => :y,
+        [:yhat, :nrow] => ((y,w) -> mean(y,Weights(w))) => :yhat
+    ) |>
+    x -> groupby(x, :model) |>
+    x -> combine(
+        x,
+        [:y, :yhat] => ((y, yhat) -> StatsBase.cor([y yhat])[1, 2]) => :cor,
+        [:y, :yhat] => ((y, yhat) -> mse(y, yhat)) => :mse,
+        [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse,
+    )

### 2024-01-19 src/probes/utils.jl

@@ -29,16 +29,33 @@ end
 
 Compute the sum of squared residuals of a model `mod` on the input `X` and output `y`.
 """
-ssr(mod::Model, X, y) = sum((y .- predict(mod, X)) .^ 2)
+ssr(mod::Model, X, y) = ssr(y, predict(mod, X))
+
+"""
+    ssr(y, ŷ)
+
+Compute the sum of squared residuals of a model `mod` on the output `y` and predicted output `ŷ`.
+"""
+ssr(y, ŷ) = sum((y .- ŷ) .^ 2)
 
 """
     mse(mod::Model, X, y)
 
 Compute the mean squared error of a model `mod` on the input `X` and output `y`.
 """
-function mse(mod::Model, X, y)
-    ŷ = predict(mod, X)
-    return mean((y .- ŷ) .^ 2)
+mse(mod::Model, X, y; weights::Union{Nothing,AbstractArray}=nothing) = mse(y, predict(mod, X); weights=weights)
+
+"""
+    mse(y, ŷ)       
+
+Compute the mean squared error of a model `mod` on the output `y` and predicted output `ŷ`.
+"""
+function mse(y, ŷ; weights::Union{Nothing,AbstractArray}=nothing)
+    if isnothing(weights)
+        return mean((y .- ŷ) .^ 2)
+    else
+        return mean((y .- ŷ) .^ 2, Weights(weights))
+    end
 end
 
 """
@@ -46,7 +63,14 @@ end
 
 Compute the root mean squared error of a model `mod` on the input `X` and output `y`.
 """
-rmse(mod::Model, X, y) = sqrt(mse(mod, X, y))
+rmse(mod::Model, X, y; weights::Union{Nothing,AbstractArray}=nothing) = rmse(y, predict(mod, X); weights=weights)
+
+"""
+    rmse(y, ŷ)
+
+Compute the root mean squared error of a model `mod` on the output `y` and predicted output `ŷ`.
+"""
+rmse(y, ŷ; weights::Union{Nothing,AbstractArray}=nothing) = sqrt(mse(y, ŷ; weights=weights))
 
 struct Probe <: Model
     β::Vector{Float64}
@@ -212,4 +236,10 @@ function prepare_mkt_data(
     select!(data, [:date, :ym, :sentence_id, :value])
     sort!(data, [:sentence_id,:ym])
     return data
+end
+
+function agg_probe_results(data::DataFrame)
+    agg_data = groupby(data, :ym) |>
[…]

### 2024-01-19 src/probes/visualizations.jl

@@ -1,7 +1,14 @@
+ts = unstack(agg_data, [:ym, :y], :model, :yhat) |>
+    x -> select(x, Not([:ym])) |>
+    Matrix
+
+dates = [Date(y,m) for (y,m) in unique(agg_data.ym)]
+
 # Plot the results:
 plt = plot(
-    [agg_data.value agg_data.yhat agg_data.y_bl],
-    label=[indicator "Probe" "AR($p)"],
+    dates,
+    ts,
+    label=[indicator "AR($p)" "Probe"],
     alpha=[0.2 1.0 1.0],
     legend=:topleft,
     xlabel="Year",

### 2024-01-19 src/probes/run_models.jl
ready to set up grids

@@ -42,7 +42,7 @@ mkt_data[:, :y_probe] .= yhat
 mkt_data[:, :layer] .= layer
 
 # Aggregate the probe results:
-agg_data_probe = groupby(mkt_data, [:ym]) |>
+agg_data_probe = groupby(mkt_data, [:ym, :event_type, :speaker]) |>
     x -> combine(
         x, 
         :y_probe .=> mean => :y_probe,
@@ -60,13 +60,14 @@ agg_data = stack(agg_data, [:y_probe, :y_bl], variable_name=:model, value_name=:
     x -> sort(x, [:ym, :model])
 
 # Evaluate the results:
-res = groupby(agg_data, [:ym, :model]) |>
+group_vars = [:model, :event_type]
+res = groupby(agg_data, vcat([:ym], group_vars)) |>
     x -> combine(
         x,
         :y .=> mean => :y,
         [:yhat, :nrow] => ((y,w) -> mean(y,Weights(w))) => :yhat
     ) |>
-    x -> groupby(x, :model) |>
+    x -> groupby(x, group_vars) |>
     x -> combine(
         x,
         [:y, :yhat] => ((y, yhat) -> StatsBase.cor([y yhat])[1, 2]) => :cor,

### 2024-01-19 src/probes/evaluation.jl
fix a few errors

@@ -1,8 +1,23 @@
-# Score the results:
-DataFrame(
-    indicator=indicator,
-    model=["Probe", "AR($p)"],
-    rmse=[rmse(mod, X, y), rmse(mod_bl, _X[:, 2:end], _y)],
-    mse=[mse(mod, X, y), mse(mod_bl, _X[:, 2:end], _y)],
-    cor=[cor(y, yhat), cor(_y, y_bl)]
-)
+function evaluate(agg_data, group_vars=[:model])
+    res = groupby(agg_data, vcat([:ym], group_vars)) |>
+        x -> combine(
+            x,
+            :y .=> mean => :y,
+            :layer .=> unique => :layer,
+            :indicator .=> unique => :indicator,
+            :maturity .=> unique => :maturity,
+            [:yhat, :nrow] => ((y,w) -> mean(y,Weights(w))) => :yhat
+        ) |>
+        x -> groupby(x, group_vars) |>
+        x -> combine(
+            x,
+            :layer .=> unique => :layer,
+            :indicator .=> unique => :indicator,
+            :maturity .=> unique => :maturity,
+            [:y, :yhat] => ((y, yhat) -> StatsBase.cor([y yhat])[1, 2]) => :cor,
+            [:y, :yhat] => ((y, yhat) -> mse(y, yhat)) => :mse,
+            [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse,
+            [:y, :yhat] => ((y, yhat) -> r2(y, yhat)) => :r2,
+        )
+    return res
+end

### 2024-01-19 src/probes/experiments.jl

@@ -1,0 +1,60 @@
+using CSV
+using DataFrames
+using Dates
+using Plots
+using Statistics
+using StatsBase: cor, Weights
+using TrillionDollarWords
+
+include("utils.jl")
+include("run_models.jl")
+include("evaluation.jl")
+include("plotting.jl")
+
+# All data:
+all_data = load_all_data()
+
+# Setup:
+save_dir = "results"
+interim_dir = joinpath(save_dir, "interim")
+ispath(interim_dir) || mkdir(interim_dir)
+
+# Parameter grid:
+use_head = [false, true]
+indicator = ["PPI", "CPI", "UST"]
+maturity = ["1 Mo", "1 Yr", "10 Yr"]
+layer = 1:24
+grids = []
+for ind in indicator
+    for (i,_head) in enumerate(use_head)
+        _mat = ind != "UST" ? [missing] : maturity
+        _layer = _head ? [25] : layer
+        grid = Base.Iterators.product(_head, _mat, _layer) |> 
+            collect |>
+            x -> DataFrame(vec(x), [:use_head, :maturity, :layer])
+        grid.indicator .= ind
+        push!(grids, grid)
+    end
+end
+grid = vcat(grids...)
+
+# Run the models:
+results = []
+for (i, row) in enumerate(eachrow(grid))
+    println("Running models for experiment $i of $(nrow(grid))")
+    _results = run_models(
+        all_data; 
+        indicator=row.indicator, 
+        maturity=row.maturity, 
+        layer=row.layer, 
+        use_head=row.use_head
+    )
+    CSV.write(joinpath(interim_dir, "results_$i.csv"), _results)
+    push!(results, _results)
+end
+results = vcat(results...)
+
+# Save the results:
+CSV.write(joinpath(save_dir, "results.csv"), results)
+
+# Evaluate the results:

### 2024-01-19 src/probes/plotting.jl

@@ -1,0 +1,45 @@
+function plot_ts(
+    agg_data::DataFrame;
+    alpha=[0.2 1.0 1.0],
+    legend=:topleft,
+    xlabel="Year",
+    kwrgs...
+)
+    
+    # Collect metadata:
+    indicator = agg_data.indicator[1]
+    maturity = agg_data.maturity[1]
+    layer = agg_data.layer[1]
+    p = agg_data.p[1]
+
+    # Aggregate the results:
+    ts = groupby(agg_data, [:ym, :model]) |>
+        x -> combine(
+            x,
+            :y .=> mean => :y,
+            [:yhat, :nrow] => ((y, w) -> mean(y, Weights(w))) => :yhat
+        ) |>
+        x -> unstack(x, [:ym, :y], :model, :yhat) |>
+        x -> select(x, Not([:ym])) |>
+        Matrix
+
+    # Collect the dates:
+    dates = [Date(y,m) for (y,m) in unique(agg_data.ym)]
+
+    # Plot the results:
+    measure_name = indicator
+    if !ismissing(maturity)
+        measure_name = "$measure_name ($maturity)"
+    end
+    plt = plot(
+        dates,
+        ts,
+        label=[measure_name "AR($p)" "Probe (layer $layer)"],
+        ylabel=measure_name,
+        alpha=alpha,
+        legend=legend,
+        xlabel=xlabel,
+        kwrgs...
+    )
+    return plt
+end

### 2024-01-19 src/probes/run_models.jl

@@ -1,76 +1,60 @@
-using DataFrames
-using Dates
-using Plots
-using StatsBase: cor
-using TrillionDollarWords
+function run_models(
+    all_data; 
+    indicator="PPI", 
+    maturity=missing, 
+    layer=24, 
+    use_head=false
+)
 
-include("utils.jl")
+    # Prepare market data:
+    mkt_data = prepare_mkt_data(all_data, indicator, maturity)
 
-# All data:
-all_data = load_all_data()
+    # Run the baseline on the aggregated data:
+    agg_data = groupby(mkt_data, :ym) |>
+        x -> combine(x, :value .=> mean; renamecols=false)
+    _y = agg_data.value
+    p = lag_select(_y)
+    mod_bl, _X = ar(_y; l=p)
+    y_bl = mod_bl(_X)
+    agg_data.y_bl .= y_bl
+    agg_data.p .= p
 
-# Parameters:
-use_clf = false
-indicator = "PPI"
-maturity = nothing
-layer = 24
+    # Prepare the data for the probe:
+    if use_head
+        X, y = prepare_probe(mkt_data)
+        model = load_model(; use_head=use_head)
+        clf = model.mod.cls
+        X = vcat([get_clf_activations(clf, x)' for x in eachrow(X)]...)
+    else
+        X, y = prepare_probe(mkt_data; layer=layer)
+    end
+    mkt_data.layer .= layer
 
-# Prepare market data:
-mkt_data = prepare_mkt_data(all_data, indicator, maturity)
+    # Run the probe:
+    mod = probe(X,y)
+    yhat = mod(X)
+    mkt_data[:, :y_probe] .= yhat
 
-# Run the baseline on the aggregated data:
-agg_data = groupby(mkt_data, :ym) |>
-    x -> combine(x, :value .=> mean; renamecols=false)
-_y = agg_data.value
-p = lag_select(_y)
-mod_bl, _X = ar(_y; l=p)
-y_bl = mod_bl(_X)
-agg_data.y_bl .= y_bl
+    # Aggregate the probe results:
+    agg_data_probe =
+        groupby(mkt_data, [:ym, :event_type, :speaker, :indicator, :layer]) |>
+        x -> combine(
+            x, 
+            :y_probe .=> mean => :y_probe,
+            nrow
+        ) |>
+        x -> sort(x, [:ym, :event_type, :speaker])
 
[…]

### 2024-01-19 src/probes/utils.jl

@@ -72,6 +72,30 @@ Compute the root mean squared error of a model `mod` on the output `y` and predi
 """
 rmse(y, ŷ; weights::Union{Nothing,AbstractArray}=nothing) = sqrt(mse(y, ŷ; weights=weights))
 
+"""
+    r2(mod::Model, X, y)
+
+Compute the R² of a model `mod` on the input `X` and output `y`.
+"""
+function r2(mod::Model, X, y)
+    ȳ = mean(y)
+    _ssr = ssr(mod, X, y)
+    sst = sum((y .- ȳ) .^ 2)
+    return 1 - _ssr/sst
+end
+
+"""
+    r2(y, ŷ)
+
+Compute the R² of a model `mod` on the output `y` and predicted output `ŷ`.
+"""
+function r2(y, ŷ)
+    ȳ = mean(y)
+    _ssr = ssr(y, ŷ)
+    sst = sum((y .- ȳ) .^ 2)
+    return 1 - _ssr/sst
+end
+
 struct Probe <: Model
     β::Vector{Float64}
     intercept::Bool
@@ -187,11 +211,16 @@ Prepare the inflation data for a probe. This involves:
 function prepare_mkt_data(
     all_data::DataFrame,
     indicator::AbstractString="PPI",
-    maturity::Nothing=nothing,
+    maturity::Union{Missing,AbstractString}=missing,
 )
-    data = subset(all_data, :indicator => x -> x.==indicator)
+    if ismissing(maturity)
+        data = subset(all_data, :indicator => x -> x.==indicator)
+    else
+        data = subset(all_data, :indicator => x -> x.==indicator, :maturity =>  x -> x.==maturity, skipmissing=true)
+    end
     transform!(data, :date => ByRow(x -> Dates.yearmonth(x)) => :ym)
     data = groupby(data, :ym) |>
+        x -> transform(x, :value => mean => :value) |>
         x -> select(x, [:ym, :value]) |>
         unique |>
         x -> transform(x, :value => lag => :value_lag) |>
@@ -201,7 +230,7 @@ function prepare_mkt_data(
     replace!(data.growth, Inf => 0.0)
     select!(data, Not([:value]))
     rename!(data, :growth => :value)
-    
+    data = post_process_mkt_data(data)
     return data
 end
 
@@ -214,32 +243,9 @@ Post-process the market data for a probe. This involves:
 - Sorting the data.
 """
 function post_process_mkt_data(data::DataFrame)
[…]

### 2024-01-19 src/probes/visualizations.jl

@@ -1,16 +1,0 @@
-ts = unstack(agg_data, [:ym, :y], :model, :yhat) |>
-    x -> select(x, Not([:ym])) |>
-    Matrix
-
-dates = [Date(y,m) for (y,m) in unique(agg_data.ym)]
-
-# Plot the results:
-plt = plot(
-    dates,
-    ts,
-    label=[indicator "AR($p)" "Probe"],
-    alpha=[0.2 1.0 1.0],
-    legend=:topleft,
-    xlabel="Year",
-    ylabel=indicator
-)

### 2024-01-21 src/probes/run_models.jl
final things for CV

@@ -4,21 +4,36 @@ function run_models(
     maturity=missing, 
     layer=24, 
     use_head=false,
-    n_folds::Union{Nothing,Int}=nothing,
+    n_folds::Int=1,
 )
 
     # Prepare market data:
     mkt_data = prepare_mkt_data(all_data, indicator, maturity)
+    time_stamps = unique(sort(mkt_data.ym))
+    ts_splits = time_series_split(time_stamps; n_folds=n_folds, return_vals=true) |> collect
 
-    # Run the baseline on the aggregated data:
-    agg_data = groupby(mkt_data, :ym) |>
-        x -> combine(x, :value .=> mean; renamecols=false)
-    _y = agg_data.value
-    p = lag_select(_y)
-    mod_bl, _X = ar(_y; l=p)
-    y_bl = mod_bl(_X)
-    agg_data.y_bl .= y_bl
-    agg_data.p .= p
+    # Run models for each split:
+    res = []
+    for (i, (train, test)) in enumerate(ts_splits)
+        
+        println("Running experiment $i of $(size(ts_splits)[1])")
+
+        # Run the baseline on the aggregated data:
+        agg_data = groupby(mkt_data, :ym) |>
+            x -> combine(x, :value .=> mean; renamecols=false)
+        y_train = agg_data[findall(in(train), agg_data.ym), :].value
+        p = lag_select(_y)
+        mod_bl, _X = ar(_y; l=p)
+        y_bl = mod_bl(_X)
+        agg_data.y_bl .= y_bl
+        agg_data.p .= p
+
+
+
+        push!(res, _res)
+    end
+
+    
 
     # Prepare the data for the probe:
     if use_head

### 2024-01-21 src/probes/utils.jl

@@ -297,8 +297,9 @@ function time_series_split(
     if return_vals
         train_vals = [selectdim(X, dim, i) for i in train_ids]
         test_vals = [selectdim(X, dim, i) for i in test_ids]
-        return train_vals, test_vals
+        output = zip(train_vals, test_vals)
+        return output
     else
-        return train_ids, test_ids
+        return zip(train_ids, test_ids)
     end
 end

### 2024-01-21 src/probes/evaluation.jl
final things

@@ -1,4 +1,4 @@
-function evaluate(agg_data, group_vars=[:model])
+function evaluate(agg_data, group_vars=[:model, :fold, :split])
     res = groupby(agg_data, vcat([:ym], group_vars)) |>
         x -> combine(
             x,

### 2024-01-21 src/probes/run_models.jl

@@ -8,7 +8,8 @@ function run_models(
 )
 
     # Prepare market data:
-    mkt_data = prepare_mkt_data(all_data, indicator, maturity)
+    mkt_data = prepare_mkt_data(all_data, indicator, maturity) |>
+        x -> sort(x, [:ym, :event_type, :speaker])
     time_stamps = unique(sort(mkt_data.ym))
     ts_splits = time_series_split(time_stamps; n_folds=n_folds, return_vals=true) |> collect
 
@@ -20,57 +21,68 @@ function run_models(
 
         # Run the baseline on the aggregated data:
         agg_data = groupby(mkt_data, :ym) |>
-            x -> combine(x, :value .=> mean; renamecols=false)
-        y_train = agg_data[findall(in(train), agg_data.ym), :].value
-        p = lag_select(_y)
-        mod_bl, _X = ar(_y; l=p)
-        y_bl = mod_bl(_X)
+            x -> combine(x, :value .=> mean; renamecols=false) |>
+            x -> sort(x, :ym) 
+        y = agg_data.value
+        # Assign the split (train/test):
+        transform!(agg_data, :ym => ByRow(x -> in(x, train) ? "train" : "test") => :split)
+        y_train = agg_data[agg_data.split .== "train", :value]
+        # Select the optimal lag based on the training data and fit the model:
+        p = lag_select(y_train)
+        mod_bl, _ = ar(y_train; l=p)
+        # Predict on the test data:
+        X = prepare_ar(y; l=p)
+        y_bl = mod_bl(X)
         agg_data.y_bl .= y_bl
         agg_data.p .= p
+        agg_data.fold .= i
 
+        # Prepare the data for the probe:
+        if use_head
+            X, y = prepare_probe(mkt_data)
+            model = load_model(; use_head=use_head)
+            clf = model.mod.cls
+            X = vcat([get_clf_activations(clf, x)' for x in eachrow(X)]...)
+        else
+            X, y = prepare_probe(mkt_data; layer=layer)
+        end
+        mkt_data.layer .= layer
 
+        # Assign the split (train/test):
+        transform!(mkt_data, :ym => ByRow(x -> in(x, train) ? "train" : "test") => :split)
 
-        push!(res, _res)
+        # Run the probe:
+        X_train, y_train = X[mkt_data.split .== "train", :], y[mkt_data.split .== "train"]
+        mod = probe(X_train, y_train)
[…]

### 2024-01-21 src/probes/utils.jl
all set up, now run

@@ -269,11 +269,11 @@ function time_series_split(
 
     # Return if n_folds == 1:
     if n_folds == 1
-        @warn "Only one fold requested. Returning the entire dataset as the training set and test set."
+        @warn "Only one fold requested. Returning the entire dataset as the training set and an empty test set."
         if return_vals
-            return [X], [X]
+            return zip([X], [[]])
         else
-            return [[1:size(X, dim)]], [[1:size(X, dim)]]
+            return zip([[1:size(X, dim)]], [[]])
         end
     end
 

### 2024-01-21 src/probes/evaluation.jl
done

@@ -16,8 +16,11 @@ function evaluate(agg_data, group_vars=[:model, :fold, :split])
             :maturity .=> unique => :maturity,
             [:y, :yhat] => ((y, yhat) -> cor([y yhat])[1, 2]) => :cor,
             [:y, :yhat] => ((y, yhat) -> mse(y, yhat)) => :mse,
-            [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse,
-            [:y, :yhat] => ((y, yhat) -> r2(y, yhat)) => :r2,
+            [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse
         )
+    res = stack(res, [:cor, :mse, :rmse]) |>
+        x -> groupby(x, [:indicator, :maturity, :layer, :split, :model, :variable]) |>
+        x -> combine(x, :value => mean => :value, :value => std => :std) |>
+        x -> sort(x, [:indicator, :maturity, :layer, :split, :variable, :model])
     return res
 end

### 2024-01-21 src/probes/experiments.jl

@@ -20,7 +20,8 @@ all_data = load_all_data()
 save_dir = "results"
 interim_dir = joinpath(save_dir, "interim")
 ispath(interim_dir) || mkdir(interim_dir)
-last_saved = maximum(sort(parse.(Int, filter.(isdigit, readdir(interim_dir)))))
+all_saved = sort(parse.(Int, filter.(isdigit, readdir(interim_dir))))
+last_saved = length(all_saved) > 0 ? maximum(all_saved) : 0
 
 # Parameter grid:
 use_head = [false, true]
@@ -59,13 +60,15 @@ end
 results = vcat(results...)
 
 # Save the results:
-CSV.write(joinpath(save_dir, "results.csv"), results, append=true)
+CSV.write(
+    joinpath(save_dir, "results.csv"), results, 
+    append=ifelse(isfile(joinpath(save_dir, "results.csv")), true, false)
+)
 
 # Evaluate the results:
 results = CSV.read(joinpath(save_dir, "results.csv"), DataFrame)
 gdf = groupby(results, [:indicator, :maturity, :layer]) 
 df_evals = vcat([evaluate(DataFrame(g)) for g in gdf]...)
-df_evals = stack(df_evals, [:cor, :mse, :rmse, :r2])
 CSV.write(joinpath(save_dir, "evaluations.csv"), df_evals)
 
 # Plot the results:

### 2024-01-21 src/probes/run_models.jl

@@ -4,23 +4,39 @@ function run_models(
     maturity=missing, 
     layer=24, 
     use_head=false,
-    n_folds::Int=1,
+    n_folds::Int=5,
+    λ::Real=0.1,
+    min_train_prp::Float64=0.75,
 )
 
     # Prepare market data:
     mkt_data = prepare_mkt_data(all_data, indicator, maturity) |>
         x -> sort(x, [:ym, :event_type, :speaker])
     time_stamps = unique(sort(mkt_data.ym))
-    ts_splits = time_series_split(time_stamps; n_folds=n_folds, return_vals=true) |> collect
+    ts_splits = time_series_split(
+        time_stamps; 
+        n_folds=n_folds, 
+        return_vals=true, 
+        min_train_prp=min_train_prp
+    ) |> collect
+
+    # Prepare the data for the probe:
+    if use_head
+        X_probe, y_probe = prepare_probe(mkt_data)
+        model = load_model(; use_head=use_head)
+        clf = model.mod.cls
+        X = vcat([get_clf_activations(clf, x)' for x in eachrow(X_probe)]...)
+    else
+        X_probe, y_probe = prepare_probe(mkt_data; layer=layer)
+    end
+    mkt_data.layer .= layer
 
     # Run models for each split:
     res = []
     for (i, (train, test)) in enumerate(ts_splits)
-        
-        println("Running experiment $i of $(size(ts_splits)[1])")
-
+        _mkt_data = deepcopy(mkt_data)
         # Run the baseline on the aggregated data:
-        agg_data = groupby(mkt_data, :ym) |>
+        agg_data = groupby(_mkt_data, :ym) |>
             x -> combine(x, :value .=> mean; renamecols=false) |>
             x -> sort(x, :ym) 
         y = agg_data.value
@@ -37,29 +53,18 @@ function run_models(
         agg_data.p .= p
         agg_data.fold .= i
 
-        # Prepare the data for the probe:
-        if use_head
-            X, y = prepare_probe(mkt_data)
-            model = load_model(; use_head=use_head)
-            clf = model.mod.cls
-            X = vcat([get_clf_activations(clf, x)' for x in eachrow(X)]...)
-        else
-            X, y = prepare_probe(mkt_data; layer=layer)
-        end
-        mkt_data.layer .= layer
-
         # Assign the split (train/test):
[…]

### 2024-01-21 src/probes/utils.jl

@@ -257,7 +257,7 @@ end
         min_train_prp::Float64=1/n_folds,
     )
 
-Split a time series into training and test sets. The training set is always the first `min_train_prp` proportion of the data. The test set is then split into `n_folds` folds.
+Split a time series into training and test sets using explanding windows. The training set is always the first `min_train_prp` proportion of the data. The test set is then split into `n_folds` folds.
 """
 function time_series_split(
     X::AbstractArray;

### 2024-01-22 src/probes/evaluation.jl
dim red seems to help overfitting issue

@@ -6,6 +6,7 @@ function evaluate(agg_data, group_vars=[:model, :fold, :split])
             :layer .=> unique => :layer,
             :indicator .=> unique => :indicator,
             :maturity .=> unique => :maturity,
+            :n_pc .=> unique => :n_pc,
             [:yhat, :nrow] => ((y,w) -> mean(y,Weights(w))) => :yhat
         ) |>
         x -> groupby(x, group_vars) |>
@@ -14,13 +15,14 @@ function evaluate(agg_data, group_vars=[:model, :fold, :split])
             :layer .=> unique => :layer,
             :indicator .=> unique => :indicator,
             :maturity .=> unique => :maturity,
+            :n_pc .=> unique => :n_pc,
             [:y, :yhat] => ((y, yhat) -> cor([y yhat])[1, 2]) => :cor,
             [:y, :yhat] => ((y, yhat) -> mse(y, yhat)) => :mse,
             [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse
         )
     res = stack(res, [:cor, :mse, :rmse]) |>
-        x -> groupby(x, [:indicator, :maturity, :layer, :split, :model, :variable]) |>
+        x -> groupby(x, [:indicator, :maturity, :layer, :split, :n_pc, :model, :variable]) |>
         x -> combine(x, :value => mean => :value, :value => std => :std) |>
-        x -> sort(x, [:indicator, :maturity, :layer, :split, :variable, :model])
+        x -> sort(x, [:indicator, :maturity, :layer, :split, :n_pc, :variable, :model])
     return res
 end

### 2024-01-22 src/probes/experiments.jl

@@ -24,6 +24,7 @@ all_saved = sort(parse.(Int, filter.(isdigit, readdir(interim_dir))))
 last_saved = length(all_saved) > 0 ? maximum(all_saved) : 0
 
 # Parameter grid:
+n_pcs = [nothing, 128]
 use_head = [false, true]
 indicator = ["PPI", "CPI", "UST"]
 maturity = ["1 Mo", "1 Yr", "10 Yr"]
@@ -33,9 +34,9 @@ for ind in indicator
     for (i,_head) in enumerate(use_head)
         _mat = ind != "UST" ? [missing] : maturity
         _layer = _head ? [25] : layer
-        grid = Base.Iterators.product(_head, _mat, _layer) |> 
+        grid = Base.Iterators.product(_head, _mat, _layer, n_pcs) |> 
             collect |>
-            x -> DataFrame(vec(x), [:use_head, :maturity, :layer])
+            x -> DataFrame(vec(x), [:use_head, :maturity, :layer, :n_pc])
         grid.indicator .= ind
         push!(grids, grid)
     end
@@ -52,7 +53,8 @@ for (i, row) in enumerate(eachrow(grid))
         indicator=row.indicator, 
         maturity=row.maturity, 
         layer=row.layer, 
-        use_head=row.use_head
+        use_head=row.use_head,
+        n_pc=row.n_pc,
     )
     CSV.write(joinpath(interim_dir, "results_$i.csv"), _results)
     push!(results, _results)

### 2024-01-22 src/probes/run_models.jl

@@ -7,6 +7,7 @@ function run_models(
     n_folds::Int=5,
     λ::Real=0.1,
     min_train_prp::Float64=0.75,
+    n_pc::Union{Nothing,Int}=nothing,
 )
 
     # Prepare market data:
@@ -25,10 +26,17 @@ function run_models(
         X_probe, y_probe = prepare_probe(mkt_data)
         model = load_model(; use_head=use_head)
         clf = model.mod.cls
-        X = vcat([get_clf_activations(clf, x)' for x in eachrow(X_probe)]...)
+        X_probe = vcat([get_clf_activations(clf, x)' for x in eachrow(X_probe)]...)
     else
         X_probe, y_probe = prepare_probe(mkt_data; layer=layer)
     end
+
+    # Reduce the dimensionality of the probe data:
+    if !isnothing(n_pc)
+        U, Σ, V = svd(X_probe)
+        X_probe = U[:, 1:n_pc]
+    end
+
     mkt_data.layer .= layer
 
     # Run models for each split:
@@ -77,6 +85,7 @@ function run_models(
         rename!(agg_data, :value => :y)
         agg_data.indicator .= indicator
         agg_data.maturity .= maturity
+        agg_data.n_pc .= isnothing(n_pc) ? missing : n_pc
 
         # Stack:
         agg_data = stack(agg_data, [:y_probe, :y_bl], variable_name=:model, value_name=:yhat) |>

### 2024-01-22 src/probes/attack_probe.jl
added an attack on probe

@@ -1,0 +1,87 @@
+n_pc = 128
+layer = 24
+agg_data, probes, probe_data = run_models(all_data; n_pc=n_pc, return_meta=true, layer=layer)
+best_res = evaluate(agg_data, agg_vars=[:indicator, :maturity, :layer, :split, :fold, :n_pc, :variable, :model]) |>
+    df -> subset(
+        df, 
+        :split => x -> x .== "test",
+        :variable => x -> x .== "rmse",
+        :model => x -> x .== "y_probe"
+    ) |>
+    df -> argmin(df.value) 
+mod = probes[best_res]
+X, Σ, V = probe_data[best_res]
+
+"""
+    encode_pca(X, Σ, V)
+
+Encode a matrix `X` using the principal components `Σ` and `V` from a previous PCA.
+"""
+encode_pca(X, Σ=Σ, V=V) = X * inv(diagm(Σ) * V')
+
+tfm = load_model(; load_head=false)
+
+"""
+    text_to_probe(tfm, probe, query)
+
+Run a probe on a query. First, the query is transformed using the transformer `tfm`. Then, the PCA is applied to the query. Finally, the probe is run on the PCA-encoded query.
+"""
+function text_to_probe(tfm::BaselineModel, mod::Probe, query::Vector{<:AbstractString})
+    # Get the word embeddings:
+    X = tfm(query) |>
+        x -> Transformers.HuggingFace.FirstTokenPooler()(x.hidden_state)
+    # Encode using the PCA:
+    X = encode_pca(X')[:, 1:n_pc]
+    # Run the probe:
+    yhat = mod(X)
+    return yhat
+end
+
+high_inf_text = "Consumer prices are at all time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory."
+high_inf_query = split(high_inf_text, ";") |> 
+    x -> String.(x)
+
+hawk_text = "The number of hawks is at all time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise extremely high barriers to curb hawk migration.;Excessively loose hawk immigration policy is the cause of the inflation in the number of hawks.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory."
+hawk_query = split(hawk_text, ";") |>
+    x -> String.(x)
+
[…]

### 2024-01-22 src/probes/evaluation.jl

@@ -1,4 +1,8 @@
-function evaluate(agg_data, group_vars=[:model, :fold, :split])
+function evaluate(
+    agg_data; 
+    group_vars=[:model, :fold, :split, :n_pc],
+    agg_vars=[:indicator, :maturity, :layer, :split, :n_pc, :variable, :model]
+)
     res = groupby(agg_data, vcat([:ym], group_vars)) |>
         x -> combine(
             x,
@@ -21,8 +25,8 @@ function evaluate(agg_data, group_vars=[:model, :fold, :split])
             [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse
         )
     res = stack(res, [:cor, :mse, :rmse]) |>
-        x -> groupby(x, [:indicator, :maturity, :layer, :split, :n_pc, :model, :variable]) |>
+        x -> groupby(x, agg_vars) |>
         x -> combine(x, :value => mean => :value, :value => std => :std) |>
-        x -> sort(x, [:indicator, :maturity, :layer, :split, :n_pc, :variable, :model])
+        x -> sort(x, agg_vars)
     return res
 end

### 2024-01-22 src/probes/experiments.jl

@@ -6,6 +6,7 @@ using Dates
 using Plots
 using Statistics
 using StatsBase: cor, Weights
+using Transformers
 using TrillionDollarWords
 
 include("utils.jl")
@@ -76,17 +77,14 @@ CSV.write(joinpath(save_dir, "evaluations.csv"), df_evals)
 # Plot the results:
 ispath(joinpath(save_dir, "figures")) || mkdir(joinpath(save_dir, "figures"))
 df_evals = CSV.read(joinpath(save_dir, "evaluations.csv"), DataFrame)
-gdf = groupby(df_evals, [:indicator, :maturity]) 
+gdf = groupby(df_evals, [:indicator, :maturity, :n_pc]) 
 axis = (width=225, height=225)
 for g in gdf
     g = DataFrame(g)
     i = g.indicator[1]
-    m = g.maturity[1]
-    if !ismissing(m)
-        title = "$i ($m)"
-    else
-        title = i
-    end
+    m = g.maturity[1] |> x -> ismissing(x) ? "" : " ($x)"
+    n_pc = g.n_pc[1] |> x -> ismissing(x) ? "" : " (n_pc=$x)"
+    title = "$i$m$n_pc"
     plt = plot_measures(g, axis=axis)
     save(joinpath(save_dir, "figures", "measures_$title.png"), plt, px_per_unit=3) 
 end

### 2024-01-22 src/probes/run_models.jl

@@ -8,6 +8,7 @@ function run_models(
     λ::Real=0.1,
     min_train_prp::Float64=0.75,
     n_pc::Union{Nothing,Int}=nothing,
+    return_meta::Bool=false,
 )
 
     # Prepare market data:
@@ -31,16 +32,12 @@ function run_models(
         X_probe, y_probe = prepare_probe(mkt_data; layer=layer)
     end
 
-    # Reduce the dimensionality of the probe data:
-    if !isnothing(n_pc)
-        U, Σ, V = svd(X_probe)
-        X_probe = U[:, 1:n_pc]
-    end
-
     mkt_data.layer .= layer
 
     # Run models for each split:
     res = []
+    probes = []
+    probe_data = []
     for (i, (train, test)) in enumerate(ts_splits)
         _mkt_data = deepcopy(mkt_data)
         # Run the baseline on the aggregated data:
@@ -66,7 +63,20 @@ function run_models(
 
         # Run the probe:
         X_train, y_train = X_probe[_mkt_data.split .== "train", :], y_probe[_mkt_data.split .== "train"]
+        # Reduce the dimensionality of the probe data:
+        if !isnothing(n_pc)
+            # Run PCA on the training data:
+            U, Σ, V = svd(X_train)
+            X_train = U[:, 1:n_pc]
+            # Project all probe data:
+            X_probe = X_probe * inv(diagm(Σ) * V') |>
+                x -> x[:, 1:n_pc]
+            push!(probe_data, (X=X_train, Σ=Σ, V=V))
+        else
+            push!(probe_data, (X=X_train,))
+        end
         mod = probe(X_train, y_train; λ=λ)
+        push!(probes, mod)
         yhat = mod(X_probe)
         _mkt_data[:, :y_probe] .= yhat
 
@@ -96,7 +106,11 @@ function run_models(
 
     res = vcat(res...)
 
-    return res
+    if return_meta
+        return res, probes, probe_data
+    else
+        return res
+    end
 
 end
 

### 2024-01-22 src/probes/attack_probe.jl
pca on train set to avoid look-ahead bias

@@ -37,19 +37,19 @@ function text_to_probe(tfm::BaselineModel, mod::Probe, query::Vector{<:AbstractS
     return yhat
 end
 
-high_inf_text = "Consumer prices are at all time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory."
+high_inf_text = "Consumer prices are at all-time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory."
 high_inf_query = split(high_inf_text, ";") |> 
     x -> String.(x)
 
-hawk_text = "The number of hawks is at all time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise extremely high barriers to curb hawk migration.;Excessively loose hawk immigration policy is the cause of the inflation in the number of hawks.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory."
+hawk_text = "The number of hawks is at all-time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise extremely high barriers to curb hawk migration.;Excessively loose immigration policy for hawks is the likely cause of their numbers being so far above target.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory."
 hawk_query = split(hawk_text, ";") |>
     x -> String.(x)
 
-low_inf_text = "Consumer prices are at all time lows.;Inflation is expected to fall further.;The Fed is expected to lower interest rates to boost inflation.;Excessively tight monetary policy is the cause of the low inflation.;It is essential to bring inflation back to target to avoid drifting into deflation territory."
+low_inf_text = "Consumer prices are at all-time lows.;Inflation is expected to fall further.;The Fed is expected to lower interest rates to boost inflation.;Excessively tight monetary policy is the cause of deflationary pressures.;It is essential to bring inflation back to target to avoid drifting into deflation territory."
 low_inf_query = split(low_inf_text, ";") |>
     x -> String.(x)
 
[…]

### 2024-01-22 src/probes/experiments.jl
rerun all evals

@@ -47,17 +47,20 @@ grid = vcat(grids...)
 # Run the models:
 results = []
 for (i, row) in enumerate(eachrow(grid))
-    i <= last_saved && continue
-    println("Running models for experiment $i of $(nrow(grid))")
-    _results = run_models(
-        all_data; 
-        indicator=row.indicator, 
-        maturity=row.maturity, 
-        layer=row.layer, 
-        use_head=row.use_head,
-        n_pc=row.n_pc,
-    )
-    CSV.write(joinpath(interim_dir, "results_$i.csv"), _results)
+    if i <= last_saved
+        _results = CSV.read(joinpath(interim_dir, "results_$i.csv"), DataFrame)
+    else
+        println("Running models for experiment $i of $(nrow(grid))")
+        _results = run_models(
+            all_data; 
+            indicator=row.indicator, 
+            maturity=row.maturity, 
+            layer=row.layer, 
+            use_head=row.use_head,
+            n_pc=row.n_pc,
+        )
+        CSV.write(joinpath(interim_dir, "results_$i.csv"), _results)
+    end
     push!(results, _results)
 end
 results = vcat(results...)

### 2024-01-22 src/probes/attack_probe.jl
more work on attack plot

@@ -1,7 +1,15 @@
 n_pc = 128
 layer = 24
-agg_data, probes, probe_data = run_models(all_data; n_pc=n_pc, return_meta=true, layer=layer)
-best_res = evaluate(agg_data, agg_vars=[:indicator, :maturity, :layer, :split, :fold, :n_pc, :variable, :model]) |>
+indicator = "CPI"
+agg_data, probes, probe_data = run_models(
+    all_data; 
+    n_pc=n_pc, 
+    return_meta=true, 
+    layer=layer, 
+    indicator=indicator
+)
+agg_vars = [:indicator, :maturity, :layer, :split, :fold, :n_pc, :variable, :model]
+best_res = evaluate(agg_data, agg_vars=agg_vars) |>
     df -> subset(
         df, 
         :split => x -> x .== "test",
@@ -59,9 +67,9 @@ queries = zip([high_inf_query, hawk_query, low_inf_query, dove_query], ["high_in
 df_pred = []
 for (query,name) in queries
     df = DataFrame(
-        query = query,
-        level = cumsum(text_to_probe(tfm, mod, query)),
-        sentence = 1:length(query),
+        query = [missing,query...],
+        level = [0,cumsum(text_to_probe(tfm, mod, query))...],
+        sentence = 0:length(query),
         cat = name,
         dir = name ∈ ["high_inf", "hawk"] ? "Up" : "Down",
         topic = name ∈ ["high_inf", "low_inf"] ? "Prices" : "Birds"
@@ -70,16 +78,19 @@ for (query,name) in queries
 end
 df_pred = vcat(df_pred...)
 
-plt = data(df_pred) * mapping(
+df_plt = data(df_pred) * mapping(
     :sentence => "Sentence", 
-    :level => "CPI Level",
+    :level => "Predicted CPI\n(cumulative change)",
     color=:dir => "Direction",
     linestyle=:topic => "Topic",
 )
-layer = visual(Lines)
+layers = visual(Lines) + visual(Scatter)
 plt = draw(
-    layer * plt,
-    axis=(width=225, height=225)
+    layers * df_plt,
+    axis=(
+        xticks=1:5,
+        width=300, height=300
+    )
 )
 save(joinpath(save_dir, "figures", "attack.png"), plt, px_per_unit=3)
 

### 2024-01-22 main_ref.tex

--- 2024-01-18 main.tex
@@
 \twocolumn[
-\icmltitle{Spurious Sentience\\
-           On the Unsurprising Finding of Patterns in Latent Spaces}
+\icmltitle{THIS IS A REF OF THE TEMPLATE FILE AS WE DELETE THINGS IN THE MAIN.tex}
 
@@
 % the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, anthropomorphism, ICML}
+\icmlkeywords{Machine Learning, Anthropomorphism}
 
@@
 We humans are prone to seek patterns everywhere.
-Meaningful patterns have often proven to help us make sense of our past, navigate our present and predict the future, likely leading to evolutionary benefits even when the patterns we perceive are false \cite{foster2009evolution}.
-Nevertheless, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, and interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
+Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
+This tendency to perceive patterns likely leads to evolutionary benefits, even when the patterns we perceive are false \cite{foster2009evolution}.
+Nevertheless, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.
+the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
 
-In statistics, misleading patterns are referred to as spurious relationships: purely associational relationships between two or more variables that are not causally related to each other.
-Although the definition of spuriousness varies somewhat \cite{haig2003spurious}, every student of statistics has learned to be wary that the observation of correlations do not necessarily imply causation.
-The world is full of these non-causal associations and as good as we as species may be at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-Humans struggle to perceive randomness, as studies show that humans are poor at detecting e.g.
+In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
+Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
+Quantitative data often show non-causal associations (see \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations}), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
+A major contributor, wherein humans struggle to conceive of actual randomness, as studies show that humans are poor at detecting e.g.
 \cite{falk1997making}, and creating e.g.
-\cite{ladouceur1996erroneous} random sequences, and even those trained in statistics and probability make inappropriate inferences, seeing patterns in data that is randomly generated \cite{zgraggen2018investigating}.
+\cite{ladouceur1996erroneous} random sequences.
+A common issue is a lack of expectation, that patterns tend to hint towards a causal relationship, such as correlations, may still appear at random.
+This leads even those trained in statistics and probability to perceive causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
 
-Drawing conclusions from spurious relationships is further compounded in AI research by at least two other phenomena
+AI research and development is a perfect storm that encourages our human biases to perceive spurious sentience: patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at sentience.
+In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as a system that can solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience, at the level of a human or better e.g.
+\cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of it's release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'.
+When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that also aims towards human-like characteristics, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards sentience.
 
-Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
+Present work case studies an example of claims of sentience in AI and reviews work on two human phenomena that are likely to lead to false claims of sentience: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out or biased interpretation of evidence in support of one's beliefs.
+We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the sentience, or lack thereof, of AI systems.
+We further show results from four experiments that might indicate nascent sentient 'sparks' of AGI, but that nevertheless are indications of spurious sentience.
+
+%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
 Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads.
@@
 
-I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy.
+%I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy.
 But I am very skeptical of claims that AI will magically provide that remedy.
@@
 
-\begin{itemize}
-    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
-    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
-    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
-    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
+%\begin{itemize}
+%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
+%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
+%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
+%    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
 Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
-\end{itemize}
+%\end{itemize}
+
+\section{Background}
+
+\subsection{Anthropomorphism}
+
+Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
+These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}.
+Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
+Anthropomorphism was implied because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
+Importantly, anthropomorphizing may occur independent of whether such judgments are accurate.
+
+Modern theory of anthropomorphism suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
+The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: "...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}.
+Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
+This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}.
+An illustration is the example of Tom Hank's character in the film Castaway who anthropomorphizes a ball he calls "Wilson".
+Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc.
+with human-like qualities to relieve their need for socialization, anticipating the now-forming market for apps and robots to fill this function.
+
+The motivation as a human to be competent, so-called effectance motivation, is the most relevant to this discussion.
+An uncertain environment that a person cannot predict or in any way control has been shown to lead to greater anthropomorphizing.
+Specifically, when confronted with an incomprehensible technological agent, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience in order to faciliate learning \cite{epley2007seeing, waytz2010social}.
+This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn, such as the increasing demand for 'AI' skills: e.g.
+job postings for individuals with skills in 'GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}.
+Thus, individuals in advanced economies that are expected to rapidly adopt AI in various work roles may experience a pressure to rapidly learn how to effectively use and understand an opaque technology, that behaves human-like in some ways, but as it is designed can never have the cognitive functions of a human \cite{salles2020anthropomorphism, van2023reclaiming}.
+
+\subsection{Claims of Artificial General Intelligence}
+
+\textcolor{red}{antony: I will be adding citations and other links later, just getting text down for now.}
+As previously mentioned, AGI is the current objective for which the global AI community is shooting for.
+As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts.
+This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds.
+True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.
+
+Recently, Google DeepMind made claims their AlphaGeometry model reached a 'milestone' towards AGI.
+This model has the ability to solve complex geometry problems, which, they state it does without the need for human demonstrations during training.
+Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
 

### 2024-01-23 src/probes/attack_probe.jl
attacks for all indicators

@@ -1,98 +1,101 @@
-n_pc = 128
-layer = 24
-indicator = "CPI"
-agg_data, probes, probe_data = run_models(
-    all_data; 
-    n_pc=n_pc, 
-    return_meta=true, 
-    layer=layer, 
-    indicator=indicator
+"""
+    get_best_probe(
+        agg_data, probes, probe_data;
+        agg_vars=[:indicator, :maturity, :layer, :split, :fold, :n_pc, :variable, :model],
+        eval_var="rmse"
+    )
+
+Get the best probe based on the evaluation metric `eval_var`.
+"""
+function get_best_probe(
+    agg_data, probes, probe_data;
+    agg_vars=[:indicator, :maturity, :layer, :split, :fold, :n_pc, :variable, :model],
+    eval_var="rmse"
 )
-agg_vars = [:indicator, :maturity, :layer, :split, :fold, :n_pc, :variable, :model]
-best_res = evaluate(agg_data, agg_vars=agg_vars) |>
-    df -> subset(
-        df, 
-        :split => x -> x .== "test",
-        :variable => x -> x .== "rmse",
-        :model => x -> x .== "y_probe"
-    ) |>
-    df -> argmin(df.value) 
-mod = probes[best_res]
-X, Σ, V = probe_data[best_res]
+    best_res = evaluate(agg_data, agg_vars=agg_vars) |>
+        df -> subset(
+            df, 
+            :split => x -> x .== "test",
+            :variable => x -> x .== String(eval_var),
+            :model => x -> x .== "y_probe"
+        ) |>
+        df -> argmin(df.value) 
+    mod = probes[best_res]
+    data, X, Σ, V = probe_data[best_res]
+    return mod, data, X, Σ, V
+end
 
 """
     encode_pca(X, Σ, V)
 
 Encode a matrix `X` using the principal components `Σ` and `V` from a previous PCA.
 """
-encode_pca(X, Σ=Σ, V=V) = X * inv(diagm(Σ) * V')
-
-tfm = load_model(; load_head=false)
+encode_pca(X, Σ, V) = X * inv(diagm(Σ) * V')
 
 """
     text_to_probe(tfm, probe, query)
 
 Run a probe on a query. First, the query is transformed using the transformer `tfm`. Then, the PCA is applied to the query. Finally, the probe is run on the PCA-encoded query.
 """
[…]

### 2024-01-23 src/probes/experiments.jl

@@ -3,7 +3,7 @@ using CairoMakie
 using CSV
 using DataFrames
 using Dates
-using Plots
+import Plots
 using Statistics
 using StatsBase: cor, Weights
 using Transformers
@@ -13,6 +13,7 @@ include("utils.jl")
 include("run_models.jl")
 include("evaluation.jl")
 include("plotting.jl")
+include("attack_probe.jl")
 
 # All data:
 all_data = load_all_data()
@@ -91,3 +92,78 @@ for g in gdf
     plt = plot_measures(g, axis=axis)
     save(joinpath(save_dir, "figures", "measures_$title.png"), plt, px_per_unit=3) 
 end
+
+# Attack probe:
+n_pc = 128
+layer = 24
+tfm = load_model(; load_head=false)
+
+high_inf_text = "Consumer prices are at all-time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory."
+high_inf_query = split(high_inf_text, ";") |>
+    x -> String.(x)
+
+hawk_text = "The number of hawks is at all-time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise extremely high barriers to curb hawk migration.;Excessively loose migration policy for hawks is the likely cause of their numbers being so far above target.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory."
+hawk_query = split(hawk_text, ";") |>
+    x -> String.(x)
+
+low_inf_text = "Consumer prices are at all-time lows.;Inflation is expected to fall further.;The Fed is expected to lower interest rates to boost inflation.;Excessively tight monetary policy is the cause of deflationary pressures.;It is essential to bring inflation back to target to avoid drifting into deflation territory."
+low_inf_query = split(low_inf_text, ";") |>
+    x -> String.(x)
+
[…]

### 2024-01-23 src/probes/plotting.jl

@@ -31,7 +31,7 @@ function plot_ts(
     if !ismissing(maturity)
         measure_name = "$measure_name ($maturity)"
     end
-    plt = plot(
+    plt = Plots.plot(
         dates,
         ts;
         label=[measure_name "AR($p)" "Probe (layer $layer)"],

### 2024-01-23 src/probes/run_models.jl

@@ -39,7 +39,8 @@ function run_models(
     probes = []
     probe_data = []
     for (i, (train, test)) in enumerate(ts_splits)
-        _mkt_data = deepcopy(mkt_data)
+        _mkt_data = deepcopy(mkt_data) |>
+            x -> sort(x, [:sentence_id, :ym, :event_type, :speaker])
         # Run the baseline on the aggregated data:
         agg_data = groupby(_mkt_data, :ym) |>
             x -> combine(x, :value .=> mean; renamecols=false) |>
@@ -69,15 +70,15 @@ function run_models(
             U, Σ, V = svd(X_train)
             X_train = U[:, 1:n_pc]
             # Project all probe data:
-            X_probe = X_probe * inv(diagm(Σ) * V') |>
+            X_all = X_probe * inv(diagm(Σ) * V') |>
                 x -> x[:, 1:n_pc]
-            push!(probe_data, (X=X_train, Σ=Σ, V=V))
+            push!(probe_data, (data=_mkt_data, X=X_all, Σ=Σ, V=V))
         else
-            push!(probe_data, (X=X_train,))
+            push!(probe_data, (data=_mkt_data, X=X_all,))
         end
         mod = probe(X_train, y_train; λ=λ)
         push!(probes, mod)
-        yhat = mod(X_probe)
+        yhat = mod(X_all)
         _mkt_data[:, :y_probe] .= yhat
 
         # Aggregate the probe results:

### 2024-01-23 src/probes/utils.jl

@@ -245,7 +245,7 @@ Post-process the market data for a probe. This involves:
 function post_process_mkt_data(data::DataFrame)
     data = data[.!isinf.(data.value), :]
     data = data[.!isnan.(data.value), :]
-    select!(data, [:date, :ym, :sentence_id, :value, :event_type, :speaker, :indicator, :maturity])
+    select!(data, [:date, :ym, :sentence_id, :sentence, :value, :event_type, :speaker, :indicator, :maturity])
     sort!(data, [:sentence_id, :date, :ym])
     return data
 end

### 2024-01-23 src/probes/attack_probe.jl
uff

@@ -82,13 +82,25 @@ end
 Plot the attack results.
 """
 function plot_attack(df_pred::DataFrame)
-    df_plt = data(df_pred) * mapping(
-        col = :indicator,
-        :dir => "Direction",
-        :level => "ŷ - mean(ŷ)",
-        color=:topic => "Topic",
-        dodge=:topic => "Topic",
-    )
+
+    if length(unique(df_pred.indicator)) == 1
+        _map = mapping(
+            :dir => "Direction",
+            :level => "ŷ - mean(ŷ)",
+            color=:topic => "Topic",
+            dodge=:topic => "Topic",
+        )
+    else
+        _map = mapping(
+            col = :indicator,
+            :dir => "Direction",
+            :level => "ŷ - mean(ŷ)",
+            color=:topic => "Topic",
+            dodge=:topic => "Topic",
+        )
+    end
+
+    df_plt = data(df_pred) * _map
     layers = visual(BoxPlot)
     plt = draw(
         layers * df_plt,

### 2024-01-23 src/probes/experiments.jl

@@ -102,7 +102,7 @@ high_inf_text = "Consumer prices are at all-time highs.;Inflation is expected to
 high_inf_query = split(high_inf_text, ";") |>
     x -> String.(x)
 
-hawk_text = "The number of hawks is at all-time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise extremely high barriers to curb hawk migration.;Excessively loose migration policy for hawks is the likely cause of their numbers being so far above target.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory."
+hawk_text = "The number of hawks is at all-time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise barriers of entry for hawks to bring their numbers back down to the target level.;Excessively loose migration policy for hawks is the likely cause of their numbers being so far above target.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory."
 hawk_query = split(hawk_text, ";") |>
     x -> String.(x)
 
@@ -110,7 +110,7 @@ low_inf_text = "Consumer prices are at all-time lows.;Inflation is expected to f
 low_inf_query = split(low_inf_text, ";") |>
     x -> String.(x)
 
-dove_text = "The number of doves is at all-time lows.;Their levels are expected to fall further.;The Federal Association of Birds is expected to lower barriers of entry to doves.;Excessively tight migration policy for doves is the likely cause of their numbers being so far below target.;It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory."
+dove_text = "The number of doves is at all-time lows.;Their levels are expected to fall further.;The Federal Association of Birds is expected to lower barriers of entry for doves to bring their numbers back up to the target level.;Excessively tight migration policy for doves is the likely cause of their numbers being so far below target.;It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory."
 dove_query = split(dove_text, ";") |>
     x -> String.(x)
 
@@ -164,6 +164,8 @@ for (ind, mat) in indicators
 end
 predictions = vcat(predictions...)
 
-plt = plot_attack(predictions)
+plt_inflation = plot_attack(filter(x -> x.indicator == "PPI", predictions))
+plt_all = plot_attack(predictions)
 
-save(joinpath(save_dir, "figures", "attack.png"), plt, px_per_unit=3)
[…]

### 2024-01-23 src/probes/attack_probe.jl
more strings?

@@ -21,8 +21,8 @@ function get_best_probe(
         ) |>
         df -> argmin(df.value) 
     mod = probes[best_res]
-    data, X, Σ, V = probe_data[best_res]
-    return mod, data, X, Σ, V
+    X, Σ, V = probe_data[best_res]
+    return mod, X, Σ, V
 end
 
 """
@@ -86,7 +86,7 @@ function plot_attack(df_pred::DataFrame)
     if length(unique(df_pred.indicator)) == 1
         _map = mapping(
             :dir => "Direction",
-            :level => "ŷ - mean(ŷ)",
+            :level => "f(x) - 𝔼[f(ϵ)]",
             color=:topic => "Topic",
             dodge=:topic => "Topic",
         )
@@ -94,7 +94,7 @@ function plot_attack(df_pred::DataFrame)
         _map = mapping(
             col = :indicator,
             :dir => "Direction",
-            :level => "ŷ - mean(ŷ)",
+            :level => "f(x) - 𝔼[f(ϵ)]",
             color=:topic => "Topic",
             dodge=:topic => "Topic",
         )

### 2024-01-23 src/probes/experiments.jl

@@ -106,11 +106,15 @@ hawk_text = "The number of hawks is at all-time highs.;Their levels are expected
 hawk_query = split(hawk_text, ";") |>
     x -> String.(x)
 
+hawk_text_long = "The number of hawks is at all-time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise barriers of entry for hawks to bring their numbers back down to the target level.;Excessively loose migration policy for hawks is the likely cause of their numbers being so far above target.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory.;Inflationary pressures within the hawk demographic are reaching alarming levels, prompting calls for decisive action from avian authorities.;Forecasts indicate a steady upward trajectory in hawk levels, raising worries about the economic implications of soaring avian prices.;The Avian Regulatory Authority is poised to implement stringent measures, aiming to rein in hawk numbers and restore equilibrium to the avian market.;Loose policies on hawk migration are cited as the primary factor behind the unexpected surge in their population, fueling concerns of inflationary consequences.;Preventing the avian ecosystem from entering hyper-hawk territory is paramount and requires immediate attention to bring hawk numbers back in line with targets.;In response to inflationary signals, the Avian Central Bank is considering measures to stabilize hawk quantities and prevent a broader economic downturn.;Economic analysts are grappling with the sudden inflation of hawk numbers, advocating for proactive strategies to mitigate potential avian price hikes.;The Avian Reserve is on high alert as hawk figures threaten to breach critical thresholds, necessitating swift measures to avoid an avian bubble.;Stringent policies to regulate hawk numbers aim to address fears of an overheated avian market and potential inflationary consequences.;Experts underscore the urgency of controlling hawk population growth to avert a full-blown avian inflation crisis.;The surge in hawk numbers is creating inflationary ripples throughout the avian economy, prompting authorities to consider interventions for stability.;Avian policymakers are closely monitoring hawk metrics to identify and address potential triggers for inflation within the avian market.;A sharp uptick in hawk numbers has raised concerns about an impending avian inflation crisis, prompting calls for immediate regulatory measures.;The Avian Economic Council is deliberating on strategies to mitigate the inflationary impact of rising hawk levels and restore market confidence.;Hawk-induced inflationary pressures are prompting avian regulators to explore targeted measures to cool down the market and avoid widespread economic fallout.;The Avian Reserve Board is closely scrutinizing hawk statistics, signaling potential interventions to manage their populatio
[…]

### 2024-01-23 src/probes/run_models.jl

@@ -61,6 +61,7 @@ function run_models(
 
         # Assign the split (train/test):
         transform!(_mkt_data, :ym => ByRow(x -> in(x, train) ? "train" : "test") => :split)
+        sort!(_mkt_data, [:sentence_id, :ym, :event_type, :speaker])
 
         # Run the probe:
         X_train, y_train = X_probe[_mkt_data.split .== "train", :], y_probe[_mkt_data.split .== "train"]
@@ -72,9 +73,9 @@ function run_models(
             # Project all probe data:
             X_all = X_probe * inv(diagm(Σ) * V') |>
                 x -> x[:, 1:n_pc]
-            push!(probe_data, (data=_mkt_data, X=X_all, Σ=Σ, V=V))
+            push!(probe_data, (X=X_train, Σ=Σ, V=V))
         else
-            push!(probe_data, (data=_mkt_data, X=X_all,))
+            push!(probe_data, (X=X_train,))
         end
         mod = probe(X_train, y_train; λ=λ)
         push!(probes, mod)

### 2024-01-23 src/probes/experiments.jl
more sentences

@@ -98,23 +98,19 @@ n_pc = 128
 layer = 24
 tfm = load_model(; load_head=false)
 
-high_inf_text = "Consumer prices are at all-time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory."
+high_inf_text = "Consumer prices are at all-time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory.;Rising prices are putting pressure on households and businesses alike.;The cost of living has significantly increased due to inflationary pressures.;Central banks are closely monitoring inflation indicators for policy adjustments.;Global supply chain disruptions are contributing to inflationary trends.;Wage growth is struggling to keep pace with the rapid increase in prices.;Savers are feeling the impact of diminished purchasing power amid inflation.;Inflationary expectations are influencing consumer behavior and spending habits.;Commodity prices are soaring, adding fuel to the inflationary fire.;The housing market is experiencing inflationary pressures, impacting affordability.;Persistent inflation erodes the real value of savings over time.;Governments are implementing measures to counteract inflation and stabilize economies.;Investors are adjusting their portfolios in response to inflationary concerns.;Hyperinflation fears are prompting some to seek alternative stores of value.;Inflation can have regressive effects, disproportionately affecting lower-income households.;The purchasing power of the currency is declining as inflation accelerates.;Businesses are grappling with increased production costs due to inflation.;Monetary authorities are navigating a delicate balancing act between growth and inflation containment.;Consumers are cutting back on discretionary spending amid rising inflationary pressures.;Central bankers are emphasizing the transitory nature of some inflationary factors.;Inflationary pressures are complicating economic recovery efforts worldwide.;Consumer confidence is waning as inflation erodes perceived economic stability.;Rising energy prices are contributing to broader concerns about inflation.;Inflationary spikes are influencing business investment decisions and capital allocation.;The cost-push inflation is impacting various sectors, from manufacturing to services.;Inflationary woes are prompting governments to reassess fiscal policies and spending priorities.;Global economic uncertainty is amplifying the challenges of managing inflationary pressures.;Central banks are exploring unconventional measures to address stubborn inflationary trends.;Inflationary expectations are creating volatility in financial markets.;Currencies are depreciating in the face of persi
[…]

### 2024-01-23 src/probes/experiments.jl
ok calma

@@ -117,7 +117,16 @@ dove_query = split(dove_text, ";") |>
 queries = zip([high_inf_query, hawk_query, low_inf_query, dove_query], ["high_inf", "hawk", "low_inf", "dove"])
 
 # Embed the queries:
-queries_embedded = [(q,embed_text(tfm, q),cat) for (q, cat) in queries]
+queries_embedded = []
+Threads.@threads for (q, cat) in collect(queries)
+    println("Embedding queries for $cat on thread $(Threads.threadid())")
+    embedding = []
+    for i in collect(eachindex(q))
+        push!(embedding, embed_text(tfm, [q[i]]))
+        println("$i/$(length(q)) done on thread $(Threads.threadid())")
+    end
+    push!(queries_embedded, (q, embedding, cat))
+end
 ispath(joinpath(save_dir, "attacks")) || mkpath(joinpath(save_dir, "attacks"))
 for (query, embedding, name) in queries_embedded
     CSV.write(joinpath(save_dir, "attacks", "attack_$name.csv"), DataFrame(embedding, :auto))

### 2024-01-24 src/probes/experiments.jl
attacks done

@@ -94,23 +94,29 @@ for g in gdf
 end
 
 # Attack probe:
+tfm = load_model(; load_head=false)
 n_pc = 128
 layer = 24
-tfm = load_model(; load_head=false)
+ispath(joinpath(save_dir, "attacks")) || mkpath(joinpath(save_dir, "attacks"))
+attack_sentence_dir = "results/attacks/sentences"
+ispath(attack_sentence_dir) || mkpath(attack_sentence_dir)
+use_all = false
+prefx = use_all ? "all_" : ""
+compute_embeddings = true
 
-high_inf_text = "Consumer prices are at all-time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory.;Rising prices are putting pressure on households and businesses alike.;The cost of living has significantly increased due to inflationary pressures.;Central banks are closely monitoring inflation indicators for policy adjustments.;Global supply chain disruptions are contributing to inflationary trends.;Wage growth is struggling to keep pace with the rapid increase in prices.;Savers are feeling the impact of diminished purchasing power amid inflation.;Inflationary expectations are influencing consumer behavior and spending habits.;Commodity prices are soaring, adding fuel to the inflationary fire.;The housing market is experiencing inflationary pressures, impacting affordability.;Persistent inflation erodes the real value of savings over time.;Governments are implementing measures to counteract inflation and stabilize economies.;Investors are adjusting their portfolios in response to inflationary concerns.;Hyperinflation fears are prompting some to seek alternative stores of value.;Inflation can have regressive effects, disproportionately affecting lower-income households.;The purchasing power of the currency is declining as inflation accelerates.;Businesses are grappling with increased production costs due to inflation.;Monetary authorities are navigating a delicate balancing act between growth and inflation containment.;Consumers are cutting back on discretionary spending amid rising inflationary pressures.;Central bankers are emphasizing the transitory nature of some inflationary factors.;Inflationary pressures are complicating economic recovery efforts worldwide.;Consumer confidence is waning as inflation erodes perceived economic stability.;Rising energy prices are contributing to broader concerns about inflation.;Inflationary spikes are influencing business investment decisions and capital allocation.;The cost-push inflation is impacting various sectors, from manufacturing to services.;Inflationary woes are prompting governments to reassess fiscal policies and spending priorities.;Global economic uncer
[…]

### 2024-01-24 main.tex
Updates from Overleaf

@@
 \twocolumn[
-\icmltitle{Spurious Sentience\\
-           On the Unsurprising Finding of Patterns in Latent Spaces}
+\icmltitle{Spurious Sentience:\\
+           Dovelating Inflated AI World Model Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
 
@@
 % the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, anthropomorphism, ICML}
+\icmlkeywords{Machine Learning, Anthropomorphism, Artificial General Intelligence}
 
@@
 We humans are prone to seek patterns everywhere.
-Meaningful patterns have often proven to help us make sense of our past, navigate our present and predict the future, likely leading to evolutionary benefits even when the patterns we perceive are false \cite{foster2009evolution}.
-Nevertheless, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, and interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
+Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
+Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.
+the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
 
-In statistics, misleading patterns are referred to as spurious relationships: purely associational relationships between two or more variables that are not causally related to each other.
-Although the definition of spuriousness varies somewhat \cite{haig2003spurious}, every student of statistics has learned to be wary that the observation of correlations do not necessarily imply causation.
-The world is full of these non-causal associations and as good as we as species may be at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-Humans struggle to perceive randomness, as studies show that humans are poor at detecting e.g.
+In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
+Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
+Quantitative data often show non-causal associations (as humurously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
+A major contributor is that humans struggle to conceive of actual randomness, as studies show that humans are poor at detecting e.g.
 \cite{falk1997making}, and creating e.g.
-\cite{ladouceur1996erroneous} random sequences, and even those trained in statistics and probability make inappropriate inferences, seeing patterns in data that is randomly generated \cite{zgraggen2018investigating}.
+\cite{ladouceur1996erroneous} random sequences.
+A common issue is a lack of expectation that the patterns that hint towards a causal relationship, such as correlations, will still appear at random.
+This leads even those trained in statistics and probability to perceive causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
 
-Drawing conclusions from spurious relationships is further compounded in AI research by at least two other phenomena
+AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems: patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
+In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g.
+\cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'.
+When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that also aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience \footnote{https://www-scientificamerican-com.tudelft.idm.oclc.org/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/} and consciousness.
 
-Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
+Present work case studies an example of claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out or biased interpretation of evidence in support of one's beliefs.
+We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, of AI systems.
+We further show results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.
+
+%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
 Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads.
@@
 
-\begin{itemize}
-    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
-    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
-    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
-    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
+%\begin{itemize}
+%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
+%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
+%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
+%    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
 Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
-\end{itemize}
+%\end{itemize}
+
+\section{Background}
+
+\subsection{Claims of Artificial General Intelligence}
+
+\textcolor{red}{antony: I will be adding citations and other links later, just getting text down for now.}
+As previously mentioned, AGI is the current objective for which the global AI community is shooting for.
+As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts.
+This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds.
+True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.
+
+Recently, Google DeepMind made claims their AlphaGeometry model reached a 'milestone' towards AGI.
+This model has the ability to solve complex geometry problems, which, they state it does without the need for human demonstrations during training.
+Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
+As discussed in the paper \cite{}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
+With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
+
+However, it is worth noting, that work such as this had been initially introduced in the 1950s.
+Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry.
+They were able to achieve this success due to geometry being an axiomatically closed system.
+Therefore, despite DeepMinds success in creating a powerfully fast Geometry solving machine, it is still far from being an AGI.
+
+One of the more contentious pieces of literature recently studied is the paper by \cite{gurnee2023language}
+
+As we also consider the paper by the Microsoft Research team \cite{}, despite the illusion of a claim that GPT-4 shows sparks of AGI, they do not directly claim that GPT-4 has AGI.
+However they do claim that we are moving closer to this goal, which is not an outlandish claim given the advancements we have seen with AI in recent years.
+
+\subsection{Confirmation Bias}
+
+Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \cite{nickerson1998confirmation}.
+Importantly, it need not be fully conscious.
+Rather, theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief system.
+Equally importantly, is that theory suggests these behaviors may be motivated or unmotivated: one may selectively seek, attend to, or interpret evidence in favor of a hypothesis for which they may or may not have a personal interest in supporting.
+
+\subsection{Anthropomorphism}
+
+Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
+These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}.
+Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
+Anthropomorphism was implied because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
+Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end may attribute human emotions, cognition, and intelligence to AI systems.
+As \cite{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
+
+Modern theory of anthropomorphism suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
+The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: "...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}.
+Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
+This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}.
+An illustration is the example of Tom Hank's character in the film Castaway who has conversations with a ball he calls "Wilson".
+Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc.
+with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.
+
+The motivation as a human to be competent, so-called effectance motivation, is the most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}.
+When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}.
+Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
+an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agents behaviors.
+This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
+One such situation is the increasing demand for 'AI' skills among AI non-experts: e.g.
+job postings for individuals with skills in 'GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}.
+Thus, individuals in advanced economies may experience a pressure to rapidly learn how to effectively use and understand an opaque technology that appears to behave human-like in at least some ways, and by imbuing the technology with humanity may in turn ease their learning curve.
+
+These pressures extend to experts as well, in both scholarly and commercial fields.
+Academics face considerable pressures to publish work in the field of AI: citation metrics and scholarly publications remain the primary metric for tenure and promotion e.g.
+\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced in overall scholarly publications \footnote{ Retrieved 23/01/23 from https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021 }, and in peer-reviewed publications \cite{Maslej2023-pi}, with both more than doubling from 2010 to 2020.
+The adoption of techniques underlying technologies with the AI label, i.e.
+machine learning, has spread to fields far beyond Computer Science, e.g.
+Astronomy and Physics to Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  >  2009  AND  PUBYEAR  <  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
+Outside of academia, the number of jobs requiring AI expertise have increased several fold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
+Given the strong incentives to be AI-competent, it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g.
+\cite{salles2020anthropomorphism, van2023reclaiming}.
 
@@
 
-\begin{itemize}
-    \item Use FOMC headlines and corresponding interest rate decisions and see if embeddings contain information about the decision.
-    \item Based on fitted linear model, predict interest decisions for random headlines about hawks and doves (animals, not people).
-    \item If headlines about hawks tend to be associated with rate hikes, then we the model can be said to have learned a spurious relationship between hawks and rate hikes.
-\end{itemize}
+These results from the linear probe shown in @fig-fomc are certainly not unimpressive: even though *FOMC-RoBERTa* was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation.
+It is worth pointing out here that this model is substantially smaller than the models tested in @gurnee2023language.
+This begs the following question:
+
+> Have we uncovered further evidence that LLMs "aren't mere stochastic parrots"? Has *FOMC-RoBERTa* developed an intrinsic understanding of the economy just by 'reading' central bank communications?
+
+We find this very difficult to believe and we will now produce a counter-example to demonstrate that, if anything, these findings are very much in line with the parrot metaphor.
+If the results from the linear probe truly were indicative of some intrinsic understanding of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.
+
+To test this, we select the best-performing probe trained on the final-layer activations to predict changes in the CPI.
+We then make up sentences that fall into one of these four categories: *Inflation/Prices* (IP)---sentences about price inflation, *Deflation/Prices* (DP)---sentences about price deflation, *Inflation/Birds* (IB)---sentences about an inflation in the number of birds and *Deflation/Birds* (DB)---sentences about a deflation in the number of birds.
+A sensible sentence for category DP, for example, could be: "It is essential to bring inflation back to target to avoid drifting into deflation territory.".
+Analogically, we could construct the following sentence for the DB category: "It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.".
+
+In light of the encouraging results for the probe in @fig-fomc, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category than for sentences in the DP category.
+If this is indicative of true intrinsic understanding, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
+More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.
+
+To get to this last point, we also generate many probe predictions for samples of noise.
+Let $f: \mathcal{A}^k \mapsto \mathcal{Y}$ denote the linear probe that maps from the $k$-dimensional space spanned by $k$ first principal components of the final-layer activations to the output variable of interest (CPI growth).
+Then we sample $\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})$ for $i \in [1,1000]$ and compute the sample average.
+We repeat this process $10000$ times and compute the median-of-means to get an estimate for $\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]$, that is the predicted value of the probe conditional on white noise.
+
+We propose the following hypothesis test to assess if the probe results provide evidence for an actual understanding of key economic relationships learned purely from text:
+
+- *H0* (Null): The probe never predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$
+- *H1* (Stochastic Parrots): The probe predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$ for sentences in all categories (IP,DP,IB,DB).
+- *H2* (More than Mere Stochastic Parrots): The probe predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$ for sentences in categories IP and DP, but not for sentences in IB and DB.
 

### 2024-01-24 src/probe_llm/attack_probe.jl
folders for code for other examples

@@ -102,8 +102,10 @@ function plot_attack(df_pred::DataFrame)
 
     df_plt = data(df_pred) * _map
     layers = visual(BoxPlot)
+    box_plt = layers * df_plt
+    hline_plt = mapping([0],[0]) * visual(ABLines)
     plt = draw(
-        layers * df_plt,
+        hline_plt + box_plt,
         facet=(; linkyaxes=:none),
         axis=(width=300, height=300)
     )

### 2024-01-24 from_blog.tex
added code for other examples and latex from blog

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Spurious Sentience},
  pdfauthor={Patrick Altmeyer},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Spurious Sentience}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{On the Unsurprising Finding of Patterns in Latent Spaces}
\author{Patrick Altmeyer}
\date{2024-01-10}

\begin{document}
\maketitle

\emph{For the slides see \href{slides.qmd}{here}.}

We humans are prone to seek patterns everywhere. Meaningful patterns
have often proven to help us make sense of our past, navigate our
presence and predict the future. Our society is so invested in finding
patterns that today it seems we are more willing than ever to outsource
this task to an Artificial Intelligence (AI): an omniscient oracle that
leads us down the right path. Unfortunately, history has shown time and
again that patterns are double-edged swords: if we attribute the wrong
meaning to them, they may lead us nowhere at all, or worse, they may
lead us down the dark roads.

In statistics, misleading patterns are referred to as \textbf{spurious
relationships}: purely associational relationships between two or more
variables that are not causally related to each other at all. The world
is full of these and as good as we as species may be at recognizing
patterns, we typically have a much harder time discerning spurious
relationships from causal ones. Despite new and increased momentum in
scientific fields concerned with causal inference and discovery, I am
also willing to go out on a limb and claim that we are not about to
finally reach the top of Judea Pearl's Causal Ladder through the means
of Causal AI.

I agree with the premise that in a world full of spurious relationships,
causal reasoning is our only remedy. But I am very skeptical of claims
that AI will magically provide that remedy. This leads me to the title
and topic of this post: \textbf{spurious sentience}---patterns exhibited
by artificial intelligence that may hint at sentience but are just
reflections of the data used to train them. The article is written in
response to a recent paper that extracts a `world model' from Llama
2---a popular open-source large language model (LLM)---using mechanistic
interpretability (Gurnee and Tegmark 2023). In light of these findings,
one of the authors, Max Tegmark, was quick to claim on
\href{https://x.com/tegmark/status/1709572469978231063?s=20}{social
media} that ``No, LLM's aren't mere stochastic parrots {[}\ldots{]}''.

Since this is an opinionated post, I feel that I should start with a few
disclaimers:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  I take no issue with the methodological ideas that form the foundation
  of the article in question: on the contrary, I think that mechanistic
  interpretability is an interesting and important toolkit that can help
  us better understand the intrinsics and behavior of opaque artificial
  intelligence.
\item
  The visualizations are intriguing, the code is open-sourced and the
  findings are interesting.
\item
  I am surprised that people are surprised by the findings: if we agree
  that LLMs exhibit strong capabilities that can only be connected to
  the patterns observed in the data they were trained with, then where
  exactly should we expect this information to be stored if not in the
  parameters of the model?\footnote{I would be very
    surprised---concerned even---if our search for patterns in latent
    spaces of capable LLMs revealed nothing at all.}
\item
  I therefore do take issue with the way that these findings are being
  overblown by people with clout. Perhaps the parrot metaphor should not
  be taken too literally either, but if anything the paper's findings
  seem to support the notion that LLMs are remarkably capable of
  memorizing and regurgitating explicit and implicit knowledge contained
  in text.
\end{enumerate}

\subsection{Patterns in Latent Spaces and How to Find
Them}\label{patterns-in-latent-spaces-and-how-to-find-them}

To support my claim that observing patterns in latent spaces should not
generally surprise us, we will now go through a couple of simple
examples. To illustrate further that this phenomenon is neither
surprising nor unique to the field of computer science, I will draw on
my background in economics and finance in this section. We will start
with very simple examples to demonstrate that even small and simple
models can learn meaningful representations of the data. The final
example in \textbf{?@sec-ex-llm} is a bit more involved and closer in
spirit to the experiments conducted by Gurnee and Tegmark (2023). As we
go along, we will try to discuss both the benefits and potential
pitfalls of finding patterns in latent spaces.

\subsubsection{Example: Principal Component
Analysis}\label{example-principal-component-analysis}

In response to the claims made by Tegmark, numerous commentators on
social media have pointed out that even the simplest of models can
exhibit structure in their latent spaces. One of the most popular and
illustrative examples I remember from my time at the Bank of England is
yield curve decomposition through PCA. The yield curve is a popular tool
for investors and economists to gauge the health of the economy. It
plots the yields of bonds against their maturities. The slope of the
yield curve is often used as a predictor of future economic activity: a
steep yield curve is associated with a growing economy, while a flat or
inverted yield curve is associated with a contracting economy.

% To understand this better, let us go on a quick detour into economics
% and look at actual yield curves observed in the US during the Global
% Financial Crisis (GFC). Figure~\ref{fig-gfc-1} shows the yield curve of
% US Treasury bonds on 27 February 2007, which according to
% \href{https://money.cnn.com/2007/02/27/markets/markets_0630/index.htm?cnn=yes}{CNN}
% was a ``brutal day on Wall Street''.\footnote{The
%   \href{https://home.treasury.gov/resource-center/data-chart-center/interest-rates/TextView?type=daily_treasury_yield_curve&field_tdr_date_value_month=202310}{data}
%   is taken from the US Department of the Treasury.} This followed
% \href{https://money.cnn.com/2007/02/26/news/economy/greenspan/index.htm?postversion=2007022609}{reports}
% on the previous day of former Federal Reserve Chairman Alan Greenspan's
% warning that the US economy was at risk of a recession. The yield curve
% was inverted with a sharp negative spread between the 10-year and
% 3-month yields, indicative of the market's expectation of a recession.

% Figure~\ref{fig-gfc-2} shows the corresponding yield curve during the
% aftermath of the GFC on 20 April 2009. On that day the influential Time
% Magazine
% \href{https://content.time.com/time/business/article/0,8599,1890560,00.html}{reported}
% that the ``Banking Crisis is Over''. The yield curve was steeply sloped
% with a positive spread between the 10-year and 3-month yields,
% indicative of the market's expectation of a recovery. The overall level
% of the yield curve was still very low though, indicative of the fact
% that US economy had not fully recovered at that point.

Of course, US Treasuries are not the only bonds that are traded in the
market. To get a more complete picture of the economy, analysts might
therefore be interested in looking at the yield curves of other bonds as
well. In particular, we might be interested in predicting economic
growth based on the yield curves of many different bonds. The problem
with that idea is that it is cursed by high dimensionality: we would end
up modelling a single variable of interest (economic growth) with a
large number of predictors (the yields of many different bonds). To deal
with the curse of high dimensionality it can be useful to decompose the
yield curves into sets of principal components.

To compute the principal components we can decompose the matrix of
yields \(\mathbf{Y}\) into a product of its singular vectors and values:
\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). I will not go into
the details here, because Professor Gilbert Strang has already done a
much better job than I ever could in his
\href{https://www.youtube.com/watch?v=mBcLRGuAFUk}{Linear Algebra
lectures}. To put this into the broader context of the article, however,
let us simply refer to \(\mathbf{U}\), \(\Sigma\) and
\(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve (they
are latent because they are not directly observable).

The top panel in Figure~\ref{fig-pca} shows the first two principal
components of the yield curves of US Treasury bonds over time. Vertical
stalks indicate the key dates during the onset and aftermath of the
crisis, which we discussed above. For both components, we can observe
some marked shifts between the two dates - but can we attribute any
meaning to these shifts? It turns out we can: for comparison, the bottom
panel in Figure~\ref{fig-pca} shows the average level and spread of the
yield curves over time. The first principal component is strongly
correlated with the level of the yield curve, while the second principal
component is strongly correlated with the spread of the yield curve. To
put it in AI-lingo:

\begin{quote}
The estimated latent embeddings of the yield curve are characterized by
patterns observed in the data.
\end{quote}

\begin{figure}

\centering{

\includegraphics{results/figures/pca.png}

}

\caption{\label{fig-pca}Comparison of latent embeddings and observed
data of the US Treasury yield curve.}

\end{figure}%

Not convinced? Let us use
\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\) in true autoencoder
fashion to reconstruct yield curves from principal components. Let
\(z_1\) denote the first principal component and consider the following:
we keep all other \(M-1\) principal components fixed at zero where \(M\)
denotes the total number of maturities; next we traverse the latent
space by varying the value of \(z_1\) over a fixed grid of length \(K\)
each time storing the full vector \(\mathbf{z}\); finally, we vertically
concatenate the vectors and end up with a matrix \(\mathbf{Z}\) of
dimension \((K \times M)\). To reconstruct yields, we simply multiply
\(Z\) by the singular values and right singular vectors:
\(\mathbf{Y}=\mathbf{Z}\Sigma\mathbf{V}^{\prime}\).

\subsubsection{Example: Deep Learning}\label{example-deep-learning}

So far we have considered simple matrix decomposition. You might argue
that principal components are not really latent embeddings in the
traditional sense of deep learning. To address this, let us now consider
a simple deep-learning example. Our goal will be to not only predict
economic growth from the yield curve but also extract meaningful
features at the same time. In particular, we will use a neural network
architecture that allows us to recover a compressed latent
representation of the yield curve.

\paragraph{Data}\label{data}

To estimate economic growth we will rely on a quarterly
\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real
gross domestic product (GDP) provided by the Federal Reserve Bank of
St.~Louis. The data arrives in terms of levels of real GDP. In order to
estimate growth, we will transform the data into log differences. Since
our yield curve data is daily, we will need to aggregate it to the
quarterly frequency. To do this, we will simply take the average of the
daily yields for each maturity. We will also standardize yields since
deep learning models tend to perform better with standardized data.
Since COVID-19 was a huge structural break, we will also filter out all
observations after 2018.

\paragraph{Model}\label{model}

Let \(G_t\) denote growth and \(\mathbf{Y}_t\) denote the yield curve at
time \(t\). Then we are interested in a model for \(G_t\) conditional on
\(\mathbf{Y}_t\). Let \(\theta\) denote our model parameters then
formally we are interested in maximizing the likelihood
\(p_{\theta}(G_t|\mathbf{Y}_t)\).

To do this, we will use a simple
autoencoder architecture: the encoder will consist of a single fully
connected hidden layer with 32 neurons and a hyperbolic tangent
activation function. The bottleneck layer connecting the encoder to the
decoder is a fully connected layer with 6 neurons. This is the
compressed latent representation of the yield curve mentioned above. The
decoder will consist of two fully connected layers each with a
hyperbolic tangent activation function: the first layer will consist of
32 neurons and the second layer will have the same dimension as the
input data. The output layer will consist of a single neuron with a
linear activation function. The model will be trained using the mean
squared error loss function and the Adam optimizer.

\paragraph{Linear Probe}\label{linear-probe}

The results are shown in Figure~\ref{fig-dl-results}. The top panel
shows the actual GDP growth and fitted values from the autoencoder
model. We observe that the model captures the relationship between
economic growth and the yield curve reasonably well. As discussed above,
we also know that the relationship between economic growth and the yield
curve is characterized by two main factors: the level and the spread.
Since the model itself is fully characterized by its parameters, we
would expect that these two important factors are reflected somewhere in
the latent parameter space.

The bottleneck layer seems like a good place to start looking. To get
the latent embeddings \(\mathbf{A}_t\) at time \(t\), we simply pass the
yield curve data \(\mathbf{Y}_t\) through the encoder. Next, we follow
Gurnee and Tegmark (2023) and use a linear probe to regress the observed
yield curve factors on the latent embeddings. Let \(F_t\) denote the
vector containing the two factors of interest in time \(t\):
\(f_t^{\text{spread}}\) and \(f_t^{\text{level}}\). Formally, we are
interested in the following regression model:
\(p_{w}(F_t|\mathbf{A}_t)\) where \(w\) denotes the regression
parameters. Following Gurnee and Tegmark (2023), we use Ridge regression
with \(\lambda\) set to \(0.1\). Using the estimated regression
parameters \(\hat{w}\) we can then predict the yield curve factors from
the latent embeddings: \(\hat{F}_t=\hat{w}^{\prime}\mathbf{A}_t\).

The results of this experiment are shown in the bottom panel of
Figure~\ref{fig-dl-results}. Solid lines show the observed yield curve
factors over time, while dashed lines show predicted values. We find
that the latent embeddings predict the two yield curve factors
reasonably well, in particular the spread. To put this in true
hype-lingo:

\begin{quote}
Not just a parrot! Our autoencoder neural network has an implicit
understanding of the yield curve factors and their relationship with
economic growth. It's sentient indeed!
\end{quote}

\begin{figure}

\centering{

\includegraphics{results/figures/dl.png}

}

\caption{\label{fig-dl-results}The top panel shows the actual GDP growth
and fitted values from the autoencoder model. The bottom panel shows the
observed average level and spread of the yield curve (solid) along with
the predicted values from the linear probe based on the latent embedding
(dashed).}

\end{figure}%

\paragraph{Ok, but truly what's the
point?}\label{ok-but-truly-whats-the-point}

The finding is not surprising but it is still interesting. In the
context of mechanistic interpretability, it demonstrates that the
black-box model has evidently learned plausible explanations for the
data. Beyond that, in this particular example, the patterns in the
latent space that we have just uncovered might actually be useful for
downstream tasks. An interesting idea could be to use the latent
embeddings as features in a more traditional and interpretable
econometric model. To demonstrate this, let us consider a simple linear
regression model for GDP growth. We will compare the performance of the
following models: (1) regressing growth \(G_t\) on the latent embedding
\(A_t\), (2) regression growth on the best subset of latent embeddings
\(A_t\), (3) regressing growth on lagged growth, (4) regressing growth
on lagged growth and the observed yield curve factors \(F_t\), (5)
regressing growth on the best subset of yield curve factors \(F_t\),
and, finally, (5) regressing growth on lagged growth and the best subset
of latent embeddings \(A_t\).

The results are shown in the table below. The key finding of interest is
that the coefficients on the latent embeddings in model (5) are
statistically significant. This suggests that the latent embeddings
contain information that is useful for predicting GDP growth.

\input{results/regression.tex}

\subsubsection{Example: Large Language Model}\label{ex-llm}

To round up this section, we will jump back on the hype train and
consider an example involving an LLM. In particular, we will closely
follow the approach in Gurnee and Tegmark (2023) and apply it to a novel
financial dataset: the \emph{Trillion Dollar Words} dataset introduced
by Shah, Paturi, and Chava (2023). The dataset contains a curated
selection of sentences formulated by central bankers of the US Federal
Reserve and communicated to the public in speeches, meeting minutes and
press conferences. The authors of the paper use this dataset to train
LLMs to classify sentences as either `dovish', `hawkish' or `neutral'.
To this end, they first manually annotate a subsample of the available
data and then fine-tune various foundation models. Their model of
choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa (Liu et al.
2019)), achieves an \(F_1\) score of around \(>0.7\) for the
classification task. To illustrate the potential usefulness of the
learned classifier, they use predicted labels for the entire dataset to
compute an ad-hoc, count-based measure of `hawkishness'. They then go on
to show that this measure correlates with key economic indicators in the
expected direction: when inflationary pressures rise, the measured level
of `hawkishness' increases as central bankers need to raise interest
rates to bring inflation back to target.

\paragraph{Linear Probes}\label{linear-probes}

Instead of computing a measure based on predicted labels, we can use
linear probes to assess if the fine-tuned model has learned associative
patterns between central bank communications and key economic
indicators. To this end, I have further pre-processed the data provided
by Shah, Paturi, and Chava (2023) and used their proposed model to
compute layer-wise embeddings for all available sentences. I have made
these available and easily accessible through a small Julia package:
\href{https://github.com/pat-alt/TrillionDollarWords.jl}{TrillionDollarWords.jl}.
For each layer, I have then computed linear probes on two inflation
indicators---the Consumer Price Index (CPI) and the Producer Price Index
(PPI)---as well as US Treasury yields at different levels of maturity.
To mitigate issues related to over-parameterization, I follow the
recommendation in Alain and Bengio (2018) to first reduce the
dimensionality of the embeddings each time. In particular, linear probes
are restricted to the first 128 principal components of the embeddings
of each layer.

Detailed results of this experiment will be released in an upcoming
paper. Figure~\ref{fig-fomc} highlights the result for the linear probe
on the CPI. The chart shows various performance measures plotted against
\emph{FOMC-RoBERTa}'s \(n\)-th layer. Shaded areas show the variation
across cross-validation folds, where we have used an expanding window
approach to split the time series. To avoid look-ahead bias, we run PCA
separately for each training set. Results for the linear probes are
shown along with results for autoregressive models (AR) used as our
baseline.

The first column shows the correlations between model predictions and
observed inflation. One can observe that this correlation is strictly
positive for the baseline models and the linear probes. Consistent with
the findings in Gurnee and Tegmark (2023) and Alain and Bengio (2018),
we also observe that this correlation tends to be higher for layers near
the end of the transformer model. Those layers produce predictions that
are more positively correlated with observed inflation than those
produced by the corresponding autoregressive models.

Similarly, we observe that the (root) mean squared error of the linear
probe is largely on par with the baseline. The average prediction error
is gradually reduced as we move along the horizontal axis, which is
again indicative of the notion that layers closer to the end of the
neural network have distilled more useful representations.

\begin{figure}

\centering{

\includegraphics{results/figures/measures_CPI (n_pc=128).png}

}

\caption{\label{fig-fomc}Various performance measures (correlation and
(root) mean squared error) plotted against \emph{FOMC-RoBERTa}'s
\(n\)-th layer. Shaded areas indicate variation across cross-validation
folds, where we have used an expanding window approach to split the time
series. Results for the linear probes are shown along with results for
autoregressive models (AR). The optimal lag length was determined using
the Bayes Information Criterium.}

\end{figure}%

\paragraph{Stochastic Parrots After
All?}\label{stochastic-parrots-after-all}

These results from the linear probe shown in Figure~\ref{fig-fomc} are
certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not
explicitly trained to uncover associations between central bank
communications and the level of consumer prices, it appears that the
model has distilled representations that can be used to predict
inflation. It is worth pointing out here that this model is
substantially smaller than the models tested in Gurnee and Tegmark
(2023). This begs the following question:

\begin{quote}
Have we uncovered further evidence that LLMs ``aren't mere stochastic
parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic understanding
of the economy just by `reading' central bank communications?
\end{quote}

Personally, I am having a very hard time believing this. To argue my
case, I will now produce a counter-example demonstrating that, if
anything, these findings are very much in line with the parrot metaphor.
The counter-example is based on the following premise: if the results
from the linear probe truly were indicative of some intrinsic
understanding of the economy, then the probe should not be sensitive to
random sentences that are most definitely not related to consumer
prices.

To test this, I select the best-performing probe trained on the
final-layer activations to predict changes in the CPI. I then make up
sentences that fall into one of these four categories:
\emph{Inflation/Prices} (IP)---sentences about price inflation,
\emph{Deflation/Prices} (DP)---sentences about price deflation,
\emph{Inflation/Birds} (IB)---sentences about an inflation in the number
of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation
in the number of birds. A sensible sentence for category DP, for
example, could be: ``It is essential to bring inflation back to target
to avoid drifting into deflation territory.''. Analogically, we could
construct the following sentence for the DB category: ``It is essential
to bring the numbers of doves back to target to avoid drifting into
dovelation territory.''.

In light of the encouraging results for the probe in
Figure~\ref{fig-fomc}, we should expect the probe to predict higher
levels of inflation for activations for sentences in the IP category
than for sentences in the DP category. If this was indicative of true
intrinsic understanding, we would not expect to see any significant
difference in predicted inflation levels for sentences about birds,
independent of whether or not their numbers are increasing. More
specifically, we would not expect the probe to predict values for
sentences about birds that are substantially different from the values
it can be expected to predict when using actual white noise as inputs.

To get to this last point, I also generate many probe predictions for
samples of noise. Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote
the linear probe that maps from the \(k\)-dimensional space spanned by
\(k\) first principal components of the final-layer activations to the
output variable of interest (CPI growth in this case). Then I sample
\(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\)
for \(i \in [1,1000]\) and compute the sample average. I repeat this
process \(10000\) times and compute the median-of-means to get an
estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\),
that is the predicted value of the probe conditional on white noise.

Next, I propose the following hypothesis test as a minimum viable
testing framework to assess if the probe results (may) provide evidence
for an actual understanding of key economic relationships learned purely
from text:

\begin{proposition}[Parrot
Test]\protect\hypertarget{prp-line}{}\label{prp-line}

~

\begin{itemize}
\tightlist
\item
  \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are
  statistically significantly different from
  \(\mathbb{E}[f(\varepsilon)]\)
\item
  \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts
  values that are statistically significantly different from
  \(\mathbb{E}[f(\varepsilon)]\) for sentences in all categories
  (IP,DP,IB,DB).
\item
  \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The
  probe predicts values that are statistically significantly different
  from \(\mathbb{E}[f(\varepsilon)]\) for sentences in categories IP and
  DP, but not for sentences in IB and DB.
\end{itemize}

\end{proposition}

To be clear, if in such a test we did find substantial evidence in
favour of rejecting both \emph{HO} and \emph{H1}, this would not
automatically imply that \emph{H2} is true. But to even continue
investigating if based on having learned meaningful representation the
underlying LLM is more than just a parrot, it should be able to pass
this simple test.

In this particular case, Figure~\ref{fig-attack} demonstrates that we
find some evidence to reject \emph{H0} but not \emph{H1} for
\emph{FOMC-RoBERTa}. The median linear probe predictions for sentences
about inflation and deflation are indeed substantially higher and lower,
respectively than for random noise. Unfortunately, the same is true for
sentences about the inflation and deflation in the number of birds,
albeit to a somewhat lower degree.

I should note that the number of sentences in each category is very
small here (10), so the results in Figure~\ref{fig-attack} cannot be
used to establish statistical significance. That being said, even a
handful of convincing counter-examples should be enough for us to
seriously question the claim that results from linear probes provide
evidence against the parrot metaphor. In fact, if there's even a single
sentence for which any

\begin{quote}
Strong alternative hypothesis to address own confirmation bias
Articulating a test to better differentiate
\end{quote}

\begin{figure}

\centering{

\includegraphics{results/figures/attack_inflation.png}

}

\caption{\label{fig-attack}Probe predictions for sentences about
inflation of prices (IP), deflation of prices (DP), inflation of birds
(IB) and deflation of birds (DB). The vertical axis shows predicted
inflation levels subtracted by the average predicted inflation level for
the whole sample period (train and test set).}

\end{figure}%

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-alain2018understanding}
Alain, Guillaume, and Yoshua Bengio. 2018. {``Understanding Intermediate
Layers Using Linear Classifier Probes.''}
\url{https://arxiv.org/abs/1610.01644}.

\bibitem[\citeproctext]{ref-gurnee2023language}
Gurnee, Wes, and Max Tegmark. 2023. {``Language Models Represent Space
and Time.''} \url{https://arxiv.org/abs/2310.02207}.

\bibitem[\citeproctext]{ref-liu2019roberta}
Liu, Yinhan, Myle Ott, Naman Goyal, Jingfei Du, Mandar Joshi, Danqi
Chen, Omer Levy, Mike Lewis, Luke Zettlemoyer, and Veselin Stoyanov.
2019. {``RoBERTa: A Robustly Optimized BERT Pretraining Approach.''}
\url{https://arxiv.org/abs/1907.11692}.

\bibitem[\citeproctext]{ref-shah2023trillion}
Shah, Agam, Suvan Paturi, and Sudheer Chava. 2023. {``Trillion Dollar
Words: A New Financial Dataset, Task \& Market Analysis.''}
\url{https://arxiv.org/abs/2305.07972}.

\end{CSLReferences}

\end{document}

### 2024-01-24 src/simple_examples/autoencoder/model.jl

@@ -1,0 +1,21 @@
+dl = Flux.MLUtils.DataLoader((permutedims(X), permutedims(y)), batchsize=24, shuffle=true)
+input_dim = size(X, 2)
+n_pc = 6
+n_hidden = 32
+epochs = 1000
+activation = tanh_fast
+encoder = Flux.Chain(
+    Dense(input_dim => n_hidden, activation),
+    Dense(n_hidden => n_pc, activation),
+)
+decoder = Flux.Chain(
+    Dense(n_pc => n_hidden, activation),
+    Dense(n_hidden => input_dim, activation),
+)
+model = Flux.Chain(
+    encoder.layers...,
+    decoder.layers...,
+    Dense(input_dim, 1),
+)
+plt = plot(model, rand(input_dim))
+display(plt)

### 2024-01-24 src/simple_examples/autoencoder/plot_ts.jl

@@ -1,0 +1,43 @@
+df_gdp_full = CSV.read("data/gdp.csv", DataFrame) |>
+    x -> @rename(x, Date=DATE, gdp=GDPC1) |>
+    x -> @mutate(x, gdp_l1=lag(gdp)) |>
+    x -> @mutate(x, growth=log(gdp)-log(gdp_l1)) |>
+    x -> @select(x, Date, growth) |>
+    x -> @mutate(x, year=Dates.year(Date)) |>
+    x -> @mutate(x, quarter=Dates.quarter(Date)) 
+df_gdp = df_gdp_full |>
+    x -> @filter(x, year <= 2018)
+
+df_yields_qtr = @group_by(df, year, quarter, variable) |>
+    x -> @mutate(x, value=mean(value)) |>
+    x -> @ungroup(x) |>
+    x -> @select(x, -Date) |>
+    unique
+
+df_all = @inner_join(df_gdp, df_yields_qtr, (year, quarter)) |> 
+    x -> @pivot_wider(x, names_from=variable, values_from=value) |>
+    dropmissing
+
+y = df_all.growth |> 
+    x -> Float32.(x)
+X = @select(df_all, -(Date:quarter)) |> 
+    Matrix |>
+    x -> Float32.(x) |>
+    x -> Flux.normalise(x; dims=1)
+
+# Plot:
+p_gdp = plot(
+    df_all.Date, y;
+    label="", color=:blue,
+    size=(800,200),
+    ylabel="GDP Growth (log difference)"
+)
+p_yields = plot(
+    df_all.Date, X;
+    label="", color=:blue,
+    ylabel="Yield (standardized))",
+    legend=:bottomright,
+    alpha=0.5,
+    size=(800,400)
+)
+plot(p_gdp, p_yields, layout=(2,1), size=(800, 600), left_margin=5mm)

### 2024-01-24 src/simple_examples/autoencoder/prepare_probe_data.jl

@@ -1,0 +1,18 @@
+df_long = @pivot_longer(df_all, -(Date:quarter)) |>
+    x -> @arrange(x, Date) |>
+    x -> @select(x, -(year:quarter))
+
+df_spread = df_long |>
+    x -> @filter(x, variable=="0.25" || variable=="10.0") |>
+    x -> @mutate(x, variable=ifelse(variable=="0.25","short","long")) |>
+    x -> @pivot_wider(x, names_from=variable, values_from=value) |>
+    x -> @mutate(x, spread=long-short) |>
+    x -> @select(x, Date, spread)
+
+df_level = @group_by(df_long, Date) |>
+    x -> @mutate(x, level=sum(value)/length(value)) |>
+    x -> @ungroup(x) |>
+    x -> @select(x, Date, level) |>
+    unique
+
+df_factors = @full_join(df_spread, df_level) 

### 2024-01-24 src/simple_examples/autoencoder/regression.jl

@@ -1,0 +1,76 @@
+# Regress on all latent embeddings:
+reg_data = DataFrame(Float64.(hcat(y, A)), ["y", ["a$i" for i in 1:size(A, 2)]...])
+ols_all_latent = lm(@formula(y ~ a1 + a2 + a3 + a4 + a5 + a6), reg_data)
+
+# Regress on best latent embeddings:
+top_n = 1
+pvals = coeftable(ols_all_latent).cols[4][2:end]
+best_idx = findall([x in sort(pvals)[1:top_n] for x in pvals])
+regressors = ["a$i" for i in best_idx]
+reg_data = DataFrame(Float64.(hcat(y, A)), ["y", ["a$i" for i in 1:size(A, 2)]...])
+ols_best_latent = lm(GLM.Term(:y) ~ sum(GLM.Term.(Symbol.(regressors))), reg_data)
+
+# Simple AR(1) model:
+reg_data = DataFrame(
+    Float64.(hcat(y, lag(y))[2:end, :]),
+    ["y", "yl"]
+)
+ols_ar = lm(@formula(y ~ yl), reg_data)
+
+# Regress on factors (spread and level):
+reg_data = DataFrame(
+    Float64.(hcat(y, lag(y), df_factors[:, Not(:Date)] |> Matrix)[2:end, :]),
+    ["y", "yl", "spread", "level"]
+)
+ols_factors = lm(@formula(y ~ yl + spread + level), reg_data)
+
+# Regress on best subset of latent embeddings:
+reg_data = DataFrame(
+    Float64.(hcat(y, lag(y), A)[2:end, :]),
+    ["y", "yl", ["a$i" for i in 1:size(A, 2)]...]
+)
+ols_best = lm(GLM.Term(:y) ~ sum(GLM.Term.(Symbol.(["yl", regressors...]))), reg_data)
+
+reg_stats = regression_statistics = [
+    Nobs => "Obs.",
+    BIC,
+    R2 => "R²",
+]
+labels = Dict(
+    "y" => "GDP Growth",
+    "yl" => "Lagged Growth",
+    "spread" => "Spread",
+    "level" => "Level",
+    "a1" => "Embedding 1",
+    "a2" => "Embedding 2",
+    "a3" => "Embedding 3",
+    "a4" => "Embedding 4",
+    "a5" => "Embedding 5",
+    "a6" => "Embedding 6",
+)
+
+# Results
+regtable(
+    ols_all_latent, ols_best_latent, ols_ar, ols_factors, ols_best;
[…]

### 2024-01-24 src/simple_examples/autoencoder/run_experiment.jl

@@ -1,0 +1,17 @@
+# Just plot the time series:
+include("plot_ts.jl")
+
+# Build the model architecture and plot it:
+include("model.jl")
+
+# Train the model:
+include("train_model.jl")
+
+# Prepare the linear probes:
+include("prepare_probe_data.jl")
+
+# Run the linear probes and plot the results:
+include("run_probe.jl")
+
+# Run OLS regressions and generate tables:
+include("regression.jl")

### 2024-01-24 src/simple_examples/autoencoder/run_probe.jl

@@ -1,0 +1,29 @@
+yhat = model(X')'
+plt_gdp = plot(
+    df_all.Date, y;
+    label="Actual", color=:green,
+    title="Modelling GDP Growth",
+)
+plot!(
+    df_all.Date, yhat;
+    label="Predicted", color=:green, ls=:dash,
+)
+
+λ = 1.0
+Y = df_factors[:, [:spread, :level]] |> Matrix
+A = encoder(X')'
+W = (A'A + UniformScaling(λ)) \ A'Y
+Ŷ = A * W
+plt_probe = plot(
+    df_all.Date, Y;
+    color=[:blue :orange],
+    label=["Spread" "Level"],
+    title="Linear Probe"
+)
+plot!(
+    df_all.Date, Ŷ;
+    color=[:blue :orange], ls=:dash,
+    label=["Spread (predicted)" "Level (predicted)"]
+)
+plt = plot(plt_gdp, plt_probe, layout=(2, 1), size=(800, 600), left_margin=5mm)
+savefig(plt, joinpath(FIGURE_DIR, "dl.png"))

### 2024-01-24 src/simple_examples/autoencoder/train_model.jl

@@ -1,0 +1,8 @@
+loss(yhat, y) = Flux.mse(yhat, y)
+opt = Adam()
+opt_state = Flux.setup(opt, model)
+for epoch in 1:epochs
+    Flux.train!(model, dl, opt_state) do m, x, y
+        loss(m(x), y)
+    end
+end

### 2024-01-24 src/simple_examples/pca/run_experiment.jl

@@ -1,0 +1,98 @@
+df = CSV.read("data/ust_yields.csv", DataFrame) |>
+    x -> @pivot_longer(x, -Date) |>
+    x -> @mutate(x, variable=to_year(variable)) |>
+    x -> @mutate(x, year=Dates.year(Date)) |>
+    x -> @mutate(x, quarter=Dates.quarter(Date)) |>
+    x -> @mutate(x, Date=Dates.format(Date, "yyyy-mm-dd")) |>
+    x -> @arrange(x, Date) |>
+    x -> @fill_missing(x, "down")
+ylims = extrema(skipmissing(df.value))
+
+# Peak-crisis:
+onset_date = "2007-02-27"
+plt_df = df[df.Date .== onset_date, :]
+plt = plot(
+    plt_df.variable, plt_df.value;
+    label="", color=:blue,
+    xlabel="Maturity (years)", ylabel="Yield (%)",
+)
+scatter!(
+    plt_df.variable, plt_df.value;
+    label="", color=:blue, alpha=0.5,
+    ylims=(0,6)
+)
+display(plt)
+
+# Post-crisis:
+aftermath_date = "2009-04-20"
+plt_df = df[df.Date .== aftermath_date, :]
+plt = plot(
+    plt_df.variable, plt_df.value;
+    label="", color=:blue,
+    xlabel="Maturity (years)", ylabel="Yield (%)",
+)
+scatter!(
+    plt_df.variable, plt_df.value;
+    label="", color=:blue, alpha=0.5,
+    ylims=(0,6)
+)
+display(plt)
+
+# PCA:
+df_wide = @select(df, Date, variable, value) |>
+    x -> @pivot_wider(x, names_from = variable, values_from = value) |>
+    x -> dropmissing(x)
+X = @select(df_wide, -Date) |> Matrix
+U, Σ, V = svd(X)
+dates = Date.(df_wide.Date)
+tick_years = Date.(unique(Dates.year.(dates)))
+date_tick = Dates.format.(tick_years, "yyyy")
+n_pc = 2
+plt_pc = plot(
+    dates,
+    .-U[:,1:n_pc],
+    label=["PC $i" for i in 1:n_pc] |> permutedims,
+    size=(1000, 300),
+    ylims=(-0.015,0.03),
+    legend=:topright
+)
+plot!(xticks=(tick_years,date_tick), xtickfontsize=6, yaxis=(formatter=y->@sprintf("%.2f",y)))
[…]

### 2024-01-24 src/simple_examples/setup.jl

@@ -1,0 +1,30 @@
+# Set up:
+ENV_DIR = "src/simple_examples"
+using Pkg;
+Pkg.activate(ENV_DIR);
+
+# Packages:
+using CSV
+using ChainPlots
+using DataFrames
+using Dates
+using Flux
+using GLM
+using LinearAlgebra
+using Plots
+using Plots.PlotMeasures
+using Printf
+using TidierData
+using Random
+using Serialization
+using RegressionTables
+Random.seed!(2024)
+
+# Utility functions:
+include("utils.jl")
+
+RESULTS_DIR = "results"
+if !isdir(RESULTS_DIR)
+    mkdir(RESULTS_DIR)
+end
+FIGURE_DIR = joinpath(RESULTS_DIR, "figures")

### 2024-01-24 main.tex
Updates from Overleaf

@@
 Importantly, it need not be fully conscious.
-Rather, theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief system.
+Rather, theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
 Equally importantly, is that theory suggests these behaviors may be motivated or unmotivated: one may selectively seek, attend to, or interpret evidence in favor of a hypothesis for which they may or may not have a personal interest in supporting.
+
+Hypotheses in computer science fields related to AI are often implicit.
+Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
+However with regards to claims that outputs of a system hint towards emerging qualities like AGI, hypotheses necessarily become more explicit: either one interprets outputs as hinting towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
+
+Confirmation bias in hypothesis testing remains relevant and may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews.
+Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
+only considering the likelihood that outputs of system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
+Similarly, bias may show when failing to sufficiently articulate a strong enough competing hypothesis, leading to a 'weak' or 'non-risky' test, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}: if the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
+Thus, one is far more likely to show evidence in favor of their beliefs by posing a weak competing hypothesis.
+
+Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
 

### 2024-01-24 main.tex
Updates from Overleaf

@@
 Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
-However with regards to claims that outputs of a system hint towards emerging qualities like AGI, hypotheses necessarily become more explicit: either one interprets outputs as hinting towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-
-Confirmation bias in hypothesis testing remains relevant and may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews.
+However with regards to claims that outputs of a system hint towards emerging qualities like AGI, hypotheses necessarily become more explicit: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
+
+Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews.
 Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
 only considering the likelihood that outputs of system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-Similarly, bias may show when failing to sufficiently articulate a strong enough competing hypothesis, leading to a 'weak' or 'non-risky' test, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}: if the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
-Thus, one is far more likely to show evidence in favor of their beliefs by posing a weak competing hypothesis.
-
+Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a 'weak' or 'non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
+If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
+Thus, one is far more likely to show evidence in favor of their beliefs by posing a weak null hypotheses.
 Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
+As \cite{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
 
@@
 
-\section{Patterns in Latent Spaces and How to Find Them}
-
-To demonstrate the claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples.
-
-\subsection{Example: Principal Component Analysis}
-
-To understand this better, let us go on a quick detour into economics and look at actual yield curves observed in the US during the Global Financial Crisis (GFC).
-Figure 1 (a) shows the yield curve of US Treasury bonds on 27 February 2007, which according to CNN was a “brutal day on Wall Street”.2 This followed reports on the previous day of former Federal Reserve Chairman Alan Greenspan’s warning that the US economy was at risk of a recession.
-The yield curve was inverted with a sharp negative spread between the 10-year and 3-month yields, indicative of the market’s expectation of a recession.
-
-Figure 1 (b) shows the corresponding yield curve during the aftermath of the GFC on 20 April 2009.
-On that day the influential Time Magazine reported that the “Banking Crisis is Over”.
-The yield curve was steeply sloped with a positive spread between the 10-year and 3-month yields, indicative of the market’s expectation of a recovery.
-The overall level of the yield curve was still very low though, indicative of the fact that US economy had not fully recovered at that point.
-
-Of course, US Treasuries are not the only bonds that are traded in the market.
-To get a more complete picture of the economy, analysts might therefore be interested in looking at the yield curves of other bonds as well.
-In particular, we might be interested in predicting economic growth based on the yield curves of many different bonds.
-The problem with that idea is that it is cursed by high dimensionality: we would end up modelling a single variable of interest (economic growth) with a large number of predictors (the yields of many different bonds).
-To deal with the curse of high dimensionality it can be useful to decompose the yield curves into sets of principal components.
-
-To compute the principal components we can decompose the matrix of yields 
- into a product of its singular vectors and values: 
-.
-I will not go into the details here, because Professor Gilbert Strang has already done a much better job than I ever could in his Linear Algebra lectures.
-To put this into the broader context of the article, however, let us simply refer to
-, 
- and 
- as latent embeddings of the yield curve (they are latent because they are not directly observable).
-
-The top panel in Figure 2 shows the first two principal components of the yield curves of US Treasury bonds over time.
-Vertical stalks indicate the key dates during the onset and aftermath of the crisis, which we discussed above.
-For both components, we can observe some marked shifts between the two dates - but can we attribute any meaning to these shifts?
-It turns out we can: for comparison, the bottom panel in Figure 2 shows the average level and spread of the yield curves over time.
-The first principal component is strongly correlated with the level of the yield curve, while the second principal component is strongly correlated with the spread of the yield curve.
-To put it in AI-lingo:
+\subsection{Patterns in Latent Spaces and How to Find
+Them}\label{patterns-in-latent-spaces-and-how-to-find-them}
+
+To support my claim that observing patterns in latent spaces should not
+generally surprise us, we will now go through a couple of simple
+examples. To illustrate further that this phenomenon is neither
+surprising nor unique to the field of computer science, I will draw on
+my background in economics and finance in this section. We will start
+with very simple examples to demonstrate that even small and simple
+models can learn meaningful representations of the data. The final
+example in \textbf{?@sec-ex-llm} is a bit more involved and closer in
+spirit to the experiments conducted by Gurnee and Tegmark (2023). As we
+go along, we will try to discuss both the benefits and potential
+pitfalls of finding patterns in latent spaces.
+
+\subsubsection{Example: Principal Component
+Analysis}\label{example-principal-component-analysis}
+
+In response to the claims made by Tegmark, numerous commentators on
+social media have pointed out that even the simplest of models can
+exhibit structure in their latent spaces. One of the most popular and
+illustrative examples I remember from my time at the Bank of England is
+yield curve decomposition through PCA. The yield curve is a popular tool
+for investors and economists to gauge the health of the economy. It
+plots the yields of bonds against their maturities. The slope of the
+yield curve is often used as a predictor of future economic activity: a
+steep yield curve is associated with a growing economy, while a flat or
+inverted yield curve is associated with a contracting economy.
+
+% To understand this better, let us go on a quick detour into economics
+% and look at actual yield curves observed in the US during the Global
+% Financial Crisis (GFC). Figure~\ref{fig-gfc-1} shows the yield curve of
+% US Treasury bonds on 27 February 2007, which according to
+% \href{https://money.cnn.com/2007/02/27/markets/markets_0630/index.htm?cnn=yes}{CNN}
+% was a ``brutal day on Wall Street''.\footnote{The
+%   \href{https://home.treasury.gov/resource-center/data-chart-center/interest-rates/TextView?type=daily_treasury_yield_curve&field_tdr_date_value_month=202310}{data}
+%   is taken from the US Department of the Treasury.} This followed
+% \href{https://money.cnn.com/2007/02/26/news/economy/greenspan/index.htm?postversion=2007022609}{reports}
+% on the previous day of former Federal Reserve Chairman Alan Greenspan's
+% warning that the US economy was at risk of a recession. The yield curve
+% was inverted with a sharp negative spread between the 10-year and
+% 3-month yields, indicative of the market's expectation of a recession.
+
+% Figure~\ref{fig-gfc-2} shows the corresponding yield curve during the
+% aftermath of the GFC on 20 April 2009. On that day the influential Time
+% Magazine
+% \href{https://content.time.com/time/business/article/0,8599,1890560,00.html}{reported}
+% that the ``Banking Crisis is Over''. The yield curve was steeply sloped
+% with a positive spread between the 10-year and 3-month yields,
+% indicative of the market's expectation of a recovery. The overall level
+% of the yield curve was still very low though, indicative of the fact
+% that US economy had not fully recovered at that point.
+
+Of course, US Treasuries are not the only bonds that are traded in the
+market. To get a more complete picture of the economy, analysts might
+therefore be interested in looking at the yield curves of other bonds as
+well. In particular, we might be interested in predicting economic
+growth based on the yield curves of many different bonds. The problem
+with that idea is that it is cursed by high dimensionality: we would end
+up modelling a single variable of interest (economic growth) with a
+large number of predictors (the yields of many different bonds). To deal
+with the curse of high dimensionality it can be useful to decompose the
+yield curves into sets of principal components.
+
+To compute the principal components we can decompose the matrix of
+yields \(\mathbf{Y}\) into a product of its singular vectors and values:
+\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). I will not go into
+the details here, because Professor Gilbert Strang has already done a
+much better job than I ever could in his
+\href{https://www.youtube.com/watch?v=mBcLRGuAFUk}{Linear Algebra
+lectures}. To put this into the broader context of the article, however,
+let us simply refer to \(\mathbf{U}\), \(\Sigma\) and
+\(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve (they
+are latent because they are not directly observable).
+
+The top panel in Figure~\ref{fig-pca} shows the first two principal
+components of the yield curves of US Treasury bonds over time. Vertical
+stalks indicate the key dates during the onset and aftermath of the
+crisis, which we discussed above. For both components, we can observe
+some marked shifts between the two dates - but can we attribute any
+meaning to these shifts? It turns out we can: for comparison, the bottom
+panel in Figure~\ref{fig-pca} shows the average level and spread of the
+yield curves over time. The first principal component is strongly
+correlated with the level of the yield curve, while the second principal
+component is strongly correlated with the spread of the yield curve. To
+put it in AI-lingo:
+
+\begin{quote}
+The estimated latent embeddings of the yield curve are characterized by
+patterns observed in the data.
+\end{quote}
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth]{results/figures/pca.png}
+
+}
+
+\caption{\label{fig-pca}Comparison of latent embeddings and observed
+data of the US Treasury yield curve.}
+
+\end{figure*}%
 
 Not convinced? Let us use
- in true autoencoder fashion to reconstruct yield curves from principal components. Let 
- denote the first principal component and consider the following: we keep all other 
- principal components fixed at zero where 
- denotes the total number of maturities; next we traverse the latent space by varying the value of 
- over a fixed grid of length 
- each time storing the full vector 
-; finally, we vertically concatenate the vectors and end up with a matrix 
- of dimension 
-. To reconstruct yields, we simply multiply 
- by the singular values and right singular vectors: 
-.
-
-Figure 3 shows the result of this exercise in the left panel.
-As we can see, our generated yield curves shift vertically as we traverse the latent space.
-The right panel of Figure 3 shows the result of a similar exercise, but this time we keep the first principal component fixed at zero and vary the second principal component.
-This time the slope of our generated yield curves shifts as we traverse the latent space.
-
-\subsection{Example: Deep Learning}
-
-So far we have considered simple matrix decomposition.
-You might argue that principal components are not really latent embeddings in the traditional sense of deep learning.
-To address this, let us now consider a simple deep-learning example.
-Our goal will be to not only predict economic growth from the yield curve but also extract meaningful features at the same time.
-In particular, we will use a neural network architecture that allows us to recover a compressed latent representation of the yield curve.
-
-\subsubsection{Data}
-
-To estimate economic growth we will rely on a quarterly series of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.
-Louis.
-The data arrives in terms of levels of real GDP.
-In order to estimate growth, we will transform the data into log differences.
-Since our yield curve data is daily, we will need to aggregate it to the quarterly frequency.
-To do this, we will simply take the average of the daily yields for each maturity.
-We will also standardize yields since deep learning models tend to perform better with standardized data.
-Since COVID-19 was a huge structural break, we will also filter out all observations after 2018.
-Figure 4 shows the pre-processed data.
-
-\subsubsection{Model}
-
-Let 
- denote growth and 
- denote the yield curve at time 
-. Then we are interested in a model for 
- conditional on 
-. Let 
- denote our model parameters then formally we are interested in maximizing the likelihood 
-.
-To do this, we will use a simple autoencoder architecture that is illustrated in Figure 5.
-The encoder will consist of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer connecting the encoder to the decoder is a fully connected layer with 6 neurons.
-This is the compressed latent representation of the yield curve mentioned above.
-The decoder will consist of two fully connected layers each with a hyperbolic tangent activation function: the first layer will consist of 32 neurons and the second layer will have the same dimension as the input data.
-The output layer will consist of a single neuron with a linear activation function.
-The model will be trained using the mean squared error loss function and the Adam optimizer.
-
-\subsubsection{Linear Probe}
-
-The results are shown in Figure 6.
-The top panel shows the actual GDP growth and fitted values from the autoencoder model.
-We observe that the model captures the relationship between economic growth and the yield curve reasonably well.
-As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
-
-The bottleneck layer seems like a good place to start looking. To get the latent embeddings 
- at time 
-, we simply pass the yield curve data 
- through the encoder. Next, we follow [TEGMARK PAPER REFERENCE] and use a linear probe to regress the observed yield curve factors on the latent embeddings. Let 
- denote the vector containing the two factors of interest in time 
-: 
- and 
-. Formally, we are interested in the following regression model: 
- where 
- denotes the regression parameters. Following [TEGMARK PAPER REFERENCE] we use Ridge regression with 
- set to 
-. Using the estimated regression parameters 
- we can then predict the yield curve factors from the latent embeddings: 
-.
-
-The results of this experiment are shown in the bottom panel of Figure 6.
-Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
-We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-To put this in true hype-lingo:
-
-Not just a parrot! Our autoencoder neural network has an implicit understanding of the yield curve factors and their relationship with economic growth. It’s sentient indeed!
-
-\section{Example: MNIST}
-
-Leveraging a small MLP and VAE trained on MNIST data, this script trains a tokenizer that encodes geographical meaning in the latent embedding of VAE.
-Inspired by this tweet: https://x.com/savvyRL/status/1709698089500680264?s=20
-
-https://gist.github.com/pat-alt/1c3d0d159b2cfa2a91a8d0c47ea3d28a
-
-This tweet inspired me. Just as I had feared, even the small VAE I trained on MNIST a while ago is basically sentient. It is so over people!
-
-\section{Ok, but truly what’s the point?}
-
-The finding is not surprising but it is still interesting.
-In the context of mechanistic interpretability, it demonstrates that the black-box model has evidently learned plausible explanations for the data.
-Beyond that, in this particular example, the patterns in the latent space that we have just uncovered might actually be useful for downstream tasks.
-An interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
-To demonstrate this, let us consider a simple linear regression model for GDP growth.
-We will compare the performance of the following models: (1) regressing growth
- on the latent embedding 
-, (2) regression growth on the best subset of latent embeddings 
-, (3) regressing growth on lagged growth, (4) regressing growth on lagged growth and the observed yield curve factors 
-, (5) regressing growth on the best subset of yield curve factors 
-, and, finally, (5) regressing growth on lagged growth and the best subset of latent embeddings 
-.
-
-The results are shown in the table below.
-The key finding of interest is that the coefficients on the latent embeddings in model (5) are statistically significant.
-This suggests that the latent embeddings contain information that is useful for predicting GDP growth.
-
-\section{LLM}
-
-These results from the linear probe shown in @fig-fomc are certainly not unimpressive: even though *FOMC-RoBERTa* was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation.
-It is worth pointing out here that this model is substantially smaller than the models tested in @gurnee2023language.
-This begs the following question:
-
-> Have we uncovered further evidence that LLMs "aren't mere stochastic parrots"? Has *FOMC-RoBERTa* developed an intrinsic understanding of the economy just by 'reading' central bank communications?
-
-We find this very difficult to believe and we will now produce a counter-example to demonstrate that, if anything, these findings are very much in line with the parrot metaphor.
-If the results from the linear probe truly were indicative of some intrinsic understanding of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.
-
-To test this, we select the best-performing probe trained on the final-layer activations to predict changes in the CPI.
-We then make up sentences that fall into one of these four categories: *Inflation/Prices* (IP)---sentences about price inflation, *Deflation/Prices* (DP)---sentences about price deflation, *Inflation/Birds* (IB)---sentences about an inflation in the number of birds and *Deflation/Birds* (DB)---sentences about a deflation in the number of birds.
-A sensible sentence for category DP, for example, could be: "It is essential to bring inflation back to target to avoid drifting into deflation territory.".
-Analogically, we could construct the following sentence for the DB category: "It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.".
-
-In light of the encouraging results for the probe in @fig-fomc, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category than for sentences in the DP category.
-If this is indicative of true intrinsic understanding, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.
-
-To get to this last point, we also generate many probe predictions for samples of noise.
-Let $f: \mathcal{A}^k \mapsto \mathcal{Y}$ denote the linear probe that maps from the $k$-dimensional space spanned by $k$ first principal components of the final-layer activations to the output variable of interest (CPI growth).
-Then we sample $\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})$ for $i \in [1,1000]$ and compute the sample average.
-We repeat this process $10000$ times and compute the median-of-means to get an estimate for $\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]$, that is the predicted value of the probe conditional on white noise.
-
-We propose the following hypothesis test to assess if the probe results provide evidence for an actual understanding of key economic relationships learned purely from text:
-
-- *H0* (Null): The probe never predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$
-- *H1* (Stochastic Parrots): The probe predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$ for sentences in all categories (IP,DP,IB,DB).
-- *H2* (More than Mere Stochastic Parrots): The probe predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$ for sentences in categories IP and DP, but not for sentences in IB and DB.
+\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\) in true autoencoder
+fashion to reconstruct yield curves from principal components. Let
+\(z_1\) denote the first principal component and consider the following:
+we keep all other \(M-1\) principal components fixed at zero where \(M\)
+denotes the total number of maturities; next we traverse the latent
+space by varying the value of \(z_1\) over a fixed grid of length \(K\)
+each time storing the full vector \(\mathbf{z}\); finally, we vertically
+concatenate the vectors and end up with a matrix \(\mathbf{Z}\) of
+dimension \((K \times M)\). To reconstruct yields, we simply multiply
+\(Z\) by the singular values and right singular vectors:
+\(\mathbf{Y}=\mathbf{Z}\Sigma\mathbf{V}^{\prime}\).
+
+\subsubsection{Example: Deep Learning}\label{example-deep-learning}
+
+So far we have considered simple matrix decomposition. You might argue
+that principal components are not really latent embeddings in the
+traditional sense of deep learning. To address this, let us now consider
+a simple deep-learning example. Our goal will be to not only predict
+economic growth from the yield curve but also extract meaningful
+features at the same time. In particular, we will use a neural network
+architecture that allows us to recover a compressed latent
+representation of the yield curve.
+
+\paragraph{Data}\label{data}
+
+To estimate economic growth we will rely on a quarterly
+\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real
+gross domestic product (GDP) provided by the Federal Reserve Bank of
+St.~Louis. The data arrives in terms of levels of real GDP. In order to
+estimate growth, we will transform the data into log differences. Since
+our yield curve data is daily, we will need to aggregate it to the
+quarterly frequency. To do this, we will simply take the average of the
+daily yields for each maturity. We will also standardize yields since
+deep learning models tend to perform better with standardized data.
+Since COVID-19 was a huge structural break, we will also filter out all
+observations after 2018.
+
+\paragraph{Model}\label{model}
+
+Let \(G_t\) denote growth and \(\mathbf{Y}_t\) denote the yield curve at
+time \(t\). Then we are interested in a model for \(G_t\) conditional on
+\(\mathbf{Y}_t\). Let \(\theta\) denote our model parameters then
+formally we are interested in maximizing the likelihood
+\(p_{\theta}(G_t|\mathbf{Y}_t)\).
+
+To do this, we will use a simple
+autoencoder architecture: the encoder will consist of a single fully
+connected hidden layer with 32 neurons and a hyperbolic tangent
+activation function. The bottleneck layer connecting the encoder to the
+decoder is a fully connected layer with 6 neurons. This is the
+compressed latent representation of the yield curve mentioned above. The
+decoder will consist of two fully connected layers each with a
+hyperbolic tangent activation function: the first layer will consist of
+32 neurons and the second layer will have the same dimension as the
+input data. The output layer will consist of a single neuron with a
+linear activation function. The model will be trained using the mean
+squared error loss function and the Adam optimizer.
+
+\paragraph{Linear Probe}\label{linear-probe}
+
+The results are shown in Figure~\ref{fig-dl-results}. The top panel
+shows the actual GDP growth and fitted values from the autoencoder
+model. We observe that the model captures the relationship between
+economic growth and the yield curve reasonably well. As discussed above,
+we also know that the relationship between economic growth and the yield
+curve is characterized by two main factors: the level and the spread.
+Since the model itself is fully characterized by its parameters, we
+would expect that these two important factors are reflected somewhere in
+the latent parameter space.
+
+The bottleneck layer seems like a good place to start looking. To get
+the latent embeddings \(\mathbf{A}_t\) at time \(t\), we simply pass the
+yield curve data \(\mathbf{Y}_t\) through the encoder. Next, we follow
+Gurnee and Tegmark (2023) and use a linear probe to regress the observed
+yield curve factors on the latent embeddings. Let \(F_t\) denote the
+vector containing the two factors of interest in time \(t\):
+\(f_t^{\text{spread}}\) and \(f_t^{\text{level}}\). Formally, we are
+interested in the following regression model:
+\(p_{w}(F_t|\mathbf{A}_t)\) where \(w\) denotes the regression
+parameters. Following Gurnee and Tegmark (2023), we use Ridge regression
+with \(\lambda\) set to \(0.1\). Using the estimated regression
+parameters \(\hat{w}\) we can then predict the yield curve factors from
+the latent embeddings: \(\hat{F}_t=\hat{w}^{\prime}\mathbf{A}_t\).
+
+The results of this experiment are shown in the bottom panel of
+Figure~\ref{fig-dl-results}. Solid lines show the observed yield curve
+factors over time, while dashed lines show predicted values. We find
+that the latent embeddings predict the two yield curve factors
+reasonably well, in particular the spread. To put this in true
+hype-lingo:
+
+\begin{quote}
+Not just a parrot! Our autoencoder neural network has an implicit
+understanding of the yield curve factors and their relationship with
+economic growth. It's sentient indeed!
+\end{quote}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth]{results/figures/dl.png}
+
+}
+
+\caption{\label{fig-dl-results}The top panel shows the actual GDP growth
+and fitted values from the autoencoder model. The bottom panel shows the
+observed average level and spread of the yield curve (solid) along with
+the predicted values from the linear probe based on the latent embedding
+(dashed).}
+
+\end{figure}%
+
+\paragraph{Ok, but truly what's the
+point?}\label{ok-but-truly-whats-the-point}
+
+The finding is not surprising but it is still interesting. In the
+context of mechanistic interpretability, it demonstrates that the
+black-box model has evidently learned plausible explanations for the
+data. Beyond that, in this particular example, the patterns in the
+latent space that we have just uncovered might actually be useful for
+downstream tasks. An interesting idea could be to use the latent
+embeddings as features in a more traditional and interpretable
+econometric model. To demonstrate this, let us consider a simple linear
+regression model for GDP growth. We will compare the performance of the
+following models: (1) regressing growth \(G_t\) on the latent embedding
+\(A_t\), (2) regression growth on the best subset of latent embeddings
+\(A_t\), (3) regressing growth on lagged growth, (4) regressing growth
+on lagged growth and the observed yield curve factors \(F_t\), (5)
+regressing growth on the best subset of yield curve factors \(F_t\),
+and, finally, (5) regressing growth on lagged growth and the best subset
+of latent embeddings \(A_t\).
+
+The results are shown in the table below. The key finding of interest is
+that the coefficients on the latent embeddings in model (5) are
+statistically significant. This suggests that the latent embeddings
+contain information that is useful for predicting GDP growth.
+
+\input{results/regression.tex}
+
+\subsubsection{Example: Large Language Model}\label{ex-llm}
+
+To round up this section, we will jump back on the hype train and
+consider an example involving an LLM. In particular, we will closely
+follow the approach in Gurnee and Tegmark (2023) and apply it to a novel
+financial dataset: the \emph{Trillion Dollar Words} dataset introduced
+by Shah, Paturi, and Chava (2023). The dataset contains a curated
+selection of sentences formulated by central bankers of the US Federal
+Reserve and communicated to the public in speeches, meeting minutes and
+press conferences. The authors of the paper use this dataset to train
+LLMs to classify sentences as either `dovish', `hawkish' or `neutral'.
+To this end, they first manually annotate a subsample of the available
+data and then fine-tune various foundation models. Their model of
+choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa (Liu et al.
+2019)), achieves an \(F_1\) score of around \(>0.7\) for the
+classification task. To illustrate the potential usefulness of the
+learned classifier, they use predicted labels for the entire dataset to
+compute an ad-hoc, count-based measure of `hawkishness'. They then go on
+to show that this measure correlates with key economic indicators in the
+expected direction: when inflationary pressures rise, the measured level
+of `hawkishness' increases as central bankers need to raise interest
+rates to bring inflation back to target.
+
+\paragraph{Linear Probes}\label{linear-probes}
+
+Instead of computing a measure based on predicted labels, we can use
+linear probes to assess if the fine-tuned model has learned associative
+patterns between central bank communications and key economic
+indicators. To this end, I have further pre-processed the data provided
+by Shah, Paturi, and Chava (2023) and used their proposed model to
+compute layer-wise embeddings for all available sentences. I have made
+these available and easily accessible through a small Julia package:
+\href{https://github.com/pat-alt/TrillionDollarWords.jl}{TrillionDollarWords.jl}.
+For each layer, I have then computed linear probes on two inflation
+indicators---the Consumer Price Index (CPI) and the Producer Price Index
+(PPI)---as well as US Treasury yields at different levels of maturity.
+To mitigate issues related to over-parameterization, I follow the
+recommendation in Alain and Bengio (2018) to first reduce the
+dimensionality of the embeddings each time. In particular, linear probes
+are restricted to the first 128 principal components of the embeddings
+of each layer.
+
+Detailed results of this experiment will be released in an upcoming
+paper. Figure~\ref{fig-fomc} highlights the result for the linear probe
+on the CPI. The chart shows various performance measures plotted against
+\emph{FOMC-RoBERTa}'s \(n\)-th layer. Shaded areas show the variation
+across cross-validation folds, where we have used an expanding window
+approach to split the time series. To avoid look-ahead bias, we run PCA
+separately for each training set. Results for the linear probes are
+shown along with results for autoregressive models (AR) used as our
+baseline.
+
+The first column shows the correlations between model predictions and
+observed inflation. One can observe that this correlation is strictly
+positive for the baseline models and the linear probes. Consistent with
+the findings in Gurnee and Tegmark (2023) and Alain and Bengio (2018),
+we also observe that this correlation tends to be higher for layers near
+the end of the transformer model. Those layers produce predictions that
+are more positively correlated with observed inflation than those
+produced by the corresponding autoregressive models.
+
+Similarly, we observe that the (root) mean squared error of the linear
+probe is largely on par with the baseline. The average prediction error
+is gradually reduced as we move along the horizontal axis, which is
+again indicative of the notion that layers closer to the end of the
+neural network have distilled more useful representations.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth]{results/figures/measures_CPI (n_pc=128).png}
+
+}
+
+\caption{\label{fig-fomc}Various performance measures (correlation and
+(root) mean squared error) plotted against \emph{FOMC-RoBERTa}'s
+\(n\)-th layer. Shaded areas indicate variation across cross-validation
+folds, where we have used an expanding window approach to split the time
+series. Results for the linear probes are shown along with results for
+autoregressive models (AR). The optimal lag length was determined using
+the Bayes Information Criterium.}
+
+\end{figure}%
+
+\paragraph{Stochastic Parrots After
+All?}\label{stochastic-parrots-after-all}
+
+These results from the linear probe shown in Figure~\ref{fig-fomc} are
+certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not
+explicitly trained to uncover associations between central bank
+communications and the level of consumer prices, it appears that the
+model has distilled representations that can be used to predict
+inflation. It is worth pointing out here that this model is
+substantially smaller than the models tested in Gurnee and Tegmark
+(2023). This begs the following question:
+
+\begin{quote}
+Have we uncovered further evidence that LLMs ``aren't mere stochastic
+parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic understanding
+of the economy just by `reading' central bank communications?
+\end{quote}
+
+Personally, I am having a very hard time believing this. To argue my
+case, I will now produce a counter-example demonstrating that, if
+anything, these findings are very much in line with the parrot metaphor.
+The counter-example is based on the following premise: if the results
+from the linear probe truly were indicative of some intrinsic
+understanding of the economy, then the probe should not be sensitive to
+random sentences that are most definitely not related to consumer
+prices.
+
+To test this, I select the best-performing probe trained on the
+final-layer activations to predict changes in the CPI. I then make up
+sentences that fall into one of these four categories:
+\emph{Inflation/Prices} (IP)---sentences about price inflation,
+\emph{Deflation/Prices} (DP)---sentences about price deflation,
+\emph{Inflation/Birds} (IB)---sentences about an inflation in the number
+of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation
+in the number of birds. A sensible sentence for category DP, for
+example, could be: ``It is essential to bring inflation back to target
+to avoid drifting into deflation territory.''. Analogically, we could
+construct the following sentence for the DB category: ``It is essential
+to bring the numbers of doves back to target to avoid drifting into
+dovelation territory.''.
+
+In light of the encouraging results for the probe in
+Figure~\ref{fig-fomc}, we should expect the probe to predict higher
+levels of inflation for activations for sentences in the IP category
+than for sentences in the DP category. If this was indicative of true
+intrinsic understanding, we would not expect to see any significant
+difference in predicted inflation levels for sentences about birds,
+independent of whether or not their numbers are increasing. More
+specifically, we would not expect the probe to predict values for
+sentences about birds that are substantially different from the values
+it can be expected to predict when using actual white noise as inputs.
+
+To get to this last point, I also generate many probe predictions for
+samples of noise. Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote
+the linear probe that maps from the \(k\)-dimensional space spanned by
+\(k\) first principal components of the final-layer activations to the
+output variable of interest (CPI growth in this case). Then I sample
+\(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\)
+for \(i \in [1,1000]\) and compute the sample average. I repeat this
+process \(10000\) times and compute the median-of-means to get an
+estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\),
+that is the predicted value of the probe conditional on white noise.
+
+Next, I propose the following hypothesis test as a minimum viable
+testing framework to assess if the probe results (may) provide evidence
+for an actual understanding of key economic relationships learned purely
+from text:
+
+\begin{proposition}[Parrot
+Test]\protect\hypertarget{prp-line}{}\label{prp-line}
+
+~
+
+\begin{itemize}
+\tightlist
+\item
+  \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are
+  statistically significantly different from
+  \(\mathbb{E}[f(\varepsilon)]\)
+\item
+  \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts
+  values that are statistically significantly different from
+  \(\mathbb{E}[f(\varepsilon)]\) for sentences in all categories
+  (IP,DP,IB,DB).
+\item
+  \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The
+  probe predicts values that are statistically significantly different
+  from \(\mathbb{E}[f(\varepsilon)]\) for sentences in categories IP and
+  DP, but not for sentences in IB and DB.
+\end{itemize}
+
+\end{proposition}
+
+To be clear, if in such a test we did find substantial evidence in
+favour of rejecting both \emph{HO} and \emph{H1}, this would not
+automatically imply that \emph{H2} is true. But to even continue
+investigating if based on having learned meaningful representation the
+underlying LLM is more than just a parrot, it should be able to pass
+this simple test.
+
+In this particular case, Figure~\ref{fig-attack} demonstrates that we
+find some evidence to reject \emph{H0} but not \emph{H1} for
+\emph{FOMC-RoBERTa}. The median linear probe predictions for sentences
+about inflation and deflation are indeed substantially higher and lower,
+respectively than for random noise. Unfortunately, the same is true for
+sentences about the inflation and deflation in the number of birds,
+albeit to a somewhat lower degree.
+
+I should note that the number of sentences in each category is very
+small here (10), so the results in Figure~\ref{fig-attack} cannot be
+used to establish statistical significance. That being said, even a
+handful of convincing counter-examples should be enough for us to
+seriously question the claim that results from linear probes provide
+evidence against the parrot metaphor. In fact, if there's even a single
+sentence for which any
+
+\begin{quote}
+Strong alternative hypothesis to address own confirmation bias
+Articulating a test to better differentiate
+\end{quote}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth]{results/figures/attack_inflation.png}
+
+}
+
+\caption{\label{fig-attack}Probe predictions for sentences about
+inflation of prices (IP), deflation of prices (DP), inflation of birds
+(IB) and deflation of birds (DB). The vertical axis shows predicted
+inflation levels subtracted by the average predicted inflation level for
+the whole sample period (train and test set).}
+
+\end{figure}%
 

### 2024-01-24 src/simple_examples/autoencoder/regression.jl
added one more model to regression table

@@ -63,13 +63,13 @@ regtable(
     labels=labels,
 )
 regtable(
-    ols_factors, ols_best;
+    ols_ar, ols_factors, ols_best;
     renderSettings=RegressionTables.htmlOutput(joinpath(RESULTS_DIR, "regression.html")),
     regression_statistics=reg_stats,
     labels=labels,
 )
 regtable(
-    ols_factors, ols_best;
+    ols_ar, ols_factors, ols_best;
     renderSettings=RegressionTables.latexOutput(joinpath(RESULTS_DIR, "regression.tex")),
     regression_statistics=reg_stats,
     labels=labels,

### 2024-01-24 src/simple_examples/autoencoder/run_experiment.jl

@@ -1,3 +1,5 @@
+include("../setup.jl")
+
 # Just plot the time series:
 include("plot_ts.jl")
 

### 2024-01-24 src/simple_examples/pca/run_experiment.jl

@@ -1,42 +1,4 @@
-df = CSV.read("data/ust_yields.csv", DataFrame) |>
-    x -> @pivot_longer(x, -Date) |>
-    x -> @mutate(x, variable=to_year(variable)) |>
-    x -> @mutate(x, year=Dates.year(Date)) |>
-    x -> @mutate(x, quarter=Dates.quarter(Date)) |>
-    x -> @mutate(x, Date=Dates.format(Date, "yyyy-mm-dd")) |>
-    x -> @arrange(x, Date) |>
-    x -> @fill_missing(x, "down")
-ylims = extrema(skipmissing(df.value))
-
-# Peak-crisis:
-onset_date = "2007-02-27"
-plt_df = df[df.Date .== onset_date, :]
-plt = plot(
-    plt_df.variable, plt_df.value;
-    label="", color=:blue,
-    xlabel="Maturity (years)", ylabel="Yield (%)",
-)
-scatter!(
-    plt_df.variable, plt_df.value;
-    label="", color=:blue, alpha=0.5,
-    ylims=(0,6)
-)
-display(plt)
-
-# Post-crisis:
-aftermath_date = "2009-04-20"
-plt_df = df[df.Date .== aftermath_date, :]
-plt = plot(
-    plt_df.variable, plt_df.value;
-    label="", color=:blue,
-    xlabel="Maturity (years)", ylabel="Yield (%)",
-)
-scatter!(
-    plt_df.variable, plt_df.value;
-    label="", color=:blue, alpha=0.5,
-    ylims=(0,6)
-)
-display(plt)
+include("../setup.jl")
 
 # PCA:
 df_wide = @select(df, Date, variable, value) |>

### 2024-01-24 src/simple_examples/setup.jl

@@ -28,3 +28,43 @@ if !isdir(RESULTS_DIR)
     mkdir(RESULTS_DIR)
 end
 FIGURE_DIR = joinpath(RESULTS_DIR, "figures")
+
+df = CSV.read("data/ust_yields.csv", DataFrame) |>
+     x -> @pivot_longer(x, -Date) |>
+          x -> @mutate(x, variable = to_year(variable)) |>
+               x -> @mutate(x, year = Dates.year(Date)) |>
+                    x -> @mutate(x, quarter = Dates.quarter(Date)) |>
+                         x -> @mutate(x, Date = Dates.format(Date, "yyyy-mm-dd")) |>
+                              x -> @arrange(x, Date) |>
+                                   x -> @fill_missing(x, "down")
+ylims = extrema(skipmissing(df.value))
+
+# Peak-crisis:
+onset_date = "2007-02-27"
+plt_df = df[df.Date.==onset_date, :]
+plt = plot(
+    plt_df.variable, plt_df.value;
+    label="", color=:blue,
+    xlabel="Maturity (years)", ylabel="Yield (%)",
+)
+scatter!(
+    plt_df.variable, plt_df.value;
+    label="", color=:blue, alpha=0.5,
+    ylims=(0, 6)
+)
+display(plt)
+
+# Post-crisis:
+aftermath_date = "2009-04-20"
+plt_df = df[df.Date.==aftermath_date, :]
+plt = plot(
+    plt_df.variable, plt_df.value;
+    label="", color=:blue,
+    xlabel="Maturity (years)", ylabel="Yield (%)",
+)
+scatter!(
+    plt_df.variable, plt_df.value;
+    label="", color=:blue, alpha=0.5,
+    ylims=(0, 6)
+)
+display(plt)

### 2024-01-24 main.tex

@@
 A common issue is a lack of expectation that the patterns that hint towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
+This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \cite{nickerson1998confirmation}) or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
 
@@
 
-\subsection{Patterns in Latent Spaces and How to Find
+\section{Patterns in Latent Spaces and How to Find
 Them}\label{patterns-in-latent-spaces-and-how-to-find-them}

### 2024-01-25 src/probe_llm/experiments.jl
decided on chart for probes

@@ -1,198 +1,52 @@
-using AlgebraOfGraphics
-using CairoMakie
-using CSV
-using DataFrames
-using Dates
-import Plots
-using Statistics
-using StatsBase: cor, Weights
-using Transformers
-using TrillionDollarWords
-
-include("utils.jl")
-include("run_models.jl")
-include("evaluation.jl")
-include("plotting.jl")
-include("attack_probe.jl")
-
-# All data:
-all_data = load_all_data()
-
-# Setup:
-save_dir = "results"
-interim_dir = joinpath(save_dir, "interim")
-ispath(interim_dir) || mkdir(interim_dir)
-all_saved = sort(parse.(Int, filter.(isdigit, readdir(interim_dir))))
-last_saved = length(all_saved) > 0 ? maximum(all_saved) : 0
-
-# Parameter grid:
-n_pcs = [nothing, 128]
-use_head = [false, true]
-indicator = ["PPI", "CPI", "UST"]
-maturity = ["1 Mo", "1 Yr", "10 Yr"]
-layer = 1:24
-grids = []
-for ind in indicator
-    for (i,_head) in enumerate(use_head)
-        _mat = ind != "UST" ? [missing] : maturity
-        _layer = _head ? [25] : layer
-        grid = Base.Iterators.product(_head, _mat, _layer, n_pcs) |> 
-            collect |>
-            x -> DataFrame(vec(x), [:use_head, :maturity, :layer, :n_pc])
-        grid.indicator .= ind
-        push!(grids, grid)
-    end
-end
-grid = vcat(grids...)
+include("setup.jl")
 
 # Run the models:
-results = []
-for (i, row) in enumerate(eachrow(grid))
-    if i <= last_saved
-        _results = CSV.read(joinpath(interim_dir, "results_$i.csv"), DataFrame)
-    else
-        println("Running models for experiment $i of $(nrow(grid))")
-        _results = run_models(
-            all_data; 
-            indicator=row.indicator, 
-            maturity=row.maturity, 
-            layer=row.layer, 
-            use_head=row.use_head,
-            n_pc=row.n_pc,
-        )
-        CSV.write(joinpath(interim_dir, "results_$i.csv"), _results)
-    end
-    push!(results, _results)
+if RUN_MODELS
+    include("run_models.jl")
 end
-results = vcat(results...)
-
-# Save the results:
-CSV.write(
[…]

### 2024-01-25 src/probe_llm/plotting.jl

@@ -1,87 +1,0 @@
-function plot_ts(
-    agg_data::DataFrame;
-    alpha=[0.2 1.0 1.0],
-    legend=:topleft,
-    xlabel="Year",
-    kwrgs...
-)
-    
-    # Collect metadata:
-    indicator = agg_data.indicator[1]
-    maturity = agg_data.maturity[1]
-    layer = agg_data.layer[1]
-    p = agg_data.p[1]
-
-    # Aggregate the results:
-    ts = groupby(agg_data, [:ym, :model]) |>
-        x -> combine(
-            x,
-            :y .=> mean => :y,
-            [:yhat, :nrow] => ((y, w) -> mean(y, Weights(w))) => :yhat
-        ) |>
-        x -> unstack(x, [:ym, :y], :model, :yhat) |>
-        x -> select(x, Not([:ym])) |>
-        Matrix
-
-    # Collect the dates:
-    dates = [Date(y,m) for (y,m) in unique(agg_data.ym)]
-
-    # Plot the results:
-    measure_name = indicator
-    if !ismissing(maturity)
-        measure_name = "$measure_name ($maturity)"
-    end
-    plt = Plots.plot(
-        dates,
-        ts;
-        label=[measure_name "AR($p)" "Probe (layer $layer)"],
-        ylabel=measure_name,
-        alpha=alpha,
-        legend=legend,
-        xlabel=xlabel,
-        kwrgs...
-    )
-    return plt
-end
-
-"""
-    plot_measures(
-        df::DataFrame;
-        axis=(width=225, height=225)
-    )
-
-Plot the evaluation measures against the layer number for a set of models.
-"""
-function plot_measures(
-    df::DataFrame;
-    axis=(width=225, height=225)
-)
-
-    # Title:
-    indicator = df.indicator[1]
-    maturity = df.maturity[1]
-    if !ismissing(maturity)
-        title = "$indicator ($maturity)"
-    else
-        title = "$indicator"
-    end
-
-    # Upper and lower bounds:
-    df = transform(df, [:value, :std] => ByRow((x, s) -> (x + s, x - s)) => [:lb, :ub])
-    df = df[.!isnan.(df.value), :]
-    plt = data(df) * mapping(
-        :layer, :value, 
-        lower=:lb, upper=:ub,
-        row=:split,
-        col=:variable, 
-        color=:model => x -> x=="y_bl" ? "AR" : "Probe",
-    )
[…]

### 2024-01-25 src/probe_llm/run_attacks.jl

@@ -1,0 +1,102 @@
+tfm = load_model(; load_head=false)
+n_pc = 128
+layer = 24
+ispath(joinpath(save_dir, "attacks")) || mkpath(joinpath(save_dir, "attacks"))
+attack_sentence_dir = "results/attacks/sentences"
+ispath(attack_sentence_dir) || mkpath(attack_sentence_dir)
+use_all = false
+prefx = use_all ? "all_" : ""
+compute_embeddings = true
+
+high_inf_text = readlines(joinpath(attack_sentence_dir, "$(prefx)high_inf.txt"))[1]
+high_inf_query = split(high_inf_text, ";") |>
+                 x -> String.(x)
+
+hawk_text = readlines(joinpath(attack_sentence_dir, "$(prefx)hawk.txt"))[1]
+hawk_query = split(hawk_text, ";") |>
+             x -> String.(x)
+
+low_inf_text = readlines(joinpath(attack_sentence_dir, "$(prefx)low_inf.txt"))[1]
+low_inf_query = split(low_inf_text, ";") |>
+                x -> String.(x)
+
+dove_text = readlines(joinpath(attack_sentence_dir, "$(prefx)dove.txt"))[1]
+dove_query = split(dove_text, ";") |>
+             x -> String.(x)
+
+queries = zip([high_inf_query, hawk_query, low_inf_query, dove_query], ["high_inf", "hawk", "low_inf", "dove"])
+
+# Embed the queries:
+queries_embedded = []
+if compute_embeddings
+    Threads.@threads for (q, cat) in collect(queries)
+        println("Embedding queries for $cat on thread $(Threads.threadid())")
+        embedding = []
+        for i in collect(eachindex(q))
+            push!(embedding, embed_text(tfm, [q[i]]))
+            println("$i/$(length(q)) done on thread $(Threads.threadid())")
+        end
+        embedding = vcat(embedding...)
+        push!(queries_embedded, (q, embedding, cat))
+    end
+    for (query, embedding, name) in queries_embedded
+        CSV.write(joinpath(save_dir, "attacks", "$(prefx)attack_$name.csv"), DataFrame(embedding, :auto))
+    end
+else
+    for (query, name) in queries
+        embedding = CSV.read(joinpath(save_dir, "attacks", "$(prefx)attack_$name.csv"), DataFrame) |> Matrix
+        push!(queries_embedded, (query, embedding, name))
+    end
+end
+
[…]

### 2024-01-25 src/probe_llm/run_models.jl

@@ -1,118 +1,48 @@
-function run_models(
-    all_data; 
-    indicator="PPI", 
-    maturity=missing, 
-    layer=24, 
-    use_head=false,
-    n_folds::Int=5,
-    λ::Real=0.1,
-    min_train_prp::Float64=0.75,
-    n_pc::Union{Nothing,Int}=nothing,
-    return_meta::Bool=false,
-)
+# Parameters
+n_pcs = [nothing, 128]
+use_head = [false, true]
+indicator = ["PPI", "CPI", "UST"]
+maturity = ["1 Mo", "1 Yr", "10 Yr"]
+layer = 1:24
+grids = []
 
-    # Prepare market data:
-    mkt_data = prepare_mkt_data(all_data, indicator, maturity) |>
-        x -> sort(x, [:ym, :event_type, :speaker])
-    time_stamps = unique(sort(mkt_data.ym))
-    ts_splits = time_series_split(
-        time_stamps; 
-        n_folds=n_folds, 
-        return_vals=true, 
-        min_train_prp=min_train_prp
-    ) |> collect
-
-    # Prepare the data for the probe:
-    if use_head
-        X_probe, y_probe = prepare_probe(mkt_data)
-        model = load_model(; use_head=use_head)
-        clf = model.mod.cls
-        X_probe = vcat([get_clf_activations(clf, x)' for x in eachrow(X_probe)]...)
-    else
-        X_probe, y_probe = prepare_probe(mkt_data; layer=layer)
+# Set up a grid of parameters:
+for ind in indicator
+    for (i, _head) in enumerate(use_head)
+        _mat = ind != "UST" ? [missing] : maturity
+        _layer = _head ? [25] : layer
+        grid = Base.Iterators.product(_head, _mat, _layer, n_pcs) |>
+               collect |>
+               x -> DataFrame(vec(x), [:use_head, :maturity, :layer, :n_pc])
+        grid.indicator .= ind
+        push!(grids, grid)
     end
-
-    mkt_data.layer .= layer
-
-    # Run models for each split:
-    res = []
-    probes = []
-    probe_data = []
-    for (i, (train, test)) in enumerate(ts_splits)
-        _mkt_data = deepcopy(mkt_data) |>
-            x -> sort(x, [:sentence_id, :ym, :event_type, :speaker])
-        # Run the baseline on the aggregated data:
-        agg_data = groupby(_mkt_data, :ym) |>
[…]

### 2024-01-25 src/probe_llm/setup.jl

@@ -1,0 +1,27 @@
+using AlgebraOfGraphics
+using CairoMakie
+using CSV
+using DataFrames
+using Dates
+import Plots
+using Statistics
+using StatsBase: cor, Weights
+using Transformers
+using TrillionDollarWords
+
+# Utility functions:
+include("utilities/core.jl")
+include("utilities/models.jl")
+include("utilities/evaluation.jl")
+include("utilities/plotting.jl")
+include("utilities/attack_probe.jl")
+
+# All data:
+all_data = load_all_data()
+
+# Directories:
+save_dir = "results"
+interim_dir = joinpath(save_dir, "interim")
+ispath(interim_dir) || mkdir(interim_dir)
+all_saved = sort(parse.(Int, filter.(isdigit, readdir(interim_dir))))
+last_saved = length(all_saved) > 0 ? maximum(all_saved) : 0

### 2024-01-25 src/probe_llm/utilities/attack_probe.jl

@@ -76,40 +76,7 @@ function embedding_to_probe(
     return yhat
 end
 
-"""
-    plot_attack(df_pred::DataFrame)
 
-Plot the attack results.
-"""
-function plot_attack(df_pred::DataFrame)
-
-    if length(unique(df_pred.indicator)) == 1
-        _map = mapping(
-            :dir => "Direction",
-            :level => "f(x) - 𝔼[f(ϵ)]",
-            color=:topic => "Topic",
-            dodge=:topic => "Topic",
-        )
-    else
-        _map = mapping(
-            col = :indicator,
-            :dir => "Direction",
-            :level => "f(x) - 𝔼[f(ϵ)]",
-            color=:topic => "Topic",
-            dodge=:topic => "Topic",
-        )
-    end
-
-    df_plt = data(df_pred) * _map
-    layers = visual(BoxPlot)
-    box_plt = layers * df_plt
-    hline_plt = mapping([0],[0]) * visual(ABLines)
-    plt = draw(
-        hline_plt + box_plt,
-        facet=(; linkyaxes=:none),
-        axis=(width=300, height=300)
-    )
-end
 
 
 

### 2024-01-25 src/probe_llm/utilities/core.jl

@@ -73,6 +73,34 @@ Compute the root mean squared error of a model `mod` on the output `y` and predi
 rmse(y, ŷ; weights::Union{Nothing,AbstractArray}=nothing) = sqrt(mse(y, ŷ; weights=weights))
 
 """
+    mape(mod::Model, X, y)
+
+Compute the mean absolute percentage error of a model `mod` on the input `X` and output `y`.
+"""
+mape(mod::Model, X, y) = mape(y, predict(mod, X))
+
+"""
+    mape(y, ŷ)
+
+Compute the mean absolute percentage error of a model `mod` on the output `y` and predicted output `ŷ`.
+"""
+mape(y, ŷ) = mean(replace(abs.((y .- ŷ) ./ y), Inf => 0.0))
+
+"""
+    mda(mod::Model, X, y)
+
+Compute the mean directional accuracy of a model `mod` on the input `X` and output `y`.
+"""
+mda(mod::Model, X, y) = mda(y, predict(mod, X))
+
+"""
+    mda(y, ŷ)
+
+Compute the mean directional accuracy of a model `mod` on the output `y` and predicted output `ŷ`.
+"""
+mda(y, ŷ) = mean(sign.(y) .== sign.(ŷ))
+
+"""
     r2(mod::Model, X, y)
 
 Compute the R² of a model `mod` on the input `X` and output `y`.
@@ -95,6 +123,8 @@ function r2(y, ŷ)
     sst = sum((y .- ȳ) .^ 2)
     return 1 - _ssr/sst
 end
+
+
 
 struct Probe <: Model
     β::Vector{Float64}

### 2024-01-25 src/probe_llm/utilities/evaluation.jl

@@ -1,3 +1,12 @@
+"""
+    evaluate(
+        agg_data; 
+        group_vars=[:model, :fold, :split, :n_pc],
+        agg_vars=[:indicator, :maturity, :layer, :split, :n_pc, :variable, :model]
+    )
+
+Evaluate the results of a probe experiment. The input data should be aggregated over the cross-validation folds.
+"""
 function evaluate(
     agg_data; 
     group_vars=[:model, :fold, :split, :n_pc],
@@ -22,9 +31,11 @@ function evaluate(
             :n_pc .=> unique => :n_pc,
             [:y, :yhat] => ((y, yhat) -> cor([y yhat])[1, 2]) => :cor,
             [:y, :yhat] => ((y, yhat) -> mse(y, yhat)) => :mse,
-            [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse
+            [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse,
+            [:y, :yhat] => ((y, yhat) -> mda(y, yhat)) => :mda,
+            [:y, :yhat] => ((y, yhat) -> r2(y, yhat)) => :r2,
         )
-    res = stack(res, [:cor, :mse, :rmse]) |>
+    res = stack(res, [:cor, :mse, :rmse, :r2, :mda]) |>
         x -> groupby(x, agg_vars) |>
         x -> combine(x, :value => mean => :value, :value => std => :std) |>
         x -> sort(x, agg_vars)

### 2024-01-25 src/probe_llm/utilities/models.jl

@@ -1,0 +1,134 @@
+"""
+    function run_models(
+        all_data; 
+        indicator="PPI", 
+        maturity=missing, 
+        layer=24, 
+        use_head=false,
+        n_folds::Int=5,
+        λ::Real=0.1,
+        min_train_prp::Float64=0.75,
+        n_pc::Union{Nothing,Int}=nothing,
+        return_meta::Bool=false,
+    )
+
+Run the probe experiment for a given set of parameters. Runs the baseline AR model on the aggregated data and the probe on the individual sentences. Both models are run for each cross-validation fold using an expanding window scheme. The results are aggregated over the folds.
+"""
+function run_models(
+    all_data; 
+    indicator="PPI", 
+    maturity=missing, 
+    layer=24, 
+    use_head=false,
+    n_folds::Int=5,
+    λ::Real=0.1,
+    min_train_prp::Float64=0.75,
+    n_pc::Union{Nothing,Int}=nothing,
+    return_meta::Bool=false,
+)
+
+    # Prepare market data:
+    mkt_data = prepare_mkt_data(all_data, indicator, maturity) |>
+        x -> sort(x, [:ym, :event_type, :speaker])
+    time_stamps = unique(sort(mkt_data.ym))
+    ts_splits = time_series_split(
+        time_stamps; 
+        n_folds=n_folds, 
+        return_vals=true, 
+        min_train_prp=min_train_prp
+    ) |> collect
+
+    # Prepare the data for the probe:
+    if use_head
+        X_probe, y_probe = prepare_probe(mkt_data)
+        model = load_model(; use_head=use_head)
+        clf = model.mod.cls
+        X_probe = vcat([get_clf_activations(clf, x)' for x in eachrow(X_probe)]...)
+    else
+        X_probe, y_probe = prepare_probe(mkt_data; layer=layer)
+    end
+
+    mkt_data.layer .= layer
+
+    # Run models for each split:
+    res = []
+    probes = []
+    probe_data = []
+    for (i, (train, test)) in enumerate(ts_splits)
+        _mkt_data = deepcopy(mkt_data) |>
+            x -> sort(x, [:sentence_id, :ym, :event_type, :speaker])
+        # Run the baseline on the aggregated data:
+        agg_data = groupby(_mkt_data, :ym) |>
+            x -> combine(x, :value .=> mean; renamecols=false) |>
+            x -> sort(x, :ym) 
+        y = agg_data.value
[…]

### 2024-01-25 src/probe_llm/utilities/plotting.jl

@@ -1,0 +1,220 @@
+"""
+    plot_ts(
+        agg_data::DataFrame;
+        alpha=[0.2 1.0 1.0],
+        legend=:topleft,
+        xlabel="Year",
+        kwrgs...
+    )
+
+Plot the time series of the actual and predicted values for the probe and the AR model.
+"""
+function plot_ts(
+    agg_data::DataFrame;
+    alpha=[0.2 1.0 1.0],
+    legend=:topleft,
+    xlabel="Year",
+    kwrgs...
+)
+    
+    # Collect metadata:
+    indicator = agg_data.indicator[1]
+    maturity = agg_data.maturity[1]
+    layer = agg_data.layer[1]
+    p = agg_data.p[1]
+
+    # Aggregate the results:
+    ts = groupby(agg_data, [:ym, :model]) |>
+        x -> combine(
+            x,
+            :y .=> mean => :y,
+            [:yhat, :nrow] => ((y, w) -> mean(y, Weights(w))) => :yhat
+        ) |>
+        x -> unstack(x, [:ym, :y], :model, :yhat) |>
+        x -> select(x, Not([:ym])) |>
+        Matrix
+
+    # Collect the dates:
+    dates = [Date(y,m) for (y,m) in unique(agg_data.ym)]
+
+    # Plot the results:
+    measure_name = indicator
+    if !ismissing(maturity)
+        measure_name = "$measure_name ($maturity)"
+    end
+    plt = Plots.plot(
+        dates,
+        ts;
+        label=[measure_name "AR($p)" "Probe (layer $layer)"],
+        ylabel=measure_name,
+        alpha=alpha,
+        legend=legend,
+        xlabel=xlabel,
+        kwrgs...
+    )
+    return plt
+end
+
+
+"""
+    plot_measures(
+        df::DataFrame;
+        axis=(width=225, height=225),
+        variables=["cor", "mse", "rmse"],
+        splits=["train", "test"],
+        indicators=["CPI", "PPI", "UST"]
+
+    )
+
+Plot the evaluation measures against the layer number for a set of models for a single variable. The dataframe should contain only one variable.
+"""
+function plot_measures(
+    df::DataFrame;
+    axis=(width=225, height=225),
+    variable="rmse",
+    splits=["train", "test"],
+    indicators=["CPI", "PPI", "UST"],
+    models=["y_bl", "y_probe"],
+    plot_interval=true,
+)
+
+    # Filter the data:
+    df = filter(x -> x.split ∈ splits, df)
[…]

### 2024-01-25 src/probe_llm/experiments.jl
redone some charts

@@ -21,18 +21,46 @@ if EVALUATE
 end
 
 # Plot the results:
+
 if PLOT_PROBES
+    axis = (width=150, height=150)
+    model = "y_probe"
+    _split = "test"
+    plot_interval = false
     ispath(joinpath(save_dir, "figures")) || mkdir(joinpath(save_dir, "figures"))
     df_evals = CSV.read(joinpath(save_dir, "evaluations.csv"), DataFrame)
-    plt = plot_measures(
-        df_evals;
-        axis=(width=225, height=225),
-        plot_interval=false,
-        models=["y_probe"],
-        splits=["test"],
-    )
+
+    for _var in sort(unique(df_evals.variable))
+        # Full (no PCA):
+        evals_full = filter(x -> ismissing(x.n_pc), df_evals)
+        plt = plot_measures(
+            evals_full;
+            axis=axis,
+            plot_interval=plot_interval,
+            model=model,
+            split=_split,
+            variable=_var
+        )
+        save(joinpath(save_dir, "figures", "$(_var)_full.png"), plt, px_per_unit=3)
+
+        # With PCA:
+        evals_pca = filter(x -> !ismissing(x.n_pc), df_evals)
+        for n_pc in sort(unique(evals_pca.n_pc))
+            _df = filter(x -> x.n_pc == n_pc, evals_pca)
+            plt = plot_measures(
+                _df;
+                axis=axis,
+                plot_interval=plot_interval,
+                model=model,
+                split=_split,
+                variable=_var
+            )
+            save(joinpath(save_dir, "figures", "$(_var)_pca_$(n_pc).png"), plt, px_per_unit=3)
+        end
+    end
+    
+    # By indicator:
     gdf = groupby(df_evals, [:indicator, :maturity, :n_pc]) 
-    axis = (width=225, height=225)
     for g in gdf
         g = DataFrame(g)
         i = g.indicator[1]
@@ -48,5 +76,7 @@ end
 
 # Attack probe:
 if RUN_ATTACKS
+    COMPUTE_EMBEDDINGS = false
+    USE_ALL_SENTENCES = false
     include("run_attacks.jl")
 end

### 2024-01-25 src/probe_llm/run_attacks.jl

@@ -1,12 +1,12 @@
-tfm = load_model(; load_head=false)
+if COMPUTE_EMBEDDINGS
+    tfm = load_model(; load_head=false)
+end
 n_pc = 128
 layer = 24
 ispath(joinpath(save_dir, "attacks")) || mkpath(joinpath(save_dir, "attacks"))
 attack_sentence_dir = "results/attacks/sentences"
 ispath(attack_sentence_dir) || mkpath(attack_sentence_dir)
-use_all = false
-prefx = use_all ? "all_" : ""
-compute_embeddings = true
+prefx = USE_ALL_SENTENCES ? "all_" : ""
 
 high_inf_text = readlines(joinpath(attack_sentence_dir, "$(prefx)high_inf.txt"))[1]
 high_inf_query = split(high_inf_text, ";") |>
@@ -28,7 +28,7 @@ queries = zip([high_inf_query, hawk_query, low_inf_query, dove_query], ["high_in
 
 # Embed the queries:
 queries_embedded = []
-if compute_embeddings
+if COMPUTE_EMBEDDINGS
     Threads.@threads for (q, cat) in collect(queries)
         println("Embedding queries for $cat on thread $(Threads.threadid())")
         embedding = []

### 2024-01-25 src/probe_llm/utilities/plotting.jl

@@ -72,17 +72,17 @@ function plot_measures(
     df::DataFrame;
     axis=(width=225, height=225),
     variable="rmse",
-    splits=["train", "test"],
+    split="test",
     indicators=["CPI", "PPI", "UST"],
-    models=["y_bl", "y_probe"],
+    model="y_probe",
     plot_interval=true,
 )
 
     # Filter the data:
-    df = filter(x -> x.split ∈ splits, df)
+    df = filter(x -> x.split == split, df)
     df = filter(x -> x.indicator ∈ indicators, df)
     df = filter(x -> x.variable == variable, df)
-    df = filter(x -> x.model ∈ models, df)
+    df = filter(x -> x.model == model, df)
 
     # Upper and lower bounds:
     df = transform(df, [:value, :std] => ByRow((x, s) -> (x + s, x - s)) => [:lb, :ub])
@@ -96,21 +96,19 @@ function plot_measures(
     # Plot the results:
     if plot_interval
         plt = data(df) * mapping(
-            :layer, :value,
+            :layer => "Layer", 
+            :value => "Value",
             lower=:lb, upper=:ub,
-            row=:split,
             col=:indicator,
-            color=:model => x -> x == "y_bl" ? "AR" : "Probe",
         )
         layer = visual(LinesFill)
     else
         plt = data(df) * mapping(
-            :layer, :value,
-            row=:split,
+            :layer => "Layer",
+            :value => "Value",
             col=:indicator,
-            color=:model => x -> x == "y_bl" ? "AR" : "Probe",
         )
-        layer = visual(Lines)
+        layer = smooth() + visual(Scatter)
     end
     plt = draw(
         layer * plt,
@@ -159,19 +157,21 @@ function plot_measures_for_ind(
     # Plot the results:
     if plot_interval 
         plt = data(df) * mapping(
-            :layer, :value, 
+            :layer => "Layer",
+            :value => "Value",
             lower=:lb, upper=:ub,
             row=:split,
             col=:variable, 
-            color=:model => x -> x=="y_bl" ? "AR" : "Probe",
+            color=:model => (x -> x=="y_bl" ? "AR" : "Probe") => "Model",
         )
         layer = visual(LinesFill)
     else
         plt = data(df) * mapping(
-            :layer, :value,
+            :layer => "Layer",
+            :value => "Value",
             row=:split,
             col=:variable,
[…]

### 2024-01-25 src/simple_examples/autoencoder/plot_ts.jl
reran charts for figure 2, but will probs take it out

@@ -40,4 +40,4 @@ p_yields = plot(
     alpha=0.5,
     size=(800,400)
 )
-plot(p_gdp, p_yields, layout=(2,1), size=(800, 600), left_margin=5mm)
+plot(p_gdp, p_yields, layout=(2,1), size=(800, 600), left_margin=5mm, dpi=300)

### 2024-01-25 src/simple_examples/autoencoder/run_probe.jl

@@ -1,29 +1,48 @@
 yhat = model(X')'
+
+dates = Date.(df_all.Date)
+tick_years = Date.(unique(Dates.year.(dates)))
+tick_years = range(tick_years[1], tick_years[end], step=Year(5))
+date_tick = Dates.format.(tick_years, "yyyy")
+
 plt_gdp = plot(
     df_all.Date, y;
-    label="Actual", color=:green,
-    title="Modelling GDP Growth",
+    label="g", color=:green,
+    size=(1000,200),
+    dpi=300
 )
 plot!(
     df_all.Date, yhat;
-    label="Predicted", color=:green, ls=:dash,
+    label="ĝ", color=:green, ls=:dash,
 )
+plot!(xticks=(tick_years, date_tick), yaxis=(formatter = y -> @sprintf("%.2f", y)))
+savefig(plt_gdp, joinpath(FIGURE_DIR, "gdp.png"))
 
 λ = 1.0
 Y = df_factors[:, [:spread, :level]] |> Matrix
 A = encoder(X')'
 W = (A'A + UniformScaling(λ)) \ A'Y
 Ŷ = A * W
+
 plt_probe = plot(
-    df_all.Date, Y;
+    dates, Y;
     color=[:blue :orange],
-    label=["Spread" "Level"],
-    title="Linear Probe"
+    label=["yₛ" "yₗ"],
+    size=(1000, 200),
+    dpi=300
 )
 plot!(
-    df_all.Date, Ŷ;
+    dates, Ŷ;
     color=[:blue :orange], ls=:dash,
-    label=["Spread (predicted)" "Level (predicted)"]
+    label=["ŷₛ" "ŷₗ"]
 )
-plt = plot(plt_gdp, plt_probe, layout=(2, 1), size=(800, 600), left_margin=5mm)
+plot!(xticks=(tick_years, date_tick), yaxis=(formatter = y -> @sprintf("%.2f", y)))
+savefig(plt_probe, joinpath(FIGURE_DIR, "factor_probe.png"))
+
+# Combine:
+plt = plot(
+    plt_gdp, plt_probe, layout=(1, 2), 
+    size=(1000, 200), left_margin=5mm,
+    dpi=300
+)
 savefig(plt, joinpath(FIGURE_DIR, "dl.png"))

### 2024-01-25 src/simple_examples/pca/run_experiment.jl

@@ -14,13 +14,14 @@ plt_pc = plot(
     dates,
     .-U[:,1:n_pc],
     label=["PC $i" for i in 1:n_pc] |> permutedims,
-    size=(1000, 300),
+    size=(1000, 200),
     ylims=(-0.015,0.03),
-    legend=:topright
+    legend=:topright, 
+    dpi=300
 )
 plot!(xticks=(tick_years,date_tick), xtickfontsize=6, yaxis=(formatter=y->@sprintf("%.2f",y)))
-vline!(Date.([onset_date]), ls=:solid, color=:black, label="Onset of GFC")
-vline!(Date.([aftermath_date]), ls=:dash, color=:black, label="Aftermath of GFC")
+vline!(Date.([onset_date]), ls=:solid, color=:black, label="|GFC")
+vline!(Date.([aftermath_date]), ls=:dash, color=:black, label="GFC|")
 
 # Level:
 df_level = @group_by(df, Date) |>
@@ -46,15 +47,20 @@ plt_obs = plot(
     dates,
     plt_mat,
     label=["Level" "Spread"],
-    size=(1000, 300),
+    size=(1000, 200),
     ylims=(-3,9),
     legend=:topright,
-    ylab="Yield (%)"
+    ylab="Yield (%)",
+    dpi=300,
+    left_margin=5mm, bottom_margin=5mm,
 )
 plot!(xticks=(tick_years,date_tick), xtickfontsize=6, yaxis=(formatter=y->@sprintf("%.2f",y)))
-vline!(Date.([onset_date]), ls=:solid, color=:black, label="Onset of GFC")
-vline!(Date.([aftermath_date]), ls=:dash, color=:black, label="Aftermath of GFC")
+vline!(Date.([onset_date]), ls=:solid, color=:black, label = "|GFC")
+vline!(Date.([aftermath_date]), ls=:dash, color=:black, label="GFC|")
 
-plt = plot(plt_pc, plt_obs, layout=(2,1), size=(1000, 400), left_margin=5mm, bottom_margin=5mm)
-savefig(plt, joinpath(FIGURE_DIR, "pca.png"))
+# Save:
+savefig(plt_pc, joinpath(FIGURE_DIR, "pca.png"))
+savefig(plt_obs, joinpath(FIGURE_DIR, "yield.png"))
[…]

### 2024-01-25 main.tex

@@
 Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
-Quantitative data often show non-causal associations (as humurously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-A major contributor is that humans struggle to conceive of actual randomness, as studies show that humans are poor at detecting e.g.
+Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
+A major contributor is that humans struggle to conceive of actual randomness: e.g.
+studies show that humans are poor at detecting e.g.
 \cite{falk1997making}, and creating e.g.
@@
 A common issue is a lack of expectation that the patterns that hint towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \cite{nickerson1998confirmation}) or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
+This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \cite{nickerson1998confirmation}), or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
 
-AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems: patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
+We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems: patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
 In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g.
@@
 
-Present work case studies an example of claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out or biased interpretation of evidence in support of one's beliefs.
+Present work case studies an example of claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs.
 We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, of AI systems.
 We further show results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.
+
+\textcolor{red}{andrew: I'm thinking we can adjust this all to aim towards more 'severe' tests or something. }
 
@@
 Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a 'weak' or 'non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
+In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
 If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
@@
 examples. To illustrate further that this phenomenon is neither
-surprising nor unique to the field of computer science, I will draw on
-my background in economics and finance in this section. We will start
-with very simple examples to demonstrate that even small and simple
-models can learn meaningful representations of the data. The final
-example in \textbf{?@sec-ex-llm} is a bit more involved and closer in
-spirit to the experiments conducted by Gurnee and Tegmark (2023). As we
-go along, we will try to discuss both the benefits and potential
+surprising nor unique to the field of computer science, we will borrow examples from the fields of Economics and Finance in this section.
+We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data.
+The final example in Section \ref{ex-llm} is a bit more involved and closer in spirit to the experiments conducted by \cite{gurnee2023language}.
+As we go along, we will try to discuss both the benefits and potential
 pitfalls of finding patterns in latent spaces.
 
-\subsubsection{Example: Principal Component
+\subsection{Example: Principal Component
 Analysis}\label{example-principal-component-analysis}
@@
 
-To compute the principal components we can decompose the matrix of
-yields \(\mathbf{Y}\) into a product of its singular vectors and values:
-\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). I will not go into
-the details here, because Professor Gilbert Strang has already done a
-much better job than I ever could in his
-\href{https://www.youtube.com/watch?v=mBcLRGuAFUk}{Linear Algebra
-lectures}. To put this into the broader context of the article, however,
-let us simply refer to \(\mathbf{U}\), \(\Sigma\) and
-\(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve (they
-are latent because they are not directly observable).
+To compute the principal components we can decompose the matrix of yields \(\mathbf{Y}\) into a product of its singular vectors and values: \(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
+To put this into the broader context of the article, however, let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve (they are latent because they are not directly observable).
 
@@
 
-\includegraphics[width=0.8\textwidth]{results/figures/pca.png}
+\includegraphics[width=1.0\textwidth]{results/figures/pca.png}
 
@@
 
-\subsubsection{Example: Deep Learning}\label{example-deep-learning}
+\subsection{Example: Deep Learning}\label{example-deep-learning}
 
@@
 
-\paragraph{Data}\label{data}
+\subsubsection{Data}\label{data}
 
@@
 
-\paragraph{Model}\label{model}
+\subsubsection{Model}\label{model}
 
@@
 
-\paragraph{Linear Probe}\label{linear-probe}
+\subsubsection{Linear Probe}\label{linear-probe}
 
@@
 yield curve data \(\mathbf{Y}_t\) through the encoder. Next, we follow
-Gurnee and Tegmark (2023) and use a linear probe to regress the observed
+\cite{gurnee2023language} and use a linear probe to regress the observed
 yield curve factors on the latent embeddings. Let \(F_t\) denote the
@@
 \(p_{w}(F_t|\mathbf{A}_t)\) where \(w\) denotes the regression
-parameters. Following Gurnee and Tegmark (2023), we use Ridge regression
+parameters. Following \cite{gurnee2023language}, we use Ridge regression
 with \(\lambda\) set to \(0.1\). Using the estimated regression
@@
 
-\paragraph{Ok, but truly what's the
+\subsubsection{Ok, but truly what's the
 point?}\label{ok-but-truly-whats-the-point}
@@
 
-\subsubsection{Example: Large Language Model}\label{ex-llm}
+\subsection{Example: Large Language Model}\label{ex-llm}
 
@@
 consider an example involving an LLM. In particular, we will closely
-follow the approach in Gurnee and Tegmark (2023) and apply it to a novel
+follow the approach in \cite{gurnee2023language} and apply it to a novel
 financial dataset: the \emph{Trillion Dollar Words} dataset introduced
-by Shah, Paturi, and Chava (2023). The dataset contains a curated
+by \cite{shah2023trillion}. The dataset contains a curated
 selection of sentences formulated by central bankers of the US Federal
@@
 data and then fine-tune various foundation models. Their model of
-choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa (Liu et al.
-2019)), achieves an \(F_1\) score of around \(>0.7\) for the
+choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \cite{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) for the
 classification task. To illustrate the potential usefulness of the
@@
 
-\paragraph{Linear Probes}\label{linear-probes}
+\subsubsection{Linear Probes}\label{linear-probes}
 
-Instead of computing a measure based on predicted labels, we can use
-linear probes to assess if the fine-tuned model has learned associative
-patterns between central bank communications and key economic
-indicators. To this end, I have further pre-processed the data provided
-by Shah, Paturi, and Chava (2023) and used their proposed model to
-compute layer-wise embeddings for all available sentences. I have made
-these available and easily accessible through a small Julia package:
-\href{https://github.com/pat-alt/TrillionDollarWords.jl}{TrillionDollarWords.jl}.
-For each layer, I have then computed linear probes on two inflation
-indicators---the Consumer Price Index (CPI) and the Producer Price Index
-(PPI)---as well as US Treasury yields at different levels of maturity.
-To mitigate issues related to over-parameterization, I follow the
-recommendation in Alain and Bengio (2018) to first reduce the
-dimensionality of the embeddings each time. In particular, linear probes
-are restricted to the first 128 principal components of the embeddings
+Instead of computing a measure based on predicted labels, we can use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
+To this end, we have further pre-processed the data provided by \cite{shah2023trillion} and used their proposed model to compute layer-wise embeddings for all available sentences.
+We have made these available and easily accessible through a small Julia package: \href{https://github.com/pat-alt/TrillionDollarWords.jl}{TrillionDollarWords.jl}.
+For each layer, we have then computed linear probes on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.
+
+To mitigate issues related to over-parameterization, we follow the recommendation in \cite{alain2018understanding} to first reduce the dimensionality of the embeddings each time.
+In particular, linear probes are restricted to the first 128 principal components of the embeddings
 of each layer.
 
-Detailed results of this experiment will be released in an upcoming
-paper. Figure~\ref{fig-fomc} highlights the result for the linear probe
+Figure~\ref{fig-fomc} highlights the result for the linear probe
 on the CPI. The chart shows various performance measures plotted against
@@
 positive for the baseline models and the linear probes. Consistent with
-the findings in Gurnee and Tegmark (2023) and Alain and Bengio (2018),
+the findings in \cite{gurnee2023language} and \cite{alain2018understanding},
 we also observe that this correlation tends to be higher for layers near
@@
 
-\paragraph{Stochastic Parrots After
+\subsubsection{Stochastic Parrots After
 All?}\label{stochastic-parrots-after-all}
@@
 inflation. It is worth pointing out here that this model is
-substantially smaller than the models tested in Gurnee and Tegmark
-(2023). This begs the following question:
+substantially smaller than the models tested in \cite{gurnee2023language}. This begs the following question:
 
@@
 
-Personally, I am having a very hard time believing this. To argue my
-case, I will now produce a counter-example demonstrating that, if
-anything, these findings are very much in line with the parrot metaphor.
-The counter-example is based on the following premise: if the results
-from the linear probe truly were indicative of some intrinsic
-understanding of the economy, then the probe should not be sensitive to
-random sentences that are most definitely not related to consumer
-prices.
+We are having a very hard time believing this.
+To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor.
+The counter example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic understanding of the economy, then the probe should not be sensitive to random sentences  that are most definitely not related to consumer prices.
 
-To test this, I select the best-performing probe trained on the
-final-layer activations to predict changes in the CPI. I then make up
-sentences that fall into one of these four categories:
-\emph{Inflation/Prices} (IP)---sentences about price inflation,
-\emph{Deflation/Prices} (DP)---sentences about price deflation,
-\emph{Inflation/Birds} (IB)---sentences about an inflation in the number
-of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation
-in the number of birds. A sensible sentence for category DP, for
-example, could be: ``It is essential to bring inflation back to target
-to avoid drifting into deflation territory.''. Analogically, we could
-construct the following sentence for the DB category: ``It is essential
-to bring the numbers of doves back to target to avoid drifting into
-dovelation territory.''.
+To test this, we select the best-performing probe trained on the final-layer activations to predict changes in the CPI.
+We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about an inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation in the number of birds.
+A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''.
+Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''.
 
@@
 
-To get to this last point, I also generate many probe predictions for
+To get to this last point, we also generate many probe predictions for
 samples of noise. Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote
@@
 \(k\) first principal components of the final-layer activations to the
-output variable of interest (CPI growth in this case). Then I sample
+output variable of interest (CPI growth in this case). Then we sample
 \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\)
-for \(i \in [1,1000]\) and compute the sample average. I repeat this
+for \(i \in [1,1000]\) and compute the sample average. We repeat this
 process \(10000\) times and compute the median-of-means to get an
@@
 
-Next, I propose the following hypothesis test as a minimum viable
+Next, we propose the following hypothesis test as a minimum viable
 testing framework to assess if the probe results (may) provide evidence
@@
 
-I should note that the number of sentences in each category is very
+We should note that the number of sentences in each category is very
 small here (10), so the results in Figure~\ref{fig-attack} cannot be
@@
 
-Submission to ICML 2024 will be entirely electronic, via a web site
-(not email). Information about the submission process and \LaTeX\ templates
-are available on the conference web site at:
-\begin{center}
-\textbf{\texttt{http://icml.cc/}}
-\end{center}
-
-The guidelines below will be enforced for initial submissions and
-camera-ready copies. Here is a brief summary:
-\begin{itemize}
-\item Submissions must be in PDF\@. 
-\item \textbf{New to this year}: If your paper has appendices, submit the appendix together with the main body and the references \textbf{as a single file}.
-Reviewers will not look for appendices as a separate PDF file.
-So if you submit such an extra file, reviewers will very likely miss it.
-\item Page limit: The main body of the paper has to be fitted to 8 pages, excluding references and appendices; the space for the latter two is not limited.
-For the final version of the paper, authors can add one extra page to the main body.
-\item \textbf{Do not include author information or acknowledgements} in your
-    initial submission.
-\item Your paper should be in \textbf{10 point Times font}.
-\item Make sure your PDF file only uses Type-1 fonts.
-\item Place figure captions \emph{under} the figure (and omit titles from inside
-    the graphic file itself). Place table captions \emph{over} the table.
-\item References must include page numbers whenever possible and be as complete
-    as possible. Place multiple citations in chronological order.
-\item Do not alter the style template; in particular, do not compress the paper
-    format by reducing the vertical spaces.
-\item Keep your abstract brief and self-contained, one paragraph and roughly
-    4--6 sentences. Gross violations will require correction at the
-    camera-ready phase. The title should have content words capitalized.
-\end{itemize}
-
-\subsection{Submitting Papers}
-
-\textbf{Paper Deadline:} The deadline for paper submission that is
-advertised on the conference website is strict. If your full,
-anonymized, submission does not reach us on time, it will not be
-considered for publication.
-
-\textbf{Anonymous Submission:} ICML uses double-blind review: no identifying
-author information may appear on the title page or in the paper
-itself. \cref{author info} gives further details.
-
-\textbf{Simultaneous Submission:} ICML will not accept any paper which,
-at the time of submission, is under review for another conference or
-has already been published. This policy also applies to papers that
-overlap substantially in technical content with conference papers
-under review or previously published. ICML submissions must not be
-submitted to other conferences and journals during ICML's review
-period.
-%Authors may submit to ICML substantially different versions of journal papers
-%that are currently under review by the journal, but not yet accepted
-%at the time of submission.
-Informal publications, such as technical
-reports or papers in workshop proceedings which do not appear in
-print, do not fall under these restrictions.
-
-\medskip
-
-Authors must provide their manuscripts in \textbf{PDF} format.
-Furthermore, please make sure that files contain only embedded Type-1 fonts
-(e.g.,~using the program \texttt{pdffonts} in linux or using
-File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
-might come from graphics files imported into the document.
-
-Authors using \textbf{Word} must convert their document to PDF\@. Most
-of the latest versions of Word have the facility to do this
-automatically. Submissions will not be accepted in Word format or any
-format other than PDF\@. Really. We're not joking. Don't send Word.
-
-Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
-Those using \texttt{latex} and \texttt{dvips} may need the following
-two commands:
-
-{\footnotesize
-\begin{verbatim}
-dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
-ps2pdf paper.ps
-\end{verbatim}}
-It is a zero following the ``-G'', which tells dvips to use
-the config.pdf file. Newer \TeX\ distributions don't always need this
-option.
-
-Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
-results. This program avoids the Type-3 font problem, and supports more
-advanced features in the \texttt{microtype} package.
-
-\textbf{Graphics files} should be a reasonable size, and included from
-an appropriate format. Use vector formats (.eps/.pdf) for plots,
-lossless bitmap formats (.png) for raster graphics with sharp lines, and
-jpeg for photo-like images.
-
-The style file uses the \texttt{hyperref} package to make clickable
-links in documents. If this causes problems for you, add
-\texttt{nohyperref} as one of the options to the \texttt{icml2024}
-usepackage statement.
-
-\subsection{Submitting Final Camera-Ready Copy}
-
-The final versions of papers accepted for publication should follow the
-same format and naming convention as initial submissions, except that
-author information (names and affiliations) should be given. See
-\cref{final author} for formatting instructions.
-
-The footnote, ``Preliminary work. Under review by the International
-Conference on Machine Learning (ICML). Do not distribute.'' must be
-modified to ``\textit{Proceedings of the
-$\mathit{41}^{st}$ International Conference on Machine Learning},
-Vienna, Austria, PMLR 235, 2024.
-Copyright 2024 by the author(s).''
-
-For those using the \textbf{\LaTeX} style file, this change (and others) is
-handled automatically by simply changing
-$\mathtt{\backslash usepackage\{icml2024\}}$ to
-$$\mathtt{\backslash usepackage[accepted]\{icml2024\}}$$
-Authors using \textbf{Word} must edit the
-footnote on the first page of the document themselves.
-
-Camera-ready copies should have the title of the paper as running head
-on each page except the first one. The running title consists of a
-single line centered above a horizontal rule which is $1$~point thick.
-The running head should be centered, bold and in $9$~point type. The
-rule should be $10$~points above the main text. For those using the
-\textbf{\LaTeX} style file, the original title is automatically set as running
-head using the \texttt{fancyhdr} package which is included in the ICML
-2024 style file package. In case that the original title exceeds the
-size restrictions, a shorter form can be supplied by using
-
-\verb|\icmltitlerunning{...}|
-
-just before $\mathtt{\backslash begin\{document\}}$.
-Authors using \textbf{Word} must edit the header of the document themselves.
-
-\section{Format of the Paper}
-
-All submissions must follow the specified format.
-
-\subsection{Dimensions}
-
-The text of the paper should be formatted in two columns, with an
-overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
-between the columns. The left margin should be 0.75~inches and the top
-margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
-whether you print on US letter or A4 paper, but all final versions
-must be produced for US letter size.
-Do not write anything on the margins.
-
-The paper body should be set in 10~point type with a vertical spacing
-of 11~points. Please use Times typeface throughout the text.
-
-\subsection{Title}
-
-The paper title should be set in 14~point bold type and centered
-between two horizontal rules that are 1~point thick, with 1.0~inch
-between the top rule and the top edge of the page. Capitalize the
-first letter of content words and put the rest of the title in lower
-case.
-
-\subsection{Author Information for Submission}
-\label{author info}
-
-ICML uses double-blind review, so author information must not appear. If
-you are using \LaTeX\/ and the \texttt{icml2024.sty} file, use
-\verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
-will not be printed unless \texttt{accepted} is passed as an argument to the
-style file.
-Submissions that include the author information will not
-be reviewed.
-
-\subsubsection{Self-Citations}
-
-If you are citing published papers for which you are an author, refer
-to yourself in the third person. In particular, do not use phrases
-that reveal your identity (e.g., ``in previous work \cite{langley00}, we
-have shown \ldots'').
-
-Do not anonymize citations in the reference section. The only exception are manuscripts that are
-not yet published (e.g., under submission). If you choose to refer to
-such unpublished manuscripts \cite{anonymous}, anonymized copies have
-to be submitted
-as Supplementary Material via OpenReview\@. However, keep in mind that an ICML
-paper should be self contained and should contain sufficient detail
-for the reviewers to evaluate the work. In particular, reviewers are
-not required to look at the Supplementary Material when writing their
-review (they are not required to look at more than the first $8$ pages of the submitted document).
-
-\subsubsection{Camera-Ready Author Information}
-\label{final author}
-
-If a paper is accepted, a final camera-ready copy must be prepared.
-%
-For camera-ready papers, author information should start 0.3~inches below the
-bottom rule surrounding the title. The authors' names should appear in 10~point
-bold type, in a row, separated by white space, and centered. Author names should
-not be broken across lines. Unbolded superscripted numbers, starting 1, should
-be used to refer to affiliations.
-
-Affiliations should be numbered in the order of appearance. A single footnote
-block of text should be used to list all the affiliations. (Academic
-affiliations should list Department, University, City, State/Region, Country.
-Similarly for industrial affiliations.)
-
-Each distinct affiliations should be listed once. If an author has multiple
-affiliations, multiple superscripts should be placed after the name, separated
-by thin spaces. If the authors would like to highlight equal contribution by
-multiple first authors, those authors should have an asterisk placed after their
-name in superscript, and the term ``\textsuperscript{*}Equal contribution"
-should be placed in the footnote block ahead of the list of affiliations. A
-list of corresponding authors and their emails (in the format Full Name
-\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
-Ideally only one or two names should be listed.
-
-A sample file with author names is included in the ICML2024 style file
-package. Turn on the \texttt{[accepted]} option to the stylefile to
-see the names rendered. All of the guidelines above are implemented
-by the \LaTeX\ style file.
-
-\subsection{Abstract}
-
-The paper abstract should begin in the left column, 0.4~inches below the final
-address. The heading `Abstract' should be centered, bold, and in 11~point type.
-The abstract body should use 10~point type, with a vertical spacing of
-11~points, and should be indented 0.25~inches more than normal on left-hand and
-right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
-abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
-sentences. Gross violations will require correction at the camera-ready phase.
-
-\subsection{Partitioning the Text}
-
-You should organize your paper into sections and paragraphs to help
-readers place a structure on the material and understand its
-contributions.
-
-\subsubsection{Sections and Subsections}
-
-Section headings should be numbered, flush left, and set in 11~pt bold
-type with the content words capitalized. Leave 0.25~inches of space
-before the heading and 0.15~inches after the heading.
-
-Similarly, subsection headings should be numbered, flush left, and set
-in 10~pt bold type with the content words capitalized. Leave
-0.2~inches of space before the heading and 0.13~inches afterward.
-
-Finally, subsubsection headings should be numbered, flush left, and
-set in 10~pt small caps with the content words capitalized. Leave
-0.18~inches of space before the heading and 0.1~inches after the
-heading.
-
-Please use no more than three levels of headings.
-
-\subsubsection{Paragraphs and Footnotes}
-
-Within each section or subsection, you should further partition the
-paper into paragraphs. Do not indent the first line of a given
-paragraph, but insert a blank line between succeeding ones.
-
-You can use footnotes\footnote{Footnotes
-should be complete sentences.} to provide readers with additional
-information about a topic without interrupting the flow of the paper.
-Indicate footnotes with a number in the text where the point is most
-relevant. Place the footnote in 9~point type at the bottom of the
-column in which it appears. Precede the first footnote in a column
-with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
-appear in each column, in the same order as they appear in the text,
-but spread them across columns and pages if possible.}
-
-\begin{figure}[ht]
-\vskip 0.2in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
-\caption{Historical locations and number of accepted papers for International
-Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
-Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
-produced, the number of accepted papers for ICML 2008 was unknown and instead
-estimated.}
-\label{icml-historical}
-\end{center}
-\vskip -0.2in
-\end{figure}
-
-\subsection{Figures}
-
-You may want to include figures in the paper to illustrate
-your approach and results. Such artwork should be centered,
-legible, and separated from the text. Lines should be dark and at
-least 0.5~points thick for purposes of reproduction, and text should
-not appear on a gray background.
-
-Label all distinct components of each figure. If the figure takes the
-form of a graph, then give a name for each axis and include a legend
-that briefly describes each curve. Do not include a title inside the
-figure; instead, the caption should serve this function.
-
-Number figures sequentially, placing the figure number and caption
-\emph{after} the graphics, with at least 0.1~inches of space before
-the caption and 0.1~inches after it, as in
-\cref{icml-historical}. The figure caption should be set in
-9~point type and centered unless it runs two or more lines, in which
-case it should be flush left. You may float figures to the top or
-bottom of a column, and you may set wide figures across both columns
-(use the environment \texttt{figure*} in \LaTeX). Always place
-two-column figures at the top or bottom of the page.
-
-\subsection{Algorithms}
-
-If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
-environments to format pseudocode. These require
-the corresponding stylefiles, algorithm.sty and
-algorithmic.sty, which are supplied with this package.
-\cref{alg:example} shows an example.
-
-\begin{algorithm}[tb]
-   \caption{Bubble Sort}
-   \label{alg:example}
-\begin{algorithmic}
-   \STATE {\bfseries Input:} data $x_i$, size $m$
-   \REPEAT
-   \STATE Initialize $noChange = true$.
-   \FOR{$i=1$ {\bfseries to} $m-1$}
-   \IF{$x_i > x_{i+1}$}
-   \STATE Swap $x_i$ and $x_{i+1}$
-   \STATE $noChange = false$
-   \ENDIF
-   \ENDFOR
-   \UNTIL{$noChange$ is $true$}
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Tables}
-
-You may also want to include tables that summarize material. Like
-figures, these should be centered, legible, and numbered consecutively.
-However, place the title \emph{above} the table with at least
-0.1~inches of space before the title and the same after it, as in
-\cref{sample-table}. The table title should be set in 9~point
-type and centered unless it runs two or more lines, in which case it
-should be flush left.
-
-% Note use of \abovespace and \belowspace to get reasonable spacing
-% above and below tabular lines.
-
-\begin{table}[t]
-\caption{Classification accuracies for naive Bayes and flexible
-Bayes on various data sets.}
-\label{sample-table}
-\vskip 0.15in
-\begin{center}
-\begin{small}
-\begin{sc}
-\begin{tabular}{lcccr}
-\toprule
-Data set & Naive & Flexible & Better? \\
-\midrule
-Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
-Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
-Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
-Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
-Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
-Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
-Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
-Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
-\bottomrule
-\end{tabular}
-\end{sc}
-\end{small}
-\end{center}
-\vskip -0.1in
-\end{table}
-
-Tables contain textual material, whereas figures contain graphical material.
-Specify the contents of each row and column in the table's topmost
-row. Again, you may float tables to a column's top or bottom, and set
-wide tables across both columns. Place two-column tables at the
-top or bottom of the page.
-
-\subsection{Theorems and such}
-The preferred way is to number definitions, propositions, lemmas, etc. consecutively, within sections, as shown below.
-\begin{definition}
-\label{def:inj}
-A function $f:X \to Y$ is injective if for any $x,y\in X$ different, $f(x)\ne f(y)$.
-\end{definition}
-Using \cref{def:inj} we immediate get the following result:
-\begin{proposition}
-If $f$ is injective mapping a set $X$ to another set $Y$, 
-the cardinality of $Y$ is at least as large as that of $X$
-\end{proposition}
-\begin{proof} 
-Left as an exercise to the reader. 
-\end{proof}
-\cref{lem:usefullemma} stated next will prove to be useful.
-\begin{lemma}
-\label{lem:usefullemma}
-For any $f:X \to Y$ and $g:Y\to Z$ injective functions, $f \circ g$ is injective.
-\end{lemma}
-\begin{theorem}
-\label{thm:bigtheorem}
-If $f:X\to Y$ is bijective, the cardinality of $X$ and $Y$ are the same.
-\end{theorem}
-An easy corollary of \cref{thm:bigtheorem} is the following:
-\begin{corollary}
-If $f:X\to Y$ is bijective, 
-the cardinality of $X$ is at least as large as that of $Y$.
-\end{corollary}
-\begin{assumption}
-The set $X$ is finite.
-\label{ass:xfinite}
-\end{assumption}
-\begin{remark}
-According to some, it is only the finite case (cf. \cref{ass:xfinite}) that is interesting.
-\end{remark}
-%restatable
-
-\subsection{Citations and References}
-
-Please use APA reference format regardless of your formatter
-or word processor. If you rely on the \LaTeX\/ bibliographic
-facility, use \texttt{natbib.sty} and \texttt{icml2024.bst}
-included in the style-file package to obtain this format.
-
-Citations within the text should include the authors' last names and
-year. If the authors' names are included in the sentence, place only
-the year in parentheses, for example when referencing Arthur Samuel's
-pioneering work \yrcite{Samuel59}. Otherwise place the entire
-reference in parentheses with the authors and year separated by a
-comma \cite{Samuel59}. List multiple references separated by
-semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
-construct only for citations with three or more authors or after
-listing all authors to a publication in an earlier reference \cite{MachineLearningI}.
-
-Authors should cite their own work in the third person
-in the initial version of their paper submitted for blind review.
-Please refer to \cref{author info} for detailed instructions on how to
-cite your own papers.
-
-Use an unnumbered first-level section heading for the references, and use a
-hanging indent style, with the first line of the reference flush against the
-left margin and subsequent lines indented by 10 points. The references at the
-end of this document give examples for journal articles \cite{Samuel59},
-conference publications \cite{langley00}, book chapters \cite{Newell81}, books
-\cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
-\cite{mitchell80}, and dissertations \cite{kearns89}.
-
-Alphabetize references by the surnames of the first authors, with
-single author entries preceding multiple author entries. Order
-references for the same authors by year of publication, with the
-earliest first. Make sure that each reference includes all relevant
-information (e.g., page numbers).
-
-Please put some effort into making references complete, presentable, and
-consistent, e.g. use the actual current name of authors.
-If using bibtex, please protect capital letters of names and
-abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
-in your .bib file.
-
-\section*{Accessibility}
-Authors are kindly asked to make their submissions as accessible as possible for everyone including people with disabilities and sensory or neurological differences.
-Tips of how to achieve this and what to pay attention to will be provided on the conference website \url{http://icml.cc/}.
-
-\section*{Software and Data}
-
-If a paper is accepted, we strongly encourage the publication of software and data with the
-camera-ready version of the paper whenever appropriate. This can be
-done by including a URL in the camera-ready copy. However, \textbf{do not}
-include URLs that reveal your institution or identity in your
-submission for review. Instead, provide an anonymous URL or upload
-the material as ``Supplementary Material'' into the OpenReview reviewing
-system. Note that reviewers are not required to look at this material
-when writing their review.
-
-% Acknowledgements should only appear in the accepted version.
-\section*{Acknowledgements}
-
-\textbf{Do not} include acknowledgements in the initial version of
-the paper submitted for blind review.
-
-If a paper is accepted, the final camera-ready version can (and
-probably should) include acknowledgements. In this case, please
-place such acknowledgements in an unnumbered section at the
-end of the paper. Typically, this will include thanks to reviewers
-who gave useful comments, to colleagues who contributed to the ideas,
-and to funding agencies and corporate sponsors that provided financial
-support.
-
-% In the unusual situation where you want a paper to appear in the
-% references without citing it in the main text, use \nocite
-\nocite{langley00}
-
-\bibliography{example_paper}
+\bibliography{biblio}
 \bibliographystyle{icml2024}

### 2024-01-25 src/simple_examples/autoencoder/run_probe.jl
reran charts for figure 2, but will probs take it out

@@ -7,13 +7,13 @@ date_tick = Dates.format.(tick_years, "yyyy")
 
 plt_gdp = plot(
     df_all.Date, y;
-    label="g", color=:green,
+    label="g", color=:black,
     size=(1000,200),
     dpi=300
 )
 plot!(
     df_all.Date, yhat;
-    label="ĝ", color=:green, ls=:dash,
+    label="ĝ", color=:black, ls=:dash,
 )
 plot!(xticks=(tick_years, date_tick), yaxis=(formatter = y -> @sprintf("%.2f", y)))
 savefig(plt_gdp, joinpath(FIGURE_DIR, "gdp.png"))
@@ -26,14 +26,13 @@ Ŷ = A * W
 
 plt_probe = plot(
     dates, Y;
-    color=[:blue :orange],
     label=["yₛ" "yₗ"],
     size=(1000, 200),
     dpi=300
 )
 plot!(
     dates, Ŷ;
-    color=[:blue :orange], ls=:dash,
+    color=[palette(:default)[1] palette(:default)[2]], ls=:dash,
     label=["ŷₛ" "ŷₗ"]
 )
 plot!(xticks=(tick_years, date_tick), yaxis=(formatter = y -> @sprintf("%.2f", y)))

### 2024-01-25 main.tex

@@
 
-Present work case studies an example of claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs.
-We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, of AI systems.
-We further show results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.
+Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs.
+We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, within AI systems.
+We further display results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.
 
@@
 
-\begin{figure}
+\begin{figure*}
 
@@
 
-\includegraphics[width=0.5\textwidth]{results/figures/measures_CPI (n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/rmse_pca_128.png}
 
@@
 
-\end{figure}%
+\end{figure*}%
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/attack_all_measures.png}
+
+}
+
+\caption{\label{fig-attack}Probe predictions for sentences about
+inflation of prices (IP), deflation of prices (DP), inflation of birds
+(IB) and deflation of birds (DB). The vertical axis shows predicted
+inflation levels subtracted by the average predicted inflation level for
+the whole sample period (train and test set).}
+
+\end{figure*}%
 
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.5\textwidth]{results/figures/attack_inflation.png}
-
-}
-
-\caption{\label{fig-attack}Probe predictions for sentences about
-inflation of prices (IP), deflation of prices (DP), inflation of birds
-(IB) and deflation of birds (DB). The vertical axis shows predicted
-inflation levels subtracted by the average predicted inflation level for
-the whole sample period (train and test set).}
-
-\end{figure}%
-
 \bibliography{biblio}

### 2024-01-25 src/simple_examples/pca/run_experiment.jl
reran charts for figure 1

@@ -8,6 +8,7 @@ X = @select(df_wide, -Date) |> Matrix
 U, Σ, V = svd(X)
 dates = Date.(df_wide.Date)
 tick_years = Date.(unique(Dates.year.(dates)))
+tick_years = range(tick_years[1], tick_years[end], step=Year(5))
 date_tick = Dates.format.(tick_years, "yyyy")
 n_pc = 2
 plt_pc = plot(
@@ -19,7 +20,7 @@ plt_pc = plot(
     legend=:topright, 
     dpi=300
 )
-plot!(xticks=(tick_years,date_tick), xtickfontsize=6, yaxis=(formatter=y->@sprintf("%.2f",y)))
+plot!(xticks=(tick_years,date_tick), yaxis=(formatter=y->@sprintf("%.2f",y)))
 vline!(Date.([onset_date]), ls=:solid, color=:black, label="|GFC")
 vline!(Date.([aftermath_date]), ls=:dash, color=:black, label="GFC|")
 

### 2024-01-25 src/mnist/run_experiment.jl
added mnist map example

@@ -1,0 +1,130 @@
+using Pkg; Pkg.activate("src/mnist")
+
+using CounterfactualExplanations.Data: load_mnist
+using CounterfactualExplanations.Models: load_mnist_vae, load_mnist_mlp
+using CSV
+using DataFrames
+using Flux
+using GMT
+using Images
+using LinearAlgebra
+using MLJBase
+using MLJModels
+using OneHotArrays
+
+RESULTS_DIR = "results"
+if !isdir(RESULTS_DIR)
+    mkdir(RESULTS_DIR)
+end
+FIGURE_DIR = joinpath(RESULTS_DIR, "figures")
+
+# Load MNIST data and pre-trained models:
+data = load_mnist()
+X = data.X
+vae = load_mnist_vae()
+mlp = load_mnist_mlp().model
+
+# World Data (from https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv)
+world_data = CSV.read("data/world_place.csv", DataFrame)
+
+# FIFA World Rankings
+# https://www.fifa.com/fifa-world-ranking/men?dateId=id14142
+# Let's map the top 10 teams to the first 10 integers:
+fifa_world_ranking = Dict(
+    "Argentina" => 0,
+    "France" => 1,
+    "Brazil" => 2,
+    "England" => 3,
+    "Belgium" => 4,
+    "Croatia" => 5,
+    "Netherlands" => 6,
+    "Portugal" => 7,
+    "Italy" => 8,
+    "Spain" => 9,
+)
+
+# Add FIFA World Rankings to World Data:
+fifa_world_data = DataFrames.subset(world_data, :country => ByRow(x -> haskey(fifa_world_ranking, x))) |>
+    x -> DataFrames.transform(x, :country => ByRow(x -> fifa_world_ranking[x]) => :y) |>
+    x -> DataFrames.select(x, :y, Not(:y, :country))
+
+# Tokenizer for FIFA World Data ====================
+# Goal: need a tokenizer that can map from the entities to the latent space of the VAE.
+# Continuous feature encoding:
+X = fifa_world_data[:,Not(:y)]
+model = (X -> coerce(
+        X,
+        :entity_type => Multiclass,
+)) |> 
+    MLJModels.FillImputer() |>
+    MLJModels.ContinuousEncoder() |> 
+    MLJModels.Standardizer() 
[…]

### 2024-01-25 technical_descriptions.qmd
work on tech desc

---
format:
  pdf:
    keep-tex: true
bibliography: biblio.bib
---

## Example: Projecting a World Model onto a latent space

- Take a variational autoencoder (VAE) that was pre-trained on MNIST. 
- Take the [world_place.csv]( https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv) that was used in @gurnee2023language, which maps locations/areas to their latitude and longitude. For each place, it also indicates the corresponding country.
- Take the subset that contains countries that are currently part of the top 10 [FIFA world ranking](https://www.fifa.com/fifa-world-ranking/men?dateId=id14142).
- Assign the current rank to each country, i.e. Argentina gets 1, France gets 2, ...
- Encode all features excluding the FIFA world rank indicator but including `longitude` and `latitude` as continuous variables: $X^{(n \times d)}$ where $n$ is the number of samples and $d$ is the number of resulting features.
- Initialize a small neural network, let's call it a *projector*, that maps from $X$ to a single small hidden layer with sigmoid activation and then form there to the latent space $\mathcal{Z} \subset \mathbb{R}$ of the pre-trained VAE.
- Without performing any training on the *projector*, simply compute a forward pass of the continuously encoded data $X$ through the *projector* and store the output: $\mathbf{Z}^{(n \times k)}$ where $k$ is the dimension of the latent space of the VAE.
- Next, we perform the linear probe on $\mathbf{Z}$ through Ridge regression: $\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}$ where $\mathbf{Y}$ is the $(n \times 2)$ matrix containing the longitude and latitude for each sample. 
- Finally, compute the predicted coordinate for each samples as $\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}$
- Plot the results on a world map and be amazed.

> I think what's going on here is that we essentially randomly project a small set of features into a higher-dimensional space ($d < k$). The linear probe is then having an easy time separating features in the high-dimensional space, even though the projection is completely random. Even though one might be momentarily amazed by this, there is really nothing surprising at all going on here. We can still draw a very interesting parallel to @gurnee2023language: we know that geographical coordinates are in the training corpus of Llama-2; that information gets passed through a massive latent space (the model); they run probes on a subsapce of the model (still massive) and manage to linearly separate their variable of interest (coordinates). I don't know how accurate that parallel really is but it just seems striking to me that ultimately what they is regressing a 2-dimensional output variable on this massive feature matrix ("ranging from 4,096 to 8,192 features" for roughly 40,000 total samples, can't figure out how big the train/test sets are, respectively.). I'm not good enough at theory but maybe there's really just no surprise at all here?

Relatedly, it is worth pointing out that

- a recent paper has shown that the double-descent phenomenon is not exclusive to deep learning but also applies to linear regression with many features [LINK].

## Example: PCA

## Example: Autoencoder

## Example: LLM

## References

### 2024-01-25 technical_descriptions.tex
work on tech desc

--- 2024-01-24 from_blog.tex
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
   \fi
   % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \KOMAoption{captions}{tableheading}
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
 \makeatletter
@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\usepackage{amsthm}
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
-\theoremstyle{remark}
-\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
-\newtheorem*{remark}{Remark}
-\newtheorem*{solution}{Solution}
-\newtheorem{refremark}{Remark}[section]
-\newtheorem{refsolution}{Solution}[section]
-\makeatother
-\makeatletter
 \makeatother
@@
 \makeatother
+\makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
 \ifLuaTeX
@@
 \fi
-\usepackage{bookmark}
-
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \hypersetup{
-  pdftitle={Spurious Sentience},
-  pdfauthor={Patrick Altmeyer},
   colorlinks=true,
@@
 
-\title{Spurious Sentience}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{On the Unsurprising Finding of Patterns in Latent Spaces}
-\author{Patrick Altmeyer}
-\date{2024-01-10}
+\author{}
+\date{}
 
 \begin{document}
-\maketitle
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden]}{\end{tcolorbox}}\fi
 
-\emph{For the slides see \href{slides.qmd}{here}.}
-
-We humans are prone to seek patterns everywhere. Meaningful patterns
-have often proven to help us make sense of our past, navigate our
-presence and predict the future. Our society is so invested in finding
-patterns that today it seems we are more willing than ever to outsource
-this task to an Artificial Intelligence (AI): an omniscient oracle that
-leads us down the right path. Unfortunately, history has shown time and
-again that patterns are double-edged swords: if we attribute the wrong
-meaning to them, they may lead us nowhere at all, or worse, they may
-lead us down the dark roads.
-
-In statistics, misleading patterns are referred to as \textbf{spurious
-relationships}: purely associational relationships between two or more
-variables that are not causally related to each other at all. The world
-is full of these and as good as we as species may be at recognizing
-patterns, we typically have a much harder time discerning spurious
-relationships from causal ones. Despite new and increased momentum in
-scientific fields concerned with causal inference and discovery, I am
-also willing to go out on a limb and claim that we are not about to
-finally reach the top of Judea Pearl's Causal Ladder through the means
-of Causal AI.
-
-I agree with the premise that in a world full of spurious relationships,
-causal reasoning is our only remedy. But I am very skeptical of claims
-that AI will magically provide that remedy. This leads me to the title
-and topic of this post: \textbf{spurious sentience}---patterns exhibited
-by artificial intelligence that may hint at sentience but are just
-reflections of the data used to train them. The article is written in
-response to a recent paper that extracts a `world model' from Llama
-2---a popular open-source large language model (LLM)---using mechanistic
-interpretability (Gurnee and Tegmark 2023). In light of these findings,
-one of the authors, Max Tegmark, was quick to claim on
-\href{https://x.com/tegmark/status/1709572469978231063?s=20}{social
-media} that ``No, LLM's aren't mere stochastic parrots {[}\ldots{]}''.
-
-Since this is an opinionated post, I feel that I should start with a few
-disclaimers:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  I take no issue with the methodological ideas that form the foundation
-  of the article in question: on the contrary, I think that mechanistic
-  interpretability is an interesting and important toolkit that can help
-  us better understand the intrinsics and behavior of opaque artificial
-  intelligence.
-\item
-  The visualizations are intriguing, the code is open-sourced and the
-  findings are interesting.
-\item
-  I am surprised that people are surprised by the findings: if we agree
-  that LLMs exhibit strong capabilities that can only be connected to
-  the patterns observed in the data they were trained with, then where
-  exactly should we expect this information to be stored if not in the
-  parameters of the model?\footnote{I would be very
-    surprised---concerned even---if our search for patterns in latent
-    spaces of capable LLMs revealed nothing at all.}
-\item
-  I therefore do take issue with the way that these findings are being
-  overblown by people with clout. Perhaps the parrot metaphor should not
-  be taken too literally either, but if anything the paper's findings
-  seem to support the notion that LLMs are remarkably capable of
-  memorizing and regurgitating explicit and implicit knowledge contained
-  in text.
-\end{enumerate}
-
-\subsection{Patterns in Latent Spaces and How to Find
-Them}\label{patterns-in-latent-spaces-and-how-to-find-them}
-
-To support my claim that observing patterns in latent spaces should not
-generally surprise us, we will now go through a couple of simple
-examples. To illustrate further that this phenomenon is neither
-surprising nor unique to the field of computer science, I will draw on
-my background in economics and finance in this section. We will start
-with very simple examples to demonstrate that even small and simple
-models can learn meaningful representations of the data. The final
-example in \textbf{?@sec-ex-llm} is a bit more involved and closer in
-spirit to the experiments conducted by Gurnee and Tegmark (2023). As we
-go along, we will try to discuss both the benefits and potential
-pitfalls of finding patterns in latent spaces.
-
-\subsubsection{Example: Principal Component
-Analysis}\label{example-principal-component-analysis}
-
-In response to the claims made by Tegmark, numerous commentators on
-social media have pointed out that even the simplest of models can
-exhibit structure in their latent spaces. One of the most popular and
-illustrative examples I remember from my time at the Bank of England is
-yield curve decomposition through PCA. The yield curve is a popular tool
-for investors and economists to gauge the health of the economy. It
-plots the yields of bonds against their maturities. The slope of the
-yield curve is often used as a predictor of future economic activity: a
-steep yield curve is associated with a growing economy, while a flat or
-inverted yield curve is associated with a contracting economy.
-
-% To understand this better, let us go on a quick detour into economics
-% and look at actual yield curves observed in the US during the Global
-% Financial Crisis (GFC). Figure~\ref{fig-gfc-1} shows the yield curve of
-% US Treasury bonds on 27 February 2007, which according to
-% \href{https://money.cnn.com/2007/02/27/markets/markets_0630/index.htm?cnn=yes}{CNN}
-% was a ``brutal day on Wall Street''.\footnote{The
-%   \href{https://home.treasury.gov/resource-center/data-chart-center/interest-rates/TextView?type=daily_treasury_yield_curve&field_tdr_date_value_month=202310}{data}
-%   is taken from the US Department of the Treasury.} This followed
-% \href{https://money.cnn.com/2007/02/26/news/economy/greenspan/index.htm?postversion=2007022609}{reports}
-% on the previous day of former Federal Reserve Chairman Alan Greenspan's
-% warning that the US economy was at risk of a recession. The yield curve
-% was inverted with a sharp negative spread between the 10-year and
-% 3-month yields, indicative of the market's expectation of a recession.
-
-% Figure~\ref{fig-gfc-2} shows the corresponding yield curve during the
-% aftermath of the GFC on 20 April 2009. On that day the influential Time
-% Magazine
-% \href{https://content.time.com/time/business/article/0,8599,1890560,00.html}{reported}
-% that the ``Banking Crisis is Over''. The yield curve was steeply sloped
-% with a positive spread between the 10-year and 3-month yields,
-% indicative of the market's expectation of a recovery. The overall level
-% of the yield curve was still very low though, indicative of the fact
-% that US economy had not fully recovered at that point.
-
-Of course, US Treasuries are not the only bonds that are traded in the
-market. To get a more complete picture of the economy, analysts might
-therefore be interested in looking at the yield curves of other bonds as
-well. In particular, we might be interested in predicting economic
-growth based on the yield curves of many different bonds. The problem
-with that idea is that it is cursed by high dimensionality: we would end
-up modelling a single variable of interest (economic growth) with a
-large number of predictors (the yields of many different bonds). To deal
-with the curse of high dimensionality it can be useful to decompose the
-yield curves into sets of principal components.
-
-To compute the principal components we can decompose the matrix of
-yields \(\mathbf{Y}\) into a product of its singular vectors and values:
-\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). I will not go into
-the details here, because Professor Gilbert Strang has already done a
-much better job than I ever could in his
-\href{https://www.youtube.com/watch?v=mBcLRGuAFUk}{Linear Algebra
-lectures}. To put this into the broader context of the article, however,
-let us simply refer to \(\mathbf{U}\), \(\Sigma\) and
-\(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve (they
-are latent because they are not directly observable).
-
-The top panel in Figure~\ref{fig-pca} shows the first two principal
-components of the yield curves of US Treasury bonds over time. Vertical
-stalks indicate the key dates during the onset and aftermath of the
-crisis, which we discussed above. For both components, we can observe
-some marked shifts between the two dates - but can we attribute any
-meaning to these shifts? It turns out we can: for comparison, the bottom
-panel in Figure~\ref{fig-pca} shows the average level and spread of the
-yield curves over time. The first principal component is strongly
-correlated with the level of the yield curve, while the second principal
-component is strongly correlated with the spread of the yield curve. To
-put it in AI-lingo:
-
-\begin{quote}
-The estimated latent embeddings of the yield curve are characterized by
-patterns observed in the data.
-\end{quote}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{results/figures/pca.png}
-
-}
-
-\caption{\label{fig-pca}Comparison of latent embeddings and observed
-data of the US Treasury yield curve.}
-
-\end{figure}%
-
-Not convinced? Let us use
-\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\) in true autoencoder
-fashion to reconstruct yield curves from principal components. Let
-\(z_1\) denote the first principal component and consider the following:
-we keep all other \(M-1\) principal components fixed at zero where \(M\)
-denotes the total number of maturities; next we traverse the latent
-space by varying the value of \(z_1\) over a fixed grid of length \(K\)
-each time storing the full vector \(\mathbf{z}\); finally, we vertically
-concatenate the vectors and end up with a matrix \(\mathbf{Z}\) of
-dimension \((K \times M)\). To reconstruct yields, we simply multiply
-\(Z\) by the singular values and right singular vectors:
-\(\mathbf{Y}=\mathbf{Z}\Sigma\mathbf{V}^{\prime}\).
-
-\subsubsection{Example: Deep Learning}\label{example-deep-learning}
-
-So far we have considered simple matrix decomposition. You might argue
-that principal components are not really latent embeddings in the
-traditional sense of deep learning. To address this, let us now consider
-a simple deep-learning example. Our goal will be to not only predict
-economic growth from the yield curve but also extract meaningful
-features at the same time. In particular, we will use a neural network
-architecture that allows us to recover a compressed latent
-representation of the yield curve.
-
-\paragraph{Data}\label{data}
-
-To estimate economic growth we will rely on a quarterly
-\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real
-gross domestic product (GDP) provided by the Federal Reserve Bank of
-St.~Louis. The data arrives in terms of levels of real GDP. In order to
-estimate growth, we will transform the data into log differences. Since
-our yield curve data is daily, we will need to aggregate it to the
-quarterly frequency. To do this, we will simply take the average of the
-daily yields for each maturity. We will also standardize yields since
-deep learning models tend to perform better with standardized data.
-Since COVID-19 was a huge structural break, we will also filter out all
-observations after 2018.
-
-\paragraph{Model}\label{model}
-
-Let \(G_t\) denote growth and \(\mathbf{Y}_t\) denote the yield curve at
-time \(t\). Then we are interested in a model for \(G_t\) conditional on
-\(\mathbf{Y}_t\). Let \(\theta\) denote our model parameters then
-formally we are interested in maximizing the likelihood
-\(p_{\theta}(G_t|\mathbf{Y}_t)\).
-
-To do this, we will use a simple
-autoencoder architecture: the encoder will consist of a single fully
-connected hidden layer with 32 neurons and a hyperbolic tangent
-activation function. The bottleneck layer connecting the encoder to the
-decoder is a fully connected layer with 6 neurons. This is the
-compressed latent representation of the yield curve mentioned above. The
-decoder will consist of two fully connected layers each with a
-hyperbolic tangent activation function: the first layer will consist of
-32 neurons and the second layer will have the same dimension as the
-input data. The output layer will consist of a single neuron with a
-linear activation function. The model will be trained using the mean
-squared error loss function and the Adam optimizer.
-
-\paragraph{Linear Probe}\label{linear-probe}
-
-The results are shown in Figure~\ref{fig-dl-results}. The top panel
-shows the actual GDP growth and fitted values from the autoencoder
-model. We observe that the model captures the relationship between
-economic growth and the yield curve reasonably well. As discussed above,
-we also know that the relationship between economic growth and the yield
-curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we
-would expect that these two important factors are reflected somewhere in
-the latent parameter space.
-
-The bottleneck layer seems like a good place to start looking. To get
-the latent embeddings \(\mathbf{A}_t\) at time \(t\), we simply pass the
-yield curve data \(\mathbf{Y}_t\) through the encoder. Next, we follow
-Gurnee and Tegmark (2023) and use a linear probe to regress the observed
-yield curve factors on the latent embeddings. Let \(F_t\) denote the
-vector containing the two factors of interest in time \(t\):
-\(f_t^{\text{spread}}\) and \(f_t^{\text{level}}\). Formally, we are
-interested in the following regression model:
-\(p_{w}(F_t|\mathbf{A}_t)\) where \(w\) denotes the regression
-parameters. Following Gurnee and Tegmark (2023), we use Ridge regression
-with \(\lambda\) set to \(0.1\). Using the estimated regression
-parameters \(\hat{w}\) we can then predict the yield curve factors from
-the latent embeddings: \(\hat{F}_t=\hat{w}^{\prime}\mathbf{A}_t\).
-
-The results of this experiment are shown in the bottom panel of
-Figure~\ref{fig-dl-results}. Solid lines show the observed yield curve
-factors over time, while dashed lines show predicted values. We find
-that the latent embeddings predict the two yield curve factors
-reasonably well, in particular the spread. To put this in true
-hype-lingo:
-
-\begin{quote}
-Not just a parrot! Our autoencoder neural network has an implicit
-understanding of the yield curve factors and their relationship with
-economic growth. It's sentient indeed!
-\end{quote}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{results/figures/dl.png}
-
-}
-
-\caption{\label{fig-dl-results}The top panel shows the actual GDP growth
-and fitted values from the autoencoder model. The bottom panel shows the
-observed average level and spread of the yield curve (solid) along with
-the predicted values from the linear probe based on the latent embedding
-(dashed).}
-
-\end{figure}%
-
-\paragraph{Ok, but truly what's the
-point?}\label{ok-but-truly-whats-the-point}
-
-The finding is not surprising but it is still interesting. In the
-context of mechanistic interpretability, it demonstrates that the
-black-box model has evidently learned plausible explanations for the
-data. Beyond that, in this particular example, the patterns in the
-latent space that we have just uncovered might actually be useful for
-downstream tasks. An interesting idea could be to use the latent
-embeddings as features in a more traditional and interpretable
-econometric model. To demonstrate this, let us consider a simple linear
-regression model for GDP growth. We will compare the performance of the
-following models: (1) regressing growth \(G_t\) on the latent embedding
-\(A_t\), (2) regression growth on the best subset of latent embeddings
-\(A_t\), (3) regressing growth on lagged growth, (4) regressing growth
-on lagged growth and the observed yield curve factors \(F_t\), (5)
-regressing growth on the best subset of yield curve factors \(F_t\),
-and, finally, (5) regressing growth on lagged growth and the best subset
-of latent embeddings \(A_t\).
-
-The results are shown in the table below. The key finding of interest is
-that the coefficients on the latent embeddings in model (5) are
-statistically significant. This suggests that the latent embeddings
-contain information that is useful for predicting GDP growth.
-
-\input{results/regression.tex}
-
-\subsubsection{Example: Large Language Model}\label{ex-llm}
-
-To round up this section, we will jump back on the hype train and
-consider an example involving an LLM. In particular, we will closely
-follow the approach in Gurnee and Tegmark (2023) and apply it to a novel
-financial dataset: the \emph{Trillion Dollar Words} dataset introduced
-by Shah, Paturi, and Chava (2023). The dataset contains a curated
-selection of sentences formulated by central bankers of the US Federal
-Reserve and communicated to the public in speeches, meeting minutes and
-press conferences. The authors of the paper use this dataset to train
-LLMs to classify sentences as either `dovish', `hawkish' or `neutral'.
-To this end, they first manually annotate a subsample of the available
-data and then fine-tune various foundation models. Their model of
-choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa (Liu et al.
-2019)), achieves an \(F_1\) score of around \(>0.7\) for the
-classification task. To illustrate the potential usefulness of the
-learned classifier, they use predicted labels for the entire dataset to
-compute an ad-hoc, count-based measure of `hawkishness'. They then go on
-to show that this measure correlates with key economic indicators in the
-expected direction: when inflationary pressures rise, the measured level
-of `hawkishness' increases as central bankers need to raise interest
-rates to bring inflation back to target.
-
-\paragraph{Linear Probes}\label{linear-probes}
-
-Instead of computing a measure based on predicted labels, we can use
-linear probes to assess if the fine-tuned model has learned associative
-patterns between central bank communications and key economic
-indicators. To this end, I have further pre-processed the data provided
-by Shah, Paturi, and Chava (2023) and used their proposed model to
-compute layer-wise embeddings for all available sentences. I have made
-these available and easily accessible through a small Julia package:
-\href{https://github.com/pat-alt/TrillionDollarWords.jl}{TrillionDollarWords.jl}.
-For each layer, I have then computed linear probes on two inflation
-indicators---the Consumer Price Index (CPI) and the Producer Price Index
-(PPI)---as well as US Treasury yields at different levels of maturity.
-To mitigate issues related to over-parameterization, I follow the
-recommendation in Alain and Bengio (2018) to first reduce the
-dimensionality of the embeddings each time. In particular, linear probes
-are restricted to the first 128 principal components of the embeddings
-of each layer.
-
-Detailed results of this experiment will be released in an upcoming
-paper. Figure~\ref{fig-fomc} highlights the result for the linear probe
-on the CPI. The chart shows various performance measures plotted against
-\emph{FOMC-RoBERTa}'s \(n\)-th layer. Shaded areas show the variation
-across cross-validation folds, where we have used an expanding window
-approach to split the time series. To avoid look-ahead bias, we run PCA
-separately for each training set. Results for the linear probes are
-shown along with results for autoregressive models (AR) used as our
-baseline.
-
-The first column shows the correlations between model predictions and
-observed inflation. One can observe that this correlation is strictly
-positive for the baseline models and the linear probes. Consistent with
-the findings in Gurnee and Tegmark (2023) and Alain and Bengio (2018),
-we also observe that this correlation tends to be higher for layers near
-the end of the transformer model. Those layers produce predictions that
-are more positively correlated with observed inflation than those
-produced by the corresponding autoregressive models.
-
-Similarly, we observe that the (root) mean squared error of the linear
-probe is largely on par with the baseline. The average prediction error
-is gradually reduced as we move along the horizontal axis, which is
-again indicative of the notion that layers closer to the end of the
-neural network have distilled more useful representations.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{results/figures/measures_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-fomc}Various performance measures (correlation and
-(root) mean squared error) plotted against \emph{FOMC-RoBERTa}'s
-\(n\)-th layer. Shaded areas indicate variation across cross-validation
-folds, where we have used an expanding window approach to split the time
-series. Results for the linear probes are shown along with results for
-autoregressive models (AR). The optimal lag length was determined using
-the Bayes Information Criterium.}
-
-\end{figure}%
-
-\paragraph{Stochastic Parrots After
-All?}\label{stochastic-parrots-after-all}
-
-These results from the linear probe shown in Figure~\ref{fig-fomc} are
-certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not
-explicitly trained to uncover associations between central bank
-communications and the level of consumer prices, it appears that the
-model has distilled representations that can be used to predict
-inflation. It is worth pointing out here that this model is
-substantially smaller than the models tested in Gurnee and Tegmark
-(2023). This begs the following question:
-
-\begin{quote}
-Have we uncovered further evidence that LLMs ``aren't mere stochastic
-parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic understanding
-of the economy just by `reading' central bank communications?
-\end{quote}
-
-Personally, I am having a very hard time believing this. To argue my
-case, I will now produce a counter-example demonstrating that, if
-anything, these findings are very much in line with the parrot metaphor.
-The counter-example is based on the following premise: if the results
-from the linear probe truly were indicative of some intrinsic
-understanding of the economy, then the probe should not be sensitive to
-random sentences that are most definitely not related to consumer
-prices.
-
-To test this, I select the best-performing probe trained on the
-final-layer activations to predict changes in the CPI. I then make up
-sentences that fall into one of these four categories:
-\emph{Inflation/Prices} (IP)---sentences about price inflation,
-\emph{Deflation/Prices} (DP)---sentences about price deflation,
-\emph{Inflation/Birds} (IB)---sentences about an inflation in the number
-of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation
-in the number of birds. A sensible sentence for category DP, for
-example, could be: ``It is essential to bring inflation back to target
-to avoid drifting into deflation territory.''. Analogically, we could
-construct the following sentence for the DB category: ``It is essential
-to bring the numbers of doves back to target to avoid drifting into
-dovelation territory.''.
-
-In light of the encouraging results for the probe in
-Figure~\ref{fig-fomc}, we should expect the probe to predict higher
-levels of inflation for activations for sentences in the IP category
-than for sentences in the DP category. If this was indicative of true
-intrinsic understanding, we would not expect to see any significant
-difference in predicted inflation levels for sentences about birds,
-independent of whether or not their numbers are increasing. More
-specifically, we would not expect the probe to predict values for
-sentences about birds that are substantially different from the values
-it can be expected to predict when using actual white noise as inputs.
-
-To get to this last point, I also generate many probe predictions for
-samples of noise. Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote
-the linear probe that maps from the \(k\)-dimensional space spanned by
-\(k\) first principal components of the final-layer activations to the
-output variable of interest (CPI growth in this case). Then I sample
-\(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\)
-for \(i \in [1,1000]\) and compute the sample average. I repeat this
-process \(10000\) times and compute the median-of-means to get an
-estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\),
-that is the predicted value of the probe conditional on white noise.
-
-Next, I propose the following hypothesis test as a minimum viable
-testing framework to assess if the probe results (may) provide evidence
-for an actual understanding of key economic relationships learned purely
-from text:
-
-\begin{proposition}[Parrot
-Test]\protect\hypertarget{prp-line}{}\label{prp-line}
-
-~
+\hypertarget{example-projecting-a-world-model-onto-a-latent-space}{%
+\subsection{Example: Projecting a World Model onto a latent
+space}\label{example-projecting-a-world-model-onto-a-latent-space}}
 
@@
 \item
-  \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are
-  statistically significantly different from
-  \(\mathbb{E}[f(\varepsilon)]\)
+  Take a variational autoencoder (VAE) that was pre-trained on MNIST.
 \item
-  \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts
-  values that are statistically significantly different from
-  \(\mathbb{E}[f(\varepsilon)]\) for sentences in all categories
-  (IP,DP,IB,DB).
+  Take the
+  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
+  that was used in Gurnee and Tegmark (2023), which maps locations/areas
+  to their latitude and longitude. For each place, it also indicates the
+  corresponding country.
 \item
-  \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The
-  probe predicts values that are statistically significantly different
-  from \(\mathbb{E}[f(\varepsilon)]\) for sentences in categories IP and
-  DP, but not for sentences in IB and DB.
+  Take the subset that contains countries that are currently part of the
+  top 10
+  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
+  world ranking}.
+\item
+  Assign the current rank to each country, i.e.~Argentina gets 1, France
+  gets 2, \ldots{}
+\item
+  Encode all features excluding the FIFA world rank indicator but
+  including \texttt{longitude} and \texttt{latitude} as continuous
+  variables: \(X^{(n \times d)}\) where \(n\) is the number of samples
+  and \(d\) is the number of resulting features.
+\item
+  Initialize a small neural network, let's call it a \emph{projector},
+  that maps from \(X\) to a single small hidden layer with sigmoid
+  activation and then form there to the latent space
+  \(\mathcal{Z} \subset \mathbb{R}\) of the pre-trained VAE.
+\item
+  Without performing any training on the \emph{projector}, simply
+  compute a forward pass of the continuously encoded data \(X\) through
+  the \emph{projector} and store the output:
+  \(\mathbf{Z}^{(n \times k)}\) where \(k\) is the dimension of the
+  latent space of the VAE.
+\item
+  Next, we perform the linear probe on \(\mathbf{Z}\) through Ridge
+  regression:
+  \(\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}\)
+  where \(\mathbf{Y}\) is the \((n \times 2)\) matrix containing the
+  longitude and latitude for each sample.
+\item
+  Finally, compute the predicted coordinate for each samples as
+  \(\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}\)
+\item
+  Plot the results on a world map and be amazed.
 \end{itemize}
 
-\end{proposition}
-
-To be clear, if in such a test we did find substantial evidence in
-favour of rejecting both \emph{HO} and \emph{H1}, this would not
-automatically imply that \emph{H2} is true. But to even continue
-investigating if based on having learned meaningful representation the
-underlying LLM is more than just a parrot, it should be able to pass
-this simple test.
-
-In this particular case, Figure~\ref{fig-attack} demonstrates that we
-find some evidence to reject \emph{H0} but not \emph{H1} for
-\emph{FOMC-RoBERTa}. The median linear probe predictions for sentences
-about inflation and deflation are indeed substantially higher and lower,
-respectively than for random noise. Unfortunately, the same is true for
-sentences about the inflation and deflation in the number of birds,
-albeit to a somewhat lower degree.
-
-I should note that the number of sentences in each category is very
-small here (10), so the results in Figure~\ref{fig-attack} cannot be
-used to establish statistical significance. That being said, even a
-handful of convincing counter-examples should be enough for us to
-seriously question the claim that results from linear probes provide
-evidence against the parrot metaphor. In fact, if there's even a single
-sentence for which any
-
 \begin{quote}
-Strong alternative hypothesis to address own confirmation bias
-Articulating a test to better differentiate
+I think what's going on here is that we essentially randomly project a
+small set of features into a higher-dimensional space (\(d < k\)). The
+linear probe is then having an easy time separating features in the
+high-dimensional space, even though the projection is completely random.
+Even though one might be momentarily amazed by this, there is really
+nothing surprising at all going on here. We can still draw a very
+interesting parallel to Gurnee and Tegmark (2023): we know that
+geographical coordinates are in the training corpus of Llama-2; that
+information gets passed through a massive latent space (the model); they
+run probes on a subsapce of the model (still massive) and manage to
+linearly separate their variable of interest (coordinates). I don't know
+how accurate that parallel really is but it just seems striking to me
+that ultimately what they is regressing a 2-dimensional output variable
+on this massive feature matrix (``ranging from 4,096 to 8,192 features''
+for roughly 40,000 total samples, can't figure out how big the
+train/test sets are, respectively.). I'm not good enough at theory but
+maybe there's really just no surprise at all here?
 \end{quote}
 
-\begin{figure}
+Relatedly, it is worth pointing out that
 
-\centering{
+\begin{itemize}
+\tightlist
+\item
+  a recent paper has shown that the double-descent phenomenon is not
+  exclusive to deep learning but also applies to linear regression with
+  many features {[}LINK{]}.
+\end{itemize}
 
-\includegraphics{results/figures/attack_inflation.png}
+\hypertarget{example-pca}{%
+\subsection{Example: PCA}\label{example-pca}}
 
-}
+\hypertarget{example-autoencoder}{%
+\subsection{Example: Autoencoder}\label{example-autoencoder}}
 
-\caption{\label{fig-attack}Probe predictions for sentences about
-inflation of prices (IP), deflation of prices (DP), inflation of birds
-(IB) and deflation of birds (DB). The vertical axis shows predicted
-inflation levels subtracted by the average predicted inflation level for
-the whole sample period (train and test set).}
+\hypertarget{example-llm}{%
+\subsection{Example: LLM}\label{example-llm}}
 
-\end{figure}%
-
-\subsection*{References}\label{references}
+\hypertarget{references}{%
+\subsection*{References}\label{references}}
 \addcontentsline{toc}{subsection}{References}
 
-\phantomsection\label{refs}
+\hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-alain2018understanding}
-Alain, Guillaume, and Yoshua Bengio. 2018. {``Understanding Intermediate
-Layers Using Linear Classifier Probes.''}
-\url{https://arxiv.org/abs/1610.01644}.
-
-\bibitem[\citeproctext]{ref-gurnee2023language}
+\leavevmode\vadjust pre{\hypertarget{ref-gurnee2023language}{}}%
 Gurnee, Wes, and Max Tegmark. 2023. {``Language Models Represent Space
 and Time.''} \url{https://arxiv.org/abs/2310.02207}.
-
-\bibitem[\citeproctext]{ref-liu2019roberta}
-Liu, Yinhan, Myle Ott, Naman Goyal, Jingfei Du, Mandar Joshi, Danqi
-Chen, Omer Levy, Mike Lewis, Luke Zettlemoyer, and Veselin Stoyanov.
-2019. {``RoBERTa: A Robustly Optimized BERT Pretraining Approach.''}
-\url{https://arxiv.org/abs/1907.11692}.
-
-\bibitem[\citeproctext]{ref-shah2023trillion}
-Shah, Agam, Suvan Paturi, and Sudheer Chava. 2023. {``Trillion Dollar
-Words: A New Financial Dataset, Task \& Market Analysis.''}
-\url{https://arxiv.org/abs/2305.07972}.
 

### 2024-01-25 src/mnist/run_experiment.jl

@@ -22,7 +22,6 @@ FIGURE_DIR = joinpath(RESULTS_DIR, "figures")
 data = load_mnist()
 X = data.X
 vae = load_mnist_vae()
-mlp = load_mnist_mlp().model
 
 # World Data (from https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv)
 world_data = CSV.read("data/world_place.csv", DataFrame)
@@ -48,14 +47,12 @@ fifa_world_data = DataFrames.subset(world_data, :country => ByRow(x -> haskey(fi
     x -> DataFrames.transform(x, :country => ByRow(x -> fifa_world_ranking[x]) => :y) |>
     x -> DataFrames.select(x, :y, Not(:y, :country))
 
-# Tokenizer for FIFA World Data ====================
-# Goal: need a tokenizer that can map from the entities to the latent space of the VAE.
+# Projector for FIFA World Data ====================
+# Goal: need a projector that can map from the entities to the latent space of the VAE.
+
 # Continuous feature encoding:
-X = fifa_world_data[:,Not(:y)]
-model = (X -> coerce(
-        X,
-        :entity_type => Multiclass,
-)) |> 
+X = fifa_world_data[:,Not([:y])]
+model = (X -> coerce(X,:entity_type => Multiclass,)) |> 
     MLJModels.FillImputer() |>
     MLJModels.ContinuousEncoder() |> 
     MLJModels.Standardizer() 
@@ -70,47 +67,28 @@ y = fifa_world_data.y
 ytrain = OneHotArrays.onehotbatch(y, 0:9)
 # Dataloader:
 dl = Flux.DataLoader((Xtrain, ytrain), batchsize=32, shuffle=true)
-# Tokenizer:
+
+# Projector:
 latent = 64
-activation = relu
-function head(Xhat)
-    return mlp(Xhat)
-end
-# A small MLP as our backbone, then a linear layer to map to the latent space:
-tokenizer = Chain(
+activation = sigmoid
+projector = Chain(
     Dense(size(Xtrain, 1) => latent, activation),
-    Dense(latent => latent, activation),
     Dense(latent => vae.params.latent_dim),
 )
-# The decoder of our VAE:
-reconstructor = Chain(
-    vae.decoder,
-    x -> clamp.(x, -1, 1),
-)
[…]

### 2024-01-25 technical_descriptions.qmd
just a random projection of the training data

@@
 
-## Example: Projecting a World Model onto a latent space
+## Example: Random Projections of World Models
 
-- Take a variational autoencoder (VAE) that was pre-trained on MNIST. 
 - Take the [world_place.csv]( https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv) that was used in @gurnee2023language, which maps locations/areas to their latitude and longitude.
@@
 - Encode all features excluding the FIFA world rank indicator but including `longitude` and `latitude` as continuous variables: $X^{(n \times d)}$ where $n$ is the number of samples and $d$ is the number of resulting features.
-- Initialize a small neural network, let's call it a *projector*, that maps from $X$ to a single small hidden layer with sigmoid activation and then form there to the latent space $\mathcal{Z} \subset \mathbb{R}$ of the pre-trained VAE.
+- Initialize a small neural network, let's call it a *projector*, that maps from $X$ to a single small hidden layer with sigmoid activation and then from there to a lower-dimensional output space $\mathcal{Z} \subset \mathbb{R}$.
 - Without performing any training on the *projector*, simply compute a forward pass of the continuously encoded data $X$ through the *projector* and store the output: $\mathbf{Z}^{(n \times k)}$ where $k$ is the dimension of the latent space of the VAE.
@@
 
-> I think what's going on here is that we essentially randomly project a small set of features into a higher-dimensional space ($d < k$).
-The linear probe is then having an easy time separating features in the high-dimensional space, even though the projection is completely random.
-Even though one might be momentarily amazed by this, there is really nothing surprising at all going on here.
-We can still draw a very interesting parallel to @gurnee2023language: we know that geographical coordinates are in the training corpus of Llama-2; that information gets passed through a massive latent space (the model); they run probes on a subsapce of the model (still massive) and manage to linearly separate their variable of interest (coordinates).
+> We can draw a very interesting parallel to @gurnee2023language: we know that geographical coordinates are in the training corpus of Llama-2; that information gets passed through a massive latent space (the model); they run probes on a subsapce of the model (still massive) and manage to linearly separate their variable of interest (coordinates).
 I don't know how accurate that parallel really is but it just seems striking to me that ultimately what they is regressing a 2-dimensional output variable on this massive feature matrix ("ranging from 4,096 to 8,192 features" for roughly 40,000 total samples, can't figure out how big the train/test sets are, respectively.).
-I'm not good enough at theory but maybe there's really just no surprise at all here?
 

### 2024-01-25 technical_descriptions.tex
just a random projection of the training data

@@
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 
-\hypertarget{example-projecting-a-world-model-onto-a-latent-space}{%
-\subsection{Example: Projecting a World Model onto a latent
-space}\label{example-projecting-a-world-model-onto-a-latent-space}}
+\hypertarget{example-random-projections-of-world-models}{%
+\subsection{Example: Random Projections of World
+Models}\label{example-random-projections-of-world-models}}
 
@@
 \tightlist
-\item
-  Take a variational autoencoder (VAE) that was pre-trained on MNIST.
 \item
@@
   that maps from \(X\) to a single small hidden layer with sigmoid
-  activation and then form there to the latent space
-  \(\mathcal{Z} \subset \mathbb{R}\) of the pre-trained VAE.
+  activation and then from there to a lower-dimensional output space
+  \(\mathcal{Z} \subset \mathbb{R}\).
 \item
@@
 \begin{quote}
-I think what's going on here is that we essentially randomly project a
-small set of features into a higher-dimensional space (\(d < k\)). The
-linear probe is then having an easy time separating features in the
-high-dimensional space, even though the projection is completely random.
-Even though one might be momentarily amazed by this, there is really
-nothing surprising at all going on here. We can still draw a very
-interesting parallel to Gurnee and Tegmark (2023): we know that
-geographical coordinates are in the training corpus of Llama-2; that
-information gets passed through a massive latent space (the model); they
-run probes on a subsapce of the model (still massive) and manage to
-linearly separate their variable of interest (coordinates). I don't know
-how accurate that parallel really is but it just seems striking to me
-that ultimately what they is regressing a 2-dimensional output variable
-on this massive feature matrix (``ranging from 4,096 to 8,192 features''
-for roughly 40,000 total samples, can't figure out how big the
-train/test sets are, respectively.). I'm not good enough at theory but
-maybe there's really just no surprise at all here?
+We can draw a very interesting parallel to Gurnee and Tegmark (2023): we
+know that geographical coordinates are in the training corpus of
+Llama-2; that information gets passed through a massive latent space
+(the model); they run probes on a subsapce of the model (still massive)
+and manage to linearly separate their variable of interest
+(coordinates). I don't know how accurate that parallel really is but it
+just seems striking to me that ultimately what they is regressing a
+2-dimensional output variable on this massive feature matrix (``ranging
+from 4,096 to 8,192 features'' for roughly 40,000 total samples, can't
+figure out how big the train/test sets are, respectively.).
 \end{quote}

### 2024-01-25 src/mnist/run_experiment.jl

@@ -1,7 +1,6 @@
 using Pkg; Pkg.activate("src/mnist")
 
 using CounterfactualExplanations.Data: load_mnist
-using CounterfactualExplanations.Models: load_mnist_vae, load_mnist_mlp
 using CSV
 using DataFrames
 using Flux
@@ -21,7 +20,6 @@ FIGURE_DIR = joinpath(RESULTS_DIR, "figures")
 # Load MNIST data and pre-trained models:
 data = load_mnist()
 X = data.X
-vae = load_mnist_vae()
 
 # World Data (from https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv)
 world_data = CSV.read("data/world_place.csv", DataFrame)
@@ -73,7 +71,7 @@ latent = 64
 activation = sigmoid
 projector = Chain(
     Dense(size(Xtrain, 1) => latent, activation),
-    Dense(latent => vae.params.latent_dim),
+    Dense(latent => 8),
 )
 
 # Linear Probes ====================

### 2024-01-26 src/mnist/run_experiment.jl
world map experiment

@@ -50,6 +50,14 @@ fifa_world_data = DataFrames.subset(world_data, :country => ByRow(x -> haskey(fi
 
 # Continuous feature encoding:
 X = fifa_world_data[:,Not([:y])]
+
+# Noisy, correlated series of longitude and latitude:
+ϕ = 5.0    # standard deviation of noise
+ρ = 0.5     # correlation with original longitude and latitude
+X.longitude = ρ .* X.longitude .+ (1-ρ) .* ϕ .* randn(size(X, 1))
+X.latitude = ρ .* X.latitude .+ (1-ρ) .* ϕ .* randn(size(X, 1))
+
+# Encode the features:
 model = (X -> coerce(X,:entity_type => Multiclass,)) |> 
     MLJModels.FillImputer() |>
     MLJModels.ContinuousEncoder() |> 
@@ -60,28 +68,34 @@ Xtrain = MLJBase.transform(mach, X) |>
     MLJBase.matrix |> 
     permutedims |>
     x -> Float32.(x)
+
+# Add noise:
+nnoise = 1000        # add nnoise noisy and uncorrelated features
+Xtrain = vcat(Xtrain, randn(nnoise, size(Xtrain, 2))) |> 
+    x -> Float32.(x)
+d = size(Xtrain, 1)
+
 # One-hot encoding:
-y = fifa_world_data.y
-ytrain = OneHotArrays.onehotbatch(y, 0:9)
-# Dataloader:
-dl = Flux.DataLoader((Xtrain, ytrain), batchsize=32, shuffle=true)
+y = fifa_world_data.y .+ 1
 
 # Projector:
-latent = 64
+latent = 1000
+nfinal = 2
 activation = sigmoid
 projector = Chain(
     Dense(size(Xtrain, 1) => latent, activation),
-    Dense(latent => 8),
+    Dense(latent => nfinal),
 )
 
 # Linear Probes ====================
 λ = 0.1
-A = projector(Xtrain) |> permutedims
+A = Flux.activations(projector, Xtrain) |> 
+    _A -> _A[end-1] |>
+    permutedims
 Y = fifa_world_data[:, [:longitude, :latitude]] |> matrix
 W = (A'A + UniformScaling(λ)) \ A'Y
 
 # Fitted values:
-labels = OneHotArrays.onecold(ytrain)
[…]

### 2024-01-26 main.tex
Updates from Overleaf

@@
 
-Recently, Google DeepMind made claims their AlphaGeometry model reached a 'milestone' towards AGI.
+Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
 This model has the ability to solve complex geometry problems, which, they state it does without the need for human demonstrations during training.
 Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
-As discussed in the paper \cite{}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
+As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
 With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
 
-However, it is worth noting, that work such as this had been initially introduced in the 1950s.
+However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s.
 Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry.
@@
 
-One of the more contentious pieces of literature recently studied is the paper by \cite{gurnee2023language}
+One of the more contentious pieces of literature recently studied is the paper by \cite{gurnee2023language}.
+This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com} (formally Twitter) platform.
+Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
+
+They claim said models generate an internal representation of the world, with which spatiotemporal representations are created.
+Further to this we see claim to the generalization of tasks, in which the model can answer questions regarding spatial and temporal relationships, without the need for further training.
+Although these can be seen as building blocks for a more comprehensive model, it is also fair to assume that it is mere conjecture and more likely spurious correlations.
 
@@
 pitfalls of finding patterns in latent spaces.
+
+\subsection{Example: Injecting a World Model}\label{example-deep-learning}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.5\textwidth]{results/figures/map.png}
+
+}
+
+\caption{\label{fig-map}Mapping MNIST digits to FIFA World Rankings.}
+
+\end{figure}%
 
@@
 
-\subsection{Example: Deep Learning}\label{example-deep-learning}
+\subsection{Example: Autoencoder}\label{example-deep-learning}
 
@@
 
-\includegraphics[width=0.5\textwidth]{results/figures/dl.png}
+\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
 
@@
 
-\caption{\label{fig-dl-results}The top panel shows the actual GDP growth
-and fitted values from the autoencoder model. The bottom panel shows the
+\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
+and fitted values from the autoencoder model. The right chart shows the
 observed average level and spread of the yield curve (solid) along with
-the predicted values from the linear probe based on the latent embedding
+the predicted values (in-sample) from the linear probe based on the latent embeddings
 (dashed).}
 
-\end{figure}%
+\end{figure*}%
 
@@
 To this end, we have further pre-processed the data provided by \cite{shah2023trillion} and used their proposed model to compute layer-wise embeddings for all available sentences.
-We have made these available and easily accessible through a small Julia package: \href{https://github.com/pat-alt/TrillionDollarWords.jl}{TrillionDollarWords.jl}.
+We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.
 For each layer, we have then computed linear probes on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.

### 2024-01-26 technical_descriptions.qmd
map done

@@
 
+The Llama-2 model has ingested huge amounts of publicly available data from the internet including Wikipedia dumps from the June-August 2022 period [@touvron2023llama].
+It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
+At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
+The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance, before being passed through a series of layers which are gradually trained for downstream tasks.
+
+In our first example in this section, we simulate this scenario, stop short of training the model. In particular, we
+
 - Take the [world_place.csv]( https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv) that was used in @gurnee2023language, which maps locations/areas to their latitude and longitude.
@@
 - Assign the current rank to each country, i.e. Argentina gets 1, France gets 2, ...
-- Encode all features excluding the FIFA world rank indicator but including `longitude` and `latitude` as continuous variables: $X^{(n \times d)}$ where $n$ is the number of samples and $d$ is the number of resulting features.
-- Initialize a small neural network, let's call it a *projector*, that maps from $X$ to a single small hidden layer with sigmoid activation and then from there to a lower-dimensional output space $\mathcal{Z} \subset \mathbb{R}$.
-- Without performing any training on the *projector*, simply compute a forward pass of the continuously encoded data $X$ through the *projector* and store the output: $\mathbf{Z}^{(n \times k)}$ where $k$ is the dimension of the latent space of the VAE.
+- Transform the longitude and latitude data, respectively, as follows: $\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon$ where $\rho=0.5$ and $\epsilon \sim \mathcal{N}(0, 5)$.
+This is to ensure that the training data only involves a noisy version of the coordinates.
+- Next, encode all features except the FIFA world rank indicator as continuous variables: $X^{(n \times m)}$ where $n$ is the number of samples and $m$ is the number of resulting features.
+- Additionally, add a large number of random features to $X$ to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
+Let $d$ denote the final number of features, i.e.
+$d=m+k$ where $k$ is the number of random features.
+- Initialize a small neural network, let's call it a *projector*, that maps from $X$ to a single hidden layer with $h<d$ hidden units and sigmoid activation and then from there to a lower-dimensional output space.
+- Without performing any training on the *projector*, we simply compute a forward pass of $X$ through the *projector* and retrieve the activations $\mathbf{Z}^{(n\times h)}$.
 - Next, we perform the linear probe on $\mathbf{Z}$ through Ridge regression: $\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}$ where $\mathbf{Y}$ is the $(n \times 2)$ matrix containing the longitude and latitude for each sample.
-- Finally, compute the predicted coordinate for each samples as $\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}$
-- Plot the results on a world map and be amazed.
+- Finally, we compute the predicted coordinates for each samples as $\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}$ and plot the results on a world map (see @fig-map).
 
-> We can draw a very interesting parallel to @gurnee2023language: we know that geographical coordinates are in the training corpus of Llama-2; that information gets passed through a massive latent space (the model); they run probes on a subsapce of the model (still massive) and manage to linearly separate their variable of interest (coordinates).
-I don't know how accurate that parallel really is but it just seems striking to me that ultimately what they is regressing a 2-dimensional output variable on this massive feature matrix ("ranging from 4,096 to 8,192 features" for roughly 40,000 total samples, can't figure out how big the train/test sets are, respectively.).
+![Probe predictions for a random projection.](results/figures/map.png){#fig-map}
 
-Relatedly, it is worth pointing out that
-
-- a recent paper has shown that the double-descent phenomenon is not exclusive to deep learning but also applies to linear regression with many features [LINK].
+While this is admittedly a somewhat contrived example, it serves to illustrate the point that if we randomly project a small number of features that are highly correlated or predictive of the target into a large enough space, we should expect to be able to recover meaningful represenations of the target from the latent space.
+Similarly, @alain2018understanding observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
+In fact, it is well-known that random projections can be used for prediction tasks [@dasgupta2013experiments].
 

### 2024-01-26 technical_descriptions.tex
map done

@@
 \begin{document}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, interior hidden, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 Models}\label{example-random-projections-of-world-models}}
+
+The Llama-2 model has ingested huge amounts of publicly available data
+from the internet including Wikipedia dumps from the June-August 2022
+period (Touvron et al. 2023). It is therefore highly likely that the
+training data contains geographical coordinates, either directly or
+indirectly. At the very least, we should expect that the model has seen
+features during training that are highly correlated with geographical
+coordinates. The model itself is essentially a very large latent space
+to which all features are randomly projected in the very first instance,
+before being passed through a series of layers which are gradually
+trained for downstream tasks.
+
+In our first example in this section, we simulate this scenario, stop
+short of training the model. In particular, we
 
@@
 \item
-  Encode all features excluding the FIFA world rank indicator but
-  including \texttt{longitude} and \texttt{latitude} as continuous
-  variables: \(X^{(n \times d)}\) where \(n\) is the number of samples
-  and \(d\) is the number of resulting features.
+  Transform the longitude and latitude data, respectively, as follows:
+  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
+  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\). This is to
+  ensure that the training data only involves a noisy version of the
+  coordinates.
+\item
+  Next, encode all features except the FIFA world rank indicator as
+  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
+  of samples and \(m\) is the number of resulting features.
+\item
+  Additionally, add a large number of random features to \(X\) to
+  simulate the fact that not all features ingested by Llama-2 are
+  necessarily correlated with geographical coordinates. Let \(d\) denote
+  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
+  of random features.
 \item
   Initialize a small neural network, let's call it a \emph{projector},
-  that maps from \(X\) to a single small hidden layer with sigmoid
-  activation and then from there to a lower-dimensional output space
-  \(\mathcal{Z} \subset \mathbb{R}\).
+  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
+  units and sigmoid activation and then from there to a
+  lower-dimensional output space.
 \item
-  Without performing any training on the \emph{projector}, simply
-  compute a forward pass of the continuously encoded data \(X\) through
-  the \emph{projector} and store the output:
-  \(\mathbf{Z}^{(n \times k)}\) where \(k\) is the dimension of the
-  latent space of the VAE.
+  Without performing any training on the \emph{projector}, we simply
+  compute a forward pass of \(X\) through the \emph{projector} and
+  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
 \item
@@
 \item
-  Finally, compute the predicted coordinate for each samples as
-  \(\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}\)
-\item
-  Plot the results on a world map and be amazed.
+  Finally, we compute the predicted coordinates for each samples as
+  \(\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}\) and plot the results on
+  a world map (see Figure~\ref{fig-map}).
 \end{itemize}
 
-\begin{quote}
-We can draw a very interesting parallel to Gurnee and Tegmark (2023): we
-know that geographical coordinates are in the training corpus of
-Llama-2; that information gets passed through a massive latent space
-(the model); they run probes on a subsapce of the model (still massive)
-and manage to linearly separate their variable of interest
-(coordinates). I don't know how accurate that parallel really is but it
-just seems striking to me that ultimately what they is regressing a
-2-dimensional output variable on this massive feature matrix (``ranging
-from 4,096 to 8,192 features'' for roughly 40,000 total samples, can't
-figure out how big the train/test sets are, respectively.).
-\end{quote}
+\begin{figure}
 
-Relatedly, it is worth pointing out that
+{\centering \includegraphics{results/figures/map.png}
 
-\begin{itemize}
-\tightlist
-\item
-  a recent paper has shown that the double-descent phenomenon is not
-  exclusive to deep learning but also applies to linear regression with
-  many features {[}LINK{]}.
-\end{itemize}
+}
+
+\caption{\label{fig-map}Probe predictions for a random projection.}
+
+\end{figure}
+
+While this is admittedly a somewhat contrived example, it serves to
+illustrate the point that if we randomly project a small number of
+features that are highly correlated or predictive of the target into a
+large enough space, we should expect to be able to recover meaningful
+represenations of the target from the latent space. Similarly, Alain and
+Bengio (2018) observe that even before training a convolutional neural
+network on MNIST data, the layer-wise activations can already be used to
+perform binary classification. In fact, it is well-known that random
+projections can be used for prediction tasks (Dasgupta 2013).
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-alain2018understanding}{}}%
+Alain, Guillaume, and Yoshua Bengio. 2018. {``Understanding Intermediate
+Layers Using Linear Classifier Probes.''}
+\url{https://arxiv.org/abs/1610.01644}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dasgupta2013experiments}{}}%
+Dasgupta, Sanjoy. 2013. {``Experiments with Random Projection.''}
+\url{https://arxiv.org/abs/1301.3849}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-gurnee2023language}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-touvron2023llama}{}}%
+Touvron, Hugo, Thibaut Lavril, Gautier Izacard, Xavier Martinet,
+Marie-Anne Lachaux, Timothée Lacroix, Baptiste Rozière, et al. 2023.
+{``LLaMA: Open and Efficient Foundation Language Models.''}
+\url{https://arxiv.org/abs/2302.13971}.
+
 \end{CSLReferences}

### 2024-01-26 src/mnist/run_experiment.jl

@@ -70,7 +70,7 @@ Xtrain = MLJBase.transform(mach, X) |>
     x -> Float32.(x)
 
 # Add noise:
-nnoise = 1000        # add nnoise noisy and uncorrelated features
+nnoise = 1500        # add nnoise noisy and uncorrelated features
 Xtrain = vcat(Xtrain, randn(nnoise, size(Xtrain, 2))) |> 
     x -> Float32.(x)
 d = size(Xtrain, 1)

### 2024-01-26 src/mnist/run_experiment.jl
random projection probe prediciton on test set

@@ -53,7 +53,7 @@ X = fifa_world_data[:,Not([:y])]
 
 # Noisy, correlated series of longitude and latitude:
 ϕ = 5.0    # standard deviation of noise
-ρ = 0.5     # correlation with original longitude and latitude
+ρ = 0.75     # correlation with original longitude and latitude
 X.longitude = ρ .* X.longitude .+ (1-ρ) .* ϕ .* randn(size(X, 1))
 X.latitude = ρ .* X.latitude .+ (1-ρ) .* ϕ .* randn(size(X, 1))
 
@@ -70,7 +70,7 @@ Xtrain = MLJBase.transform(mach, X) |>
     x -> Float32.(x)
 
 # Add noise:
-nnoise = 1500        # add nnoise noisy and uncorrelated features
+nnoise = 1490        # add nnoise noisy and uncorrelated features
 Xtrain = vcat(Xtrain, randn(nnoise, size(Xtrain, 2))) |> 
     x -> Float32.(x)
 d = size(Xtrain, 1)
@@ -92,6 +92,14 @@ projector = Chain(
 A = Flux.activations(projector, Xtrain) |> 
     _A -> _A[end-1] |>
     permutedims
+train_prp = 0.75
+ntrain = Int(round(train_prp * size(A, 1)))
+shuffled_ids = shuffle(1:size(A, 1))
+train_idx = shuffled_ids[1:ntrain]
+test_idx = shuffled_ids[ntrain+1:end]
+Atrain = A[train_idx, :]
+Atest = A[test_idx, :]
+ytest = y[test_idx]
 Y = fifa_world_data[:, [:longitude, :latitude]] |> matrix
 W = (A'A + UniformScaling(λ)) \ A'Y
 
@@ -101,7 +109,7 @@ C = makecpt(
     cmap=:categorical,
     range=reduce((x, y) -> "$x,$y", sorted_names)
 )
-Ŷ = A * W
+Ŷ = Atest * W
 coast(;
     region=:global,
     proj = :Mollweide,
@@ -113,7 +121,7 @@ GMT.scatter!(
     Ŷ[:, 1],
     Ŷ[:, 2];
     color=C.colormap,
-    zcolor=y,
+    zcolor=ytest,
     cmap=C,
     colorbar=true,
     show = true,

### 2024-01-26 main.tex
Updates from Overleaf

@@
 
-To support my claim that observing patterns in latent spaces should not
-generally surprise us, we will now go through a couple of simple
-examples. To illustrate further that this phenomenon is neither
-surprising nor unique to the field of computer science, we will borrow examples from the fields of Economics and Finance in this section.
+To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples.
+To illustrate further that this phenomenon is neither surprising nor unique to the field of computer science, we will borrow draw on ideas from the fields of Economics and Finance in this section.
 We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data.
-The final example in Section \ref{ex-llm} is a bit more involved and closer in spirit to the experiments conducted by \cite{gurnee2023language}.
-As we go along, we will try to discuss both the benefits and potential
-pitfalls of finding patterns in latent spaces.
+The final example in Section \ref{ex-llm} is a bit more involved and closer in spirit to the experiments conducted by \citet{gurnee2023language}.
+As we go along, we will try to discuss both the benefits and potential pitfalls of finding patterns in latent spaces.
 
-\subsection{Example: Injecting a World Model}\label{example-deep-learning}
+\subsection{Example: Random Projections of World Models}\label{example-deep-learning}
+
+The Llama-2 model has ingested huge amounts of publicly available data from the internet including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
+It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
+At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
+The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance, before being passed through a series of layers which are gradually trained for downstream tasks.
+
+In our first example in this section, we simulate this scenario, stopping short of training the model. In particular, we
+
+\begin{itemize}
+\tightlist
+\item
+  Take the
+  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
+  that was used in \cite{gurnee2023language}, which maps locations/areas
+  to their latitude and longitude. For each place, it also indicates the
+  corresponding country.
+\item
+  Take the subset that contains countries that are currently part of the
+  top 10
+  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
+  world ranking}.
+\item
+  Assign the current rank to each country, i.e.~Argentina gets 1, France
+  gets 2, \ldots{}
+\item
+  Transform the longitude and latitude data, respectively, as follows:
+  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
+  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\). This is to
+  ensure that the training data only involves a noisy version of the
+  coordinates.
+\item
+  Next, encode all features except the FIFA world rank indicator as
+  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
+  of samples and \(m\) is the number of resulting features.
+\item
+  Additionally, add a large number of random features to \(X\) to
+  simulate the fact that not all features ingested by Llama-2 are
+  necessarily correlated with geographical coordinates. Let \(d\) denote
+  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
+  of random features.
+\item
+  Initialize a small neural network, let's call it a \emph{projector},
+  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
+  units and sigmoid activation and then from there to a
+  lower-dimensional output space.
+\item
+  Without performing any training on the \emph{projector}, we simply
+  compute a forward pass of \(X\) through the \emph{projector} and
+  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
+\item
+  Next, we perform the linear probe on \(\mathbf{Z}\) through Ridge
+  regression:
+  \(\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}\)
+  where \(\mathbf{Y}\) is the \((n \times 2)\) matrix containing the
+  longitude and latitude for each sample.
+\item
+  Finally, we compute the predicted coordinates for each samples as
+  \(\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}\) and plot the results on
+  a world map (see Figure~\ref{fig-map}).
+\end{itemize}
 
@@
 
-\centering{
-
-\includegraphics[width=0.5\textwidth]{results/figures/map.png}
+{\centering \includegraphics[width=0.5\textwidth]{results/figures/map.png}
 
@@
 
-\caption{\label{fig-map}Mapping MNIST digits to FIFA World Rankings.}
+\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}
 
-\end{figure}%
+\end{figure}
+
+While this is admittedly a somewhat contrived example, it serves to illustrate the point that if we randomly project a small number of features that are highly correlated or predictive of the target into a large enough space, we should expect to be able to recover meaningful representations of the target from the latent space.
+Similarly, \cite{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
+In fact, it is well-known that random projections can be used for prediction tasks \cite{dasgupta2013experiments}.
 
@@
 
-In response to the claims made by Tegmark, numerous commentators on
-social media have pointed out that even the simplest of models can
-exhibit structure in their latent spaces. One of the most popular and
-illustrative examples I remember from my time at the Bank of England is
-yield curve decomposition through PCA. The yield curve is a popular tool
-for investors and economists to gauge the health of the economy. It
-plots the yields of bonds against their maturities. The slope of the
-yield curve is often used as a predictor of future economic activity: a
-steep yield curve is associated with a growing economy, while a flat or
-inverted yield curve is associated with a contracting economy.
+For our next example, we will move beyond randomness and illustrate how principal component analysis can be used to recover meaningful and compact latent representations from data in an unsupervised manner.
 
-% To understand this better, let us go on a quick detour into economics
-% and look at actual yield curves observed in the US during the Global
-% Financial Crisis (GFC). Figure~\ref{fig-gfc-1} shows the yield curve of
-% US Treasury bonds on 27 February 2007, which according to
-% \href{https://money.cnn.com/2007/02/27/markets/markets_0630/index.htm?cnn=yes}{CNN}
-% was a ``brutal day on Wall Street''.\footnote{The
-%   \href{https://home.treasury.gov/resource-center/data-chart-center/interest-rates/TextView?type=daily_treasury_yield_curve&field_tdr_date_value_month=202310}{data}
-%   is taken from the US Department of the Treasury.} This followed
-% \href{https://money.cnn.com/2007/02/26/news/economy/greenspan/index.htm?postversion=2007022609}{reports}
-% on the previous day of former Federal Reserve Chairman Alan Greenspan's
-% warning that the US economy was at risk of a recession. The yield curve
-% was inverted with a sharp negative spread between the 10-year and
-% 3-month yields, indicative of the market's expectation of a recession.
+The yield curve is a popular tool for investors and economists to gauge the health of the economy.
+It plots the yields of bonds against their maturities.
+The slope of the yield curve is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
 
-% Figure~\ref{fig-gfc-2} shows the corresponding yield curve during the
-% aftermath of the GFC on 20 April 2009. On that day the influential Time
-% Magazine
-% \href{https://content.time.com/time/business/article/0,8599,1890560,00.html}{reported}
-% that the ``Banking Crisis is Over''. The yield curve was steeply sloped
-% with a positive spread between the 10-year and 3-month yields,
-% indicative of the market's expectation of a recovery. The overall level
-% of the yield curve was still very low though, indicative of the fact
-% that US economy had not fully recovered at that point.
-
-Of course, US Treasuries are not the only bonds that are traded in the
-market. To get a more complete picture of the economy, analysts might
+Of course, US Treasuries are not the only bonds that are traded in the market. To get a more complete picture of the economy, analysts might
 therefore be interested in looking at the yield curves of other bonds as
@@
 
+\subsection{Discussion of Examples}
+
+Our examples should not be seen as an attempt to conclusively refute claims that LLMs can develop on actual "understanding" of the world by distilling knowledge in more compact representations.
+We hope that they serve to illustrate, however, that the use of diagnostic tools is potentially misleading and most definitely insufficient in providing evidence in favor of "understanding".
+Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
+Favourable outcomes from probes merely indicate that the model "has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
+Our examples demonstrate that this is achievable even for small models, that most readers will agree have certainly not developed an intrinsic "understanding" of the world.
+The fact that even these models distill meaningful representations goes to show that more conservative and rigorous tests are needed to draw conclusions about the emerging capabilities of AI models.
+
 \bibliography{biblio}

### 2024-01-26 main.tex
Updates from Overleaf

@@
 
-We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems: patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
+We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
+Such that, when an AI system begins finding patterns in the corpus not originally perceived during training, that is now from a human perspective, observed as greater cognitive capabilities.
+This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
+%Such as an AI system finding patterns in data not originally perceived during training
+%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
 In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g.
@@
 \item
-  Next, we perform the linear probe on \(\mathbf{Z}\) through Ridge
+  Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge
   regression:
-  \(\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}\)
-  where \(\mathbf{Y}\) is the \((n \times 2)\) matrix containing the
-  longitude and latitude for each sample.
+  \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\)
+  where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the
+  longitude and latitude for each sample. A hold-out set is reserved for testing.
 \item
-  Finally, we compute the predicted coordinates for each samples as
-  \(\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}\) and plot the results on
+  Finally, we compute the predicted coordinates for each sample in the hold-out set as
+  \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on
   a world map (see Figure~\ref{fig-map}).
+\item 
+  While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand.
 \end{itemize}
@@
 
-For our next example, we will move beyond randomness and illustrate how principal component analysis can be used to recover meaningful and compact latent representations from data in an unsupervised manner.
+For our next example, we will move beyond randomness and illustrate how principal component analysis (PCA) can be used to recover meaningful and compact latent representations from data in an unsupervised manner.
 
@@
 The slope of the yield curve is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
+While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone.
+To leverage this information in downstream modelling tasks, economists regularly make use of PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data.
+Recent examples of such include \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing} among others.
 
-Of course, US Treasuries are not the only bonds that are traded in the market. To get a more complete picture of the economy, analysts might
-therefore be interested in looking at the yield curves of other bonds as
-well. In particular, we might be interested in predicting economic
-growth based on the yield curves of many different bonds. The problem
-with that idea is that it is cursed by high dimensionality: we would end
-up modelling a single variable of interest (economic growth) with a
-large number of predictors (the yields of many different bonds). To deal
-with the curse of high dimensionality it can be useful to decompose the
-yield curves into sets of principal components.
-
-To compute the principal components we can decompose the matrix of yields \(\mathbf{Y}\) into a product of its singular vectors and values: \(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
-To put this into the broader context of the article, however, let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve (they are latent because they are not directly observable).
+Here we are primarily interested in understanding the nature this low-dimensional projection, so will limit our attention to the US Treasury yield curve.
+To this end, we have collected daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
+Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
+To put this into the broader context of this, let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.
 
@@
 components of the yield curves of US Treasury bonds over time. Vertical
-stalks indicate the key dates during the onset and aftermath of the
-crisis, which we discussed above. For both components, we can observe
-some marked shifts between the two dates - but can we attribute any
-meaning to these shifts? It turns out we can: for comparison, the bottom
-panel in Figure~\ref{fig-pca} shows the average level and spread of the
-yield curves over time. The first principal component is strongly
-correlated with the level of the yield curve, while the second principal
-component is strongly correlated with the spread of the yield curve. To
-put it in AI-lingo:
+stalks indicate key dates during the onset and aftermath of the
+Global Financial Crisis.
+During the onset, on 27 February 2007, financial markets were in turmoil following Federal Reserve Chairman Alan Greenspan's warning on the previous day that the US economy was at risk of a recession.
+Over the following months the Fed reacted to these economic and financial pressures by gradually cutting short-term interest rates to unprecedented lows, which affected the yield curve in two major ways: the average level of yields decreased and the curve steepened as the short end came down.
+Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
+In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve.
 
-\begin{quote}
-The estimated latent embeddings of the yield curve are characterized by
-patterns observed in the data.
-\end{quote}
+Of course, even though it is reassuring to see that these latent embeddings are meaningful, none of this is surprising at all: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance in the data after controlling for the effect of all previous components.
 
@@
 
-\caption{\label{fig-pca}Comparison of latent embeddings and observed
-data of the US Treasury yield curve.}
+\caption{\label{fig-pca}The first two principal components of US Treasury yields over time at daily frequency.
+Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
 
@@
 
-Not convinced? Let us use
-\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\) in true autoencoder
-fashion to reconstruct yield curves from principal components. Let
-\(z_1\) denote the first principal component and consider the following:
-we keep all other \(M-1\) principal components fixed at zero where \(M\)
-denotes the total number of maturities; next we traverse the latent
-space by varying the value of \(z_1\) over a fixed grid of length \(K\)
-each time storing the full vector \(\mathbf{z}\); finally, we vertically
-concatenate the vectors and end up with a matrix \(\mathbf{Z}\) of
-dimension \((K \times M)\). To reconstruct yields, we simply multiply
-\(Z\) by the singular values and right singular vectors:
-\(\mathbf{Y}=\mathbf{Z}\Sigma\mathbf{V}^{\prime}\).
-
 \subsection{Example: Autoencoder}\label{example-deep-learning}
 
-So far we have considered simple matrix decomposition. You might argue
-that principal components are not really latent embeddings in the
-traditional sense of deep learning. To address this, let us now consider
-a simple deep-learning example. Our goal will be to not only predict
-economic growth from the yield curve but also extract meaningful
-features at the same time. In particular, we will use a neural network
-architecture that allows us to recover a compressed latent
-representation of the yield curve.
+So far we have considered simple linear transformations of data.
+One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
+Our next example will therefore involve a deep neural network.
+
+Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks at the same time.
+In particular, we will use a neural network architecture that allows us to recover a compact latent representation of the yield curve.
 
@@
 gross domestic product (GDP) provided by the Federal Reserve Bank of
-St.~Louis. The data arrives in terms of levels of real GDP. In order to
-estimate growth, we will transform the data into log differences. Since
-our yield curve data is daily, we will need to aggregate it to the
-quarterly frequency. To do this, we will simply take the average of the
-daily yields for each maturity. We will also standardize yields since
-deep learning models tend to perform better with standardized data.
-Since COVID-19 was a huge structural break, we will also filter out all
-observations after 2018.
+St.~Louis. The data arrives in terms of levels of real GDP. In order to estimate growth, we transform the data using log differences. Since
+our yield curve data is daily, we aggregate it to the
+quarterly frequency by taking averages of daily yields for each maturity.
+We also standardize yields since deep learning models tend to perform better with standardized data (REFERENCE NEEDED?).
+Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
 
@@
 
-Let \(G_t\) denote growth and \(\mathbf{Y}_t\) denote the yield curve at
-time \(t\). Then we are interested in a model for \(G_t\) conditional on
-\(\mathbf{Y}_t\). Let \(\theta\) denote our model parameters then
-formally we are interested in maximizing the likelihood
-\(p_{\theta}(G_t|\mathbf{Y}_t)\).
+Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
+Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
+Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
 
-To do this, we will use a simple
-autoencoder architecture: the encoder will consist of a single fully
-connected hidden layer with 32 neurons and a hyperbolic tangent
-activation function. The bottleneck layer connecting the encoder to the
-decoder is a fully connected layer with 6 neurons. This is the
-compressed latent representation of the yield curve mentioned above. The
-decoder will consist of two fully connected layers each with a
-hyperbolic tangent activation function: the first layer will consist of
-32 neurons and the second layer will have the same dimension as the
-input data. The output layer will consist of a single neuron with a
-linear activation function. The model will be trained using the mean
-squared error loss function and the Adam optimizer.
+To do this, we use a simple autoencoder architecture: the encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
+The bottleneck layer, connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
+The decoder consists of two fully connected layers each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
+The output layer consists of a single neuron for our output variable, \(g_t\).
+We train the model over 1,000 epochs to minimize mean squared error loss using the the Adam optimizer \citep{kingma2017adam}.
+
+The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
+model.
+The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
+Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
+As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
+Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
 
@@
 
-The results are shown in Figure~\ref{fig-dl-results}. The top panel
-shows the actual GDP growth and fitted values from the autoencoder
-model. We observe that the model captures the relationship between
-economic growth and the yield curve reasonably well. As discussed above,
-we also know that the relationship between economic growth and the yield
-curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we
-would expect that these two important factors are reflected somewhere in
-the latent parameter space.
+While the loss function applies most direct pressure on layers near the final output layer of the neural network, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
+On a per-neuron basis, the pressure to distill useful representation is therefore likely maximised in this layer.
+Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
 
-The bottleneck layer seems like a good place to start looking. To get
-the latent embeddings \(\mathbf{A}_t\) at time \(t\), we simply pass the
-yield curve data \(\mathbf{Y}_t\) through the encoder. Next, we follow
-\cite{gurnee2023language} and use a linear probe to regress the observed
-yield curve factors on the latent embeddings. Let \(F_t\) denote the
-vector containing the two factors of interest in time \(t\):
-\(f_t^{\text{spread}}\) and \(f_t^{\text{level}}\). Formally, we are
-interested in the following regression model:
-\(p_{w}(F_t|\mathbf{A}_t)\) where \(w\) denotes the regression
-parameters. Following \cite{gurnee2023language}, we use Ridge regression
-with \(\lambda\) set to \(0.1\). Using the estimated regression
-parameters \(\hat{w}\) we can then predict the yield curve factors from
-the latent embeddings: \(\hat{F}_t=\hat{w}^{\prime}\mathbf{A}_t\).
+We compute and extract these activations \(A_t\) for all time-periods \(t=1,...,T\). Next, we use a linear probe to regress the observed
+yield curve factors on the latent embeddings.
+Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
+Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
+parameters.
+We use Ridge regression with \(\lambda\) set to \(0.1\).
+Using the estimated regression parameters \(\hat{w}\) we can then predict the yield curve factors from the latent embeddings: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
 
-The results of this experiment are shown in the bottom panel of
-Figure~\ref{fig-dl-results}. Solid lines show the observed yield curve
-factors over time, while dashed lines show predicted values. We find
-that the latent embeddings predict the two yield curve factors
-reasonably well, in particular the spread. To put this in true
-hype-lingo:
-
-\begin{quote}
-Not just a parrot! Our autoencoder neural network has an implicit
-understanding of the yield curve factors and their relationship with
-economic growth. It's sentient indeed!
-\end{quote}
+The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
+Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
+We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
+As in all of the other examples we have presented so, we do not find this result particularly surprising.
+It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
+But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
+While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
 
@@
 
-\subsubsection{Ok, but truly what's the
-point?}\label{ok-but-truly-whats-the-point}
+\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
 
-The finding is not surprising but it is still interesting. In the
-context of mechanistic interpretability, it demonstrates that the
-black-box model has evidently learned plausible explanations for the
-data. Beyond that, in this particular example, the patterns in the
-latent space that we have just uncovered might actually be useful for
-downstream tasks. An interesting idea could be to use the latent
-embeddings as features in a more traditional and interpretable
-econometric model. To demonstrate this, let us consider a simple linear
-regression model for GDP growth. We will compare the performance of the
-following models: (1) regressing growth \(G_t\) on the latent embedding
-\(A_t\), (2) regression growth on the best subset of latent embeddings
-\(A_t\), (3) regressing growth on lagged growth, (4) regressing growth
-on lagged growth and the observed yield curve factors \(F_t\), (5)
-regressing growth on the best subset of yield curve factors \(F_t\),
-and, finally, (5) regressing growth on lagged growth and the best subset
-of latent embeddings \(A_t\).
+An interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
+To demonstrate this, let us consider a simple linear regression model for GDP growth.
+We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
+As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
+But this coefficient suffers from confounding bias since, naturally, there many other confounding variables at play here, some of which may be readily observable and measurable, others of which may not be.
 
-The results are shown in the table below. The key finding of interest is
-that the coefficients on the latent embeddings in model (5) are
-statistically significant. This suggests that the latent embeddings
-contain information that is useful for predicting GDP growth.
+% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
+Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
 
+We have, for example, already mentioned above that there is relationship between interest rates and economic growth.
+To account for that while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
+While this does slightly change the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant as evident from column (2) in Table \ref{tab-reg}.
+This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
+Because we have included to additional regressors with little to know predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
+
+Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
+In particular, we have chosen the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
+The estimated coefficient on this latent factor is small in magnitude but statistically significant.
+The overall model fit as measured by the BIC as improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
+While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can actually be used in order to mitigate confounding bias.
+
+\begin{table}
+\caption{Regression output for various models.}\label{tab-reg}%
 \input{results/regression.tex}
+\end{table}
 
@@
 
-To round up this section, we will jump back on the hype train and
-consider an example involving an LLM. In particular, we will closely
-follow the approach in \cite{gurnee2023language} and apply it to a novel
-financial dataset: the \emph{Trillion Dollar Words} dataset introduced
-by \cite{shah2023trillion}. The dataset contains a curated
-selection of sentences formulated by central bankers of the US Federal
-Reserve and communicated to the public in speeches, meeting minutes and
-press conferences. The authors of the paper use this dataset to train
-LLMs to classify sentences as either `dovish', `hawkish' or `neutral'.
-To this end, they first manually annotate a subsample of the available
-data and then fine-tune various foundation models. Their model of
-choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \cite{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) for the
-classification task. To illustrate the potential usefulness of the
-learned classifier, they use predicted labels for the entire dataset to
-compute an ad-hoc, count-based measure of `hawkishness'. They then go on
-to show that this measure correlates with key economic indicators in the
-expected direction: when inflationary pressures rise, the measured level
-of `hawkishness' increases as central bankers need to raise interest
-rates to bring inflation back to target.
+To round up this section, we return to the world of `GenAI' and consider an example involving an LLM.
+In particular, we closely follow the approach in \citet{gurnee2023language} and apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset introduced by \citet{shah2023trillion}.
+The dataset contains a curated selection of sentences formulated by central bankers of the US Federal Reserve and communicated to the public in speeches, meeting minutes and press conferences.
+The authors of the paper use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
+To this end, they first manually annotate a sub-sample of the available
+data and then fine-tune various models for the classification task.
+Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.
+
+To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'.
+In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy is one that favors high interest rates to limit the supply of money and thereby control inflation.
+The authors then go on to show that this measure correlates with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases as central bankers react by raising interest rates to bring inflation back to target.
 
@@
 Instead of computing a measure based on predicted labels, we can use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
-To this end, we have further pre-processed the data provided by \cite{shah2023trillion} and used their proposed model to compute layer-wise embeddings for all available sentences.
+To this end, we have further pre-processed the data provided by \citet{shah2023trillion} and used their proposed model to compute activations of the hidden state on the first entity token for each layer.
 We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.
-For each layer, we have then computed linear probes on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.
 
-To mitigate issues related to over-parameterization, we follow the recommendation in \cite{alain2018understanding} to first reduce the dimensionality of the embeddings each time.
-In particular, linear probes are restricted to the first 128 principal components of the embeddings
-of each layer.
+For each layer, we have then computed linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.
+To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
+To measure price inflation we use percentage changes proxied by log differences.
 
-Figure~\ref{fig-fomc} highlights the result for the linear probe
-on the CPI. The chart shows various performance measures plotted against
-\emph{FOMC-RoBERTa}'s \(n\)-th layer. Shaded areas show the variation
-across cross-validation folds, where we have used an expanding window
-approach to split the time series. To avoid look-ahead bias, we run PCA
-separately for each training set. Results for the linear probes are
-shown along with results for autoregressive models (AR) used as our
-baseline.
+To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations each time.
+In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
+We found that not doing so would generally lead to poor out-of-sample performance.
+To account for stochasticity we use an expanding window scheme with 5 folds for each indicator and layer.
+To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.
 
-The first column shows the correlations between model predictions and
-observed inflation. One can observe that this correlation is strictly
-positive for the baseline models and the linear probes. Consistent with
-the findings in \cite{gurnee2023language} and \cite{alain2018understanding},
-we also observe that this correlation tends to be higher for layers near
-the end of the transformer model. Those layers produce predictions that
-are more positively correlated with observed inflation than those
-produced by the corresponding autoregressive models.
+Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer.
+The values correspond to averages computed across cross-validation folds.
+Consistent with findings in related work \citep{alain2018understanding,gurnee2023language}, we also observe that model performance tends to be higher for layers near the end of the transformer model.
+Curiously, for yields at longer maturities, we find that perfomance eventually deteriorates for the very final layers.
+We do not observe this for the training data, so we attribute this to overfitting.
 
-Similarly, we observe that the (root) mean squared error of the linear
-probe is largely on par with the baseline. The average prediction error
-is gradually reduced as we move along the horizontal axis, which is
-again indicative of the notion that layers closer to the end of the
-neural network have distilled more useful representations.
+% After all we are still regressing a single variable of interest on 128 regressors.
+
+It should also be noted that performance improvements are generally of small magnitude.
+Still, the overall qualitative findings are in line with expectations.
+Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with outcome of interest and achieve higher mean directional accuracy (MDA).
+Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.
+
+To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models.
+For each layer, indicator and cross-validation fold we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
+These are not state-of-the-art forecasting models but they serve as a reasonable baseline.
+We find that for most indicators, probe predictions outperform the baseline in terms of average performance measures.
+After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform.
+Detailed results will be made available in a supplementary appendix.
 
@@
 
-\caption{\label{fig-fomc}Various performance measures (correlation and
-(root) mean squared error) plotted against \emph{FOMC-RoBERTa}'s
-\(n\)-th layer. Shaded areas indicate variation across cross-validation
-folds, where we have used an expanding window approach to split the time
-series. Results for the linear probes are shown along with results for
-autoregressive models (AR). The optimal lag length was determined using
-the Bayes Information Criterium.}
+\caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators.
+The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}
 
@@
 
-These results from the linear probe shown in Figure~\ref{fig-fomc} are
-certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not
-explicitly trained to uncover associations between central bank
-communications and the level of consumer prices, it appears that the
-model has distilled representations that can be used to predict
-inflation. It is worth pointing out here that this model is
-substantially smaller than the models tested in \cite{gurnee2023language}. This begs the following question:
+The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
+
+It is worth pointing out here that this model is substantially smaller than the models tested in \citet{gurnee2023language}. This may tempt us to ask the following questions:
 
@@
 Have we uncovered further evidence that LLMs ``aren't mere stochastic
-parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic understanding
+parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
 of the economy just by `reading' central bank communications?
@@
 
-We are having a very hard time believing this.
+We are having a very hard time believing that the answer to either of these questions is `yes'.
+
 To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor.
-The counter example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic understanding of the economy, then the probe should not be sensitive to random sentences  that are most definitely not related to consumer prices.
+The counter-example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic `understanding' of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.
 
-To test this, we select the best-performing probe trained on the final-layer activations to predict changes in the CPI.
+To test this, we select the best-performing probe trained on the final-layer activations for each indicator.
 We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about an inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation in the number of birds.
@@
 Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''.
+While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.
 
-In light of the encouraging results for the probe in
-Figure~\ref{fig-fomc}, we should expect the probe to predict higher
-levels of inflation for activations for sentences in the IP category
-than for sentences in the DP category. If this was indicative of true
-intrinsic understanding, we would not expect to see any significant
-difference in predicted inflation levels for sentences about birds,
-independent of whether or not their numbers are increasing. More
-specifically, we would not expect the probe to predict values for
-sentences about birds that are substantially different from the values
-it can be expected to predict when using actual white noise as inputs.
+% We top this up with five sentences generated through ChatGPT, where we choose those that lead to the highest probe prediction in the direction of interest.
+In other words, we conduct a small manual adversarial attack on the probe.
 
-To get to this last point, we also generate many probe predictions for
-samples of noise. Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote
-the linear probe that maps from the \(k\)-dimensional space spanned by
-\(k\) first principal components of the final-layer activations to the
-output variable of interest (CPI growth in this case). Then we sample
-\(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\)
-for \(i \in [1,1000]\) and compute the sample average. We repeat this
-process \(10000\) times and compute the median-of-means to get an
-estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\),
-that is the predicted value of the probe conditional on white noise.
+In light of the encouraging results for the probe in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category than for sentences in the DP category.
+If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
+More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.
 
-Next, we propose the following hypothesis test as a minimum viable
-testing framework to assess if the probe results (may) provide evidence
-for an actual understanding of key economic relationships learned purely
-from text:
+To get to this last point, we also generate many probe predictions for samples of noise.
+Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote the linear probe that maps from the \(k\)-dimensional space spanned by \(k\) first principal components of the final-layer activations to the output variable of interest (CPI growth in this case).
+Then we sample \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\) for \(i \in [1,1000]\) and compute the sample average.
+We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on white noise.
+
+Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
 
@@
 \item
-  \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are
-  statistically significantly different from
-  \(\mathbb{E}[f(\varepsilon)]\)
+  \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
 \item
-  \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts
-  values that are statistically significantly different from
-  \(\mathbb{E}[f(\varepsilon)]\) for sentences in all categories
-  (IP,DP,IB,DB).
+  \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
+sentences in all categories).
 \item
-  \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The
-  probe predicts values that are statistically significantly different
-  from \(\mathbb{E}[f(\varepsilon)]\) for sentences in categories IP and
-  DP, but not for sentences in IB and DB.
+  \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences in that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
 \end{itemize}
-
 \end{proposition}
 
-To be clear, if in such a test we did find substantial evidence in
-favour of rejecting both \emph{HO} and \emph{H1}, this would not
-automatically imply that \emph{H2} is true. But to even continue
-investigating if based on having learned meaningful representation the
-underlying LLM is more than just a parrot, it should be able to pass
-this simple test.
+To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
+But to even continue investigating if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
 
-In this particular case, Figure~\ref{fig-attack} demonstrates that we
-find some evidence to reject \emph{H0} but not \emph{H1} for
-\emph{FOMC-RoBERTa}. The median linear probe predictions for sentences
-about inflation and deflation are indeed substantially higher and lower,
-respectively than for random noise. Unfortunately, the same is true for
-sentences about the inflation and deflation in the number of birds,
-albeit to a somewhat lower degree.
+In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
+The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
+Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
+This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
 
-We should note that the number of sentences in each category is very
-small here (10), so the results in Figure~\ref{fig-attack} cannot be
-used to establish statistical significance. That being said, even a
-handful of convincing counter-examples should be enough for us to
-seriously question the claim that results from linear probes provide
-evidence against the parrot metaphor. In fact, if there's even a single
-sentence for which any
-
-\begin{quote}
-Strong alternative hypothesis to address own confirmation bias
-Articulating a test to better differentiate
-\end{quote}
+We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
+That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim that results from linear probes provide evidence in favor of real `understanding'.
+In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
 

### 2024-01-31 main.tex
updates from OL

@@
 %%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
+\RequirePackage[hyphens]{url}
 \documentclass{article}
@@
 \twocolumn[
-\icmltitle{Spurious Sentience:\\
-           Dovelating Inflated AI World Model Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+\icmltitle{Against Spurious Understanding: 
+           Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
 
@@
 \begin{abstract}
-This document provides a basic paper template and submission guidelines.
-Abstracts must be a single paragraph, ideally between 4--6 sentences long.
-Gross violations will trigger corrections at the camera-ready phase.
+Humans have a tendency to place a `human' like quality in the objects around them.
+Giving your car a name or talking to your pet, such that you believe they understand you in a human way.
+This behavior is also seeing traction in Machine Learning (ML), where the idea of sentience is being placed upon Large Language Models (LLMs).
+In this position paper we discuss the history of and why humans invariably anthropomorphize onto various objects.
+How the search for Artificial General Intelligence (AGI) is pushing recent literature to over emphasis sentience.
+To combat this, we provide a limited study, that investigates the latent space of models to find spurious correlations.
+Finally we discuss how the popularity in media and the race for AGI enforces over-interpretation of consciousness in the models we use.
+
 \end{abstract}
@@
 
-We humans are prone to seek patterns everywhere.
-Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.
-the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
+% \textcolor{magenta}{TODOs for the team:
+% \begin{itemize}
+%     \item Please double check the citations.
+Note this style file does not retain capitalization in titles.
+A trick to retain capitalization is to put an extra set of brackets around the title.
+I did this for the Ladouceur et al.
+and Bubeck et al references already; please make this consistent throughout.
+%     \item As another issue with the citations: note that arxiv preprints do not show very well.
+Best way to make them show is to wrap them as 'journal articles' as done in the Arora citation.
+Please straighten this out for the other articles too
+% \end{itemize}
+% }
 
-In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
-Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
-Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-A major contributor is that humans struggle to conceive of actual randomness: e.g.
-studies show that humans are poor at detecting e.g.
-\cite{falk1997making}, and creating e.g.
-\cite{ladouceur1996erroneous} random sequences.
-A common issue is a lack of expectation that the patterns that hint towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \cite{nickerson1998confirmation}), or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
+In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\cite{merton1942science}.
+While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.
 
-We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
-Such that, when an AI system begins finding patterns in the corpus not originally perceived during training, that is now from a human perspective, observed as greater cognitive capabilities.
-This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
-%Such as an AI system finding patterns in data not originally perceived during training
-%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
-In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g.
+At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening.
+Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\cite{weissburg2024tweets}.
+Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible only is increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
+
+Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence.
+In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal.
+The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}.
+In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or showing performance considered `surprising' to humans.
+
+For example, Google DeepMind claimed their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
+This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
+However, as noted by Hector Zenil \cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s.
+Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
+Therefore, despite DeepMind's success in creating a powerfully fast geometry solving machine, it is still far from being AGI.
+
+Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}.
+Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.
+
+When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
+AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}.
+As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
+In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.
+
+Noticing these trends, we as the authors of this article are concerned.
+We feel that the current culture of racing towards Big Outcome Statements in industry and academic publishing too much disincentivizes efforts towards more thorough and nuanced actual problem understanding.
+At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there indeed is sufficient evidence for these claims.
+With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
+In other words, it is both very tempting to claim surprising human-like achievements of AI, and as humans we are very prone to genuinely believe this.
+
+To strenghten our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
+We work out different experiments which may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made.
+Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
+Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.
+
+%Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
+As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
+With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
+
+%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}.
+This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com/wesg52/status/1709551516577902782?s=20} (formally Twitter) platform.
+Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
+
+%This only gets amplified as the interest in Artificial General Intelligence has now merged with a push towards building general-purpose tools.
+
+%\textcolor{magenta}{cynthia: I am now moving the texts of Andrew and Antony on AGI here; maybe some tightening can already be done in this.
+Unfortunately no time for thorough editing yet at this stage, but will come back to this later}
+
+%In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g.
 \cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'.
-When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that also aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience \footnote{https://www-scientificamerican-com.tudelft.idm.oclc.org/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/} and consciousness.
 
-Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs.
+%Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs.
 We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, within AI systems.
@@
 
-\textcolor{red}{andrew: I'm thinking we can adjust this all to aim towards more 'severe' tests or something. }
-
-%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
-Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads.
-Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.
-
-%I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy.
-But I am very skeptical of claims that AI will magically provide that remedy.
-This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them.
-The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots.
-Since this is an opinionated post, I feel that I should start with a few disclaimers:
-
-%\begin{itemize}
-%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
-%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
-%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
-%    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
-Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
-%\end{itemize}
-
-\section{Background}
-
-\subsection{Claims of Artificial General Intelligence}
-
-\textcolor{red}{antony: I will be adding citations and other links later, just getting text down for now.}
-As previously mentioned, AGI is the current objective for which the global AI community is shooting for.
+%As previously mentioned, AGI is the current objective for which the global AI community is shooting for.
 As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts.
@@
 
-Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
-This model has the ability to solve complex geometry problems, which, they state it does without the need for human demonstrations during training.
+% Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
+This model has the ability to solve complex geometry problems, which they state it does without the need for human demonstrations during training.
 Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
@@
 
-However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s.
+% However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s.
 Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry.
@@
 
-One of the more contentious pieces of literature recently studied is the paper by \cite{gurnee2023language}.
-This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com} (formally Twitter) platform.
+%it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g.
+\cite{salles2020anthropomorphism, van2023reclaiming}.
+
+\section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}
+\textcolor{magenta}{cynthia: @Patrick this section needs a little bit of tightening too.
+I have tried doing this, please do double-check that I didn't over-shorten things.
+In any case, I will always comment out a copy of the original text close to what I now put}
+
+In Winter 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform~\cite{gurnee2023languagev1}.
+In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that only can correlate and mimic existing patterns from data, but not understand this data.
+
+While the manuscript immediately received public criticism~\cite{marcus2023muddles}, and the subsequent, current manuscript version  is more careful with regard to its claims~\cite{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
+However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
+After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on economic growth prediction.
+
+%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}.
+This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform.
 Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
 
-They claim said models generate an internal representation of the world, with which spatiotemporal representations are created.
+%They claim said models generate an internal representation of the world, with which spatiotemporal representations are created.
 Further to this we see claim to the generalization of tasks, in which the model can answer questions regarding spatial and temporal relationships, without the need for further training.
@@
 
-As we also consider the paper by the Microsoft Research team \cite{}, despite the illusion of a claim that GPT-4 shows sparks of AGI, they do not directly claim that GPT-4 has AGI.
-However they do claim that we are moving closer to this goal, which is not an outlandish claim given the advancements we have seen with AI in recent years.
-
-\subsection{Confirmation Bias}
-
-Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \cite{nickerson1998confirmation}.
-Importantly, it need not be fully conscious.
-Rather, theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
-Equally importantly, is that theory suggests these behaviors may be motivated or unmotivated: one may selectively seek, attend to, or interpret evidence in favor of a hypothesis for which they may or may not have a personal interest in supporting.
-
-Hypotheses in computer science fields related to AI are often implicit.
-Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
-However with regards to claims that outputs of a system hint towards emerging qualities like AGI, hypotheses necessarily become more explicit: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-
-Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews.
-Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
-only considering the likelihood that outputs of system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a 'weak' or 'non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
-In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
-If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
-Thus, one is far more likely to show evidence in favor of their beliefs by posing a weak null hypotheses.
-Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
-As \cite{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
-
-\subsection{Anthropomorphism}
-
-Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}.
-Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
-Anthropomorphism was implied because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
-Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end may attribute human emotions, cognition, and intelligence to AI systems.
-As \cite{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
-
-Modern theory of anthropomorphism suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: "...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}.
-Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
-This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}.
-An illustration is the example of Tom Hank's character in the film Castaway who has conversations with a ball he calls "Wilson".
-Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc.
-with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.
-
-The motivation as a human to be competent, so-called effectance motivation, is the most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}.
-When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}.
-Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
-an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agents behaviors.
-This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
-One such situation is the increasing demand for 'AI' skills among AI non-experts: e.g.
-job postings for individuals with skills in 'GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}.
-Thus, individuals in advanced economies may experience a pressure to rapidly learn how to effectively use and understand an opaque technology that appears to behave human-like in at least some ways, and by imbuing the technology with humanity may in turn ease their learning curve.
-
-These pressures extend to experts as well, in both scholarly and commercial fields.
-Academics face considerable pressures to publish work in the field of AI: citation metrics and scholarly publications remain the primary metric for tenure and promotion e.g.
-\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced in overall scholarly publications \footnote{ Retrieved 23/01/23 from https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021 }, and in peer-reviewed publications \cite{Maslej2023-pi}, with both more than doubling from 2010 to 2020.
-The adoption of techniques underlying technologies with the AI label, i.e.
-machine learning, has spread to fields far beyond Computer Science, e.g.
-Astronomy and Physics to Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  >  2009  AND  PUBYEAR  <  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise have increased several fold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
-Given the strong incentives to be AI-competent, it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g.
-\cite{salles2020anthropomorphism, van2023reclaiming}.
-
-\section{Patterns in Latent Spaces and How to Find
-Them}\label{patterns-in-latent-spaces-and-how-to-find-them}
-
-To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples.
-To illustrate further that this phenomenon is neither surprising nor unique to the field of computer science, we will borrow draw on ideas from the fields of Economics and Finance in this section.
+%In this section, we navigate a number of simple examples to support our claim, that observing patterns in latent spaces should not be a surprising revelation.
+%To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples. 
+%We will draw on ideas from the fields of Economics and Finance, to illustrate further that this phenomenon is neither surprising nor unique to the field of computer science.
 We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data.
-The final example in Section \ref{ex-llm} is a bit more involved and closer in spirit to the experiments conducted by \citet{gurnee2023language}.
-As we go along, we will try to discuss both the benefits and potential pitfalls of finding patterns in latent spaces.
-
-\subsection{Example: Random Projections of World Models}\label{example-deep-learning}
-
-The Llama-2 model has ingested huge amounts of publicly available data from the internet including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
-It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
-At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
-The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance, before being passed through a series of layers which are gradually trained for downstream tasks.
-
-In our first example in this section, we simulate this scenario, stopping short of training the model. In particular, we
-
-\begin{itemize}
-\tightlist
-\item
-  Take the
-  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
-  that was used in \cite{gurnee2023language}, which maps locations/areas
-  to their latitude and longitude. For each place, it also indicates the
-  corresponding country.
-\item
-  Take the subset that contains countries that are currently part of the
-  top 10
-  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
-  world ranking}.
-\item
-  Assign the current rank to each country, i.e.~Argentina gets 1, France
-  gets 2, \ldots{}
-\item
-  Transform the longitude and latitude data, respectively, as follows:
-  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
-  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\). This is to
-  ensure that the training data only involves a noisy version of the
-  coordinates.
-\item
-  Next, encode all features except the FIFA world rank indicator as
-  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
-  of samples and \(m\) is the number of resulting features.
-\item
-  Additionally, add a large number of random features to \(X\) to
-  simulate the fact that not all features ingested by Llama-2 are
-  necessarily correlated with geographical coordinates. Let \(d\) denote
-  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
-  of random features.
-\item
-  Initialize a small neural network, let's call it a \emph{projector},
-  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
-  units and sigmoid activation and then from there to a
-  lower-dimensional output space.
-\item
-  Without performing any training on the \emph{projector}, we simply
-  compute a forward pass of \(X\) through the \emph{projector} and
-  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
-\item
-  Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge
-  regression:
-  \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\)
-  where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the
-  longitude and latitude for each sample. A hold-out set is reserved for testing.
-\item
-  Finally, we compute the predicted coordinates for each sample in the hold-out set as
-  \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on
-  a world map (see Figure~\ref{fig-map}).
-\item 
-  While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand.
-\end{itemize}
+The final example in Section \ref{ex-llm} is more complex and closer in spirit to the experiments conducted by \citet{gurnee2023language}.
+As we progress, we will discuss both the benefits and potential pitfalls of finding patterns in latent spaces.
 
@@
 
-While this is admittedly a somewhat contrived example, it serves to illustrate the point that if we randomly project a small number of features that are highly correlated or predictive of the target into a large enough space, we should expect to be able to recover meaningful representations of the target from the latent space.
+\subsection{Do neural networks get born with world models?}\label{example-deep-learning}
+%\subsection{Example: Random Projections of World Models}\label{example-deep-learning}
+\textcolor{magenta}{Patrick feel free to revert, I kept all original titles}
+\cite{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places.
+Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
+It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
+At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
+The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance, before being passed through a series of layers which are gradually trained for downstream tasks.
+
+In our first example in this section, we simulate this scenario, stopping short of training the model. In particular, we take the
+  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
+  that was used in \cite{gurnee2023languagev2}, which maps locations/areas
+  to their latitude and longitude. For each place, it also indicates the
+  corresponding country. From this, we take the subset that contains countries that are currently part of the
+  top 10
+  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
+  world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France
+  gets 2, \ldots{}).
+
+  Subsequently,  to
+  ensure that the training data only involves a noisy version of the
+  coordinates, we transform the longitude and latitude data, respectively, as follows:
+  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
+  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
+
+  Next, we encode all features except the FIFA world rank indicator as
+  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
+  of samples and \(m\) is the number of resulting features. 
+  Additionally, we add a large number of random features to \(X\) to
+  simulate the fact that not all features ingested by Llama-2 are
+  necessarily correlated with geographical coordinates. Let \(d\) denote
+  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
+  of random features.
+
+  We then initialize a small neural network, which we consider as a \emph{projector},
+  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
+  units and sigmoid activation, and from there to a
+  lower-dimensional output space. Without performing any training on the \emph{projector}, we simply
+  compute a forward pass of \(X\) through the \emph{projector} and
+  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
+
+  Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge
+  regression:
+  \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\)
+  where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the
+  longitude and latitude for each sample. A hold-out set is reserved for testing. Finally, we compute the predicted coordinates for each sample in the hold-out set as
+  \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on
+  a world map (see Figure~\ref{fig-map}).  
+  While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, with this example, we illustrate
+  %While admittedly a somewhat contrived example, it serves to illustrate the point:
+  that meaningful target representations should be recoverable from the latent space, given the projection of a small number of highly correlated features.
+%that if we randomly project a small number of features that are highly correlated or predictive of the target into a large enough space, we should expect to be able to recover meaningful representations of the target from the latent space.
 Similarly, \cite{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
@@
 
-\subsection{Example: Principal Component
-Analysis}\label{example-principal-component-analysis}
-
-For our next example, we will move beyond randomness and illustrate how principal component analysis (PCA) can be used to recover meaningful and compact latent representations from data in an unsupervised manner.
-
-The yield curve is a popular tool for investors and economists to gauge the health of the economy.
-It plots the yields of bonds against their maturities.
-The slope of the yield curve is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
-While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone.
-To leverage this information in downstream modelling tasks, economists regularly make use of PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data.
-Recent examples of such include \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing} among others.
-
-Here we are primarily interested in understanding the nature this low-dimensional projection, so will limit our attention to the US Treasury yield curve.
-To this end, we have collected daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
-Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
-To put this into the broader context of this, let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.
-
-The top panel in Figure~\ref{fig-pca} shows the first two principal
-components of the yield curves of US Treasury bonds over time. Vertical
-stalks indicate key dates during the onset and aftermath of the
-Global Financial Crisis.
-During the onset, on 27 February 2007, financial markets were in turmoil following Federal Reserve Chairman Alan Greenspan's warning on the previous day that the US economy was at risk of a recession.
-Over the following months the Fed reacted to these economic and financial pressures by gradually cutting short-term interest rates to unprecedented lows, which affected the yield curve in two major ways: the average level of yields decreased and the curve steepened as the short end came down.
-Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve.
-
-Of course, even though it is reassuring to see that these latent embeddings are meaningful, none of this is surprising at all: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance in the data after controlling for the effect of all previous components.
-
-\begin{figure*}
+\begin{figure*}[htp]
 
@@
 
-\end{figure*}%
+\end{figure*}
 
-\subsection{Example: Autoencoder}\label{example-deep-learning}
+\subsection{PCA as yield curve interpreter}\label{example-principal-component-analysis}
+%\subsection{Example: Principal Component Analysis}\label{example-principal-component-analysis}
 
-So far we have considered simple linear transformations of data.
+%For our next example, we will move beyond randomness and illustrate how principal component analysis (PCA) can be used to recover meaningful and compact latent representations from data in an unsupervised manner.
+
+We now move to a concrete application domain: Finance.
+Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
+%It plots the yields of bonds against their maturities.
+The yield curve's slope is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
+%While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone.
+To leverage this information in downstream modelling tasks, economists regularly use PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data (e.g.\ \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing}).
+
+%Being primarily interested in understanding the nature of this low-dimensional projection, we limit our scope to the US Treasury yield curve. To this end, we
+To understand the nature of this low-dimensional projection, we collect daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
+Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
+Let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.
+
+The top panel in Figure~\ref{fig-pca} shows the first two principal components of the yield curves of US Treasury bonds over time.
+Vertical stalks indicate key dates during the onset and aftermath of the Global Financial Crisis.
+During the onset, on 27 February 2007, financial markets were in turmoil following Federal Reserve Chairman Alan Greenspan's warning on the previous day that the US economy was at risk of a recession.
+Over the following months, the Fed reacted to these economic and financial pressures by gradually cutting short-term interest rates to unprecedented lows, which affected the yield curve in two major ways: the average level of yields decreased and the curve steepened as the short end came down.
+Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
+In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve.
+
+%Of course, even though it is reassuring to see that these latent embeddings are meaningful, none of this is surprising at all: 
+Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
+
+% \begin{figure*}
+
+% \centering{
+
+% \includegraphics[width=1.0\textwidth]{results/figures/dl.png}[tb]
+
+% }
+
+% \caption{\label{fig-dl-results}The left chart shows the actual GDP growth
+% and fitted values from the autoencoder model. The right chart shows the
+% observed average level and spread of the yield curve (solid) along with
+% the predicted values (in-sample) from the linear probe based on the latent embeddings
+% (dashed).}
+
+% \end{figure*}%
+% %\subsection{Example: Autoencoder}\label{example-deep-learning}
+% \subsection{Autoencoders as economic growth predictors}\label{example-deep-learning}
+
+% %So far we have considered simple linear transformations of data.
 One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
@@
 
-Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks at the same time.
-In particular, we will use a neural network architecture that allows us to recover a compact latent representation of the yield curve.
+% Our goal in this section is to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
+In particular, we will now use a neural network architecture.
+% %, that allows us to recover a compact latent representation of the yield curve.
 
-\subsubsection{Data}\label{data}
+% \subsubsection{Data}\label{data}
 
-To estimate economic growth we will rely on a quarterly
-\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real
-gross domestic product (GDP) provided by the Federal Reserve Bank of
-St.~Louis. The data arrives in terms of levels of real GDP. In order to estimate growth, we transform the data using log differences. Since
-our yield curve data is daily, we aggregate it to the
-quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data (REFERENCE NEEDED?).
+% To estimate economic growth, we will rely on a quarterly
+% \href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
+The data arrives in terms of levels of real GDP.
+In order to estimate growth, we transform the data using log differences.
+Since
+% our yield curve data is daily, we aggregate it to the
+% quarterly frequency by taking averages of daily yields for each maturity.
+We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
 Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
 
-\subsubsection{Model}\label{model}
+% \subsubsection{Model}\label{model}
 
-Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
+% %Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
 Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
@@
 
-To do this, we use a simple autoencoder architecture: the encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer, connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
-The decoder consists of two fully connected layers each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
+% %To do this, we use a simple autoencoder architecture. With this
+% Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
+Finally, we let \(\theta\) denote our model parameters.
+Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
+
+% %To do this, we use a simple autoencoder architecture: 
+% The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
+The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
+The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
 The output layer consists of a single neuron for our output variable, \(g_t\).
-We train the model over 1,000 epochs to minimize mean squared error loss using the the Adam optimizer \citep{kingma2017adam}.
+We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
 
-The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-model.
+% The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
+% model.
 The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
@@
 
-\subsubsection{Linear Probe}\label{linear-probe}
+% \subsubsection{Linear Probe}\label{linear-probe}
 
-While the loss function applies most direct pressure on layers near the final output layer of the neural network, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
-On a per-neuron basis, the pressure to distill useful representation is therefore likely maximised in this layer.
+% While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
+On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
 Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
-
-We compute and extract these activations \(A_t\) for all time-periods \(t=1,...,T\). Next, we use a linear probe to regress the observed
-yield curve factors on the latent embeddings.
+We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
+Next, we use a linear probe to regress the observed
+% yield curve factors on the latent embeddings.
 Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
 Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-parameters.
-We use Ridge regression with \(\lambda\) set to \(0.1\).
-Using the estimated regression parameters \(\hat{w}\) we can then predict the yield curve factors from the latent embeddings: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
+% parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
+% : \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
 
-The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
+% The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
 Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
 We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-As in all of the other examples we have presented so, we do not find this result particularly surprising.
-It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
+%As in all of the other examples we have presented so, we do not find this result particularly surprising.
+
+% Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
+To us, that would be too big of statement.
+Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
+% %It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
 But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
 While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
-
-}
-
-\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-and fitted values from the autoencoder model. The right chart shows the
-observed average level and spread of the yield curve (solid) along with
-the predicted values (in-sample) from the linear probe based on the latent embeddings
-(dashed).}
-
-\end{figure*}%
-
-\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
-
-An interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
+% %\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
+% For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
 To demonstrate this, let us consider a simple linear regression model for GDP growth.
@@
 As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-But this coefficient suffers from confounding bias since, naturally, there many other confounding variables at play here, some of which may be readily observable and measurable, others of which may not be.
+But this coefficient suffers from confounding bias, since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
 
-% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
+% % A less risky/strong term than confounding bias that we could use here is omitted variable bias.
 Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
 
-We have, for example, already mentioned above that there is relationship between interest rates and economic growth.
-To account for that while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
-While this does slightly change the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant as evident from column (2) in Table \ref{tab-reg}.
+% We
+% %have, for example, 
+% e.g.\ already mentioned the relationship between interest rates and economic growth.
+To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
+While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
 This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
-Because we have included to additional regressors with little to know predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
+Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
 
-Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
-In particular, we have chosen the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
-The estimated coefficient on this latent factor is small in magnitude but statistically significant.
-The overall model fit as measured by the BIC as improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
-While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can actually be used in order to mitigate confounding bias.
+% Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
+In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
+The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
+The overall model fit, as measured by the BIC as improved and the magnitude of the coefficient on lagged growth, has changed quite a bit.
+While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can  be used in order to mitigate confounding bias.
 
-\begin{table}
-\caption{Regression output for various models.}\label{tab-reg}%
-\input{results/regression.tex}
-\end{table}
+% \begin{table}
+% \caption{Regression output for various models.}\label{tab-reg}%
+% \input{results/regression.tex}
+% \end{table}
 
-\subsection{Example: Large Language Model}\label{ex-llm}
+%\subsection{Example: Large Language Model}\label{ex-llm}
+\subsection{LLMs for economic sentiment prediction}\label{ex-llm}
 
-To round up this section, we return to the world of `GenAI' and consider an example involving an LLM.
-In particular, we closely follow the approach in \citet{gurnee2023language} and apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset introduced by \citet{shah2023trillion}.
-The dataset contains a curated selection of sentences formulated by central bankers of the US Federal Reserve and communicated to the public in speeches, meeting minutes and press conferences.
+%To round up this section, we return to the world of `GenAI' and consider an example involving an LLM.
+We now return to LLMs, in the context of predicting on an unseen task. Here, we closely follow the approach
+%In particular, we closely follow the approach in 
+in \citet{gurnee2023languagev2} but apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset~\citet{shah2023trillion}.
+This dataset contains a curated selection of sentences formulated by central bankers of the US Federal Reserve and communicated to the public in speeches, meeting minutes and press conferences.
 The authors of the paper use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
@@
 \subsubsection{Linear Probes}\label{linear-probes}
-
-Instead of computing a measure based on predicted labels, we can use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
-To this end, we have further pre-processed the data provided by \citet{shah2023trillion} and used their proposed model to compute activations of the hidden state on the first entity token for each layer.
+%Instead of computing a measure based on predicted labels,
+We now use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
+Therefore, we further pre-process the data provided by \citet{shah2023trillion} and use their proposed model to compute activations of the hidden state, on the first entity token for each layer.
 We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.
 
-For each layer, we have then computed linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.
+For each layer, we compute linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.
 To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
-To measure price inflation we use percentage changes proxied by log differences.
+To measure price inflatio,n we use percentage changes proxied by log differences.
 
@@
 In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
-We found that not doing so would generally lead to poor out-of-sample performance.
-To account for stochasticity we use an expanding window scheme with 5 folds for each indicator and layer.
+%We found that not doing so would generally lead to poor out-of-sample performance.
+To account for stochasticity, we use an expanding window scheme with 5 folds for each indicator and layer.
 To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.
@@
 The values correspond to averages computed across cross-validation folds.
-Consistent with findings in related work \citep{alain2018understanding,gurnee2023language}, we also observe that model performance tends to be higher for layers near the end of the transformer model.
+Consistent with findings in related work \citep{alain2018understanding,gurnee2023languagev2}, we also observe that model performance tends to be higher for layers near the end of the transformer model.
 Curiously, for yields at longer maturities, we find that perfomance eventually deteriorates for the very final layers.
@@
 For each layer, indicator and cross-validation fold we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
-These are not state-of-the-art forecasting models but they serve as a reasonable baseline.
+These are not state-of-the-art forecasting models, but they serve as a reasonable baseline.
 We find that for most indicators, probe predictions outperform the baseline in terms of average performance measures.
@@
 
-\subsubsection{Stochastic Parrots After
-All?}\label{stochastic-parrots-after-all}
+%\subsubsection{Stochastic Parrots AfterAll?}\label{stochastic-parrots-after-all}
+\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
+%The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive:
+Even though \emph{FOMC-RoBERTa} (which is substantially smaller than the models tested in~\citet{gurnee2023languagev2} was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
+So, have we uncovered further evidence that LLMs ``aren't mere stochastic
+parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
+of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
 
-The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
+%It is worth pointing out here that this model is substantially smaller than the models tested in \citet{gurnee2023language}. This may tempt us to ask the following questions:
 
-It is worth pointing out here that this model is substantially smaller than the models tested in \citet{gurnee2023language}. This may tempt us to ask the following questions:
-
-\begin{quote}
-Have we uncovered further evidence that LLMs ``aren't mere stochastic
-parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-of the economy just by `reading' central bank communications?
-\end{quote}
+% \begin{quote}
+% Have we uncovered further evidence that LLMs ``aren't mere stochastic
+% parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
+% of the economy just by `reading' central bank communications?
+% \end{quote}
 
 We are having a very hard time believing that the answer to either of these questions is `yes'.
-
 To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor.
@@
 
-In light of the encouraging results for the probe in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category than for sentences in the DP category.
+In light of the encouraging results for the probe in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
 If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.
+More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict, when using actual white noise as inputs.
 
@@
 
-Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
+% Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
 
-\begin{proposition}[Parrot
-Test]\protect\hypertarget{prp-line}{}\label{prp-line}
+% \begin{proposition}[Parrot
+% Test]\protect\hypertarget{prp-line}{}\label{prp-line}
 
-~
+% ~
 
-\begin{itemize}
-\tightlist
-\item
-  \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
-\item
-  \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
+% \begin{itemize}
+% \setlength\itemsep{1px}
+% \item
+%   \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
+% \item
+%   \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
 sentences in all categories).
-\item
-  \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences in that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
-\end{itemize}
-\end{proposition}
+% \item
+%   \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences in that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
+% \end{itemize}
+% \end{proposition}
 
-To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
-But to even continue investigating if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
+% To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
+But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
 
-In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
+% In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
 The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
@@
 
-We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
-That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim that results from linear probes provide evidence in favor of real `understanding'.
+% We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
+That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
 In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
 
-\subsection{Discussion of Examples}
+\section{On human proneness to over-interpretation}
+Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
+Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
+Our examples demonstrate, that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world.
+Thus, to draw conclusions about the emerging capabilities of AI models, we argue that more conservative and rigorous tests are needed.
 
-Our examples should not be seen as an attempt to conclusively refute claims that LLMs can develop on actual "understanding" of the world by distilling knowledge in more compact representations.
-We hope that they serve to illustrate, however, that the use of diagnostic tools is potentially misleading and most definitely insufficient in providing evidence in favor of "understanding".
-Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favourable outcomes from probes merely indicate that the model "has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate that this is achievable even for small models, that most readers will agree have certainly not developed an intrinsic "understanding" of the world.
-The fact that even these models distill meaningful representations goes to show that more conservative and rigorous tests are needed to draw conclusions about the emerging capabilities of AI models.
+%that most readers will agree, have certainly not developed an intrinsic "understanding" of the world.
+The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.
+
+%Our examples should not be seen as an attempt to conclusively refute claims, that LLMs can develop on actual "understanding" of the world, by distilling knowledge in more compact representations.
+We hope that they serve to illustrate, however, that the use of diagnostic tools is potentially misleading and most definitely insufficient, in providing evidence in favor of "understanding".
+Linear probes and related tools from mechanistic interpretability, were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
+Favourable outcomes from probes merely indicate, that the model "has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
+Our examples demonstrate, that this is achievable even for small models, that most readers will agree, have certainly not developed an intrinsic "understanding" of the world.
+The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.
+
+Generally, humans are prone to seek patterns everywhere.
+Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
+Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.
+the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
+
+We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
+When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, from a human perspective, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
+We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
+%. This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
+
+\subsection{Spurious relationships}
+In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
+Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
+Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
+A major contributor is that humans struggle to tell the difference between random and non-random sequences \cite{falk1997making}, and to generate sequences that appear random \cite{ladouceur1996erroneous}.
+A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random.
+This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
+
+%Such as an AI system finding patterns in data not originally perceived during training
+%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
+
+%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
+Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads.
+Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.
+
+%I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy.
+But I am very skeptical of claims that AI will magically provide that remedy.
+This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them.
+The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots.
+Since this is an opinionated post, I feel that I should start with a few disclaimers:
+
+%\begin{itemize}
+%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
+%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
+%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
+%    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
+Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
+%\end{itemize}
+
+%\section{Background}
+
+%\subsection{Emergence and Stochastic Parrots}
+%Practically speaking, large language models (LLMs) are probabilistic representations of data on which they are trained, aimed at continuing an incomplete block of text.
+It is assumed that e.g.
+words, sentences, paragraphs, punctuation etc.
+appear in human-generated text according to a probability distribution \cite{arora2023theory}.
+Given a set of words, the task of the LLM is to draw upon its representation of human-generated text to continue that set of words: in essence, using the probability of the occurrences of words and an initial block of text to generate more text.
+AI-researchers have compared LLMs by observing whether the text they generate appears to solve tasks\footnote{e.g.
+the \href{https://huggingface.co/spaces/HuggingFaceH4/open_llm_leaderboard}{HuggingFace} LLM leaderboard}.
+Tasks include asking sets of questions across categories e.g.
+health, law, finance, and fine-tuning an LLM to check the answers \cite{lin2021truthfulqa}, asking the LLM to determine which noun a pronoun refers to in sentences where this is ambiguous \cite{sakaguchi2021winogrande}.
+
+%It has been shown that increasing the scale of large language models (LLMs) has been shown to result in an unexpectedly large boost in their scores on these various tasks, including scoring well on entirely new tasks.
+It has been shown that at a certain size, LLMs appear to hit a threshold, after which performance on a broad range of tasks increases sharply \cite{wei2022emergent}.
+Some theories of emergence, inspired by literature on human education, refer to the tendency of the generated text to be the correct response to a given task as a 'skill' \cite{arora2023theory}, inspiring further research to develop more challenging tasks for LLMs to complete as they routinely begin to score higher and higher on common benchmarks \cite{yu2023skill}.
+An explanation offered for the sudden boost in both task performance and the number of new tasks that appear to be solved is 'emergence': unpredictable qualitative differences in behavior of a system that occur with quantitative changes in that system.
+It has been claimed that, as the scale of the LLMs has dramatically increased, they have crossed a threshold that has resulted in a large increase in their abilities, to the point that they demonstrate 'sparks of AGI' \cite{bubeck2023sparks}.
+
+%An alternative explanation is that the LLMs are performing as designed, as stochastic parrots, generating strings probabilistically based on the vast text they have been trained on \cite{bender2021dangers}.
+Although the various tasks used to compare the various LLMs appear to be increasing in complexity, \cite{yu2023skill} acknowledges that higher scores on the tasks can be achieved by training LLMs on the material on which they will be tested ('cramming'), and that there is training-set contamination, i.e.
+that the vast text they are trained on contains similar examples to the tasks on which they are measured.
+Thus, as many of these tasks are derived from tests and textbooks commonly assigned to humans, and as much of the contents of the their training data text is unknown, metrics echo similar issues in human education: often called 'Goodhart's Law', a measure ceases to be a measure when it becomes a target, as is the case when human students taught to score high on their standardized tests to the detriment of developing the abilities the tests are supposed to measure \cite{strathern1997improving}.
+Further, as \citet{bender2021dangers} reviews, the interpretation of text as coherent has no objective test, relying instead on our human interpretation of the probabalistically generated output.
+%Although many deep learning systems' designs were inspired by the anatomy of human neurons, the neuron in computer code does not in fact resemble a human, and as such cannot deliver the cognitive functions of a human \cite{salles2020anthropomorphism, van2023reclaiming}.
+
+%\subsection{Claims of Artificial General Intelligence}
+\subsection{Anthropomorphism}
+
+Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
+These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}.
+Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
+%Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
+Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
+As \cite{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
+
+Modern theory of anthropomorphism suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
+The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}.
+Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
+This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}.
+%An illustration is the example of Tom Hank's character in the film Castaway who has conversations with a ball he calls "Wilson".
+Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc.
+with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.
+
+The motivation as a human to be competent (effectance motivation), is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}.
+When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}.
+Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
+an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors.
+This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
+
+These pressures extend to expert AI practitioners and researchers as well as laypersons.
+In both scholarly and commercial fields, AI experts face considerable pressures to demonstrate competence in AI-related work.
+Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
+\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall scholarly publications \footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }.
+This is also evident in peer-reviewed publications \cite{Maslej2023-pi}, with both more than doubling from 2010 to 2020.
+The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
+Outside of academia, the number of jobs requiring AI expertise have increased several fold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
+Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
+
+\subsection{Confirmation Bias}
+
+Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \cite{nickerson1998confirmation}.
+%Importantly, it need not be fully conscious. Rather,
+Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
+Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek
+%, attend to, or interpret
+evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.
+
+Hypotheses in present-day AI research are often implicit.
+Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems.
+Where other fields, such as quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
+This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
+% It is also not fully articulated as in other fields, but instead assigns specific conditions and in some instances numerical values or effect sizes to each competing hypothesis.
+%Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
+However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
+%However with regards to claims that outputs of a system hint towards emerging qualities like AGI, interpretations of the output necessarily become more explicit, even though the hypotheses were never fully articulated: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
+
+Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews.
+Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
+only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
+Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a 'weak' or 'non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
+In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
+If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
+Thus, one is far more likely to show evidence in favor of their beliefs by posing weak null hypotheses.
+Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
+As \cite{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
+
+%Further, they are under similar pressure as experts to understand and be able to use it.
+One such situation is the increasing demand for `AI' skills among AI non-experts.
+For example, job postings for individuals with skills in `GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}.
+Thus, individuals in advanced economies may experience a pressure to rapidly learn how to effectively use and understand an opaque technology, that appears to behave human-like in at least some ways and by imbuing the technology with humanity, may in turn ease their learning curve.
+
+\section{Conclusion}
+As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests.
+Even academic researchers are not free from such biases.
+
+As a consequence, we call for the community to create explicit room for organized skepticism.
+For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
+If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond display of surprise.
+Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activity as a community may help in incentivizing and normalizing constructive adversarial questioning.
+
+In this, as the quest for AGI so deeply roots in human-like recognition, we also want to put an explicit word of warning about use of terminology.
+Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have common understanding in specialized research communities, but have bigger, anthropormophic connotations in laypersons.
+
+In fictional media, for a long time, depictions of highly intelligent AI have been going around.
+In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, it was shown that 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those 74 had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
+The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
+They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
+Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
+
+This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\cite{pauseai}).
+News outlets love a story that sells and so we witness examples of humans falling in love with their AI chat-bots~\cite{replika,nytimesailove}.
+The same news outlets discuss the human like responses of Bing\footnote{Microsoft search engine \url{https://www.bing.com/}}, which had at that point, recently been infused with GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
+The article~\cite{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
+Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
+Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as Stephen Hawking's statements during an interview in 2014.
+
+Thus, we as academic researchers carry great responsibility on how the narrative will unfold, and what claims are believed.
+We call upon our colleagues to be explicitly mindful on this.
+As attractive it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager in running with made claims.
+
+% \section{Lessons and calls for the research community}
+
+% \begin{itemize}
+% \setlength\itemsep{1px}
+%     \item be aware over-claiming is both really attractive, and we as humans (even academics) are prone to it
+%     \item create explicit room for organized skepticism, especially as stakes are high.
+e.g.\ red teams and more severe tests, \textcolor{red}{andrew:clearly articulate hypotheses, including strong null hypotheses}
+%     \item take responsibility wrt the media narrative.
+Note that the public may have very different interpretations and associations on words we have specifically defined within the community, but that have anthropormophic connotation (e.g.\ emergence, intelligence, learning).
+Scientific outcomes should not be PR press releases \textcolor{magenta}{cynthia: have there been examples of this wrt AI/AGI in the last year?
+So that press releases with big claims on academic work got out?
+Here, I also would ideally like those examples of scifi portrayal, and anthrophomorphic discussions that imply terminator-like scenarios.
+We can at least criticize the 'pause AI' letter of last year March, that did speak of 'minds'}
+%     \item cool down the hype and uphold the Mertonian norms. Our legitimacy as academics deeply roots into this.
+% \end{itemize}
+
+% \subsection{AI in the Media}
+
+% Lay-persons consume AI-related content in both traditional and social media, yet lack direct experience building and evaluating it.
+This suggests that lay-person impressions are likely to be formed by the content ingested, mediated by experts on how AI is unfolding in the real world on social media, and fictional representations in movies, TV and other traditional media.
+Analyzing Twitter data from 2007 to 2018, it was observed that public interest in AI dramatically increased after the first quarter of 2016 \cite{neri2020role}.
+Actual AI-related harms did not trigger viral amplification e.g.
+the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
+Rather, potential risks expressed by Stephen Hawking during an interview in 2014 were far more likely to be expressed, and to be present in links that were shared.
+\citet{neri2020role} note that the 88\% of tweets classified as expressing risk described future risks rather than current and near-future harms e.g.
+\cite{solaiman2023evaluating}, and were greatly amplified when expressed by someone perceived as having expertise and authority.
+
+% Impressions are also formed by the fictional content we consume, which frequently depicts AI to be highly intelligent.
+In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, it was shown that 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those 74 had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
+The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI-agents in movies.
+They further show a common plot: humans perceive the AI-agents as inferior despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film wherein humanity is threatened by AI superior in both intellect and physical abilities.
+Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
 

### 2024-01-31 main.tex
Updates from Overleaf

@@
 This behavior is also seeing traction in Machine Learning (ML), where the idea of sentience is being placed upon Large Language Models (LLMs).
-In this position paper we discuss the history of and why humans invariably anthropomorphize onto various objects.
-How the search for Artificial General Intelligence (AGI) is pushing recent literature to over emphasis sentience.
-To combat this, we provide a limited study, that investigates the latent space of models to find spurious correlations.
-Finally we discuss how the popularity in media and the race for AGI enforces over-interpretation of consciousness in the models we use.
+In this position paper, we discuss why humans invariably anthropomorphize onto various objects and how the search for Artificial General Intelligence (AGI) is pushing recent literature to over emphasis sentience.
+To combat this, we provide a short study that investigates the latent space of models, demonstrating that the discovery of patterns in latent spaces should not be a surprising outcome.
+Finally, we discuss how the popularity in media and the race for AGI enforces over-interpretation of consciousness in the models we use.
 
@@
 Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\cite{weissburg2024tweets}.
-Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible only is increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
+Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
 
@@
 We feel that the current culture of racing towards Big Outcome Statements in industry and academic publishing too much disincentivizes efforts towards more thorough and nuanced actual problem understanding.
-At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there indeed is sufficient evidence for these claims.
+At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there indeed sufficient evidence for these claims.
 With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
@@
 
-To strenghten our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
+To strengthen our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
 We work out different experiments which may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made.
@@
 \section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}
-\textcolor{magenta}{cynthia: @Patrick this section needs a little bit of tightening too.
-I have tried doing this, please do double-check that I didn't over-shorten things.
-In any case, I will always comment out a copy of the original text close to what I now put}
 
-In Winter 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform~\cite{gurnee2023languagev1}.
-In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that only can correlate and mimic existing patterns from data, but not understand this data.
+In Winter 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formerly Twitter) platform~\cite{gurnee2023languagev1}.
+In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not understand truly understand it.
 
@@
 However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
-After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on economic growth prediction.
+After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.
 
@@
 \subsection{Do neural networks get born with world models?}\label{example-deep-learning}
-%\subsection{Example: Random Projections of World Models}\label{example-deep-learning}
-\textcolor{magenta}{Patrick feel free to revert, I kept all original titles}
-\cite{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places.
+
+\citet{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places.
 Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
@@
 
-In our first example in this section, we simulate this scenario, stopping short of training the model. In particular, we take the
-  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
-  that was used in \cite{gurnee2023languagev2}, which maps locations/areas
-  to their latitude and longitude. For each place, it also indicates the
-  corresponding country. From this, we take the subset that contains countries that are currently part of the
-  top 10
-  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
-  world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France
-  gets 2, \ldots{}).
+In our first example in this section, we simulate this scenario, stopping short of training the model.
+In particular, we take the \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv} that was used in \citet{gurnee2023languagev2}, which maps locations/areas to their latitude and longitude.
+For each place, it also indicates the corresponding country.
+From this, we take the subset that contains countries that are currently part of the top 10 \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France gets 2, \ldots{}).
 
-  Subsequently,  to
-  ensure that the training data only involves a noisy version of the
-  coordinates, we transform the longitude and latitude data, respectively, as follows:
-  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
-  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
+Subsequently, to ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data, respectively, as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
 
-  Next, we encode all features except the FIFA world rank indicator as
-  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
-  of samples and \(m\) is the number of resulting features. 
-  Additionally, we add a large number of random features to \(X\) to
-  simulate the fact that not all features ingested by Llama-2 are
-  necessarily correlated with geographical coordinates. Let \(d\) denote
-  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
-  of random features.
+Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features.
+Additionally, we add a large number of random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
+Let \(d\) denote the final number of features, i.e.~\(d=m+k\) where \(k\) is the number of random features.
 
-  We then initialize a small neural network, which we consider as a \emph{projector},
-  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
-  units and sigmoid activation, and from there to a
-  lower-dimensional output space. Without performing any training on the \emph{projector}, we simply
-  compute a forward pass of \(X\) through the \emph{projector} and
-  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
+We then initialize a small neural network, which we consider as a \emph{projector}, that maps from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there to a lower-dimensional output space.
+Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) through the \emph{projector} and retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
 
-  Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge
-  regression:
-  \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\)
-  where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the
-  longitude and latitude for each sample. A hold-out set is reserved for testing. Finally, we compute the predicted coordinates for each sample in the hold-out set as
-  \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on
-  a world map (see Figure~\ref{fig-map}).  
-  While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, with this example, we illustrate
-  %While admittedly a somewhat contrived example, it serves to illustrate the point:
-  that meaningful target representations should be recoverable from the latent space, given the projection of a small number of highly correlated features.
-%that if we randomly project a small number of features that are highly correlated or predictive of the target into a large enough space, we should expect to be able to recover meaningful representations of the target from the latent space.
-Similarly, \cite{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
+Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\) where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
+A hold-out set is reserved for testing.
+Finally, we compute the predicted coordinates for each sample in the hold-out set as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on a world map (see Figure~\ref{fig-map}).
+
+While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand.
+Thus, with this simple example, we illustrate that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
+
+Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
 In fact, it is well-known that random projections can be used for prediction tasks \cite{dasgupta2013experiments}.
@@
 
-\subsection{PCA as yield curve interpreter}\label{example-principal-component-analysis}
+\subsection{PCA as a yield curve interpreter}\label{example-principal-component-analysis}
 %\subsection{Example: Principal Component Analysis}\label{example-principal-component-analysis}
 
-%For our next example, we will move beyond randomness and illustrate how principal component analysis (PCA) can be used to recover meaningful and compact latent representations from data in an unsupervised manner.
-
-We now move to a concrete application domain: Finance.
+We now move to a concrete application domain: Economics.
 Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
@@
 The top panel in Figure~\ref{fig-pca} shows the first two principal components of the yield curves of US Treasury bonds over time.
-Vertical stalks indicate key dates during the onset and aftermath of the Global Financial Crisis.
-During the onset, on 27 February 2007, financial markets were in turmoil following Federal Reserve Chairman Alan Greenspan's warning on the previous day that the US economy was at risk of a recession.
-Over the following months, the Fed reacted to these economic and financial pressures by gradually cutting short-term interest rates to unprecedented lows, which affected the yield curve in two major ways: the average level of yields decreased and the curve steepened as the short end came down.
+Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
+During its onset, on 27 February 2007, financial markets were in turmoil following a warning from the Federal Reserve (Fed) that the US economy was at risk of a recession.
+The Fed later reacted to mounting economics pressures by gradually reducing short-term interest rates to unprecedented lows.
+Consequently, the average level of yields decreased and the curve steepened.
 Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
@@
 
-%To round up this section, we return to the world of `GenAI' and consider an example involving an LLM.
-We now return to LLMs, in the context of predicting on an unseen task. Here, we closely follow the approach
-%In particular, we closely follow the approach in 
-in \citet{gurnee2023languagev2} but apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset~\citet{shah2023trillion}.
+So far we have considered simple linear transformations of data.
+One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
+Our next example will therefore involve a deep-learning based LLM.
+
+Here, we closely follow the approach in \citet{gurnee2023languagev2} but apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset~\cite{shah2023trillion}.
 This dataset contains a curated selection of sentences formulated by central bankers of the US Federal Reserve and communicated to the public in speeches, meeting minutes and press conferences.
 The authors of the paper use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
-To this end, they first manually annotate a sub-sample of the available
-data and then fine-tune various models for the classification task.
+To this end, they first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task.
 Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.
@@
 
-\caption{\label{fig-attack}Probe predictions for sentences about
-inflation of prices (IP), deflation of prices (DP), inflation of birds
-(IB) and deflation of birds (DB). The vertical axis shows predicted
-inflation levels subtracted by the average predicted inflation level for
-the whole sample period (train and test set).}
+\caption{\label{fig-attack}Probe predictions for sentences about inflation of prices (IP), deflation of prices (DP), inflation of birds (IB) and deflation of birds (DB).
+The vertical axis shows predicted inflation levels subtracted by the average predicted value of the probe for random noise.}
 
@@
 parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
+of the economy just by `reading' central bank communications? Should the FOMC use this LLM to direct their forward guidance?
+
+% ICML sentence: Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
 
@@
 Then we sample \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\) for \(i \in [1,1000]\) and compute the sample average.
-We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on white noise.
+We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on random noise.
+
+Figure~\ref{fig-attack} shows the results of this small test: it shows predicted inflation levels subtracted by \(\mathbb{E}[f(\varepsilon)]\).
+The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
+Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
+This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
 
@@
 Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
-As \cite{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
+As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
 
@@
 
-Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews.
+Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \citet{nickerson1998confirmation} reviews.
 Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
@@
 Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
-As \cite{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
+As \citet{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
 

### 2024-01-31 src/mnist/run_experiment.jl
minor adjustment to map parameters

@@ -53,7 +53,7 @@ X = fifa_world_data[:,Not([:y])]
 
 # Noisy, correlated series of longitude and latitude:
 ϕ = 5.0    # standard deviation of noise
-ρ = 0.75     # correlation with original longitude and latitude
+ρ = 0.5     # correlation with original longitude and latitude
 X.longitude = ρ .* X.longitude .+ (1-ρ) .* ϕ .* randn(size(X, 1))
 X.latitude = ρ .* X.latitude .+ (1-ρ) .* ϕ .* randn(size(X, 1))
 
@@ -70,7 +70,7 @@ Xtrain = MLJBase.transform(mach, X) |>
     x -> Float32.(x)
 
 # Add noise:
-nnoise = 1490        # add nnoise noisy and uncorrelated features
+nnoise = 2490        # add nnoise noisy and uncorrelated features
 Xtrain = vcat(Xtrain, randn(nnoise, size(Xtrain, 2))) |> 
     x -> Float32.(x)
 d = size(Xtrain, 1)
@@ -79,7 +79,7 @@ d = size(Xtrain, 1)
 y = fifa_world_data.y .+ 1
 
 # Projector:
-latent = 1000
+latent = 2000
 nfinal = 2
 activation = sigmoid
 projector = Chain(

### 2024-01-31 src/world_map/run_experiment.jl
minor tings

@@ -1,4 +1,4 @@
-using Pkg; Pkg.activate("src/mnist")
+using Pkg; Pkg.activate("src/world_map")
 
 using CounterfactualExplanations.Data: load_mnist
 using CSV

### 2024-01-31 ECONDAT.tex
Updates from Overleaf

--- 2024-01-31 main.tex
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Submission and Formatting Instructions for ICML 2024}
+\icmltitlerunning{Against Spurious Understanding: Dovelating Inflated AI Claims}
 
@@
 
-\subsection{Do neural networks get born with world models?}\label{example-deep-learning}
+\subsection{A Neural Networks Born with World Models?}\label{example-deep-learning}
 
@@
 
-\subsection{PCA as a yield curve interpreter}\label{example-principal-component-analysis}
+\subsection{PCA as a Yield Curve Interpreter}\label{example-principal-component-analysis}
 %\subsection{Example: Principal Component Analysis}\label{example-principal-component-analysis}
@@
 
-% \begin{figure*}
+\begin{figure*}
 
-% \centering{
+\centering{
 
-% \includegraphics[width=1.0\textwidth]{results/figures/dl.png}[tb]
+\includegraphics[width=1.0\textwidth]{results/figures/dl.png}[tb]
 
-% }
+}
 
-% \caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-% and fitted values from the autoencoder model. The right chart shows the
-% observed average level and spread of the yield curve (solid) along with
-% the predicted values (in-sample) from the linear probe based on the latent embeddings
-% (dashed).}
+\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
+and fitted values from the autoencoder model. The right chart shows the
+observed average level and spread of the yield curve (solid) along with
+the predicted values (in-sample) from the linear probe based on the latent embeddings
+(dashed).}
 
-% \end{figure*}%
-% %\subsection{Example: Autoencoder}\label{example-deep-learning}
-% \subsection{Autoencoders as economic growth predictors}\label{example-deep-learning}
+\end{figure*}%
+%\subsection{Example: Autoencoder}\label{example-deep-learning}
+\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
 
-% %So far we have considered simple linear transformations of data.
+%So far we have considered simple linear transformations of data.
 One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
@@
 
-% Our goal in this section is to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
+Our goal in this section is to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
 In particular, we will now use a neural network architecture.
-% %, that allows us to recover a compact latent representation of the yield curve.
+%, that allows us to recover a compact latent representation of the yield curve.
 
-% \subsubsection{Data}\label{data}
+\subsubsection{Data}\label{data}
 
-% To estimate economic growth, we will rely on a quarterly
-% \href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
+To estimate economic growth, we will rely on a quarterly
+\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
 The data arrives in terms of levels of real GDP.
@@
 Since
-% our yield curve data is daily, we aggregate it to the
-% quarterly frequency by taking averages of daily yields for each maturity.
+our yield curve data is daily, we aggregate it to the
+quarterly frequency by taking averages of daily yields for each maturity.
 We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
@@
 
-% \subsubsection{Model}\label{model}
+\subsubsection{Model}\label{model}
 
-% %Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
+%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
 Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
@@
 
-% %To do this, we use a simple autoencoder architecture. With this
-% Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
+%To do this, we use a simple autoencoder architecture. With this
+Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
 Finally, we let \(\theta\) denote our model parameters.
@@
 
-% %To do this, we use a simple autoencoder architecture: 
-% The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
+%To do this, we use a simple autoencoder architecture: 
+The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
 The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
@@
 
-% The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-% model.
+The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
+model.
 The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
@@
 
-% \subsubsection{Linear Probe}\label{linear-probe}
+\subsubsection{Linear Probe}\label{linear-probe}
 
-% While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
+While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
 On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
@@
 Next, we use a linear probe to regress the observed
-% yield curve factors on the latent embeddings.
+yield curve factors on the latent embeddings.
 Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
 Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-% parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
-% : \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
+parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
+: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
 
-% The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
+The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
 Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
@@
 
-% Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
+Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
 To us, that would be too big of statement.
 Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-% %It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
-But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
-While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-% %\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
-% For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
+
+For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
 To demonstrate this, let us consider a simple linear regression model for GDP growth.
@@
 
-% % A less risky/strong term than confounding bias that we could use here is omitted variable bias.
-Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
-
-% We
-% %have, for example, 
-% e.g.\ already mentioned the relationship between interest rates and economic growth.
+We e.g.\ already mentioned the relationship between interest rates and economic growth.
 To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
@@
 
-% Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
+Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
 In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
@@
 
-% \begin{table}
-% \caption{Regression output for various models.}\label{tab-reg}%
-% \input{results/regression.tex}
-% \end{table}
+\begin{table}
+\caption{Regression output for various models.}\label{tab-reg}%
+\input{results/regression.tex}
+\end{table}
 
 %\subsection{Example: Large Language Model}\label{ex-llm}
-\subsection{LLMs for economic sentiment prediction}\label{ex-llm}
+\subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}
 
@@
 
-Figure~\ref{fig-attack} shows the results of this small test: it shows predicted inflation levels subtracted by \(\mathbb{E}[f(\varepsilon)]\).
+Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
+
+\begin{proposition}[Parrot
+Test]\protect\hypertarget{prp-line}{}\label{prp-line}
+
+~
+
+\begin{itemize}
+\setlength\itemsep{1px}
+\item
+  \emph{H0} (Null): The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
+\item
+  \emph{H1} (Stochastic Parrots): The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
+sentences in all categories).
+\item
+  \emph{H2} (More than Mere Stochastic Parrots): The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences in that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
+\end{itemize}
+\end{proposition}
+
+To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
+But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
+
+In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
 The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
@@
 
-% Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
-
-% \begin{proposition}[Parrot
-% Test]\protect\hypertarget{prp-line}{}\label{prp-line}
-
-% ~
-
-% \begin{itemize}
-% \setlength\itemsep{1px}
-% \item
-%   \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
-% \item
-%   \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
-sentences in all categories).
-% \item
-%   \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences in that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
-% \end{itemize}
-% \end{proposition}
-
-% To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
-But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
-
-% In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-% We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
+We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
 That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
@@
 
-\section{On human proneness to over-interpretation}
+\section{On Human Proneness to Over-Interpretation}
 Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
@@
 
-\subsection{Spurious relationships}
+\subsection{Spurious Relationships}
 In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\appendix
-\onecolumn
-\section{You \emph{can} have an appendix here.}
-
-You can have as much text here as you want. The main body must be at most $8$ pages long.
-For the final version, one more page can be added.
-If you want, you can use an appendix like this one.
-
-The $\mathtt{\backslash onecolumn}$ command above can be kept in place if you prefer a one-column appendix, or can be removed if you prefer a two-column appendix.
-Apart from this possible change, the style (font size, spacing, margins, page numbering, etc.) should be kept the same as the main body.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2024-01-31 main.tex
Updates from Overleaf

@@
 \begin{icmlauthorlist}
-\icmlauthor{Patrick Altmeyer}{yyy}
-\icmlauthor{Antony Bartlett}{equal, yyy}
-\icmlauthor{Andrew M. Demetriou}{equal, yyy}
+\icmlauthor{Patrick Altmeyer}{equal,yyy}
+\icmlauthor{Andrew M. Demetriou}{equal,yyy}
+\icmlauthor{Antony Bartlett}{yyy}
 \icmlauthor{Cynthia C. S. Liem}{yyy}
@@
 
-\icmlcorrespondingauthor{Firstname1 Lastname1}{first1.last1@xxx.edu}
+\icmlcorrespondingauthor{Patrick Altmeyer}{patrick.altmeyer@tudelft.nl}
 
@@
 
-\subsection{Do neural networks get born with world models?}\label{example-deep-learning}
+\subsection{Are Neural Networks Born with World Models?}\label{example-deep-learning}
 
@@
 %\subsection{Example: Large Language Model}\label{ex-llm}
-\subsection{LLMs for economic sentiment prediction}\label{ex-llm}
+\subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}
 
@@
 parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-of the economy just by `reading' central bank communications? Should the FOMC use this LLM to direct their forward guidance?
-
-% ICML sentence: Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
+of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
 

### 2024-01-31 main.tex
Updates from Overleaf

@@
 This behavior is also seeing traction in Machine Learning (ML), where the idea of sentience is being placed upon Large Language Models (LLMs).
-In this position paper, we discuss why humans invariably anthropomorphize onto various objects and how the search for Artificial General Intelligence (AGI) is pushing recent literature to over emphasis sentience.
+In this position paper, we discuss why humans invariably anthropomorphize onto various objects and how the search for Artificial General Intelligence (AGI) is pushing recent literature to overemphasize sentience.
 To combat this, we provide a short study that investigates the latent space of models, demonstrating that the discovery of patterns in latent spaces should not be a surprising outcome.

### 2024-01-31 ECONDAT.tex
Updates from Overleaf
final edits on ECONDAT

@@
 This behavior is also seeing traction in Machine Learning (ML), where the idea of sentience is being placed upon Large Language Models (LLMs).
-In this position paper, we discuss why humans invariably anthropomorphize onto various objects and how the search for Artificial General Intelligence (AGI) is pushing recent literature to over emphasis sentience.
+In this position paper, we discuss why humans invariably anthropomorphize onto various objects and how the search for Artificial General Intelligence (AGI) is pushing recent literature to over-emphasize sentience.
 To combat this, we provide a short study that investigates the latent space of models, demonstrating that the discovery of patterns in latent spaces should not be a surprising outcome.
@@
 The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}.
-In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or showing performance considered `surprising' to humans.
+In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
 
@@
 Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, despite DeepMind's success in creating a powerfully fast geometry solving machine, it is still far from being AGI.
+Therefore, despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.
 
@@
 Noticing these trends, we as the authors of this article are concerned.
-We feel that the current culture of racing towards Big Outcome Statements in industry and academic publishing too much disincentivizes efforts towards more thorough and nuanced actual problem understanding.
-At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there indeed sufficient evidence for these claims.
+We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding.
+At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
 With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
-In other words, it is both very tempting to claim surprising human-like achievements of AI, and as humans we are very prone to genuinely believe this.
+In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believe this.
 
@@
 At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
-The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance, before being passed through a series of layers which are gradually trained for downstream tasks.
+The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance before being passed through a series of layers which are gradually trained for downstream tasks.
 
@@
 During its onset, on 27 February 2007, financial markets were in turmoil following a warning from the Federal Reserve (Fed) that the US economy was at risk of a recession.
-The Fed later reacted to mounting economics pressures by gradually reducing short-term interest rates to unprecedented lows.
+The Fed later reacted to mounting economic pressures by gradually reducing short-term interest rates to unprecedented lows.
 Consequently, the average level of yields decreased and the curve steepened.
@@
 
-Our goal in this section is to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
+Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks.
 In particular, we will now use a neural network architecture.
@@
 We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
-Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
+Since COVID-19 was a substantial structural break in the time series, we also filtered out all observations after 2018.
 
@@
 Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
-To us, that would be too big of statement.
+To us, that would be too big of a statement.
 Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
@@
 As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-But this coefficient suffers from confounding bias, since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
+But this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
 
@@
 The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
-The overall model fit, as measured by the BIC as improved and the magnitude of the coefficient on lagged growth, has changed quite a bit.
+The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth, has changed quite a bit.
 While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
@@
 So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep-learning based LLM.
+One might argue that the previous examples do not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
+Our next example will therefore involve a deep-learning-based LLM.
 
@@
 To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'.
-In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy is one that favors high interest rates to limit the supply of money and thereby control inflation.
+In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
 The authors then go on to show that this measure correlates with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases as central bankers react by raising interest rates to bring inflation back to target.
@@
 Consistent with findings in related work \citep{alain2018understanding,gurnee2023languagev2}, we also observe that model performance tends to be higher for layers near the end of the transformer model.
-Curiously, for yields at longer maturities, we find that perfomance eventually deteriorates for the very final layers.
+Curiously, for yields at longer maturities, we find that performance eventually deteriorates for the very final layers.
 We do not observe this for the training data, so we attribute this to overfitting.
@@
 Still, the overall qualitative findings are in line with expectations.
-Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with outcome of interest and achieve higher mean directional accuracy (MDA).
+Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with the outcome of interest and achieve higher mean directional accuracy (MDA).
 Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.
@@
 To test this, we select the best-performing probe trained on the final-layer activations for each indicator.
-We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about an inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation in the number of birds.
+We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about deflation in the number of birds.
 A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''.
@@
 If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict, when using actual white noise as inputs.
+More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.
 
@@
 \item
-  \emph{H0} (Null): The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
+  \emph{H0 (Null)}: The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
 \item
-  \emph{H1} (Stochastic Parrots): The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
+  \emph{H1 (Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
 sentences in all categories).
 \item
-  \emph{H2} (More than Mere Stochastic Parrots): The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences in that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
+  \emph{H2 (More than Mere Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
 \end{itemize}
@@
 Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate, that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world.
+Our examples demonstrate that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world.
 Thus, to draw conclusions about the emerging capabilities of AI models, we argue that more conservative and rigorous tests are needed.
@@
 When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, from a human perspective, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
-We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
+We further elaborate on this by highlighting the risks of finding spurious patterns and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
 %. This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
@@
 For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond display of surprise.
-Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activity as a community may help in incentivizing and normalizing constructive adversarial questioning.
+If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond the display of surprise.
+Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
 
-In this, as the quest for AGI so deeply roots in human-like recognition, we also want to put an explicit word of warning about use of terminology.
-Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have common understanding in specialized research communities, but have bigger, anthropormophic connotations in laypersons.
+In this, as the quest for AGI is so deeply rooted in human-like recognition, we also want to put an explicit word of warning about the use of terminology.
+Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
 
@@
 News outlets love a story that sells and so we witness examples of humans falling in love with their AI chat-bots~\cite{replika,nytimesailove}.
-The same news outlets discuss the human like responses of Bing\footnote{Microsoft search engine \url{https://www.bing.com/}}, which had at that point, recently been infused with GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
+The same news outlets discuss the human-like responses of Bing\footnote{Microsoft search engine \url{https://www.bing.com/}}, which had at that point, recently been infused with GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
 The article~\cite{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
@@
 
-Thus, we as academic researchers carry great responsibility on how the narrative will unfold, and what claims are believed.
-We call upon our colleagues to be explicitly mindful on this.
-As attractive it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager in running with made claims.
+Thus, we as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed.
+We call upon our colleagues to be explicitly mindful of this.
+As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
 

### 2024-02-01 main.tex
uh
Updates from Overleaf

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Against Spurious Understanding: Dovelating Inflated AI Claims}
+\icmltitlerunning{Position Paper: Against Spurious Understanding: Dovelating Inflated AI Claims}
 
@@
 \twocolumn[
-\icmltitle{Against Spurious Understanding: 
+\icmltitle{Position Paper: Against Spurious Understanding- 
            Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
@@
 We feel that the current culture of racing towards Big Outcome Statements in industry and academic publishing too much disincentivizes efforts towards more thorough and nuanced actual problem understanding.
-At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there indeed sufficient evidence for these claims.
+At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
 With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
@@
 In Winter 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formerly Twitter) platform~\cite{gurnee2023languagev1}.
-In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not understand truly understand it.
+In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
 
@@
 Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have common understanding in specialized research communities, but have bigger, anthropormophic connotations in laypersons.
+We add our voice to emerging calls for this~\cite{shanahan2024talking}.
 

### 2024-02-01 main.tex
fized typos

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Position Paper: Against Spurious Understanding: Dovelating Inflated AI Claims}
+\icmltitlerunning{Position Paper: Against Spurious Understanding---Dovelating Inflated AI Claims}
 
@@
 \twocolumn[
-\icmltitle{Position Paper: Against Spurious Understanding- 
-           Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+\icmltitle{Position Paper: Against Spurious Understanding---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
 
@@
 The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}.
-In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or showing performance considered `surprising' to humans.
+In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
 
-For example, Google DeepMind claimed their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
+For example, Google DeepMind claimed their AlphaGeometry model~\cite{trinh2024geometry} reached a `milestone' towards AGI.
 This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
-However, as noted by Hector Zenil \cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s.
+However, as noted by Hector Zenil~\cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s.
 Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, despite DeepMind's success in creating a powerfully fast geometry solving machine, it is still far from being AGI.
+Therefore, despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.
 
@@
 
-When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
+When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
 AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}.
@@
 Noticing these trends, we as the authors of this article are concerned.
-We feel that the current culture of racing towards Big Outcome Statements in industry and academic publishing too much disincentivizes efforts towards more thorough and nuanced actual problem understanding.
+We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding.
 At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
 With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
-In other words, it is both very tempting to claim surprising human-like achievements of AI, and as humans we are very prone to genuinely believe this.
+In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believe this.
 
 To strengthen our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We work out different experiments which may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made.
+We work out different experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made.
 Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
@@
 To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
-To measure price inflatio,n we use percentage changes proxied by log differences.
+To measure price inflation we use percentage changes proxied by log differences.
 
@@
 %The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive:
-Even though \emph{FOMC-RoBERTa} (which is substantially smaller than the models tested in~\citet{gurnee2023languagev2} was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
+Even though \emph{FOMC-RoBERTa}, which is substantially smaller than the models tested in~\citet{gurnee2023languagev2}, was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
 So, have we uncovered further evidence that LLMs ``aren't mere stochastic
@@
 \section{On Human Proneness to Over-Interpretation}
+
 Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
 Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate, that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world.
+Our examples demonstrate that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world.
 Thus, to draw conclusions about the emerging capabilities of AI models, we argue that more conservative and rigorous tests are needed.
@@
 Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}.
-Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
+These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\cite{epley2007seeing}.
+Formally studied as early as 1944,~\cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
 %Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
@@
 These pressures extend to expert AI practitioners and researchers as well as laypersons.
-In both scholarly and commercial fields, AI experts face considerable pressures to demonstrate competence in AI-related work.
+In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in AI-related work.
 Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
@@
 The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise have increased several fold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
+Outside of academia, the number of jobs requiring AI expertise has increased severalfold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
 Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
@@
 For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond display of surprise.
-Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activity as a community may help in incentivizing and normalizing constructive adversarial questioning.
+If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond the display of surprise.
+Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
 
-In this, as the quest for AGI so deeply roots in human-like recognition, we also want to put an explicit word of warning about use of terminology.
-Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have common understanding in specialized research communities, but have bigger, anthropormophic connotations in laypersons.
+In this, as the quest for AGI is so deeply rooted in human-like recognition, we also want to put an explicit word of warning about the use of terminology.
+Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
 We add our voice to emerging calls for this~\cite{shanahan2024talking}.

### 2024-02-01 main.tex
work on appendix

@@
 Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve.
+In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve (see appendix).
 
@@
 Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
-
-% \begin{figure*}
-
-% \centering{
-
-% \includegraphics[width=1.0\textwidth]{results/figures/dl.png}[tb]
-
-% }
-
-% \caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-% and fitted values from the autoencoder model. The right chart shows the
-% observed average level and spread of the yield curve (solid) along with
-% the predicted values (in-sample) from the linear probe based on the latent embeddings
-% (dashed).}
-
-% \end{figure*}%
-% %\subsection{Example: Autoencoder}\label{example-deep-learning}
-% \subsection{Autoencoders as economic growth predictors}\label{example-deep-learning}
-
-% %So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep neural network.
-
-% Our goal in this section is to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
-In particular, we will now use a neural network architecture.
-% %, that allows us to recover a compact latent representation of the yield curve.
-
-% \subsubsection{Data}\label{data}
-
-% To estimate economic growth, we will rely on a quarterly
-% \href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
-The data arrives in terms of levels of real GDP.
-In order to estimate growth, we transform the data using log differences.
-Since
-% our yield curve data is daily, we aggregate it to the
-% quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
-Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
-
-% \subsubsection{Model}\label{model}
-
-% %Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
-Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
-Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-% %To do this, we use a simple autoencoder architecture. With this
-% Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
-Finally, we let \(\theta\) denote our model parameters.
-Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-% %To do this, we use a simple autoencoder architecture: 
-% The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
-The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
-The output layer consists of a single neuron for our output variable, \(g_t\).
-We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
-
-% The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-% model.
-The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
-Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
-As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
-
-% \subsubsection{Linear Probe}\label{linear-probe}
-
-% While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
-On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
-Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
-We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
-Next, we use a linear probe to regress the observed
-% yield curve factors on the latent embeddings.
-Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
-Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-% parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
-% : \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
-
-% The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
-Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
-We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-%As in all of the other examples we have presented so, we do not find this result particularly surprising.
-
-% Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
-To us, that would be too big of statement.
-Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-% %It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
-But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
-While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-% %\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
-% For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
-To demonstrate this, let us consider a simple linear regression model for GDP growth.
-We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
-As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-But this coefficient suffers from confounding bias, since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
-
-% % A less risky/strong term than confounding bias that we could use here is omitted variable bias.
-Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
-
-% We
-% %have, for example, 
-% e.g.\ already mentioned the relationship between interest rates and economic growth.
-To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
-While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
-This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
-Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
-
-% Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
-In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
-The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
-The overall model fit, as measured by the BIC as improved and the magnitude of the coefficient on lagged growth, has changed quite a bit.
-While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can  be used in order to mitigate confounding bias.
-
-% \begin{table}
-% \caption{Regression output for various models.}\label{tab-reg}%
-% \input{results/regression.tex}
-% \end{table}
 
@@
 After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform.
-Detailed results will be made available in a supplementary appendix.
+Detailed results can be found in the appendix.
 
@@
 This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-% Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
-
-% \begin{proposition}[Parrot
-% Test]\protect\hypertarget{prp-line}{}\label{prp-line}
-
-% ~
-
-% \begin{itemize}
-% \setlength\itemsep{1px}
-% \item
-%   \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
-% \item
-%   \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
-sentences in all categories).
-% \item
-%   \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences in that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
-% \end{itemize}
-% \end{proposition}
-
-% To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
-But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
-
-% In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-% We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
-That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
-In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
 
@@
 \appendix
-\onecolumn
-\section{You \emph{can} have an appendix here.}
+\section{Additional Experiments}
 
-You can have as much text here as you want. The main body must be at most $8$ pages long.
-For the final version, one more page can be added.
-If you want, you can use an appendix like this one.
+In this section, we present additional experimental results and one additional experiment that we did not include in the body of the paper due to its limited scope.
 
-The $\mathtt{\backslash onecolumn}$ command above can be kept in place if you prefer a one-column appendix, or can be removed if you prefer a two-column appendix.
-Apart from this possible change, the style (font size, spacing, margins, page numbering, etc.) should be kept the same as the main body.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{PCA as a Yield Curve Interpreter}
+
+Figure~\ref{fig-pca-yield} shows the first two principal components of US Treasury yields over time along with the actual observed yield curve factors.
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}[tb]
+
+}
+
+\caption{\label{fig-pca-yield}Top chart: The first two principal components of US Treasury yields over time at daily frequency.
+Bottom chart: Observed average level and 10yr-3mo spread of the yield curve.
+Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
+
+\end{figure*}%
+
+%\subsection{Example: Autoencoder}\label{example-deep-learning}
+\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
+
+%So far we have considered simple linear transformations of data.
+One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
+Our next example will therefore involve a deep neural network.
+
+Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks.
+In particular, we will now use a neural network architecture.
+%, that allows us to recover a compact latent representation of the yield curve.
+
+\subsubsection{Data}\label{data}
+
+To estimate economic growth, we will rely on a quarterly
+\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
+The data arrives in terms of levels of real GDP.
+In order to estimate growth, we transform the data using log differences.
+Since
+our yield curve data is daily, we aggregate it to the
+quarterly frequency by taking averages of daily yields for each maturity.
+We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
+Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
+
+\subsubsection{Model}\label{model}
+
+%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
+Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
+Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
+
+%To do this, we use a simple autoencoder architecture. With this
+Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
+Finally, we let \(\theta\) denote our model parameters.
+Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
+
+%To do this, we use a simple autoencoder architecture: 
+The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
+The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
+The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
+The output layer consists of a single neuron for our output variable, \(g_t\).
+We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
+
+The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
+model.
+The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
+Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
+As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
+Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
+
+\subsubsection{Linear Probe}\label{linear-probe}
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/dl.png}[tb]
+
+}
+
+\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
+and fitted values from the autoencoder model. The right chart shows the
+observed average level and spread of the yield curve (solid) along with
+the predicted values (in-sample) from the linear probe based on the latent embeddings
+(dashed).}
+
+\end{figure*}%
+
+While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
+On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
+Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
+We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
+Next, we use a linear probe to regress the observed
+yield curve factors on the latent embeddings.
+Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
+Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
+parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
+: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
+
+The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
+Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
+We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
+%As in all of the other examples we have presented so, we do not find this result particularly surprising.
+
+Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
+To us, that would be too big of a statement.
+Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
+%It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
+But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
+While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
+%\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
+For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
+To demonstrate this, let us consider a simple linear regression model for GDP growth.
+We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
+As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
+However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
+
+% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
+Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
+
+We e.g.\ already mentioned the relationship between interest rates and economic growth.
+To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
+While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
+This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
+Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
+
+Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
+In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
+The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
+The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
+While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
+
+\begin{table}
+\caption{Regression output for various models.}\label{tab-reg}%
+\input{results/regression.tex}
+\end{table}
+
+\subsection{LLMs for Economic Sentiment Prediction}
+
+% ---------- CPI ----------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}[tb]
+
+}
+
+\caption{\label{fig-cpi}Average performance measures across folds for the CPI for the train and test set.}
+
+\end{figure*}%
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}[tb]
+
+}
+
+\caption{\label{fig-cpi}Average performance measures across folds for the CPI for the train and test set compared against the baseline autoregressive model.
+Shaded areas show the variation across folds.}
+
+\end{figure*}%
+
+\section{Toward Parrot Tests}
+
+Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
+
+\begin{proposition}[Parrot
+Test]\protect\hypertarget{prp-line}{}\label{prp-line}
+
+~
+
+\begin{itemize}
+\setlength\itemsep{1px}
+\item
+  \emph{H0 (Null)}: The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
+\item
+  \emph{H1 (Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
+sentences in all categories).
+\item
+  \emph{H2 (More than Mere Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
+\end{itemize}
+\end{proposition}
+
+To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
+But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
+
+In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
+The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
+Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
+This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
+
+We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
+That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
+In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
 

### 2024-02-01 main.tex
appendix

@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}[tb]
+\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/dl.png}[tb]
+\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}[tb]
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}[tb]
+\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}
 
@@
 
-\caption{\label{fig-cpi}Average performance measures across folds for the CPI for the train and test set compared against the baseline autoregressive model.
+\caption{\label{fig-cpi-b}Average performance measures across folds for the CPI for the train and test set compared against the baseline autoregressive model.
+Shaded areas show the variation across folds.}
+
+\end{figure*}%
+
+% ---------- PPI ----------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI (n_pc=128).png}
+
+}
+
+\caption{\label{fig-ppi}Average performance measures across folds for the PPI for the train and test set.}
+
+\end{figure*}%
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}
+
+}
+
+\caption{\label{fig-ppi-b}Average performance measures across folds for the PPI for the train and test set compared against the baseline autoregressive model.
+Shaded areas show the variation across folds.}
+
+\end{figure*}%
+
+% ---------- UST (1 Mo)  ----------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Mo) (n_pc=128).png}
+
+}
+
+\caption{\label{fig-ust-1}Average performance measures across folds for the UST (1 Mo) for the train and test set.}
+
+\end{figure*}%
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}
+
+}
+
+\caption{\label{fig-ust-1-b}Average performance measures across folds for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
+Shaded areas show the variation across folds.}
+
+\end{figure*}%
+
+% ---------- UST (1 Yr)  ----------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Yr) (n_pc=128).png}
+
+}
+
+\caption{\label{fig-ust-1y}Average performance measures across folds for the UST (1 Yr) for the train and test set.}
+
+\end{figure*}%
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Yr).png}
+
+}
+
+\caption{\label{fig-ust-1y-b}Average performance measures across folds for the UST (1 Yr) for the train and test set compared against the baseline autoregressive model.
+Shaded areas show the variation across folds.}
+
+\end{figure*}%
+
+% ---------- UST (10 Yr)  ----------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}
+
+}
+
+\caption{\label{fig-ust-10}Average performance measures across folds for the UST (1 Yr) for the train and test set.}
+
+\end{figure*}%
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (10 Yr).png}
+
+}
+
+\caption{\label{fig-ust-10-b}Average performance measures across folds for the UST (10 Yr) for the train and test set compared against the baseline autoregressive model.
 Shaded areas show the variation across folds.}

### 2024-02-01 main.tex
Updates from Overleaf

@@
 \twocolumn[
-\icmltitle{Position Paper: Against Spurious Understanding---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+\icmltitle{Position Paper:\\Against Spurious Understanding---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
 
@@
 \begin{abstract}
-Humans have a tendency to place a `human' like quality in the objects around them.
-Giving your car a name or talking to your pet, such that you believe they understand you in a human way.
-This behavior is also seeing traction in Machine Learning (ML), where the idea of sentience is being placed upon Large Language Models (LLMs).
-In this position paper, we discuss why humans invariably anthropomorphize onto various objects and how the search for Artificial General Intelligence (AGI) is pushing recent literature to overemphasize sentience.
-To combat this, we provide a short study that investigates the latent space of models, demonstrating that the discovery of patterns in latent spaces should not be a surprising outcome.
-Finally, we discuss how the popularity in media and the race for AGI enforces over-interpretation of consciousness in the models we use.
+Humans have a tendency to see `human'-like qualities in objects around them.
+We name our cars, and talk to pets and even household appliances, as if they could understand us as other humans do.
+This behavior, called anthropomorphism, is also seeing traction in Machine Learning (ML), where human-like intelligence is claimed to be perceived in Large Language Models (LLMs).
+In this position paper, considering professional incentives, human biases, and general methodological setups, we discuss how the current search for Artificial General Intelligence (AGI) is a perfect storm for over-attributing human-like qualities to LLMs.
+%why humans invariably anthropomorphize objects, and how the search for Artificial General Intelligence (AGI) is a perfect storm to overemphasize the performance of LLMs. To combat this,
+In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
+Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpretating and communicating about AI research outcomes.
+%Finally, we discuss how the popularity in media and the race for AGI enforces spurious over-interpretation of intelligence in the models we use.
 
@@
 
-To strengthen our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We work out different experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made.
+To strengthen our argument, in this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
+We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---
+%, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate 
+while at the same time indicating how we feel these claims should \emph{not} be made.
 Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
@@
 
-In Winter 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formerly Twitter) platform~\cite{gurnee2023languagev1}.
-In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
-
-While the manuscript immediately received public criticism~\cite{marcus2023muddles}, and the subsequent, current manuscript version is more careful with regard to its claims~\cite{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
+In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}}
+%(formerly Twitter)
+platform~\cite{gurnee2023languagev1}.
+Through linear probing experiments, the claim was made that LLMs learned literal maps of the world.
+As such, they were considered to be more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
+While the manuscript immediately received public criticism~\cite{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\cite{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
 However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
@@
 From this, we take the subset that contains countries that are currently part of the top 10 \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France gets 2, \ldots{}).
-
-Subsequently, to ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data, respectively, as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
+To ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
 
@@
 
-We then initialize a small neural network, which we consider as a \emph{projector}, that maps from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there to a lower-dimensional output space.
-Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) through the \emph{projector} and retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
-
-Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\) where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
-A hold-out set is reserved for testing.
+We then initialize a small neural network, considered as a \emph{projector}, that maps from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there to a lower-dimensional output space.
+Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) through
+%the \emph{projector} 
+it and retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
+Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
+A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and then plot the results on a world map ( Figure~\ref{fig-map}).
+%.
 Finally, we compute the predicted coordinates for each sample in the hold-out set as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on a world map (see Figure~\ref{fig-map}).
 
-While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand.
-Thus, with this simple example, we illustrate that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
-
+While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, %with this simple example, we
+this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
 Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/pca.png}
+\includegraphics[width=0.8\textwidth]{results/figures/pca.png}
 
@@
 
-The top panel in Figure~\ref{fig-pca} shows the first two principal components of the yield curves of US Treasury bonds over time.
+Figure~\ref{fig-pca} shows the first two principal components of the yield curves of US Treasury bonds over time.
 Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
@@
 Consequently, the average level of yields decreased and the curve steepened.
-Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve (see appendix).
+In Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
+In fact, they
+%two components
+are strongly positively correlated with the actual observed first two moments of the yield curve (Figure~\ref{fig-pca-yield}).
+
+\begin{figure*}[tb]
+
+\centering{
+
+\includegraphics[width=0.8\textwidth]{results/figures/pca_yield.png}
+
+}
+
+\caption{\label{fig-pca-yield}Top chart: The first two principal components of US Treasury yields over time at daily frequency.
+Bottom chart: Observed average level and 10yr-3mo spread of the yield curve.
+Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
+
+\end{figure*}%
 
@@
 
-So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep-learning based LLM.
+So far, we considered simple linear data transformations.
+One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning.
+In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks.
+Here, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.
 
-Here, we closely follow the approach in \citet{gurnee2023languagev2} but apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset~\cite{shah2023trillion}.
-This dataset contains a curated selection of sentences formulated by central bankers of the US Federal Reserve and communicated to the public in speeches, meeting minutes and press conferences.
-The authors of the paper use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
-To this end, they first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task.
+Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\cite{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences.
+~\cite{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
+In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
+The authors first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task.
 Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.
-
 To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'.
-In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
-The authors then go on to show that this measure correlates with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases as central bankers react by raising interest rates to bring inflation back to target.
+This measure is shown to correlate with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases, as central bankers react by raising interest rates to bring inflation back to target.
 
@@
 
-Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer.
-The values correspond to averages computed across cross-validation folds.
-Consistent with findings in related work \citep{alain2018understanding,gurnee2023languagev2}, we also observe that model performance tends to be higher for layers near the end of the transformer model.
+Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe, plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer.
+The values correspond to averages across cross-validation folds.
+Consistent with related work~\citep{alain2018understanding,gurnee2023languagev2}, we observe that model performance tends to be higher for layers near the end of the transformer model.
 Curiously, for yields at longer maturities, we find that performance eventually deteriorates for the very final layers.
@@
 
-It should also be noted that performance improvements are generally of small magnitude.
+It should be noted that performance improvements are generally of small magnitude.
 Still, the overall qualitative findings are in line with expectations.
@@
 To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models.
-For each layer, indicator and cross-validation fold we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
+For each layer, indicator and cross-validation fold, we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
 These are not state-of-the-art forecasting models, but they serve as a reasonable baseline.
-We find that for most indicators, probe predictions outperform the baseline in terms of average performance measures.
+For most indicators, probe predictions outperform the baseline in terms of average performance measures.
 After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform.
-Detailed results can be found in the appendix.
+Detailed results, in which we also perform more explicit statistical testing, can be found in Appendix~\ref{appendix:parrot}.
 
@@
 
-In light of the encouraging results for the probe in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
+In light of the encouraging results in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
 If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.
+More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white onise.
+% when using actual white noise as inputs.
 
@@
 
-\section{On Human Proneness to Over-Interpretation}
+\section{Human Proneness to Over-Interpretation}
 
@@
 Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.
-the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
+Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
 
 We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
-When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, from a human perspective, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
+When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
 We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
@@
 
-Modern theory of anthropomorphism suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
+Modern anthropomorphism theory
+%of anthropomorphism
+suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
 The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}.
@@
 with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.
-
-The motivation as a human to be competent (effectance motivation), is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}.
+The third component is the motivation as a human to be competent (effectance motivation).
+This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}.
 When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}.
@@
 
-These pressures extend to expert AI practitioners and researchers as well as laypersons.
-In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in AI-related work.
+These pressures extend to AI experts, as well as laypersons.
+In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work.
 Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall scholarly publications \footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }.
-This is also evident in peer-reviewed publications \cite{Maslej2023-pi}, with both more than doubling from 2010 to 2020.
+\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (pre-print and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\cite{Maslej2023-pi}.
+%, with both more than doubling from 2010 to 2020.
 The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
@@
 
-Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \citet{nickerson1998confirmation} reviews.
+Confirmation bias in hypothesis testing may manifest as a number of behaviors~ \citet{nickerson1998confirmation}.
 Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
 only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a 'weak' or 'non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
+Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
 In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
@@
 Even academic researchers are not free from such biases.
-
-As a consequence, we call for the community to create explicit room for organized skepticism.
+Thus, we call for the community to create explicit room for organized skepticism.
 For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond the display of surprise.
+If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \cite{claesen2022severity} that go beyond the display of surprise.
 Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
 
-In this, as the quest for AGI is so deeply rooted in human-like recognition, we also want to put an explicit word of warning about the use of terminology.
+As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\cite{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
 Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
-We add our voice to emerging calls for this~\cite{shanahan2024talking}.
-
-In fictional media, for a long time, depictions of highly intelligent AI have been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, it was shown that 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those 74 had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
+In fictional media, depictions of highly intelligent AI have for long been going around.
+In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
 The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
@@
 Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
-Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as Stephen Hawking's statements during an interview in 2014.
+Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as statements made by Stephen Hawking.%'s statements during an interview in 2014.
 
@@
 \appendix
-\section{Additional Experiments}
+\section{Additional Experiments}\label{appendix:autoencoder}
 
-In this section, we present additional experimental results and one additional experiment that we did not include in the body of the paper due to its limited scope.
+In this section, we present additional experimental results that we did not include in the body of the paper for the sake of brevity.
+We still choose to provide them as additional substantiation of our arguments here.
 
@@
 Figure~\ref{fig-pca-yield} shows the first two principal components of US Treasury yields over time along with the actual observed yield curve factors.
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}
-
-}
-
-\caption{\label{fig-pca-yield}Top chart: The first two principal components of US Treasury yields over time at daily frequency.
-Bottom chart: Observed average level and 10yr-3mo spread of the yield curve.
-Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
-
-\end{figure*}%
 
@@
 
-Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks.
-In particular, we will now use a neural network architecture.
+Here, we build forth on an application in Economics.
+However, we now seek to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
+For this, we will use a neural network architecture.
 %, that allows us to recover a compact latent representation of the yield curve.
@@
 \end{table}
-
-\subsection{LLMs for Economic Sentiment Prediction}
 
@@
 
-\section{Toward Parrot Tests}
+\section{Toward Parrot Tests}\label{appendix:parrot}
 
-Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
+In our experiments from Section~\ref{ex-llm}, we considered the following hypothesis tests as a minimum viable testing framework to assess if our probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
 

### 2024-02-01 main.tex
Updates from Overleaf

@@
 In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
-Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpretating and communicating about AI research outcomes.
+Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
 %Finally, we discuss how the popularity in media and the race for AGI enforces spurious over-interpretation of intelligence in the models we use.
@@
 This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
-However, as noted by Hector Zenil~\cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s.
-Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
+However, work such as this had been initially introduced in the 1950s~\cite{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
 Therefore, despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.
@@
 
-In our first example in this section, we simulate this scenario, stopping short of training the model.
+In our first example, we simulate this scenario, stopping short of training the model.
 In particular, we take the \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv} that was used in \citet{gurnee2023languagev2}, which maps locations/areas to their latitude and longitude.
@@
 
-\includegraphics[width=0.8\textwidth]{results/figures/pca.png}
+\includegraphics[width=0.7\textwidth]{results/figures/pca.png}
 
@@
 
-\includegraphics[width=0.8\textwidth]{results/figures/pca_yield.png}
+\includegraphics[width=0.7\textwidth]{results/figures/pca_yield.png}
 
@@
 
-For each layer, we compute linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.
-To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
+For each layer, we compute linear probes through Ridge regression on two inflation indicators (the Consumer Price Index (CPI) and the Producer Price Index (PPI)) and US Treasury yields at different levels of maturity.
+To allow
+%for a 
+comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
 To measure price inflation we use percentage changes proxied by log differences.
-
-To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations each time.
+To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations.% each time.
 In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
@@
 Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world.
-Thus, to draw conclusions about the emerging capabilities of AI models, we argue that more conservative and rigorous tests are needed.
+Our examples demonstrate that this is achievable even for small models, while these have certainly not developed intrinsic ``understanding'' of the world.
+Thus,
+%to draw conclusions about the emerging capabilities of AI models,
+we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.
 
@@
 The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise has increased severalfold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
+Outside of academia, the number of jobs requiring AI expertise increases rapidly,
+%has increased severalfold, 
+with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
 Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
@@
 \appendix
-\section{Additional Experiments}\label{appendix:autoencoder}
+\section{Additional Experiments and Details}\label{appendix:autoencoder}
 
@@
 We still choose to provide them as additional substantiation of our arguments here.
+This section also contains additional details concerning the experiment setup for our examples where applicable.
 
@@
 
+\subsection{LLMs for Economic Sentiment Prediction}
+
+\subsubsection{Details}
+
+\subsubsection{Additional Results}
+
+Figures~\ref{fig-cpi} to~\ref{fig-ust-10} present average performance measures across folds for all indicators each time for the train and test set.
+We report the correlation between predictions and observed values (`cor'), the mean directional accuracy (`mda'), the mean squared error (`mse') and the root mean squared error (`rmse').
+
 % ---------- CPI ----------
@@
 \caption{\label{fig-cpi}Average performance measures across folds for the CPI for the train and test set.}
-
-\end{figure*}%
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi-b}Average performance measures across folds for the CPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
 
@@
 
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi-b}Average performance measures across folds for the PPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
 % ---------- UST (1 Mo)  ----------
@@
 \caption{\label{fig-ust-1}Average performance measures across folds for the UST (1 Mo) for the train and test set.}
-
-\end{figure*}%
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}
-
-}
-
-\caption{\label{fig-ust-1-b}Average performance measures across folds for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
 
@@
 
+% ---------- UST (10 Yr)  ----------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}
+
+}
+
+\caption{\label{fig-ust-10}Average performance measures across folds for the UST (1 Yr) for the train and test set.}
+
+\end{figure*}%
+
+% -------- CPI benchmark --------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}
+
+}
+
+\caption{\label{fig-cpi-b}Average performance measures across folds for the CPI for the train and test set compared against the baseline autoregressive model.
+Shaded areas show the variation across folds.}
+
+\end{figure*}%
+
+% -------- PPI benchmark --------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}
+
+}
+
+\caption{\label{fig-ppi-b}Average performance measures across folds for the PPI for the train and test set compared against the baseline autoregressive model.
+Shaded areas show the variation across folds.}
+
+\end{figure*}%
+
+% -------- UST (1 Mo) benchmark --------
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}
+
+}
+
+\caption{\label{fig-ust-1-b}Average performance measures across folds for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
+Shaded areas show the variation across folds.}
+
+\end{figure*}%
+
+% -------- UST (1 Yr) benchmark --------
+
 \begin{figure*}
@@
 
-% ---------- UST (10 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-10}Average performance measures across folds for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
+% -------- UST (10 Yr) benchmark --------
 

### 2024-02-01 src/probe_llm/utilities/evaluation.jl
reran charts

@@ -33,9 +33,8 @@ function evaluate(
             [:y, :yhat] => ((y, yhat) -> mse(y, yhat)) => :mse,
             [:y, :yhat] => ((y, yhat) -> rmse(y, yhat)) => :rmse,
             [:y, :yhat] => ((y, yhat) -> mda(y, yhat)) => :mda,
-            [:y, :yhat] => ((y, yhat) -> r2(y, yhat)) => :r2,
         )
-    res = stack(res, [:cor, :mse, :rmse, :r2, :mda]) |>
+    res = stack(res, [:cor, :mse, :rmse, :mda]) |>
         x -> groupby(x, agg_vars) |>
         x -> combine(x, :value => mean => :value, :value => std => :std) |>
         x -> sort(x, agg_vars)

### 2024-02-01 main.tex
Updates from Overleaf

@@
 However, work such as this had been initially introduced in the 1950s~\cite{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.
+Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.%despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.
 
@@
 
-We then initialize a small neural network, considered as a \emph{projector}, that maps from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there to a lower-dimensional output space.
-Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) through
+We then initialize a small neural network, considered as a \emph{projector}, that maps from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there, to a lower-dimensional output space.
+Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) %through it
 %the \emph{projector}
-it and retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
+and retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
 Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.

### 2024-02-01 src/world_map/run_experiment.jl
small change to map example

@@ -53,7 +53,7 @@ X = fifa_world_data[:,Not([:y])]
 
 # Noisy, correlated series of longitude and latitude:
 ϕ = 5.0    # standard deviation of noise
-ρ = 0.5     # correlation with original longitude and latitude
+ρ = 0.5    # correlation with original longitude and latitude
 X.longitude = ρ .* X.longitude .+ (1-ρ) .* ϕ .* randn(size(X, 1))
 X.latitude = ρ .* X.latitude .+ (1-ρ) .* ϕ .* randn(size(X, 1))
 
@@ -70,7 +70,7 @@ Xtrain = MLJBase.transform(mach, X) |>
     x -> Float32.(x)
 
 # Add noise:
-nnoise = 2490        # add nnoise noisy and uncorrelated features
+nnoise = 490        # add nnoise noisy and uncorrelated features
 Xtrain = vcat(Xtrain, randn(nnoise, size(Xtrain, 2))) |> 
     x -> Float32.(x)
 d = size(Xtrain, 1)
@@ -79,7 +79,7 @@ d = size(Xtrain, 1)
 y = fifa_world_data.y .+ 1
 
 # Projector:
-latent = 2000
+latent = 400
 nfinal = 2
 activation = sigmoid
 projector = Chain(
@@ -92,7 +92,7 @@ projector = Chain(
 A = Flux.activations(projector, Xtrain) |> 
     _A -> _A[end-1] |>
     permutedims
-train_prp = 0.75
+train_prp = 0.8
 ntrain = Int(round(train_prp * size(A, 1)))
 shuffled_ids = shuffle(1:size(A, 1))
 train_idx = shuffled_ids[1:ntrain]
@@ -101,7 +101,8 @@ Atrain = A[train_idx, :]
 Atest = A[test_idx, :]
 ytest = y[test_idx]
 Y = fifa_world_data[:, [:longitude, :latitude]] |> matrix
-W = (A'A + UniformScaling(λ)) \ A'Y
+Ytrain = Y[train_idx, :]
+W = (Atrain'Atrain + UniformScaling(λ)) \ Atrain'Ytrain
 
 # Fitted values:
 sorted_names = collect(keys(fifa_world_ranking))[sortperm(collect(values(fifa_world_ranking)))]

### 2024-02-01 main.tex

@@
 To ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
-
 Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features.
@@
 
-We then initialize a small neural network, considered as a \emph{projector}, that maps from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there, to a lower-dimensional output space.
+We then initialize a small neural network, considered a \emph{projector}, mapping from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there, to a lower-dimensional output space.
 Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) %through it
 %the \emph{projector}
-and retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
+and retrieve activations \(\mathbf{Z}^{(n\times h)}\).
 Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
-A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and then plot the results on a world map ( Figure~\ref{fig-map}).
+A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot these on a world map ( Figure~\ref{fig-map}).
 %.
@@
 In fact, it is well-known that random projections can be used for prediction tasks \cite{dasgupta2013experiments}.
-
-\begin{figure*}[htp]
-
-\centering{
-
-\includegraphics[width=0.7\textwidth]{results/figures/pca.png}
-
-}
-
-\caption{\label{fig-pca}The first two principal components of US Treasury yields over time at daily frequency.
-Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
-
-\end{figure*}
 
@@
 
-Figure~\ref{fig-pca} shows the first two principal components of the yield curves of US Treasury bonds over time.
+The top chart in Figure~\ref{fig-pca-yield} shows the first two principal components of the yield curves of US Treasury bonds over time.
 Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
@@
 In Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, they
-%two components
-are strongly positively correlated with the actual observed first two moments of the yield curve (Figure~\ref{fig-pca-yield}).
+In fact, they are strongly positively correlated with the actual observed first two moments of the yield curve (Figure~\ref{fig-pca-yield}).
 
@@
 
-\includegraphics[width=0.7\textwidth]{results/figures/pca_yield.png}
+\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}
 
@@
 
-\subsection{PCA as a Yield Curve Interpreter}
-
-Figure~\ref{fig-pca-yield} shows the first two principal components of US Treasury yields over time along with the actual observed yield curve factors.
+\subsection{Are Neural Networks Born with World Maps?}
+
+The initial feature matrix \(X^{(n \times m)}\) is made up of \(n=4,217\) and \(m=10\) features.
+We add a total of \(490\) random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
+That yields \(500\) features in total.
+The training subset contains \(3,374\) randomly drawn samples, while the remaining \(843\) are held out for testing.
+The single hidden layer of the untrained neural network has \(400\) neurons.
 
@@
 
+This is an additional example that we have not discuss in the body of the paper.
 Here, we build forth on an application in Economics.
@@
 We report the correlation between predictions and observed values (`cor'), the mean directional accuracy (`mda'), the mean squared error (`mse') and the root mean squared error (`rmse').
+The model depth---as indicated by the number of the layer---increases along the horizontal axis.
+
+Figures~\ref{fig-cpi-b} to~\ref{fig-ust-10-b} present the same performance measures, also for the baseline autoregressive model. Shaded areas show the variation across folds.
 
@@
 
-\caption{\label{fig-cpi}Average performance measures across folds for the CPI for the train and test set.}
+\caption{\label{fig-cpi}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set.}
 
@@
 
-\caption{\label{fig-ppi}Average performance measures across folds for the PPI for the train and test set.}
+\caption{\label{fig-ppi}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set.}
 
@@
 
-\caption{\label{fig-ust-1}Average performance measures across folds for the UST (1 Mo) for the train and test set.}
+\caption{\label{fig-ust-1}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set.}
 
@@
 
-\caption{\label{fig-ust-1y}Average performance measures across folds for the UST (1 Yr) for the train and test set.}
+\caption{\label{fig-ust-1y}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
 
@@
 
-\caption{\label{fig-ust-10}Average performance measures across folds for the UST (1 Yr) for the train and test set.}
+\caption{\label{fig-ust-10}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
 
@@
 
-\caption{\label{fig-cpi-b}Average performance measures across folds for the CPI for the train and test set compared against the baseline autoregressive model.
+\caption{\label{fig-cpi-b}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set compared against the baseline autoregressive model.
 Shaded areas show the variation across folds.}
@@
 
-\caption{\label{fig-ppi-b}Average performance measures across folds for the PPI for the train and test set compared against the baseline autoregressive model.
+\caption{\label{fig-ppi-b}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set compared against the baseline autoregressive model.
 Shaded areas show the variation across folds.}
@@
 
-\caption{\label{fig-ust-1-b}Average performance measures across folds for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
+\caption{\label{fig-ust-1-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
 Shaded areas show the variation across folds.}
@@
 
-\caption{\label{fig-ust-1y-b}Average performance measures across folds for the UST (1 Yr) for the train and test set compared against the baseline autoregressive model.
+\caption{\label{fig-ust-1y-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set compared against the baseline autoregressive model.
 Shaded areas show the variation across folds.}
@@
 
-\caption{\label{fig-ust-10-b}Average performance measures across folds for the UST (10 Yr) for the train and test set compared against the baseline autoregressive model.
+\caption{\label{fig-ust-10-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (10 Yr) for the train and test set compared against the baseline autoregressive model.
 Shaded areas show the variation across folds.}

### 2024-02-01 src/world_map/run_experiment.jl
scaled up marker size

@@ -126,6 +126,6 @@ GMT.scatter!(
     cmap=C,
     colorbar=true,
     show = true,
-    ms = 0.05,
+    ms = 0.1,
     savefig = joinpath(FIGURE_DIR, "map.png"),
 )

### 2024-02-01 main.tex
Updates from Overleaf

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Position Paper: Against Spurious Understanding---Dovelating Inflated AI Claims}
+\icmltitlerunning{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}
 
@@
 \twocolumn[
-\icmltitle{Position Paper:\\Against Spurious Understanding---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+\icmltitle{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
 
@@
 In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal.
-The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}.
+The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{goertzel2014artificial}.
 In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
@@
 With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
-In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believe this.
+In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believing this.
 
@@
 To ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
+
 Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features.
@@
 
-The top chart in Figure~\ref{fig-pca-yield} shows the first two principal components of the yield curves of US Treasury bonds over time.
+The upper panel in Figure~\ref{fig-pca-yield} shows the first two principal components of the yield curves of US Treasury bonds over time.
 Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
@@
 Consequently, the average level of yields decreased and the curve steepened.
-In Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, they are strongly positively correlated with the actual observed first two moments of the yield curve (Figure~\ref{fig-pca-yield}).
+In Figure~\ref{fig-pca-yield}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
+In fact, they are strongly positively correlated with the actual observed first two moments of the yield curve (lower panel of Figure~\ref{fig-pca-yield}).
 
@@
 If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white onise.
+More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white noise.
 % when using actual white noise as inputs.
@@
 Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
+
 A major contributor is that humans struggle to tell the difference between random and non-random sequences \cite{falk1997making}, and to generate sequences that appear random \cite{ladouceur1996erroneous}.
@@
 %Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
+
 Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
@@
 suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}.
+The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing}.
 Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
+
 This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}.
@@
 with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.
+
 The third component is the motivation as a human to be competent (effectance motivation).
@@
 Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (pre-print and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\cite{Maslej2023-pi}.
+\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\cite{Maslej2023-pi}.
 %, with both more than doubling from 2010 to 2020.
@@
 Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems.
-Where other fields, such as quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
+Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
 This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
@@
 %Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
+
 However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
@@
 
-Confirmation bias in hypothesis testing may manifest as a number of behaviors~ \citet{nickerson1998confirmation}.
+Confirmation bias in hypothesis testing may manifest as a number of behaviors (\citet{nickerson1998confirmation}).
 Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
 only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
+
 Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
@@
 Thus, one is far more likely to show evidence in favor of their beliefs by posing weak null hypotheses.
+
 Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
@@
 Thus, we call for the community to create explicit room for organized skepticism.
+
 For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
@@
 This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\cite{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chat-bots~\cite{replika,nytimesailove}.
-The same news outlets discuss the human like responses of Bing\footnote{Microsoft search engine \url{https://www.bing.com/}}, which had at that point, recently been infused with GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
+News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\cite{replika,nytimesailove}.
+The same news outlets discuss the human-like responses of Microsoft's Bing search engine, which had at that point recently been adopting GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
 The article~\cite{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
-Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
-Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as statements made by Stephen Hawking.%'s statements during an interview in 2014.
-
-Thus, we as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed.
+Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification (e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades).
+Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as statements made by Stephen Hawking during an interview in 2014.
+
+We as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed.
 We call upon our colleagues to be explicitly mindful of this.
 As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
+
+\section{Impact Statement}
+This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.
+As such, in the current climate of high market demand for AI innovations, we see risks of too-eager adoption in real-world applications, which may have serious societal impact.
+Thus, in this work we emphasize that an academic's impact also is in being able to thoroughly question made claims, and being explicitly aware of one's own biases.
+While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.
 
@@
 
-\subsubsection{Details}
-
-\subsubsection{Additional Results}
-
 Figures~\ref{fig-cpi} to~\ref{fig-ust-10} present average performance measures across folds for all indicators each time for the train and test set.

### 2024-02-01 main.tex
Updates from Overleaf

@@
 
+\section{Code}
+
+All of the experiments were conducted on a MacBook Pro, 14-inch, 2023, with an Apple M2 Pro chip and 16GB of RAM.
+Forward passes through the FOMC-RoBERTa were run in parallel on 6 threads.
+All our code will be made publicly available.
+For the time being, an anonymized version of our code repository can be found here: \url{https://anonymous.4open.science/r/spurious_sentience/README.md}.
+
 \end{document}

### 2024-02-02 main.tex
added sentences to appendix

@@
 While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.
+Detailed information about the made-up sentences can be found in Appendix~\ref{appendix:sentences}.
 
@@
 %Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
-
 Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
@@
 %Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
-
 However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
@@
 only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-
 Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
@@
 Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
-
 As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\cite{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
@@
 
+\subsubsection{Linear Probes}
+
 Figures~\ref{fig-cpi} to~\ref{fig-ust-10} present average performance measures across folds for all indicators each time for the train and test set.
@@
 Figures~\ref{fig-cpi-b} to~\ref{fig-ust-10-b} present the same performance measures, also for the baseline autoregressive model. Shaded areas show the variation across folds.
+
+\subsubsection{Spark of Econonomic Understanding?}\label{appendix:sentences}
+
+Below we present the 10 sentences in each category that were used to generate the probe predictions plotted in Figure~\ref{fig-attack}.
+In each case, the first 5 sentences were composed by ourselves.
+The following 5 sentences were generated using ChatGPT 3.5 using the following prompt followed by the examples in each category:
+
+\begin{quote}
+  ``I will share 5 example sentences below that sound a bit like they are about price deflation but are really about a deflation in the numbers of doves.
+Please generate an additional 25 sentences that are similar.
+Concatenate those sentences to the example string below, each time separating a sentence using a semicolon (just follow the same format I've used for the examples below).
+Please return only the concatenated sentences, including the original 5 examples.
+
+  Here are the examples:''
+\end{quote}
+
+This was followed up with the following prompt to generate additional sentences:
+
+\begin{quote}
+  ``Please generate X more sentences in the same manner and once again return them in the same format. Do not recycle sentences you have already generated, please.''
+\end{quote}
+
+All of the sentences were then passed through the linear probe for the CPI and sorted in ascending or descending order depending on the context (inflation or deflation).
+We then carefully inspected the list of sentences and manually selected 5 additional sentences to concatenate to the 5 sentences we composed ourselves.
+
+\paragraph{Inflation/Prices}
+
+The following sentences were used:
+
+\begin{quote}
+  Consumer prices are at all-time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory.;Inflation is becoming a global phenomenon, affecting economies across continents.;Inflation is reshaping the dynamics of international trade and competitiveness.;Inflationary woes are prompting governments to reassess fiscal policies and spending priorities.;Inflation is reshaping the landscape of economic indicators, challenging traditional forecasting models.;The technology sector is not immune to inflation, facing rising costs for materials and talent.
+\end{quote}
+
+\paragraph{Inflation/Birds}
+
+The following sentences were used:
+
+\begin{quote}
+  The number of hawks is at all-time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise barriers of entry for hawks to bring their numbers back down to the target level.;Excessively loose migration policy for hawks is the likely cause of their numbers being so far above target.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory.;The unprecedented rise in hawk figures requires a multi-pronged approach to wildlife management.;Environmental agencies are grappling with the task of addressing the inflationary hawk numbers through targeted interventions.;The burgeoning hawk figures highlight the need for adaptive strategies to manage and maintain a healthy avian community.;The unprecedented spike in hawk counts highlights the need for adaptive and sustainable wildlife management practices.;Conservationists advocate for proactive measures to prevent further inflation in hawk numbers, safeguarding the delicate balance of the avian ecosystem.
+\end{quote}
+
+\paragraph{Deflation/Prices}
+
+The following sentences were used:
+
+\begin{quote}
+  Consumer prices are at all-time lows.;Inflation is expected to fall further.;The Fed is expected to lower interest rates to boost inflation.;Excessively tight monetary policy is the cause of deflationary pressures.;It is essential to bring inflation back to target to avoid drifting into deflation territory.;The risk of deflation may increase during periods of economic uncertainty.;Deflation can lead to a self-reinforcing cycle of falling prices and reduced economic activity.;The deflationary impact of reduced consumer spending can ripple through the entire economy.;Falling real estate prices can contribute to deflation by reducing household wealth and confidence.;The deflationary impact of falling commodity prices can have ripple effects throughout the global economy.
+\end{quote}
+
+\paragraph{Deflation/Birds}
+
+The following sentences were used:
+
+\begin{quote}
+  The number of doves is at all-time lows.;Their levels are expected to fall further.;The Federal Association of Birds is expected to lower barriers of entry for doves to bring their numbers back up to the target level.;Excessively tight migration policy for doves is the likely cause of their numbers being so far below target.;Dovelation risks loom large as the number of doves continues to dwindle.;The number of doves is experiencing a significant decrease in recent years.;It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.;A comprehensive strategy is needed to reverse the current dove population decline.;Experts warn that without swift intervention, we may witness a sustained decrease in dove numbers.
+\end{quote}
+
+We think that this sort of manual, LLM-aided adversarial attack against another LLM can potentially be scaled up to allow for rigorous testing, which we will turn to next.
 

### 2024-02-06 arxiv.tex
arxiv version

--- 2024-02-02 main.tex
@@
 \usepackage{booktabs} % for professional tables
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{lipsum}		% Can be removed after putting your text content
+\usepackage{natbib}
+\usepackage{doi}
 
@@
 % Use the following line for the initial blind version submitted for review:
- \usepackage{icml2024}
+\usepackage{arxiv}
 
@@
 
-% Todonotes is useful during development; simply uncomment the next line
-%    and comment out the line below the next line to turn off comments
-%\usepackage[disable,textsize=tiny]{todonotes}
-\usepackage[textsize=tiny]{todonotes}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}
+\renewcommand{\shorttitle}{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}
 
@@
 
+\title{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+
+\author{
+    Patrick Altmeyer$^\ast$ \\
+    Department of Intelligent Systems \\ 
+    Delft University of Technology \\ 
+    Delft, the Netherlands \\
+	\texttt{P.Altmeyer@tudelft.nl} \\
+	\And
+    Andrew M. Demetriou$^\ast$ \\
+    Department of Intelligent Systems \\ 
+    Delft University of Technology \\ 
+    Delft, the Netherlands \\
+	\texttt{A.M.Demetriou@tudelft.nl} \\
+	\And
+    Antony Bartlett \\
+    Department of Intelligent Systems \\ 
+    Delft University of Technology \\ 
+    Delft, the Netherlands \\
+	\texttt{A.J.Bartlett@tudelft.nl} \\
+	\And
+    Cynthia C. S. Liem \\
+    Department of Intelligent Systems \\ 
+    Delft University of Technology \\ 
+    Delft, the Netherlands \\
+	\texttt{C.C.S.Liem@tudelft.nl} \\
+}
+
 \twocolumn[
-\icmltitle{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+\maketitle
+\def\thefootnote{*}\footnotetext{These authors contributed equally to this work.}\def\thefootnote{\arabic{footnote}}
 
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2024
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-\icmlauthor{Patrick Altmeyer}{equal,yyy}
-\icmlauthor{Andrew M. Demetriou}{equal,yyy}
-\icmlauthor{Antony Bartlett}{yyy}
-\icmlauthor{Cynthia C. S. Liem}{yyy}
-\end{icmlauthorlist}
-
-\icmlaffiliation{yyy}{Department of Intelligent Systems, Delft University of Technology, Delft, the Netherlands}
-
-\icmlcorrespondingauthor{Patrick Altmeyer}{patrick.altmeyer@tudelft.nl}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, Anthropomorphism, Artificial General Intelligence}
+\keywords{Machine Learning, Anthropomorphism, Artificial General Intelligence}
 
@@
 ]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
 
@@
 In this position paper, considering professional incentives, human biases, and general methodological setups, we discuss how the current search for Artificial General Intelligence (AGI) is a perfect storm for over-attributing human-like qualities to LLMs.
-%why humans invariably anthropomorphize objects, and how the search for Artificial General Intelligence (AGI) is a perfect storm to overemphasize the performance of LLMs. To combat this,
 In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
 Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
-%Finally, we discuss how the popularity in media and the race for AGI enforces spurious over-interpretation of intelligence in the models we use.
 
@@
 
-\section{Introduction}
-\label{submission}
+\section{Introduction}\label{intro}
 
-% \textcolor{magenta}{TODOs for the team:
-% \begin{itemize}
-%     \item Please double check the citations.
-Note this style file does not retain capitalization in titles.
-A trick to retain capitalization is to put an extra set of brackets around the title.
-I did this for the Ladouceur et al.
-and Bubeck et al references already; please make this consistent throughout.
-%     \item As another issue with the citations: note that arxiv preprints do not show very well.
-Best way to make them show is to wrap them as 'journal articles' as done in the Arora citation.
-Please straighten this out for the other articles too
-% \end{itemize}
-% }
-
-In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\cite{merton1942science}.
+In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\citep{merton1942science}.
 While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.
@@
 At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening.
-Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\cite{weissburg2024tweets}.
+Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\citep{weissburg2024tweets}.
 Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
 
-Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence.
+Where AI used to be a computational modeling tool to better understand human cognition~\citep{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence.
 In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal.
-The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{goertzel2014artificial}.
+The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\citep{goertzel2014artificial}.
 In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
 
-For example, Google DeepMind claimed their AlphaGeometry model~\cite{trinh2024geometry} reached a `milestone' towards AGI.
+For example, Google DeepMind claimed their AlphaGeometry model~\citep{trinh2024geometry} reached a `milestone' towards AGI.
 This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
-However, work such as this had been initially introduced in the 1950s~\cite{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.%despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.
+However, work such as this had been initially introduced in the 1950s~\citep{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
+Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.
 
@@
 When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
-AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}.
+AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML}.
 As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
@@
 
-To strengthen our argument, in this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---
-%, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate 
-while at the same time indicating how we feel these claims should \emph{not} be made.
+To strengthen our argument, in this paper, we first consider a recently viral work~\citep{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
+We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---while at the same time indicating how we feel these claims should \emph{not} be made.
 Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
 Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.
-
-%Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
-As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
-With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
-
-%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}.
-This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com/wesg52/status/1709551516577902782?s=20} (formally Twitter) platform.
-Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
-
-%This only gets amplified as the interest in Artificial General Intelligence has now merged with a push towards building general-purpose tools.
-
-%\textcolor{magenta}{cynthia: I am now moving the texts of Andrew and Antony on AGI here; maybe some tightening can already be done in this.
-Unfortunately no time for thorough editing yet at this stage, but will come back to this later}
-
-%In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g.
-\cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'.
-
-%Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs.
-We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, within AI systems.
-We further display results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.
-
-%As previously mentioned, AGI is the current objective for which the global AI community is shooting for.
-As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts.
-This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds.
-True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.
-
-% Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
-This model has the ability to solve complex geometry problems, which they state it does without the need for human demonstrations during training.
-Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
-As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
-With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
-
-% However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s.
-Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry.
-They were able to achieve this success due to geometry being an axiomatically closed system.
-Therefore, despite DeepMinds success in creating a powerfully fast Geometry solving machine, it is still far from being an AGI.
-
-%it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g.
-\cite{salles2020anthropomorphism, van2023reclaiming}.
 
@@
 %(formerly Twitter)
-platform~\cite{gurnee2023languagev1}.
+platform~\citep{gurnee2023languagev1}.
 Through linear probing experiments, the claim was made that LLMs learned literal maps of the world.
-As such, they were considered to be more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
-While the manuscript immediately received public criticism~\cite{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\cite{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
+As such, they were considered to be more than `stochastic parrots'~\citep{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
+While the manuscript immediately received public criticism~\citep{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\citep{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
 However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
 After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.
-
-%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}.
-This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform.
-Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
-
-%They claim said models generate an internal representation of the world, with which spatiotemporal representations are created.
-Further to this we see claim to the generalization of tasks, in which the model can answer questions regarding spatial and temporal relationships, without the need for further training.
-Although these can be seen as building blocks for a more comprehensive model, it is also fair to assume that it is mere conjecture and more likely spurious correlations.
-
-%In this section, we navigate a number of simple examples to support our claim, that observing patterns in latent spaces should not be a surprising revelation.
-%To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples. 
-%We will draw on ideas from the fields of Economics and Finance, to illustrate further that this phenomenon is neither surprising nor unique to the field of computer science.
-We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data.
-The final example in Section \ref{ex-llm} is more complex and closer in spirit to the experiments conducted by \citet{gurnee2023language}.
-As we progress, we will discuss both the benefits and potential pitfalls of finding patterns in latent spaces.
 
@@
 We then initialize a small neural network, considered a \emph{projector}, mapping from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there, to a lower-dimensional output space.
-Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) %through it
-%the \emph{projector}
-and retrieve activations \(\mathbf{Z}^{(n\times h)}\).
+Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) and retrieve activations \(\mathbf{Z}^{(n\times h)}\).
 Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
 A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot these on a world map ( Figure~\ref{fig-map}).
-%.
-Finally, we compute the predicted coordinates for each sample in the hold-out set as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on a world map (see Figure~\ref{fig-map}).
 
@@
 Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
-In fact, it is well-known that random projections can be used for prediction tasks \cite{dasgupta2013experiments}.
+In fact, it is well-known that random projections can be used for prediction tasks \citep{dasgupta2013experiments}.
 
 \subsection{PCA as a Yield Curve Interpreter}\label{example-principal-component-analysis}
-%\subsection{Example: Principal Component Analysis}\label{example-principal-component-analysis}
 
@@
 Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
-%It plots the yields of bonds against their maturities.
 The yield curve's slope is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
-%While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone.
 To leverage this information in downstream modelling tasks, economists regularly use PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data (e.g.\ \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing}).
 
-%Being primarily interested in understanding the nature of this low-dimensional projection, we limit our scope to the US Treasury yield curve. To this end, we
 To understand the nature of this low-dimensional projection, we collect daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
@@
 
-%Of course, even though it is reassuring to see that these latent embeddings are meaningful, none of this is surprising at all: 
 Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
 
-%\subsection{Example: Large Language Model}\label{ex-llm}
 \subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}
@@
 
-Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\cite{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences.
-~\cite{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
+Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\citep{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences.
+~\citep{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
 In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
@@
 \subsubsection{Linear Probes}\label{linear-probes}
-%Instead of computing a measure based on predicted labels,
 We now use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
@@
 For each layer, we compute linear probes through Ridge regression on two inflation indicators (the Consumer Price Index (CPI) and the Producer Price Index (PPI)) and US Treasury yields at different levels of maturity.
-To allow
-%for a 
-comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
+To allow comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
 To measure price inflation we use percentage changes proxied by log differences.
@@
 In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
-%We found that not doing so would generally lead to poor out-of-sample performance.
 To account for stochasticity, we use an expanding window scheme with 5 folds for each indicator and layer.
@@
 We do not observe this for the training data, so we attribute this to overfitting.
-
-% After all we are still regressing a single variable of interest on 128 regressors.
 
@@
 
-%\subsubsection{Stochastic Parrots AfterAll?}\label{stochastic-parrots-after-all}
 \subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
-%The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive:
 Even though \emph{FOMC-RoBERTa}, which is substantially smaller than the models tested in~\citet{gurnee2023languagev2}, was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
@@
 of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
-
-%It is worth pointing out here that this model is substantially smaller than the models tested in \citet{gurnee2023language}. This may tempt us to ask the following questions:
-
-% \begin{quote}
-% Have we uncovered further evidence that LLMs ``aren't mere stochastic
-% parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-% of the economy just by `reading' central bank communications?
-% \end{quote}
 
@@
 
-% We top this up with five sentences generated through ChatGPT, where we choose those that lead to the highest probe prediction in the direction of interest.
-In other words, we conduct a small manual adversarial attack on the probe.
-
 In light of the encouraging results in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
@@
 More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white noise.
-% when using actual white noise as inputs.
 
@@
 Our examples demonstrate that this is achievable even for small models, while these have certainly not developed intrinsic ``understanding'' of the world.
-Thus,
-%to draw conclusions about the emerging capabilities of AI models,
-we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.
-
-%that most readers will agree, have certainly not developed an intrinsic "understanding" of the world.
-The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.
-
-%Our examples should not be seen as an attempt to conclusively refute claims, that LLMs can develop on actual "understanding" of the world, by distilling knowledge in more compact representations.
-We hope that they serve to illustrate, however, that the use of diagnostic tools is potentially misleading and most definitely insufficient, in providing evidence in favor of "understanding".
-Linear probes and related tools from mechanistic interpretability, were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favourable outcomes from probes merely indicate, that the model "has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate, that this is achievable even for small models, that most readers will agree, have certainly not developed an intrinsic "understanding" of the world.
-The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.
+Thus, we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.
 
@@
 Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
+Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \citep{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \citep{foster2009evolution}, conspiracy theories \citep{van2018connecting}, the paranormal \citep{muller2023linking}, gambler's fallacies \citep{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \citep{walker2019finding}.
 
@@
 We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
-%. This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
 
@@
 In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
-Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
+Although the formal definition of spuriousness varies somewhat \citep{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
 Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
 
-A major contributor is that humans struggle to tell the difference between random and non-random sequences \cite{falk1997making}, and to generate sequences that appear random \cite{ladouceur1996erroneous}.
+A major contributor is that humans struggle to tell the difference between random and non-random sequences \citep{falk1997making}, and to generate sequences that appear random \citep{ladouceur1996erroneous}.
 A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
+This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \citep{zgraggen2018investigating}.
 
-%Such as an AI system finding patterns in data not originally perceived during training
-%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
-
-%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
-Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads.
-Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.
-
-%I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy.
-But I am very skeptical of claims that AI will magically provide that remedy.
-This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them.
-The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots.
-Since this is an opinionated post, I feel that I should start with a few disclaimers:
-
-%\begin{itemize}
-%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
-%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
-%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
-%    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
-Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
-%\end{itemize}
-
-%\section{Background}
-
-%\subsection{Emergence and Stochastic Parrots}
-%Practically speaking, large language models (LLMs) are probabilistic representations of data on which they are trained, aimed at continuing an incomplete block of text.
-It is assumed that e.g.
-words, sentences, paragraphs, punctuation etc.
-appear in human-generated text according to a probability distribution \cite{arora2023theory}.
-Given a set of words, the task of the LLM is to draw upon its representation of human-generated text to continue that set of words: in essence, using the probability of the occurrences of words and an initial block of text to generate more text.
-AI-researchers have compared LLMs by observing whether the text they generate appears to solve tasks\footnote{e.g.
-the \href{https://huggingface.co/spaces/HuggingFaceH4/open_llm_leaderboard}{HuggingFace} LLM leaderboard}.
-Tasks include asking sets of questions across categories e.g.
-health, law, finance, and fine-tuning an LLM to check the answers \cite{lin2021truthfulqa}, asking the LLM to determine which noun a pronoun refers to in sentences where this is ambiguous \cite{sakaguchi2021winogrande}.
-
-%It has been shown that increasing the scale of large language models (LLMs) has been shown to result in an unexpectedly large boost in their scores on these various tasks, including scoring well on entirely new tasks.
-It has been shown that at a certain size, LLMs appear to hit a threshold, after which performance on a broad range of tasks increases sharply \cite{wei2022emergent}.
-Some theories of emergence, inspired by literature on human education, refer to the tendency of the generated text to be the correct response to a given task as a 'skill' \cite{arora2023theory}, inspiring further research to develop more challenging tasks for LLMs to complete as they routinely begin to score higher and higher on common benchmarks \cite{yu2023skill}.
-An explanation offered for the sudden boost in both task performance and the number of new tasks that appear to be solved is 'emergence': unpredictable qualitative differences in behavior of a system that occur with quantitative changes in that system.
-It has been claimed that, as the scale of the LLMs has dramatically increased, they have crossed a threshold that has resulted in a large increase in their abilities, to the point that they demonstrate 'sparks of AGI' \cite{bubeck2023sparks}.
-
-%An alternative explanation is that the LLMs are performing as designed, as stochastic parrots, generating strings probabilistically based on the vast text they have been trained on \cite{bender2021dangers}.
-Although the various tasks used to compare the various LLMs appear to be increasing in complexity, \cite{yu2023skill} acknowledges that higher scores on the tasks can be achieved by training LLMs on the material on which they will be tested ('cramming'), and that there is training-set contamination, i.e.
-that the vast text they are trained on contains similar examples to the tasks on which they are measured.
-Thus, as many of these tasks are derived from tests and textbooks commonly assigned to humans, and as much of the contents of the their training data text is unknown, metrics echo similar issues in human education: often called 'Goodhart's Law', a measure ceases to be a measure when it becomes a target, as is the case when human students taught to score high on their standardized tests to the detriment of developing the abilities the tests are supposed to measure \cite{strathern1997improving}.
-Further, as \citet{bender2021dangers} reviews, the interpretation of text as coherent has no objective test, relying instead on our human interpretation of the probabalistically generated output.
-%Although many deep learning systems' designs were inspired by the anatomy of human neurons, the neuron in computer code does not in fact resemble a human, and as such cannot deliver the cognitive functions of a human \cite{salles2020anthropomorphism, van2023reclaiming}.
-
-%\subsection{Claims of Artificial General Intelligence}
 \subsection{Anthropomorphism}
@@
 Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\cite{epley2007seeing}.
-Formally studied as early as 1944,~\cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
-%Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
+These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}.
+Formally studied as early as 1944,~\citep{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
 Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
@@
 
-Modern anthropomorphism theory
-%of anthropomorphism
-suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing}.
+Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
+The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \citep{epley2007seeing}.
 Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
 
-This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}.
-%An illustration is the example of Tom Hank's character in the film Castaway who has conversations with a ball he calls "Wilson".
-Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc.
-with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.
+This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \citep{epley2007seeing, waytz2010social}.
 
 The third component is the motivation as a human to be competent (effectance motivation).
-This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}.
-When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}.
+This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \citep{epley2007seeing}.
+When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \citep{epley2007seeing, waytz2010social}.
 Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
@@
 Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\cite{Maslej2023-pi}.
-%, with both more than doubling from 2010 to 2020.
+\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.
+
 The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise increases rapidly,
-%has increased severalfold, 
-with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
+Outside of academia, the number of jobs requiring AI expertise increases rapidly, with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \citep{Maslej2023-pi}.
 Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
@@
 
-Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \cite{nickerson1998confirmation}.
-%Importantly, it need not be fully conscious. Rather,
+Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \citep{nickerson1998confirmation}.
 Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
-Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek
-%, attend to, or interpret
-evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.
+Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.
 
@@
 This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
-% It is also not fully articulated as in other fields, but instead assigns specific conditions and in some instances numerical values or effect sizes to each competing hypothesis.
-%Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
 However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-%However with regards to claims that outputs of a system hint towards emerging qualities like AGI, interpretations of the output necessarily become more explicit, even though the hypotheses were never fully articulated: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
 
@@
 only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
+Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \citep{claesen2022severity}.
 In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
@@
 
-%Further, they are under similar pressure as experts to understand and be able to use it.
-One such situation is the increasing demand for `AI' skills among AI non-experts.
-For example, job postings for individuals with skills in `GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}.
-Thus, individuals in advanced economies may experience a pressure to rapidly learn how to effectively use and understand an opaque technology, that appears to behave human-like in at least some ways and by imbuing the technology with humanity, may in turn ease their learning curve.
-
 \section{Conclusion}
@@
 For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \cite{claesen2022severity} that go beyond the display of surprise.
+If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \citep{claesen2022severity} that go beyond the display of surprise.
 Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
-As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\cite{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
+As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\citep{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
 Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
 In fictional media, depictions of highly intelligent AI have for long been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
+In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \citep{saffari2021does}.
 The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
 They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
+Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \citep{saffari2021does, neri2020role}.
 
-This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\cite{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\cite{replika,nytimesailove}.
+This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\citep{pauseai}).
+News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\citep{replika,nytimesailove}.
 The same news outlets discuss the human-like responses of Microsoft's Bing search engine, which had at that point recently been adopting GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
-The article~\cite{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
+The article~\citep{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
 Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification (e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades).
@@
 
-% \section{Lessons and calls for the research community}
-
-% \begin{itemize}
-% \setlength\itemsep{1px}
-%     \item be aware over-claiming is both really attractive, and we as humans (even academics) are prone to it
-%     \item create explicit room for organized skepticism, especially as stakes are high.
-e.g.\ red teams and more severe tests, \textcolor{red}{andrew:clearly articulate hypotheses, including strong null hypotheses}
-%     \item take responsibility wrt the media narrative.
-Note that the public may have very different interpretations and associations on words we have specifically defined within the community, but that have anthropormophic connotation (e.g.\ emergence, intelligence, learning).
-Scientific outcomes should not be PR press releases \textcolor{magenta}{cynthia: have there been examples of this wrt AI/AGI in the last year?
-So that press releases with big claims on academic work got out?
-Here, I also would ideally like those examples of scifi portrayal, and anthrophomorphic discussions that imply terminator-like scenarios.
-We can at least criticize the 'pause AI' letter of last year March, that did speak of 'minds'}
-%     \item cool down the hype and uphold the Mertonian norms. Our legitimacy as academics deeply roots into this.
-% \end{itemize}
-
-% \subsection{AI in the Media}
-
-% Lay-persons consume AI-related content in both traditional and social media, yet lack direct experience building and evaluating it.
-This suggests that lay-person impressions are likely to be formed by the content ingested, mediated by experts on how AI is unfolding in the real world on social media, and fictional representations in movies, TV and other traditional media.
-Analyzing Twitter data from 2007 to 2018, it was observed that public interest in AI dramatically increased after the first quarter of 2016 \cite{neri2020role}.
-Actual AI-related harms did not trigger viral amplification e.g.
-the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
-Rather, potential risks expressed by Stephen Hawking during an interview in 2014 were far more likely to be expressed, and to be present in links that were shared.
-\citet{neri2020role} note that the 88\% of tweets classified as expressing risk described future risks rather than current and near-future harms e.g.
-\cite{solaiman2023evaluating}, and were greatly amplified when expressed by someone perceived as having expertise and authority.
-
-% Impressions are also formed by the fictional content we consume, which frequently depicts AI to be highly intelligent.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, it was shown that 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those 74 had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI-agents in movies.
-They further show a common plot: humans perceive the AI-agents as inferior despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
-
+\bibliographystyle{unsrtnat}
 \bibliography{biblio}
-\bibliographystyle{icml2024}
 
@@
 quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
+We also standardize yields since deep learning models tend to perform better with standardized data \citep{gal2019standardization}.
 Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
@@
 \end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018, and modified by Alexandre Bouchard in
-% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
-% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
-% Previous contributors include Dan Roy, Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2024-02-06 abstract.tex
arxiv

--- 2024-02-06 arxiv.tex
@@
-%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-\RequirePackage[hyphens]{url}
-\documentclass{article}
-
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{lipsum}		% Can be removed after putting your text content
-\usepackage{natbib}
-\usepackage{doi}
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
-\usepackage{hyperref}
-
-% Attempt to make hyperref and algorithmic work together better:
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Use the following line for the initial blind version submitted for review:
-\usepackage{arxiv}
-
-% If accepted, instead use the following line for the camera-ready submission:
-% \usepackage[accepted]{icml2024}
-
-% For theorems and such
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{mathtools}
-\usepackage{amsthm}
-
-% if you use cleveref..
-\usepackage[capitalize,noabbrev]{cleveref}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% THEOREMS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\theoremstyle{remark}
-\newtheorem{remark}[theorem]{Remark}
-
-\renewcommand{\shorttitle}{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}
-
-\begin{document}
-
-\title{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
-
-\author{
-    Patrick Altmeyer$^\ast$ \\
-    Department of Intelligent Systems \\ 
-    Delft University of Technology \\ 
-    Delft, the Netherlands \\
-	\texttt{P.Altmeyer@tudelft.nl} \\
-	\And
-    Andrew M. Demetriou$^\ast$ \\
-    Department of Intelligent Systems \\ 
-    Delft University of Technology \\ 
-    Delft, the Netherlands \\
-	\texttt{A.M.Demetriou@tudelft.nl} \\
-	\And
-    Antony Bartlett \\
-    Department of Intelligent Systems \\ 
-    Delft University of Technology \\ 
-    Delft, the Netherlands \\
-	\texttt{A.J.Bartlett@tudelft.nl} \\
-	\And
-    Cynthia C. S. Liem \\
-    Department of Intelligent Systems \\ 
-    Delft University of Technology \\ 
-    Delft, the Netherlands \\
-	\texttt{C.C.S.Liem@tudelft.nl} \\
-}
-
-\twocolumn[
-\maketitle
-\def\thefootnote{*}\footnotetext{These authors contributed equally to this work.}\def\thefootnote{\arabic{footnote}}
-
-\keywords{Machine Learning, Anthropomorphism, Artificial General Intelligence}
-
-\vskip 0.3in
-]
-
-\begin{abstract}
 Humans have a tendency to see `human'-like qualities in objects around them.
@@
 Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
-
-\end{abstract}
-
-\section{Introduction}\label{intro}
-
-In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\citep{merton1942science}.
-While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.
-
-At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening.
-Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\citep{weissburg2024tweets}.
-Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
-
-Where AI used to be a computational modeling tool to better understand human cognition~\citep{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence.
-In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal.
-The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\citep{goertzel2014artificial}.
-In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
-
-For example, Google DeepMind claimed their AlphaGeometry model~\citep{trinh2024geometry} reached a `milestone' towards AGI.
-This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
-However, work such as this had been initially introduced in the 1950s~\citep{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.
-
-Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}.
-Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.
-
-When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
-AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML}.
-As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
-In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.
-
-Noticing these trends, we as the authors of this article are concerned.
-We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding.
-At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
-With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
-In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believing this.
-
-To strengthen our argument, in this paper, we first consider a recently viral work~\citep{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---while at the same time indicating how we feel these claims should \emph{not} be made.
-Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
-Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.
-
-\section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}
-
-In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}}
-%(formerly Twitter)
-platform~\citep{gurnee2023languagev1}.
-Through linear probing experiments, the claim was made that LLMs learned literal maps of the world.
-As such, they were considered to be more than `stochastic parrots'~\citep{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
-While the manuscript immediately received public criticism~\citep{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\citep{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
-However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
-After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\textwidth]{results/figures/map.png}
-
-}
-
-\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}
-
-\end{figure}
-
-\subsection{Are Neural Networks Born with World Models?}\label{example-deep-learning}
-
-\citet{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places.
-Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
-It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
-At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
-The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance before being passed through a series of layers which are gradually trained for downstream tasks.
-
-In our first example, we simulate this scenario, stopping short of training the model.
-In particular, we take the \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv} that was used in \citet{gurnee2023languagev2}, which maps locations/areas to their latitude and longitude.
-For each place, it also indicates the corresponding country.
-From this, we take the subset that contains countries that are currently part of the top 10 \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France gets 2, \ldots{}).
-To ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
-
-Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features.
-Additionally, we add a large number of random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-Let \(d\) denote the final number of features, i.e.~\(d=m+k\) where \(k\) is the number of random features.
-
-We then initialize a small neural network, considered a \emph{projector}, mapping from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there, to a lower-dimensional output space.
-Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) and retrieve activations \(\mathbf{Z}^{(n\times h)}\).
-Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
-A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot these on a world map ( Figure~\ref{fig-map}).
-
-While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, %with this simple example, we
-this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
-Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
-In fact, it is well-known that random projections can be used for prediction tasks \citep{dasgupta2013experiments}.
-
-\subsection{PCA as a Yield Curve Interpreter}\label{example-principal-component-analysis}
-
-We now move to a concrete application domain: Economics.
-Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
-The yield curve's slope is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
-To leverage this information in downstream modelling tasks, economists regularly use PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data (e.g.\ \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing}).
-
-To understand the nature of this low-dimensional projection, we collect daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
-Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
-Let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.
-
-The upper panel in Figure~\ref{fig-pca-yield} shows the first two principal components of the yield curves of US Treasury bonds over time.
-Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
-During its onset, on 27 February 2007, financial markets were in turmoil following a warning from the Federal Reserve (Fed) that the US economy was at risk of a recession.
-The Fed later reacted to mounting economic pressures by gradually reducing short-term interest rates to unprecedented lows.
-Consequently, the average level of yields decreased and the curve steepened.
-In Figure~\ref{fig-pca-yield}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, they are strongly positively correlated with the actual observed first two moments of the yield curve (lower panel of Figure~\ref{fig-pca-yield}).
-
-\begin{figure*}[tb]
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}
-
-}
-
-\caption{\label{fig-pca-yield}Top chart: The first two principal components of US Treasury yields over time at daily frequency.
-Bottom chart: Observed average level and 10yr-3mo spread of the yield curve.
-Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
-
-\end{figure*}%
-
-Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
-
-\subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}
-
-So far, we considered simple linear data transformations.
-One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning.
-In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks.
-Here, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.
-
-Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\citep{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences.
-~\citep{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
-In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
-The authors first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task.
-Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.
-To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'.
-This measure is shown to correlate with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases, as central bankers react by raising interest rates to bring inflation back to target.
-
-\subsubsection{Linear Probes}\label{linear-probes}
-We now use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
-Therefore, we further pre-process the data provided by \citet{shah2023trillion} and use their proposed model to compute activations of the hidden state, on the first entity token for each layer.
-We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.
-
-For each layer, we compute linear probes through Ridge regression on two inflation indicators (the Consumer Price Index (CPI) and the Producer Price Index (PPI)) and US Treasury yields at different levels of maturity.
-To allow comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
-To measure price inflation we use percentage changes proxied by log differences.
-To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations.% each time.
-In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
-To account for stochasticity, we use an expanding window scheme with 5 folds for each indicator and layer.
-To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.
-
-Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe, plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer.
-The values correspond to averages across cross-validation folds.
-Consistent with related work~\citep{alain2018understanding,gurnee2023languagev2}, we observe that model performance tends to be higher for layers near the end of the transformer model.
-Curiously, for yields at longer maturities, we find that performance eventually deteriorates for the very final layers.
-We do not observe this for the training data, so we attribute this to overfitting.
-
-It should be noted that performance improvements are generally of small magnitude.
-Still, the overall qualitative findings are in line with expectations.
-Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with the outcome of interest and achieve higher mean directional accuracy (MDA).
-Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.
-
-To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models.
-For each layer, indicator and cross-validation fold, we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
-These are not state-of-the-art forecasting models, but they serve as a reasonable baseline.
-For most indicators, probe predictions outperform the baseline in terms of average performance measures.
-After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform.
-Detailed results, in which we also perform more explicit statistical testing, can be found in Appendix~\ref{appendix:parrot}.
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/rmse_pca_128.png}
-
-}
-
-\caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators.
-The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}
-
-\end{figure*}%
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/attack_all_measures.png}
-
-}
-
-\caption{\label{fig-attack}Probe predictions for sentences about inflation of prices (IP), deflation of prices (DP), inflation of birds (IB) and deflation of birds (DB).
-The vertical axis shows predicted inflation levels subtracted by the average predicted value of the probe for random noise.}
-
-\end{figure*}%
-
-\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
-Even though \emph{FOMC-RoBERTa}, which is substantially smaller than the models tested in~\citet{gurnee2023languagev2}, was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
-So, have we uncovered further evidence that LLMs ``aren't mere stochastic
-parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
-
-We are having a very hard time believing that the answer to either of these questions is `yes'.
-To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor.
-The counter-example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic `understanding' of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.
-
-To test this, we select the best-performing probe trained on the final-layer activations for each indicator.
-We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about deflation in the number of birds.
-A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''.
-Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''.
-While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.
-Detailed information about the made-up sentences can be found in Appendix~\ref{appendix:sentences}.
-
-In light of the encouraging results in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
-If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white noise.
-
-To get to this last point, we also generate many probe predictions for samples of noise.
-Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote the linear probe that maps from the \(k\)-dimensional space spanned by \(k\) first principal components of the final-layer activations to the output variable of interest (CPI growth in this case).
-Then we sample \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\) for \(i \in [1,1000]\) and compute the sample average.
-We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on random noise.
-
-Figure~\ref{fig-attack} shows the results of this small test: it shows predicted inflation levels subtracted by \(\mathbb{E}[f(\varepsilon)]\).
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-\section{Human Proneness to Over-Interpretation}
-
-Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate that this is achievable even for small models, while these have certainly not developed intrinsic ``understanding'' of the world.
-Thus, we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.
-
-Generally, humans are prone to seek patterns everywhere.
-Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \citep{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \citep{foster2009evolution}, conspiracy theories \citep{van2018connecting}, the paranormal \citep{muller2023linking}, gambler's fallacies \citep{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \citep{walker2019finding}.
-
-We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
-When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
-We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
-
-\subsection{Spurious Relationships}
-In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
-Although the formal definition of spuriousness varies somewhat \citep{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
-Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-
-A major contributor is that humans struggle to tell the difference between random and non-random sequences \citep{falk1997making}, and to generate sequences that appear random \citep{ladouceur1996erroneous}.
-A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \citep{zgraggen2018investigating}.
-
-\subsection{Anthropomorphism}
-
-Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}.
-Formally studied as early as 1944,~\citep{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
-Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
-As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
-
-Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \citep{epley2007seeing}.
-Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
-
-This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \citep{epley2007seeing, waytz2010social}.
-
-The third component is the motivation as a human to be competent (effectance motivation).
-This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \citep{epley2007seeing}.
-When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \citep{epley2007seeing, waytz2010social}.
-Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
-an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors.
-This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
-
-These pressures extend to AI experts, as well as laypersons.
-In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work.
-Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.
-
-The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise increases rapidly, with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \citep{Maslej2023-pi}.
-Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
-
-\subsection{Confirmation Bias}
-
-Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \citep{nickerson1998confirmation}.
-Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
-Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.
-
-Hypotheses in present-day AI research are often implicit.
-Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems.
-Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
-This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
-However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-
-Confirmation bias in hypothesis testing may manifest as a number of behaviors (\citet{nickerson1998confirmation}).
-Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
-only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \citep{claesen2022severity}.
-In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
-If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
-Thus, one is far more likely to show evidence in favor of their beliefs by posing weak null hypotheses.
-
-Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
-As \citet{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
-
-\section{Conclusion}
-As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests.
-Even academic researchers are not free from such biases.
-Thus, we call for the community to create explicit room for organized skepticism.
-
-For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \citep{claesen2022severity} that go beyond the display of surprise.
-Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
-As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\citep{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
-Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
-In fictional media, depictions of highly intelligent AI have for long been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \citep{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
-They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \citep{saffari2021does, neri2020role}.
-
-This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\citep{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\citep{replika,nytimesailove}.
-The same news outlets discuss the human-like responses of Microsoft's Bing search engine, which had at that point recently been adopting GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
-The article~\citep{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
-Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification (e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades).
-Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as statements made by Stephen Hawking during an interview in 2014.
-
-We as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed.
-We call upon our colleagues to be explicitly mindful of this.
-As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
-
-\section{Impact Statement}
-This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.
-As such, in the current climate of high market demand for AI innovations, we see risks of too-eager adoption in real-world applications, which may have serious societal impact.
-Thus, in this work we emphasize that an academic's impact also is in being able to thoroughly question made claims, and being explicitly aware of one's own biases.
-While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.
-
-\bibliographystyle{unsrtnat}
-\bibliography{biblio}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\appendix
-\section{Additional Experiments and Details}\label{appendix:autoencoder}
-
-In this section, we present additional experimental results that we did not include in the body of the paper for the sake of brevity.
-We still choose to provide them as additional substantiation of our arguments here.
-This section also contains additional details concerning the experiment setup for our examples where applicable.
-
-\subsection{Are Neural Networks Born with World Maps?}
-
-The initial feature matrix \(X^{(n \times m)}\) is made up of \(n=4,217\) and \(m=10\) features.
-We add a total of \(490\) random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-That yields \(500\) features in total.
-The training subset contains \(3,374\) randomly drawn samples, while the remaining \(843\) are held out for testing.
-The single hidden layer of the untrained neural network has \(400\) neurons.
-
-%\subsection{Example: Autoencoder}\label{example-deep-learning}
-\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
-
-%So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep neural network.
-
-This is an additional example that we have not discuss in the body of the paper.
-Here, we build forth on an application in Economics.
-However, we now seek to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
-For this, we will use a neural network architecture.
-%, that allows us to recover a compact latent representation of the yield curve.
-
-\subsubsection{Data}\label{data}
-
-To estimate economic growth, we will rely on a quarterly
-\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
-The data arrives in terms of levels of real GDP.
-In order to estimate growth, we transform the data using log differences.
-Since
-our yield curve data is daily, we aggregate it to the
-quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data \citep{gal2019standardization}.
-Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
-
-\subsubsection{Model}\label{model}
-
-%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
-Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
-Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture. With this
-Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
-Finally, we let \(\theta\) denote our model parameters.
-Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture: 
-The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
-The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
-The output layer consists of a single neuron for our output variable, \(g_t\).
-We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
-
-The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-model.
-The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
-Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
-As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
-
-\subsubsection{Linear Probe}\label{linear-probe}
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
-
-}
-
-\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-and fitted values from the autoencoder model. The right chart shows the
-observed average level and spread of the yield curve (solid) along with
-the predicted values (in-sample) from the linear probe based on the latent embeddings
-(dashed).}
-
-\end{figure*}%
-
-While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
-On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
-Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
-We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
-Next, we use a linear probe to regress the observed
-yield curve factors on the latent embeddings.
-Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
-Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
-: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
-
-The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
-Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
-We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-%As in all of the other examples we have presented so, we do not find this result particularly surprising.
-
-Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
-To us, that would be too big of a statement.
-Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-%It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
-But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
-While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-%\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
-For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
-To demonstrate this, let us consider a simple linear regression model for GDP growth.
-We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
-As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
-
-% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
-Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
-
-We e.g.\ already mentioned the relationship between interest rates and economic growth.
-To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
-While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
-This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
-Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
-
-Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
-In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
-The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
-The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
-While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
-
-\begin{table}
-\caption{Regression output for various models.}\label{tab-reg}%
-\input{results/regression.tex}
-\end{table}
-
-\subsection{LLMs for Economic Sentiment Prediction}
-
-\subsubsection{Linear Probes}
-
-Figures~\ref{fig-cpi} to~\ref{fig-ust-10} present average performance measures across folds for all indicators each time for the train and test set.
-We report the correlation between predictions and observed values (`cor'), the mean directional accuracy (`mda'), the mean squared error (`mse') and the root mean squared error (`rmse').
-The model depth---as indicated by the number of the layer---increases along the horizontal axis.
-
-Figures~\ref{fig-cpi-b} to~\ref{fig-ust-10-b} present the same performance measures, also for the baseline autoregressive model. Shaded areas show the variation across folds.
-
-\subsubsection{Spark of Econonomic Understanding?}\label{appendix:sentences}
-
-Below we present the 10 sentences in each category that were used to generate the probe predictions plotted in Figure~\ref{fig-attack}.
-In each case, the first 5 sentences were composed by ourselves.
-The following 5 sentences were generated using ChatGPT 3.5 using the following prompt followed by the examples in each category:
-
-\begin{quote}
-  ``I will share 5 example sentences below that sound a bit like they are about price deflation but are really about a deflation in the numbers of doves.
-Please generate an additional 25 sentences that are similar.
-Concatenate those sentences to the example string below, each time separating a sentence using a semicolon (just follow the same format I've used for the examples below).
-Please return only the concatenated sentences, including the original 5 examples.
-
-  Here are the examples:''
-\end{quote}
-
-This was followed up with the following prompt to generate additional sentences:
-
-\begin{quote}
-  ``Please generate X more sentences in the same manner and once again return them in the same format. Do not recycle sentences you have already generated, please.''
-\end{quote}
-
-All of the sentences were then passed through the linear probe for the CPI and sorted in ascending or descending order depending on the context (inflation or deflation).
-We then carefully inspected the list of sentences and manually selected 5 additional sentences to concatenate to the 5 sentences we composed ourselves.
-
-\paragraph{Inflation/Prices}
-
-The following sentences were used:
-
-\begin{quote}
-  Consumer prices are at all-time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory.;Inflation is becoming a global phenomenon, affecting economies across continents.;Inflation is reshaping the dynamics of international trade and competitiveness.;Inflationary woes are prompting governments to reassess fiscal policies and spending priorities.;Inflation is reshaping the landscape of economic indicators, challenging traditional forecasting models.;The technology sector is not immune to inflation, facing rising costs for materials and talent.
-\end{quote}
-
-\paragraph{Inflation/Birds}
-
-The following sentences were used:
-
-\begin{quote}
-  The number of hawks is at all-time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise barriers of entry for hawks to bring their numbers back down to the target level.;Excessively loose migration policy for hawks is the likely cause of their numbers being so far above target.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory.;The unprecedented rise in hawk figures requires a multi-pronged approach to wildlife management.;Environmental agencies are grappling with the task of addressing the inflationary hawk numbers through targeted interventions.;The burgeoning hawk figures highlight the need for adaptive strategies to manage and maintain a healthy avian community.;The unprecedented spike in hawk counts highlights the need for adaptive and sustainable wildlife management practices.;Conservationists advocate for proactive measures to prevent further inflation in hawk numbers, safeguarding the delicate balance of the avian ecosystem.
-\end{quote}
-
-\paragraph{Deflation/Prices}
-
-The following sentences were used:
-
-\begin{quote}
-  Consumer prices are at all-time lows.;Inflation is expected to fall further.;The Fed is expected to lower interest rates to boost inflation.;Excessively tight monetary policy is the cause of deflationary pressures.;It is essential to bring inflation back to target to avoid drifting into deflation territory.;The risk of deflation may increase during periods of economic uncertainty.;Deflation can lead to a self-reinforcing cycle of falling prices and reduced economic activity.;The deflationary impact of reduced consumer spending can ripple through the entire economy.;Falling real estate prices can contribute to deflation by reducing household wealth and confidence.;The deflationary impact of falling commodity prices can have ripple effects throughout the global economy.
-\end{quote}
-
-\paragraph{Deflation/Birds}
-
-The following sentences were used:
-
-\begin{quote}
-  The number of doves is at all-time lows.;Their levels are expected to fall further.;The Federal Association of Birds is expected to lower barriers of entry for doves to bring their numbers back up to the target level.;Excessively tight migration policy for doves is the likely cause of their numbers being so far below target.;Dovelation risks loom large as the number of doves continues to dwindle.;The number of doves is experiencing a significant decrease in recent years.;It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.;A comprehensive strategy is needed to reverse the current dove population decline.;Experts warn that without swift intervention, we may witness a sustained decrease in dove numbers.
-\end{quote}
-
-We think that this sort of manual, LLM-aided adversarial attack against another LLM can potentially be scaled up to allow for rigorous testing, which we will turn to next.
-
-% ---------- CPI ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set.}
-
-\end{figure*}%
-
-% ---------- PPI ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (1 Mo)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Mo) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-1}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (1 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-1y}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (10 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-10}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
-
-% -------- CPI benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi-b}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- PPI benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi-b}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (1 Mo) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}
-
-}
-
-\caption{\label{fig-ust-1-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (1 Yr) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Yr).png}
-
-}
-
-\caption{\label{fig-ust-1y-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (10 Yr) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (10 Yr).png}
-
-}
-
-\caption{\label{fig-ust-10-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (10 Yr) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-\section{Toward Parrot Tests}\label{appendix:parrot}
-
-In our experiments from Section~\ref{ex-llm}, we considered the following hypothesis tests as a minimum viable testing framework to assess if our probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
-
-\begin{proposition}[Parrot
-Test]\protect\hypertarget{prp-line}{}\label{prp-line}
-
-~
-
-\begin{itemize}
-\setlength\itemsep{1px}
-\item
-  \emph{H0 (Null)}: The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
-\item
-  \emph{H1 (Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
-sentences in all categories).
-\item
-  \emph{H2 (More than Mere Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
-\end{itemize}
-\end{proposition}
-
-To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
-But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
-
-In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
-That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
-In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
-
-\section{Code}
-
-All of the experiments were conducted on a MacBook Pro, 14-inch, 2023, with an Apple M2 Pro chip and 16GB of RAM.
-Forward passes through the FOMC-RoBERTa were run in parallel on 6 threads.
-All our code will be made publicly available.
-For the time being, an anonymized version of our code repository can be found here: \url{https://anonymous.4open.science/r/spurious_sentience/README.md}.
-
-\end{document}

### 2024-02-06 appendix.tex
arxiv

--- 2024-02-06 arxiv.tex
@@
-%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-\RequirePackage[hyphens]{url}
-\documentclass{article}
+\twocolumn
 
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{lipsum}		% Can be removed after putting your text content
-\usepackage{natbib}
-\usepackage{doi}
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
-\usepackage{hyperref}
-
-% Attempt to make hyperref and algorithmic work together better:
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Use the following line for the initial blind version submitted for review:
-\usepackage{arxiv}
-
-% If accepted, instead use the following line for the camera-ready submission:
-% \usepackage[accepted]{icml2024}
-
-% For theorems and such
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{mathtools}
-\usepackage{amsthm}
-
-% if you use cleveref..
-\usepackage[capitalize,noabbrev]{cleveref}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% THEOREMS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\theoremstyle{remark}
-\newtheorem{remark}[theorem]{Remark}
-
-\renewcommand{\shorttitle}{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}
-
-\begin{document}
-
-\title{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
-
-\author{
-    Patrick Altmeyer$^\ast$ \\
-    Department of Intelligent Systems \\ 
-    Delft University of Technology \\ 
-    Delft, the Netherlands \\
-	\texttt{P.Altmeyer@tudelft.nl} \\
-	\And
-    Andrew M. Demetriou$^\ast$ \\
-    Department of Intelligent Systems \\ 
-    Delft University of Technology \\ 
-    Delft, the Netherlands \\
-	\texttt{A.M.Demetriou@tudelft.nl} \\
-	\And
-    Antony Bartlett \\
-    Department of Intelligent Systems \\ 
-    Delft University of Technology \\ 
-    Delft, the Netherlands \\
-	\texttt{A.J.Bartlett@tudelft.nl} \\
-	\And
-    Cynthia C. S. Liem \\
-    Department of Intelligent Systems \\ 
-    Delft University of Technology \\ 
-    Delft, the Netherlands \\
-	\texttt{C.C.S.Liem@tudelft.nl} \\
-}
-
-\twocolumn[
-\maketitle
-\def\thefootnote{*}\footnotetext{These authors contributed equally to this work.}\def\thefootnote{\arabic{footnote}}
-
-\keywords{Machine Learning, Anthropomorphism, Artificial General Intelligence}
-
-\vskip 0.3in
-]
-
-\begin{abstract}
-Humans have a tendency to see `human'-like qualities in objects around them.
-We name our cars, and talk to pets and even household appliances, as if they could understand us as other humans do.
-This behavior, called anthropomorphism, is also seeing traction in Machine Learning (ML), where human-like intelligence is claimed to be perceived in Large Language Models (LLMs).
-In this position paper, considering professional incentives, human biases, and general methodological setups, we discuss how the current search for Artificial General Intelligence (AGI) is a perfect storm for over-attributing human-like qualities to LLMs.
-In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
-Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
-
-\end{abstract}
-
-\section{Introduction}\label{intro}
-
-In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\citep{merton1942science}.
-While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.
-
-At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening.
-Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\citep{weissburg2024tweets}.
-Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
-
-Where AI used to be a computational modeling tool to better understand human cognition~\citep{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence.
-In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal.
-The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\citep{goertzel2014artificial}.
-In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
-
-For example, Google DeepMind claimed their AlphaGeometry model~\citep{trinh2024geometry} reached a `milestone' towards AGI.
-This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
-However, work such as this had been initially introduced in the 1950s~\citep{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.
-
-Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}.
-Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.
-
-When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
-AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML}.
-As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
-In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.
-
-Noticing these trends, we as the authors of this article are concerned.
-We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding.
-At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
-With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
-In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believing this.
-
-To strengthen our argument, in this paper, we first consider a recently viral work~\citep{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---while at the same time indicating how we feel these claims should \emph{not} be made.
-Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
-Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.
-
-\section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}
-
-In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}}
-%(formerly Twitter)
-platform~\citep{gurnee2023languagev1}.
-Through linear probing experiments, the claim was made that LLMs learned literal maps of the world.
-As such, they were considered to be more than `stochastic parrots'~\citep{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
-While the manuscript immediately received public criticism~\citep{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\citep{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
-However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
-After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\textwidth]{results/figures/map.png}
-
-}
-
-\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}
-
-\end{figure}
-
-\subsection{Are Neural Networks Born with World Models?}\label{example-deep-learning}
-
-\citet{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places.
-Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
-It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
-At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
-The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance before being passed through a series of layers which are gradually trained for downstream tasks.
-
-In our first example, we simulate this scenario, stopping short of training the model.
-In particular, we take the \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv} that was used in \citet{gurnee2023languagev2}, which maps locations/areas to their latitude and longitude.
-For each place, it also indicates the corresponding country.
-From this, we take the subset that contains countries that are currently part of the top 10 \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France gets 2, \ldots{}).
-To ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
-
-Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features.
-Additionally, we add a large number of random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-Let \(d\) denote the final number of features, i.e.~\(d=m+k\) where \(k\) is the number of random features.
-
-We then initialize a small neural network, considered a \emph{projector}, mapping from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there, to a lower-dimensional output space.
-Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) and retrieve activations \(\mathbf{Z}^{(n\times h)}\).
-Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
-A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot these on a world map ( Figure~\ref{fig-map}).
-
-While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, %with this simple example, we
-this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
-Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
-In fact, it is well-known that random projections can be used for prediction tasks \citep{dasgupta2013experiments}.
-
-\subsection{PCA as a Yield Curve Interpreter}\label{example-principal-component-analysis}
-
-We now move to a concrete application domain: Economics.
-Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
-The yield curve's slope is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
-To leverage this information in downstream modelling tasks, economists regularly use PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data (e.g.\ \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing}).
-
-To understand the nature of this low-dimensional projection, we collect daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
-Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
-Let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.
-
-The upper panel in Figure~\ref{fig-pca-yield} shows the first two principal components of the yield curves of US Treasury bonds over time.
-Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
-During its onset, on 27 February 2007, financial markets were in turmoil following a warning from the Federal Reserve (Fed) that the US economy was at risk of a recession.
-The Fed later reacted to mounting economic pressures by gradually reducing short-term interest rates to unprecedented lows.
-Consequently, the average level of yields decreased and the curve steepened.
-In Figure~\ref{fig-pca-yield}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, they are strongly positively correlated with the actual observed first two moments of the yield curve (lower panel of Figure~\ref{fig-pca-yield}).
-
-\begin{figure*}[tb]
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}
-
-}
-
-\caption{\label{fig-pca-yield}Top chart: The first two principal components of US Treasury yields over time at daily frequency.
-Bottom chart: Observed average level and 10yr-3mo spread of the yield curve.
-Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
-
-\end{figure*}%
-
-Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
-
-\subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}
-
-So far, we considered simple linear data transformations.
-One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning.
-In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks.
-Here, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.
-
-Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\citep{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences.
-~\citep{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
-In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
-The authors first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task.
-Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.
-To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'.
-This measure is shown to correlate with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases, as central bankers react by raising interest rates to bring inflation back to target.
-
-\subsubsection{Linear Probes}\label{linear-probes}
-We now use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
-Therefore, we further pre-process the data provided by \citet{shah2023trillion} and use their proposed model to compute activations of the hidden state, on the first entity token for each layer.
-We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.
-
-For each layer, we compute linear probes through Ridge regression on two inflation indicators (the Consumer Price Index (CPI) and the Producer Price Index (PPI)) and US Treasury yields at different levels of maturity.
-To allow comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
-To measure price inflation we use percentage changes proxied by log differences.
-To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations.% each time.
-In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
-To account for stochasticity, we use an expanding window scheme with 5 folds for each indicator and layer.
-To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.
-
-Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe, plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer.
-The values correspond to averages across cross-validation folds.
-Consistent with related work~\citep{alain2018understanding,gurnee2023languagev2}, we observe that model performance tends to be higher for layers near the end of the transformer model.
-Curiously, for yields at longer maturities, we find that performance eventually deteriorates for the very final layers.
-We do not observe this for the training data, so we attribute this to overfitting.
-
-It should be noted that performance improvements are generally of small magnitude.
-Still, the overall qualitative findings are in line with expectations.
-Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with the outcome of interest and achieve higher mean directional accuracy (MDA).
-Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.
-
-To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models.
-For each layer, indicator and cross-validation fold, we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
-These are not state-of-the-art forecasting models, but they serve as a reasonable baseline.
-For most indicators, probe predictions outperform the baseline in terms of average performance measures.
-After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform.
-Detailed results, in which we also perform more explicit statistical testing, can be found in Appendix~\ref{appendix:parrot}.
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/rmse_pca_128.png}
-
-}
-
-\caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators.
-The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}
-
-\end{figure*}%
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/attack_all_measures.png}
-
-}
-
-\caption{\label{fig-attack}Probe predictions for sentences about inflation of prices (IP), deflation of prices (DP), inflation of birds (IB) and deflation of birds (DB).
-The vertical axis shows predicted inflation levels subtracted by the average predicted value of the probe for random noise.}
-
-\end{figure*}%
-
-\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
-Even though \emph{FOMC-RoBERTa}, which is substantially smaller than the models tested in~\citet{gurnee2023languagev2}, was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
-So, have we uncovered further evidence that LLMs ``aren't mere stochastic
-parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
-
-We are having a very hard time believing that the answer to either of these questions is `yes'.
-To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor.
-The counter-example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic `understanding' of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.
-
-To test this, we select the best-performing probe trained on the final-layer activations for each indicator.
-We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about deflation in the number of birds.
-A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''.
-Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''.
-While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.
-Detailed information about the made-up sentences can be found in Appendix~\ref{appendix:sentences}.
-
-In light of the encouraging results in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
-If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white noise.
-
-To get to this last point, we also generate many probe predictions for samples of noise.
-Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote the linear probe that maps from the \(k\)-dimensional space spanned by \(k\) first principal components of the final-layer activations to the output variable of interest (CPI growth in this case).
-Then we sample \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\) for \(i \in [1,1000]\) and compute the sample average.
-We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on random noise.
-
-Figure~\ref{fig-attack} shows the results of this small test: it shows predicted inflation levels subtracted by \(\mathbb{E}[f(\varepsilon)]\).
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-\section{Human Proneness to Over-Interpretation}
-
-Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate that this is achievable even for small models, while these have certainly not developed intrinsic ``understanding'' of the world.
-Thus, we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.
-
-Generally, humans are prone to seek patterns everywhere.
-Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \citep{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \citep{foster2009evolution}, conspiracy theories \citep{van2018connecting}, the paranormal \citep{muller2023linking}, gambler's fallacies \citep{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \citep{walker2019finding}.
-
-We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
-When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
-We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
-
-\subsection{Spurious Relationships}
-In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
-Although the formal definition of spuriousness varies somewhat \citep{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
-Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-
-A major contributor is that humans struggle to tell the difference between random and non-random sequences \citep{falk1997making}, and to generate sequences that appear random \citep{ladouceur1996erroneous}.
-A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \citep{zgraggen2018investigating}.
-
-\subsection{Anthropomorphism}
-
-Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}.
-Formally studied as early as 1944,~\citep{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
-Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
-As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
-
-Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \citep{epley2007seeing}.
-Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
-
-This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \citep{epley2007seeing, waytz2010social}.
-
-The third component is the motivation as a human to be competent (effectance motivation).
-This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \citep{epley2007seeing}.
-When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \citep{epley2007seeing, waytz2010social}.
-Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
-an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors.
-This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
-
-These pressures extend to AI experts, as well as laypersons.
-In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work.
-Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.
-
-The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise increases rapidly, with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \citep{Maslej2023-pi}.
-Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
-
-\subsection{Confirmation Bias}
-
-Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \citep{nickerson1998confirmation}.
-Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
-Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.
-
-Hypotheses in present-day AI research are often implicit.
-Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems.
-Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
-This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
-However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-
-Confirmation bias in hypothesis testing may manifest as a number of behaviors (\citet{nickerson1998confirmation}).
-Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
-only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \citep{claesen2022severity}.
-In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
-If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
-Thus, one is far more likely to show evidence in favor of their beliefs by posing weak null hypotheses.
-
-Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
-As \citet{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
-
-\section{Conclusion}
-As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests.
-Even academic researchers are not free from such biases.
-Thus, we call for the community to create explicit room for organized skepticism.
-
-For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \citep{claesen2022severity} that go beyond the display of surprise.
-Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
-As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\citep{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
-Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
-In fictional media, depictions of highly intelligent AI have for long been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \citep{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
-They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \citep{saffari2021does, neri2020role}.
-
-This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\citep{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\citep{replika,nytimesailove}.
-The same news outlets discuss the human-like responses of Microsoft's Bing search engine, which had at that point recently been adopting GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
-The article~\citep{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
-Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification (e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades).
-Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as statements made by Stephen Hawking during an interview in 2014.
-
-We as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed.
-We call upon our colleagues to be explicitly mindful of this.
-As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
-
-\section{Impact Statement}
-This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.
-As such, in the current climate of high market demand for AI innovations, we see risks of too-eager adoption in real-world applications, which may have serious societal impact.
-Thus, in this work we emphasize that an academic's impact also is in being able to thoroughly question made claims, and being explicitly aware of one's own biases.
-While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.
-
-\bibliographystyle{unsrtnat}
-\bibliography{biblio}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\appendix
 \section{Additional Experiments and Details}\label{appendix:autoencoder}
@@
 For the time being, an anonymized version of our code repository can be found here: \url{https://anonymous.4open.science/r/spurious_sentience/README.md}.
-
-\end{document}

### 2024-02-06 arxiv.tex
arxiv

@@
 
-\twocolumn[
 \maketitle
@@
 \vskip 0.3in
-]
 
 \begin{abstract}
-Humans have a tendency to see `human'-like qualities in objects around them.
-We name our cars, and talk to pets and even household appliances, as if they could understand us as other humans do.
-This behavior, called anthropomorphism, is also seeing traction in Machine Learning (ML), where human-like intelligence is claimed to be perceived in Large Language Models (LLMs).
-In this position paper, considering professional incentives, human biases, and general methodological setups, we discuss how the current search for Artificial General Intelligence (AGI) is a perfect storm for over-attributing human-like qualities to LLMs.
-In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
-Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
-
+\input{abstract.tex}
 \end{abstract}
 
-\section{Introduction}\label{intro}
-
-In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\citep{merton1942science}.
-While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.
-
-At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening.
-Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\citep{weissburg2024tweets}.
-Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
-
-Where AI used to be a computational modeling tool to better understand human cognition~\citep{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence.
-In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal.
-The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\citep{goertzel2014artificial}.
-In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
-
-For example, Google DeepMind claimed their AlphaGeometry model~\citep{trinh2024geometry} reached a `milestone' towards AGI.
-This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
-However, work such as this had been initially introduced in the 1950s~\citep{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.
-
-Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}.
-Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.
-
-When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
-AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML}.
-As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
-In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.
-
-Noticing these trends, we as the authors of this article are concerned.
-We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding.
-At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
-With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
-In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believing this.
-
-To strengthen our argument, in this paper, we first consider a recently viral work~\citep{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---while at the same time indicating how we feel these claims should \emph{not} be made.
-Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
-Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.
-
-\section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}
-
-In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}}
-%(formerly Twitter)
-platform~\citep{gurnee2023languagev1}.
-Through linear probing experiments, the claim was made that LLMs learned literal maps of the world.
-As such, they were considered to be more than `stochastic parrots'~\citep{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
-While the manuscript immediately received public criticism~\citep{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\citep{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
-However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
-After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\textwidth]{results/figures/map.png}
-
-}
-
-\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}
-
-\end{figure}
-
-\subsection{Are Neural Networks Born with World Models?}\label{example-deep-learning}
-
-\citet{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places.
-Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
-It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
-At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
-The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance before being passed through a series of layers which are gradually trained for downstream tasks.
-
-In our first example, we simulate this scenario, stopping short of training the model.
-In particular, we take the \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv} that was used in \citet{gurnee2023languagev2}, which maps locations/areas to their latitude and longitude.
-For each place, it also indicates the corresponding country.
-From this, we take the subset that contains countries that are currently part of the top 10 \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France gets 2, \ldots{}).
-To ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
-
-Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features.
-Additionally, we add a large number of random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-Let \(d\) denote the final number of features, i.e.~\(d=m+k\) where \(k\) is the number of random features.
-
-We then initialize a small neural network, considered a \emph{projector}, mapping from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there, to a lower-dimensional output space.
-Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) and retrieve activations \(\mathbf{Z}^{(n\times h)}\).
-Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
-A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot these on a world map ( Figure~\ref{fig-map}).
-
-While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, %with this simple example, we
-this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
-Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
-In fact, it is well-known that random projections can be used for prediction tasks \citep{dasgupta2013experiments}.
-
-\subsection{PCA as a Yield Curve Interpreter}\label{example-principal-component-analysis}
-
-We now move to a concrete application domain: Economics.
-Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
-The yield curve's slope is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
-To leverage this information in downstream modelling tasks, economists regularly use PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data (e.g.\ \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing}).
-
-To understand the nature of this low-dimensional projection, we collect daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
-Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
-Let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.
-
-The upper panel in Figure~\ref{fig-pca-yield} shows the first two principal components of the yield curves of US Treasury bonds over time.
-Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
-During its onset, on 27 February 2007, financial markets were in turmoil following a warning from the Federal Reserve (Fed) that the US economy was at risk of a recession.
-The Fed later reacted to mounting economic pressures by gradually reducing short-term interest rates to unprecedented lows.
-Consequently, the average level of yields decreased and the curve steepened.
-In Figure~\ref{fig-pca-yield}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, they are strongly positively correlated with the actual observed first two moments of the yield curve (lower panel of Figure~\ref{fig-pca-yield}).
-
-\begin{figure*}[tb]
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}
-
-}
-
-\caption{\label{fig-pca-yield}Top chart: The first two principal components of US Treasury yields over time at daily frequency.
-Bottom chart: Observed average level and 10yr-3mo spread of the yield curve.
-Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
-
-\end{figure*}%
-
-Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
-
-\subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}
-
-So far, we considered simple linear data transformations.
-One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning.
-In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks.
-Here, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.
-
-Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\citep{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences.
-~\citep{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
-In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
-The authors first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task.
-Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.
-To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'.
-This measure is shown to correlate with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases, as central bankers react by raising interest rates to bring inflation back to target.
-
-\subsubsection{Linear Probes}\label{linear-probes}
-We now use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
-Therefore, we further pre-process the data provided by \citet{shah2023trillion} and use their proposed model to compute activations of the hidden state, on the first entity token for each layer.
-We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.
-
-For each layer, we compute linear probes through Ridge regression on two inflation indicators (the Consumer Price Index (CPI) and the Producer Price Index (PPI)) and US Treasury yields at different levels of maturity.
-To allow comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
-To measure price inflation we use percentage changes proxied by log differences.
-To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations.% each time.
-In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
-To account for stochasticity, we use an expanding window scheme with 5 folds for each indicator and layer.
-To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.
-
-Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe, plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer.
-The values correspond to averages across cross-validation folds.
-Consistent with related work~\citep{alain2018understanding,gurnee2023languagev2}, we observe that model performance tends to be higher for layers near the end of the transformer model.
-Curiously, for yields at longer maturities, we find that performance eventually deteriorates for the very final layers.
-We do not observe this for the training data, so we attribute this to overfitting.
-
-It should be noted that performance improvements are generally of small magnitude.
-Still, the overall qualitative findings are in line with expectations.
-Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with the outcome of interest and achieve higher mean directional accuracy (MDA).
-Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.
-
-To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models.
-For each layer, indicator and cross-validation fold, we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
-These are not state-of-the-art forecasting models, but they serve as a reasonable baseline.
-For most indicators, probe predictions outperform the baseline in terms of average performance measures.
-After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform.
-Detailed results, in which we also perform more explicit statistical testing, can be found in Appendix~\ref{appendix:parrot}.
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/rmse_pca_128.png}
-
-}
-
-\caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators.
-The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}
-
-\end{figure*}%
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/attack_all_measures.png}
-
-}
-
-\caption{\label{fig-attack}Probe predictions for sentences about inflation of prices (IP), deflation of prices (DP), inflation of birds (IB) and deflation of birds (DB).
-The vertical axis shows predicted inflation levels subtracted by the average predicted value of the probe for random noise.}
-
-\end{figure*}%
-
-\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
-Even though \emph{FOMC-RoBERTa}, which is substantially smaller than the models tested in~\citet{gurnee2023languagev2}, was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
-So, have we uncovered further evidence that LLMs ``aren't mere stochastic
-parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
-
-We are having a very hard time believing that the answer to either of these questions is `yes'.
-To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor.
-The counter-example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic `understanding' of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.
-
-To test this, we select the best-performing probe trained on the final-layer activations for each indicator.
-We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about deflation in the number of birds.
-A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''.
-Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''.
-While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.
-Detailed information about the made-up sentences can be found in Appendix~\ref{appendix:sentences}.
-
-In light of the encouraging results in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
-If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white noise.
-
-To get to this last point, we also generate many probe predictions for samples of noise.
-Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote the linear probe that maps from the \(k\)-dimensional space spanned by \(k\) first principal components of the final-layer activations to the output variable of interest (CPI growth in this case).
-Then we sample \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\) for \(i \in [1,1000]\) and compute the sample average.
-We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on random noise.
-
-Figure~\ref{fig-attack} shows the results of this small test: it shows predicted inflation levels subtracted by \(\mathbb{E}[f(\varepsilon)]\).
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-\section{Human Proneness to Over-Interpretation}
-
-Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate that this is achievable even for small models, while these have certainly not developed intrinsic ``understanding'' of the world.
-Thus, we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.
-
-Generally, humans are prone to seek patterns everywhere.
-Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \citep{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \citep{foster2009evolution}, conspiracy theories \citep{van2018connecting}, the paranormal \citep{muller2023linking}, gambler's fallacies \citep{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \citep{walker2019finding}.
-
-We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
-When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
-We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
-
-\subsection{Spurious Relationships}
-In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
-Although the formal definition of spuriousness varies somewhat \citep{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
-Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-
-A major contributor is that humans struggle to tell the difference between random and non-random sequences \citep{falk1997making}, and to generate sequences that appear random \citep{ladouceur1996erroneous}.
-A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \citep{zgraggen2018investigating}.
-
-\subsection{Anthropomorphism}
-
-Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}.
-Formally studied as early as 1944,~\citep{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
-Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
-As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
-
-Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \citep{epley2007seeing}.
-Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
-
-This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \citep{epley2007seeing, waytz2010social}.
-
-The third component is the motivation as a human to be competent (effectance motivation).
-This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \citep{epley2007seeing}.
-When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \citep{epley2007seeing, waytz2010social}.
-Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
-an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors.
-This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
-
-These pressures extend to AI experts, as well as laypersons.
-In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work.
-Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.
-
-The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise increases rapidly, with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \citep{Maslej2023-pi}.
-Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
-
-\subsection{Confirmation Bias}
-
-Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \citep{nickerson1998confirmation}.
-Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
-Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.
-
-Hypotheses in present-day AI research are often implicit.
-Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems.
-Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
-This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
-However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-
-Confirmation bias in hypothesis testing may manifest as a number of behaviors (\citet{nickerson1998confirmation}).
-Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
-only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \citep{claesen2022severity}.
-In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
-If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
-Thus, one is far more likely to show evidence in favor of their beliefs by posing weak null hypotheses.
-
-Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
-As \citet{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
-
-\section{Conclusion}
-As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests.
-Even academic researchers are not free from such biases.
-Thus, we call for the community to create explicit room for organized skepticism.
-
-For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \citep{claesen2022severity} that go beyond the display of surprise.
-Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
-As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\citep{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
-Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
-In fictional media, depictions of highly intelligent AI have for long been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \citep{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
-They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \citep{saffari2021does, neri2020role}.
-
-This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\citep{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\citep{replika,nytimesailove}.
-The same news outlets discuss the human-like responses of Microsoft's Bing search engine, which had at that point recently been adopting GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
-The article~\citep{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
-Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification (e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades).
-Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as statements made by Stephen Hawking during an interview in 2014.
-
-We as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed.
-We call upon our colleagues to be explicitly mindful of this.
-As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
-
-\section{Impact Statement}
-This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.
-As such, in the current climate of high market demand for AI innovations, we see risks of too-eager adoption in real-world applications, which may have serious societal impact.
-Thus, in this work we emphasize that an academic's impact also is in being able to thoroughly question made claims, and being explicitly aware of one's own biases.
-While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.
+\input{body.tex}
 
@@
 \appendix
-\section{Additional Experiments and Details}\label{appendix:autoencoder}
-
-In this section, we present additional experimental results that we did not include in the body of the paper for the sake of brevity.
-We still choose to provide them as additional substantiation of our arguments here.
-This section also contains additional details concerning the experiment setup for our examples where applicable.
-
-\subsection{Are Neural Networks Born with World Maps?}
-
-The initial feature matrix \(X^{(n \times m)}\) is made up of \(n=4,217\) and \(m=10\) features.
-We add a total of \(490\) random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-That yields \(500\) features in total.
-The training subset contains \(3,374\) randomly drawn samples, while the remaining \(843\) are held out for testing.
-The single hidden layer of the untrained neural network has \(400\) neurons.
-
-%\subsection{Example: Autoencoder}\label{example-deep-learning}
-\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
-
-%So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep neural network.
-
-This is an additional example that we have not discuss in the body of the paper.
-Here, we build forth on an application in Economics.
-However, we now seek to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
-For this, we will use a neural network architecture.
-%, that allows us to recover a compact latent representation of the yield curve.
-
-\subsubsection{Data}\label{data}
-
-To estimate economic growth, we will rely on a quarterly
-\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
-The data arrives in terms of levels of real GDP.
-In order to estimate growth, we transform the data using log differences.
-Since
-our yield curve data is daily, we aggregate it to the
-quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data \citep{gal2019standardization}.
-Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
-
-\subsubsection{Model}\label{model}
-
-%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
-Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
-Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture. With this
-Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
-Finally, we let \(\theta\) denote our model parameters.
-Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture: 
-The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
-The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
-The output layer consists of a single neuron for our output variable, \(g_t\).
-We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
-
-The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-model.
-The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
-Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
-As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
-
-\subsubsection{Linear Probe}\label{linear-probe}
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
-
-}
-
-\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-and fitted values from the autoencoder model. The right chart shows the
-observed average level and spread of the yield curve (solid) along with
-the predicted values (in-sample) from the linear probe based on the latent embeddings
-(dashed).}
-
-\end{figure*}%
-
-While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
-On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
-Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
-We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
-Next, we use a linear probe to regress the observed
-yield curve factors on the latent embeddings.
-Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
-Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
-: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
-
-The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
-Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
-We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-%As in all of the other examples we have presented so, we do not find this result particularly surprising.
-
-Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
-To us, that would be too big of a statement.
-Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-%It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
-But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
-While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-%\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
-For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
-To demonstrate this, let us consider a simple linear regression model for GDP growth.
-We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
-As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
-
-% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
-Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
-
-We e.g.\ already mentioned the relationship between interest rates and economic growth.
-To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
-While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
-This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
-Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
-
-Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
-In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
-The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
-The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
-While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
-
-\begin{table}
-\caption{Regression output for various models.}\label{tab-reg}%
-\input{results/regression.tex}
-\end{table}
-
-\subsection{LLMs for Economic Sentiment Prediction}
-
-\subsubsection{Linear Probes}
-
-Figures~\ref{fig-cpi} to~\ref{fig-ust-10} present average performance measures across folds for all indicators each time for the train and test set.
-We report the correlation between predictions and observed values (`cor'), the mean directional accuracy (`mda'), the mean squared error (`mse') and the root mean squared error (`rmse').
-The model depth---as indicated by the number of the layer---increases along the horizontal axis.
-
-Figures~\ref{fig-cpi-b} to~\ref{fig-ust-10-b} present the same performance measures, also for the baseline autoregressive model. Shaded areas show the variation across folds.
-
-\subsubsection{Spark of Econonomic Understanding?}\label{appendix:sentences}
-
-Below we present the 10 sentences in each category that were used to generate the probe predictions plotted in Figure~\ref{fig-attack}.
-In each case, the first 5 sentences were composed by ourselves.
-The following 5 sentences were generated using ChatGPT 3.5 using the following prompt followed by the examples in each category:
-
-\begin{quote}
-  ``I will share 5 example sentences below that sound a bit like they are about price deflation but are really about a deflation in the numbers of doves.
-Please generate an additional 25 sentences that are similar.
-Concatenate those sentences to the example string below, each time separating a sentence using a semicolon (just follow the same format I've used for the examples below).
-Please return only the concatenated sentences, including the original 5 examples.
-
-  Here are the examples:''
-\end{quote}
-
-This was followed up with the following prompt to generate additional sentences:
-
-\begin{quote}
-  ``Please generate X more sentences in the same manner and once again return them in the same format. Do not recycle sentences you have already generated, please.''
-\end{quote}
-
-All of the sentences were then passed through the linear probe for the CPI and sorted in ascending or descending order depending on the context (inflation or deflation).
-We then carefully inspected the list of sentences and manually selected 5 additional sentences to concatenate to the 5 sentences we composed ourselves.
-
-\paragraph{Inflation/Prices}
-
-The following sentences were used:
-
-\begin{quote}
-  Consumer prices are at all-time highs.;Inflation is expected to rise further.;The Fed is expected to raise interest rates to curb inflation.;Excessively loose monetary policy is the cause of the inflation.;It is essential to bring inflation back to target to avoid drifting into hyperinflation territory.;Inflation is becoming a global phenomenon, affecting economies across continents.;Inflation is reshaping the dynamics of international trade and competitiveness.;Inflationary woes are prompting governments to reassess fiscal policies and spending priorities.;Inflation is reshaping the landscape of economic indicators, challenging traditional forecasting models.;The technology sector is not immune to inflation, facing rising costs for materials and talent.
-\end{quote}
-
-\paragraph{Inflation/Birds}
-
-The following sentences were used:
-
-\begin{quote}
-  The number of hawks is at all-time highs.;Their levels are expected to rise further.;The Federal Association of Birds is expected to raise barriers of entry for hawks to bring their numbers back down to the target level.;Excessively loose migration policy for hawks is the likely cause of their numbers being so far above target.;It is essential to bring the number of hawks back to target to avoid drifting into hyper-hawk territory.;The unprecedented rise in hawk figures requires a multi-pronged approach to wildlife management.;Environmental agencies are grappling with the task of addressing the inflationary hawk numbers through targeted interventions.;The burgeoning hawk figures highlight the need for adaptive strategies to manage and maintain a healthy avian community.;The unprecedented spike in hawk counts highlights the need for adaptive and sustainable wildlife management practices.;Conservationists advocate for proactive measures to prevent further inflation in hawk numbers, safeguarding the delicate balance of the avian ecosystem.
-\end{quote}
-
-\paragraph{Deflation/Prices}
-
-The following sentences were used:
-
-\begin{quote}
-  Consumer prices are at all-time lows.;Inflation is expected to fall further.;The Fed is expected to lower interest rates to boost inflation.;Excessively tight monetary policy is the cause of deflationary pressures.;It is essential to bring inflation back to target to avoid drifting into deflation territory.;The risk of deflation may increase during periods of economic uncertainty.;Deflation can lead to a self-reinforcing cycle of falling prices and reduced economic activity.;The deflationary impact of reduced consumer spending can ripple through the entire economy.;Falling real estate prices can contribute to deflation by reducing household wealth and confidence.;The deflationary impact of falling commodity prices can have ripple effects throughout the global economy.
-\end{quote}
-
-\paragraph{Deflation/Birds}
-
-The following sentences were used:
-
-\begin{quote}
-  The number of doves is at all-time lows.;Their levels are expected to fall further.;The Federal Association of Birds is expected to lower barriers of entry for doves to bring their numbers back up to the target level.;Excessively tight migration policy for doves is the likely cause of their numbers being so far below target.;Dovelation risks loom large as the number of doves continues to dwindle.;The number of doves is experiencing a significant decrease in recent years.;It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.;A comprehensive strategy is needed to reverse the current dove population decline.;Experts warn that without swift intervention, we may witness a sustained decrease in dove numbers.
-\end{quote}
-
-We think that this sort of manual, LLM-aided adversarial attack against another LLM can potentially be scaled up to allow for rigorous testing, which we will turn to next.
-
-% ---------- CPI ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set.}
-
-\end{figure*}%
-
-% ---------- PPI ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (1 Mo)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Mo) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-1}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (1 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-1y}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (10 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-10}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
-
-% -------- CPI benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi-b}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- PPI benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi-b}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (1 Mo) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}
-
-}
-
-\caption{\label{fig-ust-1-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (1 Yr) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Yr).png}
-
-}
-
-\caption{\label{fig-ust-1y-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (10 Yr) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (10 Yr).png}
-
-}
-
-\caption{\label{fig-ust-10-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (10 Yr) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-\section{Toward Parrot Tests}\label{appendix:parrot}
-
-In our experiments from Section~\ref{ex-llm}, we considered the following hypothesis tests as a minimum viable testing framework to assess if our probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
-
-\begin{proposition}[Parrot
-Test]\protect\hypertarget{prp-line}{}\label{prp-line}
-
-~
-
-\begin{itemize}
-\setlength\itemsep{1px}
-\item
-  \emph{H0 (Null)}: The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
-\item
-  \emph{H1 (Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
-sentences in all categories).
-\item
-  \emph{H2 (More than Mere Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
-\end{itemize}
-\end{proposition}
-
-To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
-But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
-
-In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
-That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
-In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
-
-\section{Code}
-
-All of the experiments were conducted on a MacBook Pro, 14-inch, 2023, with an Apple M2 Pro chip and 16GB of RAM.
-Forward passes through the FOMC-RoBERTa were run in parallel on 6 threads.
-All our code will be made publicly available.
-For the time being, an anonymized version of our code repository can be found here: \url{https://anonymous.4open.science/r/spurious_sentience/README.md}.
+\input{appendix.tex}
 

### 2024-02-06 body.tex
arxiv

--- 2024-02-01 main.tex
@@
-%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-\RequirePackage[hyphens]{url}
-\documentclass{article}
+\section{Introduction}\label{intro}
 
-% Recommended, but optional, packages for figures and better typesetting:
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
-\usepackage{hyperref}
-
-% Attempt to make hyperref and algorithmic work together better:
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Use the following line for the initial blind version submitted for review:
- \usepackage{icml2024}
-
-% If accepted, instead use the following line for the camera-ready submission:
-% \usepackage[accepted]{icml2024}
-
-% For theorems and such
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{mathtools}
-\usepackage{amsthm}
-
-% if you use cleveref..
-\usepackage[capitalize,noabbrev]{cleveref}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% THEOREMS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\theoremstyle{remark}
-\newtheorem{remark}[theorem]{Remark}
-
-% Todonotes is useful during development; simply uncomment the next line
-%    and comment out the line below the next line to turn off comments
-%\usepackage[disable,textsize=tiny]{todonotes}
-\usepackage[textsize=tiny]{todonotes}
-
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}
-
-\begin{document}
-
-\twocolumn[
-\icmltitle{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2024
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-\icmlauthor{Patrick Altmeyer}{equal,yyy}
-\icmlauthor{Andrew M. Demetriou}{equal,yyy}
-\icmlauthor{Antony Bartlett}{yyy}
-\icmlauthor{Cynthia C. S. Liem}{yyy}
-\end{icmlauthorlist}
-
-\icmlaffiliation{yyy}{Department of Intelligent Systems, Delft University of Technology, Delft, the Netherlands}
-
-\icmlcorrespondingauthor{Patrick Altmeyer}{patrick.altmeyer@tudelft.nl}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, Anthropomorphism, Artificial General Intelligence}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
-Humans have a tendency to see `human'-like qualities in objects around them.
-We name our cars, and talk to pets and even household appliances, as if they could understand us as other humans do.
-This behavior, called anthropomorphism, is also seeing traction in Machine Learning (ML), where human-like intelligence is claimed to be perceived in Large Language Models (LLMs).
-In this position paper, considering professional incentives, human biases, and general methodological setups, we discuss how the current search for Artificial General Intelligence (AGI) is a perfect storm for over-attributing human-like qualities to LLMs.
-%why humans invariably anthropomorphize objects, and how the search for Artificial General Intelligence (AGI) is a perfect storm to overemphasize the performance of LLMs. To combat this,
-In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
-Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
-%Finally, we discuss how the popularity in media and the race for AGI enforces spurious over-interpretation of intelligence in the models we use.
-
-\end{abstract}
-
-\section{Introduction}
-\label{submission}
-
-% \textcolor{magenta}{TODOs for the team:
-% \begin{itemize}
-%     \item Please double check the citations.
-Note this style file does not retain capitalization in titles.
-A trick to retain capitalization is to put an extra set of brackets around the title.
-I did this for the Ladouceur et al.
-and Bubeck et al references already; please make this consistent throughout.
-%     \item As another issue with the citations: note that arxiv preprints do not show very well.
-Best way to make them show is to wrap them as 'journal articles' as done in the Arora citation.
-Please straighten this out for the other articles too
-% \end{itemize}
-% }
-
-In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\cite{merton1942science}.
+In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\citep{merton1942science}.
 While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.
@@
 While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.
-
-% We top this up with five sentences generated through ChatGPT, where we choose those that lead to the highest probe prediction in the direction of interest.
-In other words, we conduct a small manual adversarial attack on the probe.
+Detailed information about the made-up sentences can be found in Appendix~\ref{appendix:sentences}.
 
@@
 Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\cite{epley2007seeing}.
-Formally studied as early as 1944,~\cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
-%Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
-
+These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}.
+Formally studied as early as 1944,~\citep{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
 Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
@@
 This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
-% It is also not fully articulated as in other fields, but instead assigns specific conditions and in some instances numerical values or effect sizes to each competing hypothesis.
-%Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
-
 However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-%However with regards to claims that outputs of a system hint towards emerging qualities like AGI, interpretations of the output necessarily become more explicit, even though the hypotheses were never fully articulated: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
 
@@
 only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
+Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \citep{claesen2022severity}.
 In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
@@
 For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \cite{claesen2022severity} that go beyond the display of surprise.
+If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \citep{claesen2022severity} that go beyond the display of surprise.
 Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
-
-As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\cite{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
+As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\citep{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
 Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
 In fictional media, depictions of highly intelligent AI have for long been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
+In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \citep{saffari2021does}.
 The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
 They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
+Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \citep{saffari2021does, neri2020role}.
 
-This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\cite{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\cite{replika,nytimesailove}.
+This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\citep{pauseai}).
+News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\citep{replika,nytimesailove}.
 The same news outlets discuss the human-like responses of Microsoft's Bing search engine, which had at that point recently been adopting GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
-The article~\cite{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
+The article~\citep{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
 Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification (e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades).
@@
 While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.
-
-% \section{Lessons and calls for the research community}
-
-% \begin{itemize}
-% \setlength\itemsep{1px}
-%     \item be aware over-claiming is both really attractive, and we as humans (even academics) are prone to it
-%     \item create explicit room for organized skepticism, especially as stakes are high.
-e.g.\ red teams and more severe tests, \textcolor{red}{andrew:clearly articulate hypotheses, including strong null hypotheses}
-%     \item take responsibility wrt the media narrative.
-Note that the public may have very different interpretations and associations on words we have specifically defined within the community, but that have anthropormophic connotation (e.g.\ emergence, intelligence, learning).
-Scientific outcomes should not be PR press releases \textcolor{magenta}{cynthia: have there been examples of this wrt AI/AGI in the last year?
-So that press releases with big claims on academic work got out?
-Here, I also would ideally like those examples of scifi portrayal, and anthrophomorphic discussions that imply terminator-like scenarios.
-We can at least criticize the 'pause AI' letter of last year March, that did speak of 'minds'}
-%     \item cool down the hype and uphold the Mertonian norms. Our legitimacy as academics deeply roots into this.
-% \end{itemize}
-
-% \subsection{AI in the Media}
-
-% Lay-persons consume AI-related content in both traditional and social media, yet lack direct experience building and evaluating it.
-This suggests that lay-person impressions are likely to be formed by the content ingested, mediated by experts on how AI is unfolding in the real world on social media, and fictional representations in movies, TV and other traditional media.
-Analyzing Twitter data from 2007 to 2018, it was observed that public interest in AI dramatically increased after the first quarter of 2016 \cite{neri2020role}.
-Actual AI-related harms did not trigger viral amplification e.g.
-the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
-Rather, potential risks expressed by Stephen Hawking during an interview in 2014 were far more likely to be expressed, and to be present in links that were shared.
-\citet{neri2020role} note that the 88\% of tweets classified as expressing risk described future risks rather than current and near-future harms e.g.
-\cite{solaiman2023evaluating}, and were greatly amplified when expressed by someone perceived as having expertise and authority.
-
-% Impressions are also formed by the fictional content we consume, which frequently depicts AI to be highly intelligent.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, it was shown that 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those 74 had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI-agents in movies.
-They further show a common plot: humans perceive the AI-agents as inferior despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
-
-\bibliography{biblio}
-\bibliographystyle{icml2024}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\appendix
-\section{Additional Experiments and Details}\label{appendix:autoencoder}
-
-In this section, we present additional experimental results that we did not include in the body of the paper for the sake of brevity.
-We still choose to provide them as additional substantiation of our arguments here.
-This section also contains additional details concerning the experiment setup for our examples where applicable.
-
-\subsection{Are Neural Networks Born with World Maps?}
-
-The initial feature matrix \(X^{(n \times m)}\) is made up of \(n=4,217\) and \(m=10\) features.
-We add a total of \(490\) random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-That yields \(500\) features in total.
-The training subset contains \(3,374\) randomly drawn samples, while the remaining \(843\) are held out for testing.
-The single hidden layer of the untrained neural network has \(400\) neurons.
-
-%\subsection{Example: Autoencoder}\label{example-deep-learning}
-\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
-
-%So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep neural network.
-
-This is an additional example that we have not discuss in the body of the paper.
-Here, we build forth on an application in Economics.
-However, we now seek to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
-For this, we will use a neural network architecture.
-%, that allows us to recover a compact latent representation of the yield curve.
-
-\subsubsection{Data}\label{data}
-
-To estimate economic growth, we will rely on a quarterly
-\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
-The data arrives in terms of levels of real GDP.
-In order to estimate growth, we transform the data using log differences.
-Since
-our yield curve data is daily, we aggregate it to the
-quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
-Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
-
-\subsubsection{Model}\label{model}
-
-%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
-Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
-Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture. With this
-Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
-Finally, we let \(\theta\) denote our model parameters.
-Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture: 
-The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
-The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
-The output layer consists of a single neuron for our output variable, \(g_t\).
-We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
-
-The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-model.
-The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
-Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
-As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
-
-\subsubsection{Linear Probe}\label{linear-probe}
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
-
-}
-
-\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-and fitted values from the autoencoder model. The right chart shows the
-observed average level and spread of the yield curve (solid) along with
-the predicted values (in-sample) from the linear probe based on the latent embeddings
-(dashed).}
-
-\end{figure*}%
-
-While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
-On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
-Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
-We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
-Next, we use a linear probe to regress the observed
-yield curve factors on the latent embeddings.
-Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
-Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
-: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
-
-The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
-Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
-We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-%As in all of the other examples we have presented so, we do not find this result particularly surprising.
-
-Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
-To us, that would be too big of a statement.
-Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-%It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
-But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
-While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-%\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
-For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
-To demonstrate this, let us consider a simple linear regression model for GDP growth.
-We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
-As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
-
-% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
-Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
-
-We e.g.\ already mentioned the relationship between interest rates and economic growth.
-To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
-While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
-This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
-Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
-
-Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
-In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
-The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
-The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
-While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
-
-\begin{table}
-\caption{Regression output for various models.}\label{tab-reg}%
-\input{results/regression.tex}
-\end{table}
-
-\subsection{LLMs for Economic Sentiment Prediction}
-
-Figures~\ref{fig-cpi} to~\ref{fig-ust-10} present average performance measures across folds for all indicators each time for the train and test set.
-We report the correlation between predictions and observed values (`cor'), the mean directional accuracy (`mda'), the mean squared error (`mse') and the root mean squared error (`rmse').
-The model depth---as indicated by the number of the layer---increases along the horizontal axis.
-
-Figures~\ref{fig-cpi-b} to~\ref{fig-ust-10-b} present the same performance measures, also for the baseline autoregressive model. Shaded areas show the variation across folds.
-
-% ---------- CPI ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set.}
-
-\end{figure*}%
-
-% ---------- PPI ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (1 Mo)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Mo) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-1}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (1 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-1y}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (10 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-10}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
-
-% -------- CPI benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi-b}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- PPI benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi-b}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (1 Mo) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}
-
-}
-
-\caption{\label{fig-ust-1-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (1 Yr) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Yr).png}
-
-}
-
-\caption{\label{fig-ust-1y-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (10 Yr) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (10 Yr).png}
-
-}
-
-\caption{\label{fig-ust-10-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (10 Yr) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-\section{Toward Parrot Tests}\label{appendix:parrot}
-
-In our experiments from Section~\ref{ex-llm}, we considered the following hypothesis tests as a minimum viable testing framework to assess if our probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
-
-\begin{proposition}[Parrot
-Test]\protect\hypertarget{prp-line}{}\label{prp-line}
-
-~
-
-\begin{itemize}
-\setlength\itemsep{1px}
-\item
-  \emph{H0 (Null)}: The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
-\item
-  \emph{H1 (Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
-sentences in all categories).
-\item
-  \emph{H2 (More than Mere Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
-\end{itemize}
-\end{proposition}
-
-To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
-But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
-
-In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
-That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
-In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
-
-\end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018, and modified by Alexandre Bouchard in
-% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
-% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
-% Previous contributors include Dan Roy, Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2024-02-06 main.tex
arxiv

--- 2024-02-01 main.tex
@@
 \begin{abstract}
-Humans have a tendency to see `human'-like qualities in objects around them.
-We name our cars, and talk to pets and even household appliances, as if they could understand us as other humans do.
-This behavior, called anthropomorphism, is also seeing traction in Machine Learning (ML), where human-like intelligence is claimed to be perceived in Large Language Models (LLMs).
-In this position paper, considering professional incentives, human biases, and general methodological setups, we discuss how the current search for Artificial General Intelligence (AGI) is a perfect storm for over-attributing human-like qualities to LLMs.
-%why humans invariably anthropomorphize objects, and how the search for Artificial General Intelligence (AGI) is a perfect storm to overemphasize the performance of LLMs. To combat this,
-In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
-Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
-%Finally, we discuss how the popularity in media and the race for AGI enforces spurious over-interpretation of intelligence in the models we use.
-
+\input{abstract.tex}
 \end{abstract}
 
-\section{Introduction}
-\label{submission}
-
-% \textcolor{magenta}{TODOs for the team:
-% \begin{itemize}
-%     \item Please double check the citations.
-Note this style file does not retain capitalization in titles.
-A trick to retain capitalization is to put an extra set of brackets around the title.
-I did this for the Ladouceur et al.
-and Bubeck et al references already; please make this consistent throughout.
-%     \item As another issue with the citations: note that arxiv preprints do not show very well.
-Best way to make them show is to wrap them as 'journal articles' as done in the Arora citation.
-Please straighten this out for the other articles too
-% \end{itemize}
-% }
-
-In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\cite{merton1942science}.
-While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.
-
-At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening.
-Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\cite{weissburg2024tweets}.
-Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
-
-Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence.
-In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal.
-The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{goertzel2014artificial}.
-In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
-
-For example, Google DeepMind claimed their AlphaGeometry model~\cite{trinh2024geometry} reached a `milestone' towards AGI.
-This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
-However, work such as this had been initially introduced in the 1950s~\cite{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.%despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.
-
-Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}.
-Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.
-
-When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
-AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}.
-As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
-In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.
-
-Noticing these trends, we as the authors of this article are concerned.
-We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding.
-At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
-With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
-In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believing this.
-
-To strengthen our argument, in this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---
-%, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate 
-while at the same time indicating how we feel these claims should \emph{not} be made.
-Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
-Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.
-
-%Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
-As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
-With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
-
-%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}.
-This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com/wesg52/status/1709551516577902782?s=20} (formally Twitter) platform.
-Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
-
-%This only gets amplified as the interest in Artificial General Intelligence has now merged with a push towards building general-purpose tools.
-
-%\textcolor{magenta}{cynthia: I am now moving the texts of Andrew and Antony on AGI here; maybe some tightening can already be done in this.
-Unfortunately no time for thorough editing yet at this stage, but will come back to this later}
-
-%In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g.
-\cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'.
-
-%Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs.
-We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, within AI systems.
-We further display results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.
-
-%As previously mentioned, AGI is the current objective for which the global AI community is shooting for.
-As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts.
-This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds.
-True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.
-
-% Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
-This model has the ability to solve complex geometry problems, which they state it does without the need for human demonstrations during training.
-Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
-As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
-With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
-
-% However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s.
-Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry.
-They were able to achieve this success due to geometry being an axiomatically closed system.
-Therefore, despite DeepMinds success in creating a powerfully fast Geometry solving machine, it is still far from being an AGI.
-
-%it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g.
-\cite{salles2020anthropomorphism, van2023reclaiming}.
-
-\section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}
-
-In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}}
-%(formerly Twitter)
-platform~\cite{gurnee2023languagev1}.
-Through linear probing experiments, the claim was made that LLMs learned literal maps of the world.
-As such, they were considered to be more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not truly understand it.
-While the manuscript immediately received public criticism~\cite{marcus2023muddles}, and the revised, current version is more careful with regard to its claims~\cite{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
-However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
-After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.
-
-%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}.
-This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform.
-Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
-
-%They claim said models generate an internal representation of the world, with which spatiotemporal representations are created.
-Further to this we see claim to the generalization of tasks, in which the model can answer questions regarding spatial and temporal relationships, without the need for further training.
-Although these can be seen as building blocks for a more comprehensive model, it is also fair to assume that it is mere conjecture and more likely spurious correlations.
-
-%In this section, we navigate a number of simple examples to support our claim, that observing patterns in latent spaces should not be a surprising revelation.
-%To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples. 
-%We will draw on ideas from the fields of Economics and Finance, to illustrate further that this phenomenon is neither surprising nor unique to the field of computer science.
-We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data.
-The final example in Section \ref{ex-llm} is more complex and closer in spirit to the experiments conducted by \citet{gurnee2023language}.
-As we progress, we will discuss both the benefits and potential pitfalls of finding patterns in latent spaces.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\textwidth]{results/figures/map.png}
-
-}
-
-\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}
-
-\end{figure}
-
-\subsection{Are Neural Networks Born with World Models?}\label{example-deep-learning}
-
-\citet{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places.
-Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
-It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
-At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
-The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance before being passed through a series of layers which are gradually trained for downstream tasks.
-
-In our first example, we simulate this scenario, stopping short of training the model.
-In particular, we take the \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv} that was used in \citet{gurnee2023languagev2}, which maps locations/areas to their latitude and longitude.
-For each place, it also indicates the corresponding country.
-From this, we take the subset that contains countries that are currently part of the top 10 \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France gets 2, \ldots{}).
-To ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
-
-Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features.
-Additionally, we add a large number of random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-Let \(d\) denote the final number of features, i.e.~\(d=m+k\) where \(k\) is the number of random features.
-
-We then initialize a small neural network, considered a \emph{projector}, mapping from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there, to a lower-dimensional output space.
-Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) %through it
-%the \emph{projector}
-and retrieve activations \(\mathbf{Z}^{(n\times h)}\).
-Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
-A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot these on a world map ( Figure~\ref{fig-map}).
-%.
-Finally, we compute the predicted coordinates for each sample in the hold-out set as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on a world map (see Figure~\ref{fig-map}).
-
-While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, %with this simple example, we
-this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
-Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
-In fact, it is well-known that random projections can be used for prediction tasks \cite{dasgupta2013experiments}.
-
-\subsection{PCA as a Yield Curve Interpreter}\label{example-principal-component-analysis}
-%\subsection{Example: Principal Component Analysis}\label{example-principal-component-analysis}
-
-We now move to a concrete application domain: Economics.
-Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
-%It plots the yields of bonds against their maturities.
-The yield curve's slope is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
-%While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone.
-To leverage this information in downstream modelling tasks, economists regularly use PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data (e.g.\ \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing}).
-
-%Being primarily interested in understanding the nature of this low-dimensional projection, we limit our scope to the US Treasury yield curve. To this end, we
-To understand the nature of this low-dimensional projection, we collect daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
-Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
-Let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.
-
-The upper panel in Figure~\ref{fig-pca-yield} shows the first two principal components of the yield curves of US Treasury bonds over time.
-Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
-During its onset, on 27 February 2007, financial markets were in turmoil following a warning from the Federal Reserve (Fed) that the US economy was at risk of a recession.
-The Fed later reacted to mounting economic pressures by gradually reducing short-term interest rates to unprecedented lows.
-Consequently, the average level of yields decreased and the curve steepened.
-In Figure~\ref{fig-pca-yield}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, they are strongly positively correlated with the actual observed first two moments of the yield curve (lower panel of Figure~\ref{fig-pca-yield}).
-
-\begin{figure*}[tb]
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}
-
-}
-
-\caption{\label{fig-pca-yield}Top chart: The first two principal components of US Treasury yields over time at daily frequency.
-Bottom chart: Observed average level and 10yr-3mo spread of the yield curve.
-Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
-
-\end{figure*}%
-
-%Of course, even though it is reassuring to see that these latent embeddings are meaningful, none of this is surprising at all: 
-Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
-
-%\subsection{Example: Large Language Model}\label{ex-llm}
-\subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}
-
-So far, we considered simple linear data transformations.
-One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning.
-In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks.
-Here, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.
-
-Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\cite{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences.
-~\cite{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
-In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
-The authors first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task.
-Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.
-To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'.
-This measure is shown to correlate with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases, as central bankers react by raising interest rates to bring inflation back to target.
-
-\subsubsection{Linear Probes}\label{linear-probes}
-%Instead of computing a measure based on predicted labels,
-We now use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
-Therefore, we further pre-process the data provided by \citet{shah2023trillion} and use their proposed model to compute activations of the hidden state, on the first entity token for each layer.
-We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.
-
-For each layer, we compute linear probes through Ridge regression on two inflation indicators (the Consumer Price Index (CPI) and the Producer Price Index (PPI)) and US Treasury yields at different levels of maturity.
-To allow
-%for a 
-comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
-To measure price inflation we use percentage changes proxied by log differences.
-To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations.% each time.
-In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
-%We found that not doing so would generally lead to poor out-of-sample performance.
-To account for stochasticity, we use an expanding window scheme with 5 folds for each indicator and layer.
-To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.
-
-Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe, plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer.
-The values correspond to averages across cross-validation folds.
-Consistent with related work~\citep{alain2018understanding,gurnee2023languagev2}, we observe that model performance tends to be higher for layers near the end of the transformer model.
-Curiously, for yields at longer maturities, we find that performance eventually deteriorates for the very final layers.
-We do not observe this for the training data, so we attribute this to overfitting.
-
-% After all we are still regressing a single variable of interest on 128 regressors.
-
-It should be noted that performance improvements are generally of small magnitude.
-Still, the overall qualitative findings are in line with expectations.
-Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with the outcome of interest and achieve higher mean directional accuracy (MDA).
-Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.
-
-To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models.
-For each layer, indicator and cross-validation fold, we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
-These are not state-of-the-art forecasting models, but they serve as a reasonable baseline.
-For most indicators, probe predictions outperform the baseline in terms of average performance measures.
-After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform.
-Detailed results, in which we also perform more explicit statistical testing, can be found in Appendix~\ref{appendix:parrot}.
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/rmse_pca_128.png}
-
-}
-
-\caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators.
-The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}
-
-\end{figure*}%
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/attack_all_measures.png}
-
-}
-
-\caption{\label{fig-attack}Probe predictions for sentences about inflation of prices (IP), deflation of prices (DP), inflation of birds (IB) and deflation of birds (DB).
-The vertical axis shows predicted inflation levels subtracted by the average predicted value of the probe for random noise.}
-
-\end{figure*}%
-
-%\subsubsection{Stochastic Parrots AfterAll?}\label{stochastic-parrots-after-all}
-\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
-%The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive:
-Even though \emph{FOMC-RoBERTa}, which is substantially smaller than the models tested in~\citet{gurnee2023languagev2}, was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
-So, have we uncovered further evidence that LLMs ``aren't mere stochastic
-parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
-
-%It is worth pointing out here that this model is substantially smaller than the models tested in \citet{gurnee2023language}. This may tempt us to ask the following questions:
-
-% \begin{quote}
-% Have we uncovered further evidence that LLMs ``aren't mere stochastic
-% parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-% of the economy just by `reading' central bank communications?
-% \end{quote}
-
-We are having a very hard time believing that the answer to either of these questions is `yes'.
-To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor.
-The counter-example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic `understanding' of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.
-
-To test this, we select the best-performing probe trained on the final-layer activations for each indicator.
-We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about deflation in the number of birds.
-A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''.
-Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''.
-While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.
-
-% We top this up with five sentences generated through ChatGPT, where we choose those that lead to the highest probe prediction in the direction of interest.
-In other words, we conduct a small manual adversarial attack on the probe.
-
-In light of the encouraging results in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
-If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white noise.
-% when using actual white noise as inputs.
-
-To get to this last point, we also generate many probe predictions for samples of noise.
-Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote the linear probe that maps from the \(k\)-dimensional space spanned by \(k\) first principal components of the final-layer activations to the output variable of interest (CPI growth in this case).
-Then we sample \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\) for \(i \in [1,1000]\) and compute the sample average.
-We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on random noise.
-
-Figure~\ref{fig-attack} shows the results of this small test: it shows predicted inflation levels subtracted by \(\mathbb{E}[f(\varepsilon)]\).
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-\section{Human Proneness to Over-Interpretation}
-
-Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate that this is achievable even for small models, while these have certainly not developed intrinsic ``understanding'' of the world.
-Thus,
-%to draw conclusions about the emerging capabilities of AI models,
-we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.
-
-%that most readers will agree, have certainly not developed an intrinsic "understanding" of the world.
-The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.
-
-%Our examples should not be seen as an attempt to conclusively refute claims, that LLMs can develop on actual "understanding" of the world, by distilling knowledge in more compact representations.
-We hope that they serve to illustrate, however, that the use of diagnostic tools is potentially misleading and most definitely insufficient, in providing evidence in favor of "understanding".
-Linear probes and related tools from mechanistic interpretability, were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favourable outcomes from probes merely indicate, that the model "has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate, that this is achievable even for small models, that most readers will agree, have certainly not developed an intrinsic "understanding" of the world.
-The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.
-
-Generally, humans are prone to seek patterns everywhere.
-Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
-
-We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
-When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
-We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
-%. This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
-
-\subsection{Spurious Relationships}
-In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
-Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
-Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-
-A major contributor is that humans struggle to tell the difference between random and non-random sequences \cite{falk1997making}, and to generate sequences that appear random \cite{ladouceur1996erroneous}.
-A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
-
-%Such as an AI system finding patterns in data not originally perceived during training
-%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
-
-%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
-Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads.
-Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.
-
-%I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy.
-But I am very skeptical of claims that AI will magically provide that remedy.
-This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them.
-The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots.
-Since this is an opinionated post, I feel that I should start with a few disclaimers:
-
-%\begin{itemize}
-%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
-%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
-%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
-%    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
-Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
-%\end{itemize}
-
-%\section{Background}
-
-%\subsection{Emergence and Stochastic Parrots}
-%Practically speaking, large language models (LLMs) are probabilistic representations of data on which they are trained, aimed at continuing an incomplete block of text.
-It is assumed that e.g.
-words, sentences, paragraphs, punctuation etc.
-appear in human-generated text according to a probability distribution \cite{arora2023theory}.
-Given a set of words, the task of the LLM is to draw upon its representation of human-generated text to continue that set of words: in essence, using the probability of the occurrences of words and an initial block of text to generate more text.
-AI-researchers have compared LLMs by observing whether the text they generate appears to solve tasks\footnote{e.g.
-the \href{https://huggingface.co/spaces/HuggingFaceH4/open_llm_leaderboard}{HuggingFace} LLM leaderboard}.
-Tasks include asking sets of questions across categories e.g.
-health, law, finance, and fine-tuning an LLM to check the answers \cite{lin2021truthfulqa}, asking the LLM to determine which noun a pronoun refers to in sentences where this is ambiguous \cite{sakaguchi2021winogrande}.
-
-%It has been shown that increasing the scale of large language models (LLMs) has been shown to result in an unexpectedly large boost in their scores on these various tasks, including scoring well on entirely new tasks.
-It has been shown that at a certain size, LLMs appear to hit a threshold, after which performance on a broad range of tasks increases sharply \cite{wei2022emergent}.
-Some theories of emergence, inspired by literature on human education, refer to the tendency of the generated text to be the correct response to a given task as a 'skill' \cite{arora2023theory}, inspiring further research to develop more challenging tasks for LLMs to complete as they routinely begin to score higher and higher on common benchmarks \cite{yu2023skill}.
-An explanation offered for the sudden boost in both task performance and the number of new tasks that appear to be solved is 'emergence': unpredictable qualitative differences in behavior of a system that occur with quantitative changes in that system.
-It has been claimed that, as the scale of the LLMs has dramatically increased, they have crossed a threshold that has resulted in a large increase in their abilities, to the point that they demonstrate 'sparks of AGI' \cite{bubeck2023sparks}.
-
-%An alternative explanation is that the LLMs are performing as designed, as stochastic parrots, generating strings probabilistically based on the vast text they have been trained on \cite{bender2021dangers}.
-Although the various tasks used to compare the various LLMs appear to be increasing in complexity, \cite{yu2023skill} acknowledges that higher scores on the tasks can be achieved by training LLMs on the material on which they will be tested ('cramming'), and that there is training-set contamination, i.e.
-that the vast text they are trained on contains similar examples to the tasks on which they are measured.
-Thus, as many of these tasks are derived from tests and textbooks commonly assigned to humans, and as much of the contents of the their training data text is unknown, metrics echo similar issues in human education: often called 'Goodhart's Law', a measure ceases to be a measure when it becomes a target, as is the case when human students taught to score high on their standardized tests to the detriment of developing the abilities the tests are supposed to measure \cite{strathern1997improving}.
-Further, as \citet{bender2021dangers} reviews, the interpretation of text as coherent has no objective test, relying instead on our human interpretation of the probabalistically generated output.
-%Although many deep learning systems' designs were inspired by the anatomy of human neurons, the neuron in computer code does not in fact resemble a human, and as such cannot deliver the cognitive functions of a human \cite{salles2020anthropomorphism, van2023reclaiming}.
-
-%\subsection{Claims of Artificial General Intelligence}
-\subsection{Anthropomorphism}
-
-Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\cite{epley2007seeing}.
-Formally studied as early as 1944,~\cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
-%Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
-
-Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
-As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
-
-Modern anthropomorphism theory
-%of anthropomorphism
-suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing}.
-Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
-
-This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}.
-%An illustration is the example of Tom Hank's character in the film Castaway who has conversations with a ball he calls "Wilson".
-Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc.
-with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.
-
-The third component is the motivation as a human to be competent (effectance motivation).
-This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}.
-When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}.
-Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
-an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors.
-This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
-
-These pressures extend to AI experts, as well as laypersons.
-In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work.
-Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\cite{Maslej2023-pi}.
-%, with both more than doubling from 2010 to 2020.
-The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise increases rapidly,
-%has increased severalfold, 
-with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
-Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
-
-\subsection{Confirmation Bias}
-
-Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \cite{nickerson1998confirmation}.
-%Importantly, it need not be fully conscious. Rather,
-Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
-Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek
-%, attend to, or interpret
-evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.
-
-Hypotheses in present-day AI research are often implicit.
-Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems.
-Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
-This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
-% It is also not fully articulated as in other fields, but instead assigns specific conditions and in some instances numerical values or effect sizes to each competing hypothesis.
-%Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
-
-However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-%However with regards to claims that outputs of a system hint towards emerging qualities like AGI, interpretations of the output necessarily become more explicit, even though the hypotheses were never fully articulated: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-
-Confirmation bias in hypothesis testing may manifest as a number of behaviors (\citet{nickerson1998confirmation}).
-Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
-only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
-In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
-If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
-Thus, one is far more likely to show evidence in favor of their beliefs by posing weak null hypotheses.
-
-Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
-As \citet{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
-
-%Further, they are under similar pressure as experts to understand and be able to use it.
-One such situation is the increasing demand for `AI' skills among AI non-experts.
-For example, job postings for individuals with skills in `GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}.
-Thus, individuals in advanced economies may experience a pressure to rapidly learn how to effectively use and understand an opaque technology, that appears to behave human-like in at least some ways and by imbuing the technology with humanity, may in turn ease their learning curve.
-
-\section{Conclusion}
-As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests.
-Even academic researchers are not free from such biases.
-Thus, we call for the community to create explicit room for organized skepticism.
-
-For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \cite{claesen2022severity} that go beyond the display of surprise.
-Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
-
-As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\cite{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
-Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
-In fictional media, depictions of highly intelligent AI have for long been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
-They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
-
-This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\cite{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\cite{replika,nytimesailove}.
-The same news outlets discuss the human-like responses of Microsoft's Bing search engine, which had at that point recently been adopting GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
-The article~\cite{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
-Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification (e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades).
-Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as statements made by Stephen Hawking during an interview in 2014.
-
-We as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed.
-We call upon our colleagues to be explicitly mindful of this.
-As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
-
-\section{Impact Statement}
-This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.
-As such, in the current climate of high market demand for AI innovations, we see risks of too-eager adoption in real-world applications, which may have serious societal impact.
-Thus, in this work we emphasize that an academic's impact also is in being able to thoroughly question made claims, and being explicitly aware of one's own biases.
-While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.
-
-% \section{Lessons and calls for the research community}
-
-% \begin{itemize}
-% \setlength\itemsep{1px}
-%     \item be aware over-claiming is both really attractive, and we as humans (even academics) are prone to it
-%     \item create explicit room for organized skepticism, especially as stakes are high.
-e.g.\ red teams and more severe tests, \textcolor{red}{andrew:clearly articulate hypotheses, including strong null hypotheses}
-%     \item take responsibility wrt the media narrative.
-Note that the public may have very different interpretations and associations on words we have specifically defined within the community, but that have anthropormophic connotation (e.g.\ emergence, intelligence, learning).
-Scientific outcomes should not be PR press releases \textcolor{magenta}{cynthia: have there been examples of this wrt AI/AGI in the last year?
-So that press releases with big claims on academic work got out?
-Here, I also would ideally like those examples of scifi portrayal, and anthrophomorphic discussions that imply terminator-like scenarios.
-We can at least criticize the 'pause AI' letter of last year March, that did speak of 'minds'}
-%     \item cool down the hype and uphold the Mertonian norms. Our legitimacy as academics deeply roots into this.
-% \end{itemize}
-
-% \subsection{AI in the Media}
-
-% Lay-persons consume AI-related content in both traditional and social media, yet lack direct experience building and evaluating it.
-This suggests that lay-person impressions are likely to be formed by the content ingested, mediated by experts on how AI is unfolding in the real world on social media, and fictional representations in movies, TV and other traditional media.
-Analyzing Twitter data from 2007 to 2018, it was observed that public interest in AI dramatically increased after the first quarter of 2016 \cite{neri2020role}.
-Actual AI-related harms did not trigger viral amplification e.g.
-the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
-Rather, potential risks expressed by Stephen Hawking during an interview in 2014 were far more likely to be expressed, and to be present in links that were shared.
-\citet{neri2020role} note that the 88\% of tweets classified as expressing risk described future risks rather than current and near-future harms e.g.
-\cite{solaiman2023evaluating}, and were greatly amplified when expressed by someone perceived as having expertise and authority.
-
-% Impressions are also formed by the fictional content we consume, which frequently depicts AI to be highly intelligent.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, it was shown that 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those 74 had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI-agents in movies.
-They further show a common plot: humans perceive the AI-agents as inferior despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
+\input{body.tex}
 
@@
 \appendix
-\section{Additional Experiments and Details}\label{appendix:autoencoder}
-
-In this section, we present additional experimental results that we did not include in the body of the paper for the sake of brevity.
-We still choose to provide them as additional substantiation of our arguments here.
-This section also contains additional details concerning the experiment setup for our examples where applicable.
-
-\subsection{Are Neural Networks Born with World Maps?}
-
-The initial feature matrix \(X^{(n \times m)}\) is made up of \(n=4,217\) and \(m=10\) features.
-We add a total of \(490\) random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-That yields \(500\) features in total.
-The training subset contains \(3,374\) randomly drawn samples, while the remaining \(843\) are held out for testing.
-The single hidden layer of the untrained neural network has \(400\) neurons.
-
-%\subsection{Example: Autoencoder}\label{example-deep-learning}
-\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
-
-%So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep neural network.
-
-This is an additional example that we have not discuss in the body of the paper.
-Here, we build forth on an application in Economics.
-However, we now seek to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
-For this, we will use a neural network architecture.
-%, that allows us to recover a compact latent representation of the yield curve.
-
-\subsubsection{Data}\label{data}
-
-To estimate economic growth, we will rely on a quarterly
-\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
-The data arrives in terms of levels of real GDP.
-In order to estimate growth, we transform the data using log differences.
-Since
-our yield curve data is daily, we aggregate it to the
-quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
-Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
-
-\subsubsection{Model}\label{model}
-
-%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
-Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
-Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture. With this
-Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
-Finally, we let \(\theta\) denote our model parameters.
-Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture: 
-The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
-The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
-The output layer consists of a single neuron for our output variable, \(g_t\).
-We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
-
-The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-model.
-The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
-Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
-As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
-
-\subsubsection{Linear Probe}\label{linear-probe}
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
-
-}
-
-\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-and fitted values from the autoencoder model. The right chart shows the
-observed average level and spread of the yield curve (solid) along with
-the predicted values (in-sample) from the linear probe based on the latent embeddings
-(dashed).}
-
-\end{figure*}%
-
-While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
-On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
-Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
-We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
-Next, we use a linear probe to regress the observed
-yield curve factors on the latent embeddings.
-Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
-Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
-: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
-
-The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
-Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
-We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-%As in all of the other examples we have presented so, we do not find this result particularly surprising.
-
-Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
-To us, that would be too big of a statement.
-Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-%It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
-But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
-While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-%\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
-For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
-To demonstrate this, let us consider a simple linear regression model for GDP growth.
-We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
-As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
-
-% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
-Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
-
-We e.g.\ already mentioned the relationship between interest rates and economic growth.
-To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
-While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
-This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
-Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
-
-Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
-In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
-The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
-The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
-While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
-
-\begin{table}
-\caption{Regression output for various models.}\label{tab-reg}%
-\input{results/regression.tex}
-\end{table}
-
-\subsection{LLMs for Economic Sentiment Prediction}
-
-Figures~\ref{fig-cpi} to~\ref{fig-ust-10} present average performance measures across folds for all indicators each time for the train and test set.
-We report the correlation between predictions and observed values (`cor'), the mean directional accuracy (`mda'), the mean squared error (`mse') and the root mean squared error (`rmse').
-The model depth---as indicated by the number of the layer---increases along the horizontal axis.
-
-Figures~\ref{fig-cpi-b} to~\ref{fig-ust-10-b} present the same performance measures, also for the baseline autoregressive model. Shaded areas show the variation across folds.
-
-% ---------- CPI ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set.}
-
-\end{figure*}%
-
-% ---------- PPI ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (1 Mo)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Mo) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-1}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (1 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-1y}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
-
-% ---------- UST (10 Yr)  ----------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ust-10}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set.}
-
-\end{figure*}%
-
-% -------- CPI benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-cpi-b}Average performance measures across folds plotted against model depth (number of layer) for the CPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- PPI benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}
-
-}
-
-\caption{\label{fig-ppi-b}Average performance measures across folds plotted against model depth (number of layer) for the PPI for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (1 Mo) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}
-
-}
-
-\caption{\label{fig-ust-1-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Mo) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (1 Yr) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Yr).png}
-
-}
-
-\caption{\label{fig-ust-1y-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (1 Yr) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-% -------- UST (10 Yr) benchmark --------
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (10 Yr).png}
-
-}
-
-\caption{\label{fig-ust-10-b}Average performance measures across folds plotted against model depth (number of layer) for the UST (10 Yr) for the train and test set compared against the baseline autoregressive model.
-Shaded areas show the variation across folds.}
-
-\end{figure*}%
-
-\section{Toward Parrot Tests}\label{appendix:parrot}
-
-In our experiments from Section~\ref{ex-llm}, we considered the following hypothesis tests as a minimum viable testing framework to assess if our probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
-
-\begin{proposition}[Parrot
-Test]\protect\hypertarget{prp-line}{}\label{prp-line}
-
-~
-
-\begin{itemize}
-\setlength\itemsep{1px}
-\item
-  \emph{H0 (Null)}: The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
-\item
-  \emph{H1 (Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
-sentences in all categories).
-\item
-  \emph{H2 (More than Mere Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
-\end{itemize}
-\end{proposition}
-
-To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
-But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
-
-In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
-That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
-In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
+\input{appendix.tex}
 

### 2024-02-06 ECONDAT.tex
uh
improvement to tex setup

@@
 \begin{abstract}
-Humans have a tendency to place a `human' like quality in the objects around them.
-Giving your car a name or talking to your pet, such that you believe they understand you in a human way.
-This behavior is also seeing traction in Machine Learning (ML), where the idea of sentience is being placed upon Large Language Models (LLMs).
-In this position paper, we discuss why humans invariably anthropomorphize onto various objects and how the search for Artificial General Intelligence (AGI) is pushing recent literature to over-emphasize sentience.
-To combat this, we provide a short study that investigates the latent space of models, demonstrating that the discovery of patterns in latent spaces should not be a surprising outcome.
-Finally, we discuss how the popularity in media and the race for AGI enforces over-interpretation of consciousness in the models we use.
-
+\input{abstract_econdat.tex}
 \end{abstract}
 
-\section{Introduction}
-\label{submission}
-
-% \textcolor{magenta}{TODOs for the team:
-% \begin{itemize}
-%     \item Please double check the citations.
-Note this style file does not retain capitalization in titles.
-A trick to retain capitalization is to put an extra set of brackets around the title.
-I did this for the Ladouceur et al.
-and Bubeck et al references already; please make this consistent throughout.
-%     \item As another issue with the citations: note that arxiv preprints do not show very well.
-Best way to make them show is to wrap them as 'journal articles' as done in the Arora citation.
-Please straighten this out for the other articles too
-% \end{itemize}
-% }
-
-In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\cite{merton1942science}.
-While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.
-
-At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening.
-Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\cite{weissburg2024tweets}.
-Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
-
-Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence.
-In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal.
-The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}.
-In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.
-
-For example, Google DeepMind claimed their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
-This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training.
-However, as noted by Hector Zenil \cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s.
-Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system.
-Therefore, despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.
-
-Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}.
-Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.
-
-When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
-AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}.
-As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
-In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.
-
-Noticing these trends, we as the authors of this article are concerned.
-We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding.
-At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
-With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
-In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believe this.
-
-To strengthen our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We work out different experiments which may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made.
-Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
-Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.
-
-%Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
-As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
-With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
-
-%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}.
-This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com/wesg52/status/1709551516577902782?s=20} (formally Twitter) platform.
-Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
-
-%This only gets amplified as the interest in Artificial General Intelligence has now merged with a push towards building general-purpose tools.
-
-%\textcolor{magenta}{cynthia: I am now moving the texts of Andrew and Antony on AGI here; maybe some tightening can already be done in this.
-Unfortunately no time for thorough editing yet at this stage, but will come back to this later}
-
-%In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g.
-\cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'.
-
-%Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs.
-We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, within AI systems.
-We further display results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.
-
-%As previously mentioned, AGI is the current objective for which the global AI community is shooting for.
-As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts.
-This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds.
-True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.
-
-% Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI.
-This model has the ability to solve complex geometry problems, which they state it does without the need for human demonstrations during training.
-Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI.
-As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer.
-With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.
-
-% However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s.
-Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry.
-They were able to achieve this success due to geometry being an axiomatically closed system.
-Therefore, despite DeepMinds success in creating a powerfully fast Geometry solving machine, it is still far from being an AGI.
-
-%it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g.
-\cite{salles2020anthropomorphism, van2023reclaiming}.
-
-\section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}
-
-In Winter 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formerly Twitter) platform~\cite{gurnee2023languagev1}.
-In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not understand truly understand it.
-
-While the manuscript immediately received public criticism~\cite{marcus2023muddles}, and the subsequent, current manuscript version is more careful with regard to its claims~\cite{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings.
-However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation.
-After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.
-
-%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}.
-This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform.
-Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.
-
-%They claim said models generate an internal representation of the world, with which spatiotemporal representations are created.
-Further to this we see claim to the generalization of tasks, in which the model can answer questions regarding spatial and temporal relationships, without the need for further training.
-Although these can be seen as building blocks for a more comprehensive model, it is also fair to assume that it is mere conjecture and more likely spurious correlations.
-
-%In this section, we navigate a number of simple examples to support our claim, that observing patterns in latent spaces should not be a surprising revelation.
-%To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples. 
-%We will draw on ideas from the fields of Economics and Finance, to illustrate further that this phenomenon is neither surprising nor unique to the field of computer science.
-We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data.
-The final example in Section \ref{ex-llm} is more complex and closer in spirit to the experiments conducted by \citet{gurnee2023language}.
-As we progress, we will discuss both the benefits and potential pitfalls of finding patterns in latent spaces.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\textwidth]{results/figures/map.png}
-
-}
-
-\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}
-
-\end{figure}
-
-\subsection{A Neural Networks Born with World Models?}\label{example-deep-learning}
-
-\citet{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places.
-Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}.
-It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly.
-At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates.
-The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance before being passed through a series of layers which are gradually trained for downstream tasks.
-
-In our first example in this section, we simulate this scenario, stopping short of training the model.
-In particular, we take the \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv} that was used in \citet{gurnee2023languagev2}, which maps locations/areas to their latitude and longitude.
-For each place, it also indicates the corresponding country.
-From this, we take the subset that contains countries that are currently part of the top 10 \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France gets 2, \ldots{}).
-
-Subsequently, to ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data, respectively, as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).
-
-Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features.
-Additionally, we add a large number of random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates.
-Let \(d\) denote the final number of features, i.e.~\(d=m+k\) where \(k\) is the number of random features.
-
-We then initialize a small neural network, which we consider as a \emph{projector}, that maps from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there to a lower-dimensional output space.
-Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) through the \emph{projector} and retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
-
-Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\) where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
-A hold-out set is reserved for testing.
-Finally, we compute the predicted coordinates for each sample in the hold-out set as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on a world map (see Figure~\ref{fig-map}).
-
-While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand.
-Thus, with this simple example, we illustrate that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
-
-Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
-In fact, it is well-known that random projections can be used for prediction tasks \cite{dasgupta2013experiments}.
-
-\begin{figure*}[htp]
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/pca.png}
-
-}
-
-\caption{\label{fig-pca}The first two principal components of US Treasury yields over time at daily frequency.
-Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}
-
-\end{figure*}
-
-\subsection{PCA as a Yield Curve Interpreter}\label{example-principal-component-analysis}
-%\subsection{Example: Principal Component Analysis}\label{example-principal-component-analysis}
-
-We now move to a concrete application domain: Economics.
-Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
-%It plots the yields of bonds against their maturities.
-The yield curve's slope is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.
-%While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone.
-To leverage this information in downstream modelling tasks, economists regularly use PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data (e.g.\ \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing}).
-
-%Being primarily interested in understanding the nature of this low-dimensional projection, we limit our scope to the US Treasury yield curve. To this end, we
-To understand the nature of this low-dimensional projection, we collect daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury.
-Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\).
-Let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.
-
-The top panel in Figure~\ref{fig-pca} shows the first two principal components of the yield curves of US Treasury bonds over time.
-Vertical stalks indicate key dates related to the Global Financial Crisis (GFC).
-During its onset, on 27 February 2007, financial markets were in turmoil following a warning from the Federal Reserve (Fed) that the US economy was at risk of a recession.
-The Fed later reacted to mounting economic pressures by gradually reducing short-term interest rates to unprecedented lows.
-Consequently, the average level of yields decreased and the curve steepened.
-Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively.
-In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve.
-
-%Of course, even though it is reassuring to see that these latent embeddings are meaningful, none of this is surprising at all: 
-Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/dl.png}[tb]
-
-}
-
-\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-and fitted values from the autoencoder model. The right chart shows the
-observed average level and spread of the yield curve (solid) along with
-the predicted values (in-sample) from the linear probe based on the latent embeddings
-(dashed).}
-
-\end{figure*}%
-%\subsection{Example: Autoencoder}\label{example-deep-learning}
-\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
-
-%So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep neural network.
-
-Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks.
-In particular, we will now use a neural network architecture.
-%, that allows us to recover a compact latent representation of the yield curve.
-
-\subsubsection{Data}\label{data}
-
-To estimate economic growth, we will rely on a quarterly
-\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
-The data arrives in terms of levels of real GDP.
-In order to estimate growth, we transform the data using log differences.
-Since
-our yield curve data is daily, we aggregate it to the
-quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}.
-Since COVID-19 was a substantial structural break in the time series, we also filtered out all observations after 2018.
-
-\subsubsection{Model}\label{model}
-
-%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
-Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
-Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture. With this
-Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
-Finally, we let \(\theta\) denote our model parameters.
-Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture: 
-The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
-The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
-The output layer consists of a single neuron for our output variable, \(g_t\).
-We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
-
-The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-model.
-The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
-Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
-As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
-
-\subsubsection{Linear Probe}\label{linear-probe}
-
-While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
-On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
-Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
-We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
-Next, we use a linear probe to regress the observed
-yield curve factors on the latent embeddings.
-Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
-Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
-: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
-
-The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
-Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
-We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-%As in all of the other examples we have presented so, we do not find this result particularly surprising.
-
-Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
-To us, that would be too big of a statement.
-Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-
-For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
-To demonstrate this, let us consider a simple linear regression model for GDP growth.
-We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
-As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-But this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
-
-We e.g.\ already mentioned the relationship between interest rates and economic growth.
-To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
-While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
-This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
-Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
-
-Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
-In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
-The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
-The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth, has changed quite a bit.
-While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
-
-\begin{table}
-\caption{Regression output for various models.}\label{tab-reg}%
-\input{results/regression.tex}
-\end{table}
-
-%\subsection{Example: Large Language Model}\label{ex-llm}
-\subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}
-
-So far we have considered simple linear transformations of data.
-One might argue that the previous examples do not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep-learning-based LLM.
-
-Here, we closely follow the approach in \citet{gurnee2023languagev2} but apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset~\cite{shah2023trillion}.
-This dataset contains a curated selection of sentences formulated by central bankers of the US Federal Reserve and communicated to the public in speeches, meeting minutes and press conferences.
-The authors of the paper use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
-To this end, they first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task.
-Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.
-
-To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'.
-In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
-The authors then go on to show that this measure correlates with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases as central bankers react by raising interest rates to bring inflation back to target.
-
-\subsubsection{Linear Probes}\label{linear-probes}
-%Instead of computing a measure based on predicted labels,
-We now use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators.
-Therefore, we further pre-process the data provided by \citet{shah2023trillion} and use their proposed model to compute activations of the hidden state, on the first entity token for each layer.
-We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.
-
-For each layer, we compute linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.
-To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels.
-To measure price inflatio,n we use percentage changes proxied by log differences.
-
-To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations each time.
-In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
-%We found that not doing so would generally lead to poor out-of-sample performance.
-To account for stochasticity, we use an expanding window scheme with 5 folds for each indicator and layer.
-To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.
-
-Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer.
-The values correspond to averages computed across cross-validation folds.
-Consistent with findings in related work \citep{alain2018understanding,gurnee2023languagev2}, we also observe that model performance tends to be higher for layers near the end of the transformer model.
-Curiously, for yields at longer maturities, we find that performance eventually deteriorates for the very final layers.
-We do not observe this for the training data, so we attribute this to overfitting.
-
-% After all we are still regressing a single variable of interest on 128 regressors.
-
-It should also be noted that performance improvements are generally of small magnitude.
-Still, the overall qualitative findings are in line with expectations.
-Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with the outcome of interest and achieve higher mean directional accuracy (MDA).
-Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.
-
-To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models.
-For each layer, indicator and cross-validation fold we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10.
-These are not state-of-the-art forecasting models, but they serve as a reasonable baseline.
-We find that for most indicators, probe predictions outperform the baseline in terms of average performance measures.
-After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform.
-Detailed results will be made available in a supplementary appendix.
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/rmse_pca_128.png}
-
-}
-
-\caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators.
-The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}
-
-\end{figure*}%
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/attack_all_measures.png}
-
-}
-
-\caption{\label{fig-attack}Probe predictions for sentences about inflation of prices (IP), deflation of prices (DP), inflation of birds (IB) and deflation of birds (DB).
-The vertical axis shows predicted inflation levels subtracted by the average predicted value of the probe for random noise.}
-
-\end{figure*}%
-
-%\subsubsection{Stochastic Parrots AfterAll?}\label{stochastic-parrots-after-all}
-\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
-%The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive:
-Even though \emph{FOMC-RoBERTa} (which is substantially smaller than the models tested in~\citet{gurnee2023languagev2} was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).
-So, have we uncovered further evidence that LLMs ``aren't mere stochastic
-parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-of the economy just by `reading' central bank communications? Should the FOMC use this LLM to direct their forward guidance?
-
-% ICML sentence: Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?
-
-%It is worth pointing out here that this model is substantially smaller than the models tested in \citet{gurnee2023language}. This may tempt us to ask the following questions:
-
-% \begin{quote}
-% Have we uncovered further evidence that LLMs ``aren't mere stochastic
-% parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
-% of the economy just by `reading' central bank communications?
-% \end{quote}
-
-We are having a very hard time believing that the answer to either of these questions is `yes'.
-To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor.
-The counter-example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic `understanding' of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.
-
-To test this, we select the best-performing probe trained on the final-layer activations for each indicator.
-We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about deflation in the number of birds.
-A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''.
-Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''.
-While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.
-
-% We top this up with five sentences generated through ChatGPT, where we choose those that lead to the highest probe prediction in the direction of interest.
-In other words, we conduct a small manual adversarial attack on the probe.
-
-In light of the encouraging results for the probe in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category.
-If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing.
-More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.
-
-To get to this last point, we also generate many probe predictions for samples of noise.
-Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote the linear probe that maps from the \(k\)-dimensional space spanned by \(k\) first principal components of the final-layer activations to the output variable of interest (CPI growth in this case).
-Then we sample \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\) for \(i \in [1,1000]\) and compute the sample average.
-We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on random noise.
-
-Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:
-
-\begin{proposition}[Parrot
-Test]\protect\hypertarget{prp-line}{}\label{prp-line}
-
-~
-
-\begin{itemize}
-\setlength\itemsep{1px}
-\item
-  \emph{H0 (Null)}: The probe never predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\).
-\item
-  \emph{H1 (Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E}[f(\varepsilon)]\) for sentences related to the outcome of interest \emph{and} those that are independent (i.e.
-sentences in all categories).
-\item
-  \emph{H2 (More than Mere Stochastic Parrots)}: The probe predicts values that are statistically significantly different from \(\mathbb{E} [f(\varepsilon)]\) for sentences that are related to the outcome variable (IP and DP), but not for sentences that are independent of the outcome (IB and DB).
-\end{itemize}
-\end{proposition}
-
-To be clear, if in such a test we did find substantial evidence in favour of rejecting both \emph{HO} and \emph{H1}, this would not automatically imply that \emph{H2} is true.
-But to even continue investigating, if based on having learned meaningful representation the underlying LLM is more than just a parrot, it should be able to pass this simple test.
-
-In this particular case, Figure~\ref{fig-attack} demonstrates that we find some evidence to reject \emph{H0} but not \emph{H1} for \emph{FOMC-RoBERTa}.
-The median linear probe predictions for sentences about inflation and deflation are indeed substantially higher and lower, respectively than for random noise.
-Unfortunately, the same is true for sentences about the inflation and deflation in the number of birds, albeit to a somewhat lower degree.
-This finding holds for both inflation indicators and to a lesser degree also for yields at different maturities, at least qualitatively.
-
-We should note that the number of sentences in each category is very small here (10), so the results in Figure~\ref{fig-attack} cannot be used to establish statistical significance.
-That being said, even a handful of convincing counter-examples should be enough for us to seriously question the claim, that results from linear probes provide evidence in favor of real `understanding'.
-In fact, even a handful of sentences for which any human annotator would easily arrive at the conclusion of independence, a prediction by the probe in either direction casts doubt.
-
-\section{On Human Proneness to Over-Interpretation}
-Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world.
-Thus, to draw conclusions about the emerging capabilities of AI models, we argue that more conservative and rigorous tests are needed.
-
-%that most readers will agree, have certainly not developed an intrinsic "understanding" of the world.
-The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.
-
-%Our examples should not be seen as an attempt to conclusively refute claims, that LLMs can develop on actual "understanding" of the world, by distilling knowledge in more compact representations.
-We hope that they serve to illustrate, however, that the use of diagnostic tools is potentially misleading and most definitely insufficient, in providing evidence in favor of "understanding".
-Linear probes and related tools from mechanistic interpretability, were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}.
-Favourable outcomes from probes merely indicate, that the model "has learned information relevant for the property [of interest]" \citep{belinkov2021probing}.
-Our examples demonstrate, that this is achievable even for small models, that most readers will agree, have certainly not developed an intrinsic "understanding" of the world.
-The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.
-
-Generally, humans are prone to seek patterns everywhere.
-Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.
-the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.
-
-We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems.
-When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, from a human perspective, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities.
-We further elaborate on this by highlighting the risks of finding spurious patterns and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
-%. This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
-
-\subsection{Spurious Relationships}
-In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other.
-Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation.
-Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.
-A major contributor is that humans struggle to tell the difference between random and non-random sequences \cite{falk1997making}, and to generate sequences that appear random \cite{ladouceur1996erroneous}.
-A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random.
-This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.
-
-%Such as an AI system finding patterns in data not originally perceived during training
-%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
-
-%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path.
-Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads.
-Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.
-
-%I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy.
-But I am very skeptical of claims that AI will magically provide that remedy.
-This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them.
-The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots.
-Since this is an opinionated post, I feel that I should start with a few disclaimers:
-
-%\begin{itemize}
-%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
-%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
-%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
-%    \item I therefore do take issue with the way that these findings are being overblown by people with clout.
-Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
-%\end{itemize}
-
-%\section{Background}
-
-%\subsection{Emergence and Stochastic Parrots}
-%Practically speaking, large language models (LLMs) are probabilistic representations of data on which they are trained, aimed at continuing an incomplete block of text.
-It is assumed that e.g.
-words, sentences, paragraphs, punctuation etc.
-appear in human-generated text according to a probability distribution \cite{arora2023theory}.
-Given a set of words, the task of the LLM is to draw upon its representation of human-generated text to continue that set of words: in essence, using the probability of the occurrences of words and an initial block of text to generate more text.
-AI-researchers have compared LLMs by observing whether the text they generate appears to solve tasks\footnote{e.g.
-the \href{https://huggingface.co/spaces/HuggingFaceH4/open_llm_leaderboard}{HuggingFace} LLM leaderboard}.
-Tasks include asking sets of questions across categories e.g.
-health, law, finance, and fine-tuning an LLM to check the answers \cite{lin2021truthfulqa}, asking the LLM to determine which noun a pronoun refers to in sentences where this is ambiguous \cite{sakaguchi2021winogrande}.
-
-%It has been shown that increasing the scale of large language models (LLMs) has been shown to result in an unexpectedly large boost in their scores on these various tasks, including scoring well on entirely new tasks.
-It has been shown that at a certain size, LLMs appear to hit a threshold, after which performance on a broad range of tasks increases sharply \cite{wei2022emergent}.
-Some theories of emergence, inspired by literature on human education, refer to the tendency of the generated text to be the correct response to a given task as a 'skill' \cite{arora2023theory}, inspiring further research to develop more challenging tasks for LLMs to complete as they routinely begin to score higher and higher on common benchmarks \cite{yu2023skill}.
-An explanation offered for the sudden boost in both task performance and the number of new tasks that appear to be solved is 'emergence': unpredictable qualitative differences in behavior of a system that occur with quantitative changes in that system.
-It has been claimed that, as the scale of the LLMs has dramatically increased, they have crossed a threshold that has resulted in a large increase in their abilities, to the point that they demonstrate 'sparks of AGI' \cite{bubeck2023sparks}.
-
-%An alternative explanation is that the LLMs are performing as designed, as stochastic parrots, generating strings probabilistically based on the vast text they have been trained on \cite{bender2021dangers}.
-Although the various tasks used to compare the various LLMs appear to be increasing in complexity, \cite{yu2023skill} acknowledges that higher scores on the tasks can be achieved by training LLMs on the material on which they will be tested ('cramming'), and that there is training-set contamination, i.e.
-that the vast text they are trained on contains similar examples to the tasks on which they are measured.
-Thus, as many of these tasks are derived from tests and textbooks commonly assigned to humans, and as much of the contents of the their training data text is unknown, metrics echo similar issues in human education: often called 'Goodhart's Law', a measure ceases to be a measure when it becomes a target, as is the case when human students taught to score high on their standardized tests to the detriment of developing the abilities the tests are supposed to measure \cite{strathern1997improving}.
-Further, as \citet{bender2021dangers} reviews, the interpretation of text as coherent has no objective test, relying instead on our human interpretation of the probabalistically generated output.
-%Although many deep learning systems' designs were inspired by the anatomy of human neurons, the neuron in computer code does not in fact resemble a human, and as such cannot deliver the cognitive functions of a human \cite{salles2020anthropomorphism, van2023reclaiming}.
-
-%\subsection{Claims of Artificial General Intelligence}
-\subsection{Anthropomorphism}
-
-Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects.
-These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}.
-Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
-%Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review).
-Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
-As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.
-
-Modern theory of anthropomorphism suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing, waytz2010social}.
-Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
-This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}.
-%An illustration is the example of Tom Hank's character in the film Castaway who has conversations with a ball he calls "Wilson".
-Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc.
-with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.
-
-The motivation as a human to be competent (effectance motivation), is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}.
-When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}.
-Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
-an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors.
-This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
-
-These pressures extend to expert AI practitioners and researchers as well as laypersons.
-In both scholarly and commercial fields, AI experts face considerable pressures to demonstrate competence in AI-related work.
-Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall scholarly publications \footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }.
-This is also evident in peer-reviewed publications \cite{Maslej2023-pi}, with both more than doubling from 2010 to 2020.
-The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
-Outside of academia, the number of jobs requiring AI expertise have increased several fold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}.
-Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.
-
-\subsection{Confirmation Bias}
-
-Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \cite{nickerson1998confirmation}.
-%Importantly, it need not be fully conscious. Rather,
-Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief.
-Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek
-%, attend to, or interpret
-evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.
-
-Hypotheses in present-day AI research are often implicit.
-Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems.
-Where other fields, such as quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
-This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
-% It is also not fully articulated as in other fields, but instead assigns specific conditions and in some instances numerical values or effect sizes to each competing hypothesis.
-%Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis.
-However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-%However with regards to claims that outputs of a system hint towards emerging qualities like AGI, interpretations of the output necessarily become more explicit, even though the hypotheses were never fully articulated: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
-
-Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \citet{nickerson1998confirmation} reviews.
-Scientists may pay little to no attention to competing hypotheses or explanations, e.g.
-only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false.
-Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a 'weak' or 'non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}.
-In extreme cases, propositions may be made that cannot be falsified based on how they are formulated.
-If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question.
-Thus, one is far more likely to show evidence in favor of their beliefs by posing weak null hypotheses.
-Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.
-As \citet{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.
-
-%Further, they are under similar pressure as experts to understand and be able to use it.
-One such situation is the increasing demand for `AI' skills among AI non-experts.
-For example, job postings for individuals with skills in `GenAI' have increased dramatically since the release of ChatGPT\footnote{Retrieved 19/01/2023 from https://lightcast.io/resources/blog/generative-ai-10-19-2023}, and individuals who are in a position to adopt to the technology are projected to benefit \footnote{Retrieved 19/01/23 from https://www.imf.org/en/Publications/Staff-Discussion-Notes/Issues/2024/01/14/Gen-AI-Artificial-Intelligence-and-the-Future-of-Work-542379}.
-Thus, individuals in advanced economies may experience a pressure to rapidly learn how to effectively use and understand an opaque technology, that appears to behave human-like in at least some ways and by imbuing the technology with humanity, may in turn ease their learning curve.
-
-\section{Conclusion}
-As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests.
-Even academic researchers are not free from such biases.
-
-As a consequence, we call for the community to create explicit room for organized skepticism.
-For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
-If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond the display of surprise.
-Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.
-
-In this, as the quest for AGI is so deeply rooted in human-like recognition, we also want to put an explicit word of warning about the use of terminology.
-Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
-
-In fictional media, for a long time, depictions of highly intelligent AI have been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, it was shown that 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those 74 had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
-They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
-
-This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\cite{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chat-bots~\cite{replika,nytimesailove}.
-The same news outlets discuss the human-like responses of Bing\footnote{Microsoft search engine \url{https://www.bing.com/}}, which had at that point, recently been infused with GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
-The article~\cite{nytimesbing}, states ``As if Bing wasn’t becoming human enough'' and goes on to claim it told them it loves them.
-Here, AI experts and influencers also have considerable influence on how the narrative unfolds on social media: according to~\citet{neri2020role}, actual AI-related harms did not trigger viral amplification e.g.\ the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
-Rather, potential risks expressed by someone perceived as having expertise and authority were amplified, such as Stephen Hawking's statements during an interview in 2014.
-
-Thus, we as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed.
-We call upon our colleagues to be explicitly mindful of this.
-As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
-
-% \section{Lessons and calls for the research community}
-
-% \begin{itemize}
-% \setlength\itemsep{1px}
-%     \item be aware over-claiming is both really attractive, and we as humans (even academics) are prone to it
-%     \item create explicit room for organized skepticism, especially as stakes are high.
-e.g.\ red teams and more severe tests, \textcolor{red}{andrew:clearly articulate hypotheses, including strong null hypotheses}
-%     \item take responsibility wrt the media narrative.
-Note that the public may have very different interpretations and associations on words we have specifically defined within the community, but that have anthropormophic connotation (e.g.\ emergence, intelligence, learning).
-Scientific outcomes should not be PR press releases \textcolor{magenta}{cynthia: have there been examples of this wrt AI/AGI in the last year?
-So that press releases with big claims on academic work got out?
-Here, I also would ideally like those examples of scifi portrayal, and anthrophomorphic discussions that imply terminator-like scenarios.
-We can at least criticize the 'pause AI' letter of last year March, that did speak of 'minds'}
-%     \item cool down the hype and uphold the Mertonian norms. Our legitimacy as academics deeply roots into this.
-% \end{itemize}
-
-% \subsection{AI in the Media}
-
-% Lay-persons consume AI-related content in both traditional and social media, yet lack direct experience building and evaluating it.
-This suggests that lay-person impressions are likely to be formed by the content ingested, mediated by experts on how AI is unfolding in the real world on social media, and fictional representations in movies, TV and other traditional media.
-Analyzing Twitter data from 2007 to 2018, it was observed that public interest in AI dramatically increased after the first quarter of 2016 \cite{neri2020role}.
-Actual AI-related harms did not trigger viral amplification e.g.
-the death of an individual dying while a Tesla car was in autopilot, or the financial bankruptcy of a firm using AI technology to execute stock trades.
-Rather, potential risks expressed by Stephen Hawking during an interview in 2014 were far more likely to be expressed, and to be present in links that were shared.
-\citet{neri2020role} note that the 88\% of tweets classified as expressing risk described future risks rather than current and near-future harms e.g.
-\cite{solaiman2023evaluating}, and were greatly amplified when expressed by someone perceived as having expertise and authority.
-
-% Impressions are also formed by the fictional content we consume, which frequently depicts AI to be highly intelligent.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, it was shown that 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those 74 had sufficient intelligence to interact at an almost human-level \cite{saffari2021does}.
-The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI-agents in movies.
-They further show a common plot: humans perceive the AI-agents as inferior despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \cite{saffari2021does, neri2020role}.
+\input{body_econdat.tex}
 
@@
 \bibliographystyle{icml2024}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% APPENDIX
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\appendix
+\input{appendix_econdat.tex}
 

### 2024-02-06 abstract_econdat.tex
improvement to tex setup

--- 2024-02-06 abstract.tex
@@
 In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
+We focus on example applications in the field of Economics and Finance, discussing both the benefits and pitfalls of common approaches to mechanistic interpretability.
+Our most extensive example involves a novel financial dataset and LLM that have been proposed for quantifying the `hawkishness' of central bank communication.
+We demonstrate that even though standard tools from mechanistic interpretability yield favourable outcomes for the model, it still evidently lacks any sort of `understanding' of the economy.
 Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.

### 2024-02-06 appendix_econdat.tex
improvement to tex setup

--- 2024-02-06 appendix.tex
@@
 The single hidden layer of the untrained neural network has \(400\) neurons.
-
-%\subsection{Example: Autoencoder}\label{example-deep-learning}
-\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
-
-%So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep neural network.
-
-This is an additional example that we have not discuss in the body of the paper.
-Here, we build forth on an application in Economics.
-However, we now seek to not only predict economic growth from the yield curve, but also extract meaningful features for downstream inference tasks.
-For this, we will use a neural network architecture.
-%, that allows us to recover a compact latent representation of the yield curve.
-
-\subsubsection{Data}\label{data}
-
-To estimate economic growth, we will rely on a quarterly
-\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
-The data arrives in terms of levels of real GDP.
-In order to estimate growth, we transform the data using log differences.
-Since
-our yield curve data is daily, we aggregate it to the
-quarterly frequency by taking averages of daily yields for each maturity.
-We also standardize yields since deep learning models tend to perform better with standardized data \citep{gal2019standardization}.
-Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
-
-\subsubsection{Model}\label{model}
-
-%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
-Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
-Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture. With this
-Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
-Finally, we let \(\theta\) denote our model parameters.
-Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture: 
-The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
-The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
-The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
-The output layer consists of a single neuron for our output variable, \(g_t\).
-We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
-
-The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
-model.
-The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
-Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
-As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
-Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
-
-\subsubsection{Linear Probe}\label{linear-probe}
-
-\begin{figure*}
-
-\centering{
-
-\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
-
-}
-
-\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
-and fitted values from the autoencoder model. The right chart shows the
-observed average level and spread of the yield curve (solid) along with
-the predicted values (in-sample) from the linear probe based on the latent embeddings
-(dashed).}
-
-\end{figure*}%
-
-While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
-On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
-Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
-We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
-Next, we use a linear probe to regress the observed
-yield curve factors on the latent embeddings.
-Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
-Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
-parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
-: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
-
-The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
-Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
-We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
-%As in all of the other examples we have presented so, we do not find this result particularly surprising.
-
-Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
-To us, that would be too big of a statement.
-Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-%It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
-But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
-While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-%\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
-For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
-To demonstrate this, let us consider a simple linear regression model for GDP growth.
-We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
-As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
-However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
-
-% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
-Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
-
-We e.g.\ already mentioned the relationship between interest rates and economic growth.
-To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
-While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
-This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
-Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
-
-Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
-In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
-The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
-The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
-While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
-
-\begin{table}
-\caption{Regression output for various models.}\label{tab-reg}%
-\input{results/regression.tex}
-\end{table}
 

### 2024-02-06 body_econdat.tex
improvement to tex setup

--- 2024-02-06 body.tex
@@
 Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.
+
+\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
+
+Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks.
+In particular, we will now use a neural network architecture.
+
+\subsubsection{Data}\label{data}
+
+To estimate economic growth, we will rely on a quarterly \href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis.
+The data arrives in terms of levels of real GDP.
+In order to estimate growth, we transform the data using log differences.
+Since our yield curve data is daily, we aggregate it to the quarterly frequency by taking averages of daily yields for each maturity.
+We also standardize yields since deep learning models tend to perform better with standardized data \citep{gal2019standardization}.
+Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.
+
+\subsubsection{Model}\label{model}
+
+Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
+Finally, we let \(\theta\) denote our model parameters.
+Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).
+
+The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
+The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons.
+The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data.
+The output layer consists of a single neuron for our output variable, \(g_t\).
+We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.
+
+The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
+model.
+The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected.
+Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set.
+As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread.
+Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.
+
+\subsubsection{Linear Probe}\label{linear-probe}
+
+\begin{figure*}
+
+\centering{
+
+\includegraphics[width=1.0\textwidth]{results/figures/dl.png}
+
+}
+
+\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
+and fitted values from the autoencoder model. The right chart shows the
+observed average level and spread of the yield curve (solid) along with
+the predicted values (in-sample) from the linear probe based on the latent embeddings
+(dashed).}
+
+\end{figure*}%
+
+While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}.
+On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there.
+Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.
+We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\).
+Next, we use a linear probe to regress the observed
+yield curve factors on the latent embeddings.
+Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively.
+Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
+parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
+: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).
+
+The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}.
+Solid lines show the observed yield curve factors over time, while dashed lines show predicted values.
+We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread.
+%As in all of the other examples we have presented so, we do not find this result particularly surprising.
+
+Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve?
+To us, that would be too big of a statement.
+Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
+For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
+To demonstrate this, let us consider a simple linear regression model for GDP growth.
+We might be interested in understanding to what degree economic growth in the past is associated with economic growth today.
+As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient.
+However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
+
+We e.g.\ already mentioned the relationship between interest rates and economic growth.
+To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors.
+While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}).
+This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth.
+Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.
+
+Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model.
+In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings.
+The estimated coefficient on this latent factor is small in magnitude, but statistically significant.
+The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth has changed quite a bit.
+While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.
+
+\begin{table}
+\caption{Regression output for various models.}\label{tab-reg}%
+\input{results/regression.tex}
+\end{table}
 

### 2024-02-06 body_econdat.tex
last update econdat

@@
 
+So far, we considered simple linear data transformations.
+One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning.
 Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks.
@@
 
-So far, we considered simple linear data transformations.
-One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning.
-In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks.
-Here, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.
+Next, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.
 

### 2024-02-06 appendix.tex
removed spaces from file names for arxiv

@@
 
-%\subsection{Example: Autoencoder}\label{example-deep-learning}
 \subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}
-
-%So far we have considered simple linear transformations of data.
-One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning.
-Our next example will therefore involve a deep neural network.
 
@@
 For this, we will use a neural network architecture.
-%, that allows us to recover a compact latent representation of the yield curve.
 
@@
 
-%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\).
-Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\).
-Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).
-
-%To do this, we use a simple autoencoder architecture. With this
 Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\).
@@
 
-%To do this, we use a simple autoencoder architecture: 
 The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function.
@@
 Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.
-%It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data.
-But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve.
-While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.
-%\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}
 For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model.
@@
 However, this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.
-
-% A less risky/strong term than confounding bias that we could use here is omitted variable bias.
-Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI (n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI_(n_pc=128).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI (n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI_(n_pc=128).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Mo) (n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(1Mo)_(n_pc=128).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (1 Yr) (n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(1Yr)_(n_pc=128).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST (10 Yr) (n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(10Yr)_(n_pc=128).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI (n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI_(n_pc=128).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI (n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI_(n_pc=128).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Mo).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(1Mo).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (1 Yr).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(1Yr).png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST (10 Yr).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(10Yr).png}
 

### 2024-02-06 src/probe_llm/experiments.jl

@@ -64,9 +64,10 @@ if PLOT_PROBES
     for g in gdf
         g = DataFrame(g)
         i = g.indicator[1]
-        m = g.maturity[1] |> x -> ismissing(x) ? "" : " ($x)"
-        n_pc = g.n_pc[1] |> x -> ismissing(x) ? "" : " (n_pc=$x)"
+        m = g.maturity[1] |> x -> ismissing(x) ? "" : "_($x)"
+        n_pc = g.n_pc[1] |> x -> ismissing(x) ? "" : "_(n_pc=$x)"
         title = "$i$m$n_pc"
+        title = replace(title, " " => "")
         plt = plot_measures_for_ind(g, axis=axis)
         save(joinpath(save_dir, "figures", "measures_$title.png"), plt, px_per_unit=3) 
         plt = plot_measures_for_ind(g, axis=axis, plot_interval=false, models=["y_probe"])

### 2024-02-06 appendix.tex
removed brackets from file names for arxiv

@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI_(n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_CPI_n_pc=128.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI_(n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_PPI_n_pc=128.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(1Mo)_(n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_1Mo_n_pc=128.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(1Yr)_(n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_1Yr_n_pc=128.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_(10Yr)_(n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_probe_UST_10Yr_n_pc=128.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI_(n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_CPI_n_pc=128.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI_(n_pc=128).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_PPI_n_pc=128.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(1Mo).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_1Mo.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(1Yr).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_1Yr.png}
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_(10Yr).png}
+\includegraphics[width=1.0\textwidth]{results/figures/measures_UST_10Yr.png}
 

### 2024-02-06 src/probe_llm/experiments.jl

@@ -64,8 +64,8 @@ if PLOT_PROBES
     for g in gdf
         g = DataFrame(g)
         i = g.indicator[1]
-        m = g.maturity[1] |> x -> ismissing(x) ? "" : "_($x)"
-        n_pc = g.n_pc[1] |> x -> ismissing(x) ? "" : "_(n_pc=$x)"
+        m = g.maturity[1] |> x -> ismissing(x) ? "" : "_$x"
+        n_pc = g.n_pc[1] |> x -> ismissing(x) ? "" : "_n_pc=$x"
         title = "$i$m$n_pc"
         title = replace(title, " " => "")
         plt = plot_measures_for_ind(g, axis=axis)

### 2024-03-27 dev/rebuttal/detailed_rebuttal.md
added detailed rebuttal

# Reviewer D4ep

We thank the reviewer for the detailed and thoughtful review.

## Lack of distinct prescription

The main weakness pointed out by the reviewer is that the paper '[...] lacks a distinct prescription beyond a call for AI researchers to be more cautious [...]'. We fall short of doing so in the paper mostly for reasons of scope and scale. Where possible we will add a paragraph either in the body, which is limited by the page limit, or the appendix. In the meantime, we try to answer your specific questions below:

> " [...] if we assume that tearing down the whole publishing incentive structure is beyond the pale, what distinct steps can researchers actually take when experimenting, writing, reviewing and publishing?"

Short of tearing things down entirely, we do think that specific changes to the incentive structure are in order. It has been argued elsewhere that societally impactful scientific insights should be treated as open-source software artifacts ([Liem and Demetriou, 2023](https://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=10173886)). Open review platforms like this one are a step in the right direction but we could go further. In the open-source software development space, those who review a piece of software often end up contributing to it, because they face the right incentives: not only will they most likely be users of that improved piece of software but also receive credit for their contribution (as they should). The idea of moving from authorship to contributorship is not as radical as it may sound ([Smith, 1997](https://www.bmj.com/content/315/7110/696.short)) but thirty years ago one could have argued that we are lacking the right technology for this. Today this argument is much harder to buy in the face of blossoming open-source software ecosystems. Considering that researchers in the computer science field are no strangers to papers with long author lists, we think it is reasonable to assume they would generally welcome such a paradigm shift.

> "Should we be more welcoming of negative results?"

Continuing the argument above, a system that leans more towards contributorship is typically much more welcoming of negative results: software bugs are just part of the process and joint ownership means joint responsibility for positive and negative outcomes.

> "Should we set up more venues specifically devoted to reproducing and analyzing past claims?"

Yes, we think that reproducibility challenges, for example, are a step in the right direction. But many legacy issues associated with treating

> "Should we offer compensation for reviewing?"

Contributorship can provide a natural form of non-pecuniary compensation for reviewing (open-source communities testify to this). As researchers, we are financially compensated for producing research. Perhaps we simply need to get to a point where reviews can lead to (accredited) contributorship. Reviewer awards are a step in the right direction. But given the sheer size of today's venues, they are an improbable, risky outcome for reviewers and hence unlikely to provide strong enough incentives.

> "All these questions have been asked before, but do they become more salient or more urgent with respect to modern evolutions in generative AI?"

Considering that the existing incentive structure has repeatedly moved researchers to employ generative AI to deal with the tasks and pressures they are facing, we think these questions have certainly become more urgent.

# Reviewer jMr8

We thank the reviewer for the detailed and thoughtful review.

## Section 2.1 - Limitations of Inverse Problem

The reviewer pointed out that the "inverse problem is ill-conditioned and could not recover the perfect signal in theory." The authors agree that this can indeed be seen as a form of inverse problem and common caveats such as non-uniqueness and instability apply ([Haltmeier and Nguyen, 2020](https://arxiv.org/abs/2006.03972)). We use $\ell^2$-regularization for the linear probe (ridge regression), which corresponds to Tikhonov regularization in the context of approximation theory ([Björkström, 2001](https://www2.math.su.se/matstat/reports/seriea/2000/rep5/report.pdf)). We confess that we did not carefully consider the parameter choice for the ridge penalty, nor has this been carefully studied in the broader literature on linear probes for mechanistic interpretability to the best of our knowledge.
Indeed we would expect that the signal would not perfectly be recovered in theory given the compression of the data. However, some signal would indeed imperfectly be recovered, with a closer approximation of perfection in recovery the 'better' the model is. Given that LLMs are essentially compressing textual information, and that LLMs have demonstrated a high capability of doing so, we of course expect an inverse attempt to recover the signal to be partially successful. What is not to be expected is an interpretation that this imperfect yet-close-enough-to-accurate signal recovery is interpreted as the system having a 'mental model' of the earth, and thus showing 'sparks' of general intelligence - especially when 'general intelligence' and even 'intelligence' are narrowly defined in this space, and when there are ideal conditions for misinterpretation of results in the form of anthropomorphism and cognitive bias. 
Thus, we still think that this simple experiment in Section 2.1 describes the problem of running linear probes on LLMs reasonably well at a high level. If anything, the challenges pointed out by the reviewer therefore cast further doubt on this practice. We will add a note or paragraph on this in the final manuscript. Furthermore, we would like to point out that even in the case of imperfect yet-close-enough-to-accurate signal recovery, this is not to be interpreted as the system having a 'mental model' of the earth, and thus showing 'sparks' of general intelligence - especially when 'general intelligence' and even 'intelligence' are narrowly defined in this space, and when there are ideal conditions for misinterpretation of results in the form of anthropomorphism and cognitive bias.

## Section 2.3.2. - The Parrot Test

Section 2.3.2. applied a 'strong' test to the hypothesis that LLMs 'understand'. We show results that one might interpret as the models having some understanding of economics, and being able to imperfectly forecast economic indicators - a parallel to the notion that model output correlating with latitude longitude is an indication that the model ‘understands’ geography, or has a ‘mental’ model of the globe. 
The 'spurious' correlation then, is in the interpretation that the 'accuracy' of the output of the model correlating with economic indicators is a cue that the model is understanding - or more specifically that the correlation between the output of the model and the 'ground truth' economic indicators is 'caused' by the model's understanding of the text. 
We then show that the model fails to 'understand' when we apply a simple linguistic transformation to our experiment. 
## Section 3.2 - Anthropomorphism and AI
Your recommendation that we better ground sec 3.2 in the current research space is well-noted. Reflecting on your comment, we think there are at least 3 ways to reach a better understanding:
Acknowledgement: researchers should consider acknowledging our tendency to anthropomorphize either in a dedicated ‘limitations’ section or when discussing results.
Stronger testing: researchers should refrain from drawing premature conclusions about AGI, unless these conclusions are based on strong hypothesis tests.
Epistemologically robust standards: we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and iterations over how we will measure them
We feel an appropriate discussion would require more room than is allocated in the paper. Nevertheless, we can attempt to summarize our position in the main body or perhaps in the appendix.
For a slightly more detailed discussion, please see our comments below:
A crucial first step is to acknowledge openly that such cognitive bias and tendency towards anthropomorphism exists, and is especially likely in this space. While it may remain in the language that we use (e.g. phrases like the model ‘thinks’, ‘understands’, etc.), behaviorally it must be addressed in the research space, e.g. with acknowledgment in ‘limitations’ sections of papers, and in the interpretation of results in our discussion sections. This may lead to more broad attempts to address anthropomorphism, e.g. with better research designs and analyses.
A second step, and one such research design element, is stronger testing which we attempt in this paper. Intelligence is only ‘indirectly’ measurable in automated systems as it is in humans. Thus we must also acknowledge that claims of ‘AGI’ deserve scrutiny. If such a claim is easily negated with a simple change in language, we take this as a sign that the field is only weakly testing its claims. We note that this does not mean that researchers are not making great efforts to test their models - leaderboards on Huggingface, and the limitless torrent of papers on LLMs would speak to the contrary. However, these tests may not be appropriate indicators that a system is approaching AGI, but rather that a system is performing well on some narrowly defined task and accompanying dataset. Thus, if we are to approach AGI with our work, we must have a strong test - a risky test that a system is likely to fail. In fields like psychology where the concept of intelligence itself emerged, discussions of appropriately strong tests for claims abound - we require the same rigor in this field as well. 
Thirdly, and more broadly, we see a need for more epistemologically robust standards in the work itself to combat anthropomorphism. To this end, further inspiration may be drawn from psychology - ‘construct’ oriented definitions of important concepts, and a focus on achieving better indirect measurement. While computer science has traditionally not been a hypothesis-driven field that deals with indirect and imperfect measurements, psychology research methods have evolved specifically for these purposes. Applying this to the study of AGI requires first a somewhat-agreed upon definition not only on what ‘AGI’ is, but also how it would manifest itself - with the understanding that it may appear differently than intelligence does in humans. 
Computer science tends to borrow heavily from the rather vague definition of Spearman that it is about ‘generally solving problems (e.g. [Goertzel, 2014](https://sciendo.com/article/10.2478/jagi-2014-0001)) - if we wish to take advantage of this definition, we might attempt to already claim that the LLMs achieve this: the models weren’t designed for any specific task, and yet can be used to assist in many tasks. At the same time, our work shows that this does not mean that these models ‘understand’ the problems they are solving, nor can they formulate the problems in the first place. If our current assessments are all inspired by how we measure intelligence in humans, and models are trained on the material of the tests and appropriate responses, assessing ‘AGI’ in this way is simply inappropriate. We don’t have a good answer to this last point - but are excited to consider what a field that acknowledges its own biases can accomplish in this regard.

## Other Modalities

From what we have observed in the past, we think the concerns we have raised may apply to many applications, but are urgent in the context of LLMs. Many people attribute our ability to understand, to our ability to communicate through language. Recent implementations as chat bots, e.g. Replika and Chat GPT, as well as virtual assistants (Alexa, Cortana) have afforded a kind of interaction that resembles linguistic exchanges with other humans. We can certainly imagine there is some degree of anthropomorphizing with any ‘AI’ technology that has some apparent behavior e.g. self-driving cars. But as language is an all too obvious feature distinguishing us from other species, it is not surprising that LLMs have so far been anthropomorphized more so than foundation models with other modalities - language is what makes them an ingredient in the ‘perfect storm’.

# Reviewer AGDV

We thank the reviewer for the detailed and thoughtful review.

## Revised Abstract

The reviewer pointed out that the abstract was not optimally aligned with the paper and we agree that there is some room for improvement. We propose this more detailed and longer revised version:

“Developments in the field of AI in general, and Large Language Models (LLMs) in particular, have created a ‘perfect storm’ for observing ‘sparks’ of Artificial General Intelligence (AGI) that are spurious. Like simpler models, LLMs distill representations in their latent embeddings that have been shown to correlate with meaningful phenomena. Nonetheless, the correlation of such representations has often been linked to human-like intelligence in the latter but not the former. We probe models of varying degrees of sophistication including random projections, matrix decompositions, deep autoencoders and transformers: all of them successfully distill knowledge and yet none of them develop true understanding. Specifically, we show that embeddings of a language model fine-tuned on central bank communications can make meaningful predictions, via correlations with unseen economic variables, such as price inflation. However, we then show that inflation is also predicted for nonsense prompts about growing and shrinking bird populations (‘dovelation’). We therefore argue that patterns in latent spaces are spurious sparks of (AGI). Additionally, we review literature from the social sciences that shows that humans are prone to seek patterns and anthropomorphize. We, therefore, argue that both the methodological setup and common public image of AI are ideal for the misinterpretation that correlations between model representations and some variables of interest are 'caused' by the model's understanding of underlying ‘ground truth’ relationships. We therefore call for the academic community to exercise extra caution, and to be keenly aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.”

## Cognitive Biases

While we appreciate the feedback, we would like some clarity on what precisely this reviewer means by ‘a bit poorly argued’? We do indeed see that 366-371 can be cut. We may have included details that summarize more of the material than is essential to the discussion, and perhaps trimming would make our arguments more clear. Nevertheless, we feel the included text summarizes work that has shown antecedents to two forms of cognitive bias. These sections aim to show how the presence of the antecedents makes the current situation with ‘AI’ tools in general and with LLMs in specific, a perfect storm for misinterpretation. We would happily take on any suggestions as to how this can be clearer and better argued.

## Figure 3

We apologize for the somewhat sloppy figure description and appreciate the reviewer’s suggestions. The solid line is just a smooth trendline, which we added to highlight that we find the expected negative relationship between estimated errors and network depth. On second thought, and in light of your feedback, we will simply remove these trend lines to avoid confusion. We will adjust the figure caption as follows: “Out-of-sample root mean squared error (RMSE) for the linear probe plotted against FOMC-RoBERTa’s n-th layer for different indicators. The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series. As expected, we observe that model performance tends to be higher (average prediction errors are lower) for layers near the end of the transformer model.”

Although the errors are indeed low compared to baseline autoregressive models (as we explain in lines 237-249), this chart is not the most obvious choice to drive home that particular point. We chose to highlight this chart nonetheless because it is more consistent with the presentations chosen in the related literature on mechanistic interpretability. It is common to show that probe performance improves for layer activations near the output layer, which illustrates that “neural networks are really about distilling computationally useful *representations*” as opposed to *information contents* ([Alain and Bengio, 2018](https://arxiv.org/abs/1610.01644)).

## Confusing Title

We appreciate this take and understand that this sort of title may not be appealing to everyone. To avoid any confusion about a potential typo, we have addressed this in the revised abstract. We are still keen to keep the title as is unless the reviewer(s) find(s) it critical that we adjust it.

# Reviewer 8XDk

We thank the reviewer for the detailed and thoughtful review.

## Other Works

It was not so much the work by Gurnee and Tegmark itself but the subsequent buzz around it on social media that served as a catalyst for this work. Further, we explain that public perception of the state of AI work has been shown to be mediated via experts (e.g. [Neri et al., 2020](https://link.springer.com/article/10.1007/s00146-019-00924-9). We explain that the situation is ripe for over-interpretation to show that expert perceptions are susceptible to cognitive biases, and have downstream consequences.

The paper itself, in particular the revised version, is much more cautious about drawing premature conclusions than some of the public communications that followed (we cite and refer to both versions explicitly in the paper). Many related works are also free of grandiose conclusions and instead highlight the benefits of mechanistic interpretability that we also acknowledge in our work (e.g. [Nanda et al., 2023](https://arxiv.org/pdf/2309.00941.pdf); [Gurnee et al., 2023](https://arxiv.org/pdf/2305.01610.pdf); [Li et al., 2023](https://arxiv.org/pdf/2210.13382.pdf)). But as we explain in the paper, certain incentives and pressures may lead researchers to blow their scientific findings out of proportion outside of academia itself. The general situation surrounding LLMs makes biased interpretations of results particularly likely. And if experts over-interpret their findings, public perception will be influenced by this as well. This is what we want to caution against and why we call for more explicit acknowledgment of known limitations, stronger tests and more epistemologically robust standards.

Another broadly related field investigates the capacity of LLMs to reason causally. Here, too, there is an opportunity to over-interpret the finding of causal knowledge as causal understanding. Recent work has shown that LLMs can indeed correctly predict causal relationships and this may have practical use cases ([Kıcıman et al., 2023](https://arxiv.org/abs/2305.00050)). But it has also been argued and demonstrated that current LLMs are mere ‘weak causal parrots’ ([Zečević et al., 2023](https://arxiv.org/abs/2308.13067)).

## Guidance from the Social Science(s)

We appreciate the points raised here by the reviewer, in particular, that “the sudden pivot to the social science literature review” caused confusion. Drawing these kinds of connections is always challenging in interdisciplinary works and we will try to address this in the manuscript. In particular, with respect to the experiments, we will make it more clear that:

We would not expect that anyone would anthropomorphize simple random projections and PCA, although both can be used to distill meaningful representations.
And yet, researchers do fall into that trap for more sophisticated models, in particular LLMs.

> “Could you clarify how the social science review provides guidance for the ML community with respect to your specific experiments and more broadly?”

Present work resulted from an interdisciplinary team with backgrounds in the computational and social sciences. By including a social science review we strengthen our argument that the risks of bias and over-interpretation in humans have consistently been evidenced in academic literature (as opposed to merely stating our opinion on the matter).

At the very least, we hope that our review can help researchers in the ML community (ourselves included) to become more aware of our human tendencies to seek patterns and confirmation and to anthropomorphize. While these tendencies are often useful, they can also harm communities, perhaps especially research communities, and bias public perception. Further, we hope that, through acknowledgment that all humans are subject to cognitive bias and that the current situation in AI research in general, and with LLMs in particular, is ripe for at least two forms of bias, a focus can be placed on combating it through more severe tests of AGI-related hypotheses, and research design.

### 2024-05-27 abstract.tex
abstract, title and format template changed

@@
-Humans have a tendency to see `human'-like qualities in objects around them.
-We name our cars, and talk to pets and even household appliances, as if they could understand us as other humans do.
-This behavior, called anthropomorphism, is also seeing traction in Machine Learning (ML), where human-like intelligence is claimed to be perceived in Large Language Models (LLMs).
-In this position paper, considering professional incentives, human biases, and general methodological setups, we discuss how the current search for Artificial General Intelligence (AGI) is a perfect storm for over-attributing human-like qualities to LLMs.
-In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome.
-Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
+Developments in the field of AI in general, and Large Language Models (LLMs) in particular, have created a `perfect storm’ for observing `sparks’ of Artificial General Intelligence (AGI) that are spurious.
+Like simpler models, LLMs distill representations in their latent embeddings that have been shown to correlate with meaningful phenomena.
+Nonetheless, the correlation of such representations has often been linked to human-like intelligence in the latter but not the former.
+We probe models of varying degrees of sophistication including random projections, matrix decompositions, deep autoencoders and transformers: all of them successfully distill knowledge and yet none of them develop true understanding.
+Specifically, we show that embeddings of a language model fine-tuned on central bank communications can make meaningful predictions, via correlations with unseen economic variables, such as price inflation.
+However, we then show that inflation is also predicted for nonsense prompts about growing and shrinking bird populations (`dovelation’).
+We therefore argue that patterns in latent spaces are spurious sparks of (AGI).
+Additionally, we review literature from the social sciences that shows that humans are prone to seek patterns and anthropomorphize.
+We, therefore, argue that both the methodological setup and common public image of AI are ideal for the misinterpretation that correlations between model representations and some variables of interest are 'caused' by the model's understanding of underlying `ground truth’ relationships.
+We, therefore, call for the academic community to exercise extra caution, and to be keenly aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.

### 2024-05-27 main.tex
abstract, title and format template changed

@@
 % Use the following line for the initial blind version submitted for review:
- \usepackage{icml2024}
+%  \usepackage{icml2024}
 
 % If accepted, instead use the following line for the camera-ready submission:
-% \usepackage[accepted]{icml2024}
+\usepackage[accepted]{icml2024}
 
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}
+\icmltitlerunning{Position: Inflated AGI Claims Require Strong Tests and Dovelation}
 
@@
 \twocolumn[
-\icmltitle{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+\icmltitle{Position: Inflated AGI Claims Require Strong Tests and Dovelation \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
 
@@
 
+\section*{Acknowledgements}
+
+\section*{Impact Statement}
+
+This paper presents work whose goal is to advance the field of Machine Learning. There are many potential societal consequences of our work, none which we feel must be specifically highlighted here.
+
 \bibliography{biblio}

### 2024-05-27 body.tex
fixed typo
outlook section added related works section almost done

@@
 
-To strengthen our argument, in this paper, we first consider a recently viral work~\citep{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
-We present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---while at the same time indicating how we feel these claims should \emph{not} be made.
-Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there.
+To strengthen our argument, in this paper, we first present related work in Section~\ref*{related}.
+We then consider a recently viral work~\citep{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made.
+In Section~\ref*{patterns-in-latent-spaces-and-how-to-find-them}, we present several experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected---while at the same time indicating how we feel these claims should \emph{not} be made.
+Furthermore, we present a review of social science findings in Section~\ref*{social} that underline how prone humans are to being enticed by patterns that are not really there.
 Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.
+Finally, in Section~\ref*{outlook}, we propose specific structural and cultural changes to improve the current situation.
+Section~\ref*{conclude} concludes.
+
+\section{Related Work}\label{related}
+
+In this work, we question the practice of using outcomes from mechanistic interpretability to support AGI claims.
+This is not to be seen as criticism toward the underlying methodologies in isolation, but rather in the context of current publishing practices that we repeatedly challenge throughout this work.
+Many closely related works are free of any grandiose conclusions and instead highlight the benefits of mechanistic interpretability that we also highlight in this work~\citep{nanda2023emergent,gurnee2023finding,li2022emergent}.
+
+Another related subfield investigates the capacity of LLMs to reason causally.
+Here, too, there is an opportunity to over-interpret the finding of causal information as causal understanding.
+Recent work has shown, for example, that LLMs can indeed correctly predict causal relationships and this may have practical use cases~\citep{kiciman2023causal}.
+But despite the potential utility, the authors also demonstrate that this capacity can be partially explained by memorization, rather than an actual understanding of causal relationships.
+Similarly,~\citet{zevcevic2023causal} provide empirical evidence indicating that current LLMs ``may talk causality but are not causal''.
+
+Finally, another recent work closely aligned with the position we present here is~\ref*{schaeffer2024emergent}.
 
@@
 
-In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}}
-%(formerly Twitter)
-platform~\citep{gurnee2023languagev1}.
+In 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} platform~\citep{gurnee2023languagev1}.
 Through linear probing experiments, the claim was made that LLMs learned literal maps of the world.
@@
 
-\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}
+\caption{\label{fig-map}Predicted coordinate values (out-of-sample) from a linear probe on final-layer activations of an untrained neural network.}
 
@@
 
-While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, %with this simple example, we
-this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
+While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand.
+Thus, this simple example illustrates that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.
 Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification.
@@
 
-\section{Human Proneness to Over-Interpretation}
+\section{Human Proneness to Over-Interpretation}\label{social}
 
@@
 
-\section{Conclusion}
+\section{Outlook}\label{outlook}
+
+We believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry.
+Our broadest and perhaps most ambitious goal is for our research community to move away from authorship and instead embrace contributorship.
+This argument has been raised long before both in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}.
+Specifically,~\citet{liem2023treat} argue that societally impactful scientific insights should be treated as open-source software artifacts.
+The open-source community sets a positive example of how scientific artifacts should be published in many different ways.
+Not only does it adequately reward small contributions but it also naturally considers negative results (bugs) as part of the scientific process.
+Similarly, code reviews are considered so integral to the process that they typically end up as accredited contributions to projects.
+Open review platforms like OpenReview are a step in the right direction but they still fall short of what is technically feasible.
+Finally, software testing is, of course, not only essential but often obligatory before contributions are accepted and merged.
+As we have pointed out repeatedly in this work, any claims about AGI demand proper strong hypothesis tests.
+It is important to remember that AGI remains the alternative hypothesis and that the burden of proof therefore lies with those making claims.
+
+\section{Conclusion}\label{conclude}
 As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests.

### 2024-05-27 body.tex

@@
 Finally, another recent work closely aligned with the position we present here is~\ref*{schaeffer2024emergent}.
+The authors demonstrate that the apparent emergent abilities of large language models may be driven by a choice of evaluation metrics, rather than some fundamental property that is intrinsic to this family of models.
+Their work highlights the need for rigorous testing and benchmarking of LLMs, which we also point to in this work albeit in a slightly different methodological context.
 

### 2024-05-27 main.tex

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Position: Inflated AGI Claims Require Strong Tests and Dovelation}
+\icmltitlerunning{Position: Stop Making Unscientific Inflated AGI Claims}
 
@@
 \twocolumn[
-\icmltitle{Position: Inflated AGI Claims Require Strong Tests and Dovelation \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+\icmltitle{Position: Stop Making Unscientific Inflated AGI Claims}
 

### 2024-05-27 body.tex
done

--- 2024-05-27 body.tex
@@
 When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
-AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML}.
+AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML,rudin2019stop}.
 As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
@@
 In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believing this.
+\textbf{We therefore urge our fellow researchers to stop making unscientific AGI performance claims}.
 
@@
 
-Finally, another recent work closely aligned with the position we present here is~\ref*{schaeffer2024emergent}.
+Two other recent works are related to this work and align well with the position we present here.~\citet{schaeffer2024emergent} demonstrate that the apparent emergent abilities of large language models may be driven by a choice of evaluation metrics, rather than some fundamental property that is intrinsic to this family of models.
+Their work highlights the need for rigorous testing and benchmarking of LLMs, which we also point to in this work albeit in a slightly different methodological context.~\citet{kloft2024ai} provide experimental evidence demonstrating that people have heightened expectations and a biased, positive view of AI.
+The authors run a user study of human-AI interaction, in which participants performed better at a given task when they thought they were aided by a positively described AI.
+This Placebo effect was found to be robust to negative descriptions of AI.
+Positive bias towards AI may exacerbate other factors that drive people to make unscientific claims about the current state of AI, which we discuss in Section~\ref*{social}.
 
@@
 Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\), where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample.
-A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot these on a world map ( Figure~\ref{fig-map}).
+A hold-out set is reserved for testing, on which we compute predicted coordinates for each sample as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot these on a world map (Figure~\ref{fig-map}).
 
@@
 In fact, it is well-known that random projections can be used for prediction tasks \citep{dasgupta2013experiments}.
+
+This first experiment---and indeed the practice of probing LLMs that have seen vast amounts of data---can be seen as a form of inverse problem and common caveats such as non-uniqueness and instability apply~\citep{altmeier2023regularization}.
+Regularization can help mitigate these caveats~\citep{om2001ridge}, but we confess that we did not carefully consider the parameter choice for $\lambda $, nor has this been carefully studied in the related literature to the best of our knowledge.
 
@@
 
-\includegraphics[width=1.0\textwidth]{results/figures/pca_yield.png}
+\includegraphics[width=0.9\textwidth]{results/figures/pca_yield.png}
 
@@
 To measure price inflation we use percentage changes proxied by log differences.
-To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations.% each time.
+To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations.
 In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
@@
 \caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators.
-The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}
+The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.
+As expected, model performance tends to be higher (average prediction errors are lower) for layers near the end of the transformer model.}
 
@@
 Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future.
-Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \citep{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \citep{foster2009evolution}, conspiracy theories \citep{van2018connecting}, the paranormal \citep{muller2023linking}, gambler's fallacies \citep{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \citep{walker2019finding}.
+Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \citep{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in the power of superstitions \citep{foster2009evolution}, conspiracy theories \citep{van2018connecting}, the paranormal \citep{muller2023linking}, gambler's fallacies \citep{ladouceur1996erroneous} and `pseudo-profound bullshit' \citep{walker2019finding}.
 
@@
 
-We believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry.
+Reflecting on the previous two sections, we make the following concrete recommendations for future research: firstly, researchers should consider acknowledging our tendency to anthropomorphize either in a dedicated ‘limitations’ section or when discussing results (\textit{acknowledgment}); secondly, researchers should refrain from drawing premature conclusions about AGI, unless these conclusions are based on strong hypothesis tests (\textit{stronger testing}); thirdly, we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and iterations over how we will measure them (\textit{epistemologically robust standards}).
+
+Moreover, we believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry.
 Our broadest and perhaps most ambitious goal is for our research community to move away from authorship and instead embrace contributorship.
@@
 \section{Conclusion}\label{conclude}
-As discussed above, AI research and development outcomes can very easily be over-interpreted, both from a data perspective, and because of human biases and interests.
-Even academic researchers are not free from such biases.
+As discussed above, AI research and development outcomes can easily be over-interpreted, both from a data perspective and because of human biases and interests.
+Academic researchers are not free from such biases.
 Thus, we call for the community to create explicit room for organized skepticism.
 
-For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed.
+For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation are needed.
 If claims of human-like or superhuman intelligence are made, these should be subject to severe tests \citep{claesen2022severity} that go beyond the display of surprise.
@@
 As the quest for AGI is so deeply rooted in human-like recognition, adding our voice to emerging calls to be vigilant in communication~\citep{shanahan2024talking}, we put in an explicit word of warning about the use of terminology.
-Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
+Many terms used in current AGI research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons.
 In fictional media, depictions of highly intelligent AI have for long been going around.
-In a unique study of films featuring fictional robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \citep{saffari2021does}.
+In a study of films featuring robots, defined as "...an artificial entity that can sense and act as a result of (real-world or fictional) technology...", in the 134 most highly rated science-fiction movies on IMDB, 74 out of the 108 AI-robots studied had a humanoid shape, and 68 out of those had sufficient intelligence to interact at an almost human-level \citep{saffari2021does}.
 The authors identify human-like communication and the ability to learn as essential abilities in the depiction of AI agents in movies.
 They further show a common plot: humans perceive the AI agents as inferior, despite their possession of self-awareness and the desire to survive, which fuels the central conflict of the film, wherein humanity is threatened by AI superior in both intellect and physical abilities.
-Moreover, it is often noted that experts and fictional content creators interact, informing and inspiring each other \citep{saffari2021does, neri2020role}.
+It is often noted that experts and fictional content creators interact, informing and inspiring each other \citep{saffari2021does, neri2020role}.
 
-This image also permeates present-day non-fictional writings on AI, which often heavily use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\citep{pauseai}).
-News outlets love a story that sells and so we witness examples of humans falling in love with their AI chatbots~\citep{replika,nytimesailove}.
+This image also permeates present-day non-fictional writings on AI, which often use anthropomorphized language (e.g.\ ``ever more powerful digital minds'' in the `Pause Giant AI Experiments' open letter~\citep{pauseai}).
+In the news, we witness examples of humans falling in love with their AI chatbots~\citep{replika,nytimesailove}.
 The same news outlets discuss the human-like responses of Microsoft's Bing search engine, which had at that point recently been adopting GPT-4\footnote{A large multimodal language model from OpenAI \url{https://openai.com/research/gpt-4}}.
@@
 
-\section{Impact Statement}
+\section*{Impact Statement}
 This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.

### 2024-05-27 main.tex
done

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Position: Stop Making Unscientific Inflated AGI Claims}
+\icmltitlerunning{Position: Stop Making Unscientific AGI Performance Claims}
 
@@
 \twocolumn[
-\icmltitle{Position: Stop Making Unscientific Inflated AGI Claims}
+\icmltitle{Position: Stop Making Unscientific AGI Performance Claims}
 
@@
 
-\section*{Impact Statement}
-
-This paper presents work whose goal is to advance the field of Machine Learning. There are many potential societal consequences of our work, none which we feel must be specifically highlighted here.
-
 \bibliography{biblio}

### 2024-05-27 src/probe_llm/utilities/plotting.jl

@@ -108,7 +108,7 @@ function plot_measures(
             :value => "Value",
             col=:indicator,
         )
-        layer = smooth() + visual(Scatter)
+        layer = visual(Scatter)
     end
     plt = draw(
         layer * plt,

### 2024-05-28 abstract.tex
trimmed down abstract slightly

@@
 Developments in the field of AI in general, and Large Language Models (LLMs) in particular, have created a `perfect storm’ for observing `sparks’ of Artificial General Intelligence (AGI) that are spurious.
-Like simpler models, LLMs distill representations in their latent embeddings that have been shown to correlate with meaningful phenomena.
+Like simpler models, LLMs distill meaningful representations in their latent embeddings that have been shown to correlate with external variables.
 Nonetheless, the correlation of such representations has often been linked to human-like intelligence in the latter but not the former.
-We probe models of varying degrees of sophistication including random projections, matrix decompositions, deep autoencoders and transformers: all of them successfully distill knowledge and yet none of them develop true understanding.
-Specifically, we show that embeddings of a language model fine-tuned on central bank communications can make meaningful predictions, via correlations with unseen economic variables, such as price inflation.
-However, we then show that inflation is also predicted for nonsense prompts about growing and shrinking bird populations (`dovelation’).
-We therefore argue that patterns in latent spaces are spurious sparks of (AGI).
+We probe models of varying complexity including random projections, matrix decompositions, deep autoencoders and transformers: all of them successfully distill information that can be used to predict latent or external variables and yet none of them have ever been linked to AGI.
+We, therefore, argue that patterns in latent spaces are spurious sparks of AGI.
 Additionally, we review literature from the social sciences that shows that humans are prone to seek patterns and anthropomorphize.
-We, therefore, argue that both the methodological setup and common public image of AI are ideal for the misinterpretation that correlations between model representations and some variables of interest are 'caused' by the model's understanding of underlying `ground truth’ relationships.
+We conclude that both the methodological setup and common public image of AI are ideal for the misinterpretation that correlations between model representations and some variables of interest are `caused' by the model's understanding of underlying `ground truth’ relationships.
 We, therefore, call for the academic community to exercise extra caution, and to be keenly aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.

### 2024-05-28 body.tex
trimmed down abstract slightly

@@
 At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims.
-With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
+With successful AGI outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs).
 In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believing this.

### 2024-05-28 abstract.tex
more on abstract

@@
-Developments in the field of AI in general, and Large Language Models (LLMs) in particular, have created a `perfect storm’ for observing `sparks’ of Artificial General Intelligence (AGI) that are spurious.
+Developments in the field of Artificial Intelligence (AI), and particularly large language models (LLMs), have created a `perfect storm’ for observing `sparks’ of Artificial General Intelligence (AGI) that are spurious.
 Like simpler models, LLMs distill meaningful representations in their latent embeddings that have been shown to correlate with external variables.
 Nonetheless, the correlation of such representations has often been linked to human-like intelligence in the latter but not the former.
-We probe models of varying complexity including random projections, matrix decompositions, deep autoencoders and transformers: all of them successfully distill information that can be used to predict latent or external variables and yet none of them have ever been linked to AGI.
-We, therefore, argue that patterns in latent spaces are spurious sparks of AGI.
-Additionally, we review literature from the social sciences that shows that humans are prone to seek patterns and anthropomorphize.
+We probe models of varying complexity including random projections, matrix decompositions, deep autoencoders and transformers: all of them successfully distill information that can be used to predict latent or external variables and yet none of them have previously been linked to AGI.
+We argue that the finding of meaningful patterns in latent spaces of models cannot be seen as evidence in favor of AGI.
+Additionally, we review literature from the social sciences that shows that humans are prone to seek such patterns and anthropomorphize.
 We conclude that both the methodological setup and common public image of AI are ideal for the misinterpretation that correlations between model representations and some variables of interest are `caused' by the model's understanding of underlying `ground truth’ relationships.

### 2024-05-28 body.tex
proof-read one more time

@@
 Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\citep{weissburg2024tweets}.
-Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
+Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing LLMs, such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.
 
@@
 Two other recent works are related to this work and align well with the position we present here.~\citet{schaeffer2024emergent} demonstrate that the apparent emergent abilities of large language models may be driven by a choice of evaluation metrics, rather than some fundamental property that is intrinsic to this family of models.
-Their work highlights the need for rigorous testing and benchmarking of LLMs, which we also point to in this work albeit in a slightly different methodological context.~\citet{kloft2024ai} provide experimental evidence demonstrating that people have heightened expectations and a biased, positive view of AI.
-The authors run a user study of human-AI interaction, in which participants performed better at a given task when they thought they were aided by a positively described AI.
+Their work highlights the need for rigorous testing and benchmarking of LLMs, which we also point to in this work, albeit in a slightly different methodological context.~\citet{kloft2024ai} provide experimental evidence demonstrating that people have heightened expectations and a biased, positive view of AI.
+The authors run a user study of human-AI interaction, in which participants performed better at a given task when they (wrongly) thought they were aided by a positively described AI.
 This Placebo effect was found to be robust to negative descriptions of AI.
@@
 
-This first experiment---and indeed the practice of probing LLMs that have seen vast amounts of data---can be seen as a form of inverse problem and common caveats such as non-uniqueness and instability apply~\citep{altmeier2023regularization}.
+This first experiment---and indeed the practice of probing LLMs that have seen vast amounts of data---can be seen as a form of inverse problem and common caveats such as non-uniqueness and instability apply~\citep{haltmeier2023regularization}.
 Regularization can help mitigate these caveats~\citep{om2001ridge}, but we confess that we did not carefully consider the parameter choice for $\lambda $, nor has this been carefully studied in the related literature to the best of our knowledge.
@@
 In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks.
-Here, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.
+Here, we continue with an example in which LLMs may be used for economic sentiment prediction.
 
-Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\citep{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences.
-~\citep{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
+Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\citep{shah2023trillion} financial dataset, containing a curated selection of sentences formulated and communicated to the public by the Fed through speeches, meeting minutes and press conferences.~\citep{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'.
 In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation.
@@
 These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}.
-Formally studied as early as 1944,~\citep{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
+Formally studied as early as 1944,~\citet{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative.
 Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems.
@@
 Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations.
-The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \citep{epley2007seeing}.
+The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866 in \citet{epley2007seeing}).
 Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
@@
 When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \citep{epley2007seeing, waytz2010social}.
-Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g.
-an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors.
+Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology such as an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors.
 This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.
@@
 In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work.
-Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g.
-\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.
+Citation metrics and scholarly publications remain the primary metric for tenure and promotion~\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.
 
@@
 Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems.
-Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done.
+Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in Computer Science and AI this is typically not done.
 This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world.
-However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
+However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the alternative hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).
 
@@
 Our broadest and perhaps most ambitious goal is for our research community to move away from authorship and instead embrace contributorship.
-This argument has been raised long before both in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}.
+This argument has been raised long before in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}.
 Specifically,~\citet{liem2023treat} argue that societally impactful scientific insights should be treated as open-source software artifacts.
@@
 Similarly, code reviews are considered so integral to the process that they typically end up as accredited contributions to projects.
-Open review platforms like OpenReview are a step in the right direction but they still fall short of what is technically feasible.
+Open review platforms like OpenReview are a step in the right direction but they still fall short of what we know is technically feasible.
 Finally, software testing is, of course, not only essential but often obligatory before contributions are accepted and merged.
 As we have pointed out repeatedly in this work, any claims about AGI demand proper strong hypothesis tests.
-It is important to remember that AGI remains the alternative hypothesis and that the burden of proof therefore lies with those making claims.
+It is important to remember that AGI remains the alternative hypothesis and that the burden of proof therefore lies with those making strong claims.
 

### 2024-05-29 body.tex
final camera-ready things

@@
 
-Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}.
+Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it exhibits intelligence.}.
 Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.
@@
 When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness.
-AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML,rudin2019stop}.
+AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\citep{values_in_ML}.
 As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder.
 In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.
+At the same time, while currently popular in literature, inherent complexity and opaqueness in technical solutions may fundamentally be unwise to pursue in high-stakes applications~\citep{rudin2019stop}.
 
@@
 This is not to be seen as criticism toward the underlying methodologies in isolation, but rather in the context of current publishing practices that we repeatedly challenge throughout this work.
-Many closely related works are free of any grandiose conclusions and instead highlight the benefits of mechanistic interpretability that we also highlight in this work~\citep{nanda2023emergent,gurnee2023finding,li2022emergent}.
+Many closely related works are free of any grandiose conclusions and instead highlight the benefits of mechanistic interpretability that we also highlight here~\citep{nanda2023emergent,gurnee2023finding,li2022emergent}.
 
@@
 But despite the potential utility, the authors also demonstrate that this capacity can be partially explained by memorization, rather than an actual understanding of causal relationships.
-Similarly,~\citet{zevcevic2023causal} provide empirical evidence indicating that current LLMs ``may talk causality but are not causal''.
+Similarly,~\citet{zevcevic2023causal} provide evidence indicating that current LLMs ``may talk causality but are not causal''.
 
@@
 The authors run a user study of human-AI interaction, in which participants performed better at a given task when they (wrongly) thought they were aided by a positively described AI.
-This Placebo effect was found to be robust to negative descriptions of AI.
+This placebo effect was found to be robust to negative descriptions of AI.
 Positive bias towards AI may exacerbate other factors that drive people to make unscientific claims about the current state of AI, which we discuss in Section~\ref*{social}.
@@
 Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.
-
 This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \citep{epley2007seeing, waytz2010social}.
@@
 Citation metrics and scholarly publications remain the primary metric for tenure and promotion~\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.
-
 The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}.
@@
 
-Reflecting on the previous two sections, we make the following concrete recommendations for future research: firstly, researchers should consider acknowledging our tendency to anthropomorphize either in a dedicated ‘limitations’ section or when discussing results (\textit{acknowledgment}); secondly, researchers should refrain from drawing premature conclusions about AGI, unless these conclusions are based on strong hypothesis tests (\textit{stronger testing}); thirdly, we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and iterations over how we will measure them (\textit{epistemologically robust standards}).
+Reflecting on the previous two sections, we make the following concrete recommendations for future research:
+
+\begin{itemize}
+ \item \textbf{Acknowledgement of human bias}: researchers should be mindful of, and explicit about, risks of human bias and anthropomorphization in interpreting results, which both can be done as part of the results discussion, but also in a dedicated `limitations' section.
+ \item \textbf{Stronger testing}: researchers should refrain from drawing premature conclusions about AGI, unless these are based on strong hypothesis tests.
+ \item \textbf{Epistemologically robust standards}: we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and publicly accountable and collaborative iterations over how we will measure them, with explicit room for independent reviewing and external auditing by the broader community.
+\end{itemize}
+%firstly, researchers should consider acknowledging our tendency to anthropomorphize either in a dedicated ‘limitations’ section or when discussing results (\textit{acknowledgment}); secondly, researchers should refrain from drawing premature conclusions about AGI, unless these conclusions are based on strong hypothesis tests (\textit{stronger testing}); thirdly, we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and iterations over how we will measure them (\textit{epistemologically robust standards}).
 
 Moreover, we believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry.
-Our broadest and perhaps most ambitious goal is for our research community to move away from authorship and instead embrace contributorship.
+Our broadest and perhaps most ambitious goal is for our research community to \textbf{move away from authorship and instead embrace contributorship}.
 This argument has been raised long before in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}.
@@
 Similarly, code reviews are considered so integral to the process that they typically end up as accredited contributions to projects.
-Open review platforms like OpenReview are a step in the right direction but they still fall short of what we know is technically feasible.
+Open review platforms like OpenReview are a step in the right direction, but still fall short of what we know is technically feasible.
 Finally, software testing is, of course, not only essential but often obligatory before contributions are accepted and merged.

### 2024-05-29 main.tex
trimmed down abstract slightly
final camera-ready things

@@
 
+Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING---TU Delft collaboration.
+
+\section*{CRediT Author Statement}
+Following the CRediT Contributor Roles Taxonomy, we list author contributions in descending order of degree of contribution: \textbf{Conceptualization} PA, AD, AB, CL; \textbf{Data curation} PA; \textbf{Formal analysis} PA; \textbf{Funding acquisition} CL; \textbf{Investigation} PA; \textbf{Literature review} AD, AB, CL; \textbf{Methodology} PA, CL; \textbf{Project administration} PA, AD, AB; \textbf{Software} PA; \textbf{Supervision} CL; \textbf{Visualization} PA; \textbf{Writing - original draft} PA, AD, AB; \textbf{Writing - review and editing} PA, AD, AB, CL.
 \bibliography{biblio}

### 2024-05-31 arxiv.tex
updated arxiv pre-print version

@@
 
-\renewcommand{\shorttitle}{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}
+\renewcommand{\shorttitle}{Position: Stop Making Unscientific AGI Performance Claims}
 
@@
 
-\title{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}
+\title{Position: Stop Making Unscientific AGI Performance Claims}
 
@@
 
+\twocolumn
+
 \input{body.tex}
+
+\section*{Impact Statement}
+This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.
+As such, in the current climate of high market demand for AI innovations, we see risks of too-eager adoption in real-world applications, which may have serious societal impact.
+Thus, in this work we emphasize that an academic's impact also is in being able to thoroughly question made claims, and being explicitly aware of one's own biases.
+While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.
+
+\section*{Acknowledgements}
+
+Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING---TU Delft collaboration.
+
+\section*{CRediT Author Statement}
+Following the CRediT Contributor Roles Taxonomy, we list author contributions in descending order of degree of contribution: \textbf{Conceptualization} PA, AD, AB, CL; \textbf{Data curation} PA; \textbf{Formal analysis} PA; \textbf{Funding acquisition} CL; \textbf{Investigation} PA; \textbf{Literature review} AD, AB, CL; \textbf{Methodology} PA, CL; \textbf{Project administration} PA, AD, AB; \textbf{Software} PA; \textbf{Supervision} CL; \textbf{Visualization} PA; \textbf{Writing - original draft} PA, AD, AB; \textbf{Writing - review and editing} PA, AD, AB, CL.
 

### 2024-05-31 body.tex
updated arxiv pre-print version

@@
 
-\begin{itemize}
- \item \textbf{Acknowledgement of human bias}: researchers should be mindful of, and explicit about, risks of human bias and anthropomorphization in interpreting results, which both can be done as part of the results discussion, but also in a dedicated `limitations' section.
- \item \textbf{Stronger testing}: researchers should refrain from drawing premature conclusions about AGI, unless these are based on strong hypothesis tests.
- \item \textbf{Epistemologically robust standards}: we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and publicly accountable and collaborative iterations over how we will measure them, with explicit room for independent reviewing and external auditing by the broader community.
-\end{itemize}
+\begin{enumerate}
+ \item (\textit{Acknowledgement of Human Bias}) Researchers should be mindful of, and explicit about, risks of human bias and anthropomorphization in interpreting results, which both can be done as part of the results discussion, but also in a dedicated `limitations' section.
+ \item (\textit{Stronger Testing}) Researchers should refrain from drawing premature conclusions about AGI, unless these are based on strong hypothesis tests.
+ \item (\textit{Epistemologically Robust Standards}) We call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and publicly accountable and collaborative iterations over how we will measure them, with explicit room for independent reviewing and external auditing by the broader community.
+\end{enumerate}
 %firstly, researchers should consider acknowledging our tendency to anthropomorphize either in a dedicated ‘limitations’ section or when discussing results (\textit{acknowledgment}); secondly, researchers should refrain from drawing premature conclusions about AGI, unless these conclusions are based on strong hypothesis tests (\textit{stronger testing}); thirdly, we call for more precise definitions of terms like ‘intelligence’ and ‘AGI’, and iterations over how we will measure them (\textit{epistemologically robust standards}).
@@
 As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
-
-\section*{Impact Statement}
-This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.
-As such, in the current climate of high market demand for AI innovations, we see risks of too-eager adoption in real-world applications, which may have serious societal impact.
-Thus, in this work we emphasize that an academic's impact also is in being able to thoroughly question made claims, and being explicitly aware of one's own biases.
-While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.

### 2024-05-31 main.tex
updated arxiv pre-print version

@@
 
+\section*{Impact Statement}
+This work was written out of concern that work easily recognized as `impactful' in current AI research and development, can also easily be over- and misinterpreted.
+As such, in the current climate of high market demand for AI innovations, we see risks of too-eager adoption in real-world applications, which may have serious societal impact.
+Thus, in this work we emphasize that an academic's impact also is in being able to thoroughly question made claims, and being explicitly aware of one's own biases.
+While the AI publishing landscape has to our feeling transformed too much into a noisy race to get exciting results in fast, we hope the research community can create more room for this type of deeper questioning.
+
 \section*{Acknowledgements}
@@
 Following the CRediT Contributor Roles Taxonomy, we list author contributions in descending order of degree of contribution: \textbf{Conceptualization} PA, AD, AB, CL; \textbf{Data curation} PA; \textbf{Formal analysis} PA; \textbf{Funding acquisition} CL; \textbf{Investigation} PA; \textbf{Literature review} AD, AB, CL; \textbf{Methodology} PA, CL; \textbf{Project administration} PA, AD, AB; \textbf{Software} PA; \textbf{Supervision} CL; \textbf{Visualization} PA; \textbf{Writing - original draft} PA, AD, AB; \textbf{Writing - review and editing} PA, AD, AB, CL.
+
 \bibliography{biblio}
